### 2024-04-17 submission/manuscript.Rmd
bulleted outline of intro and figures

Title: Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

Gaps in knowledge addressed: 
* Are other tissue-specific genes dysregulated, or only testis, germline genes?
* Curating a robust list of male and female germline genes
	* Should talk about 2-cell genes vs germline genes - way to systematically categorize? 
* Mechanism behind long-term germline gene misexpression
	* Recent evidence suggests loss of KDM5C in ESCs express some germline genes
		* Unclear if catalytic activity is required for long-term silencing
		* Unclear if their dysregulation lasts throughout life or  the same between brain or not
	* When in development does it begin? - Recent evidence suggests some germline genes expressed in 5CKO ESCs but unclear if their dysregulation lasts throughout differentiation and if the identity of germline genes are different compared to the brain
* Are there functional consequences to germline gene misexpression?

Introduction:
* Chromatin regulators are important for cellular identity
	* H3K4me1-3 linked to active gene promoters and enhancers
	* Surprisingly, mutations in Chromatin regulators lead to many NDDs (including many H3K4 regulators)
		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-specific genes
		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention

* Loss of KDM5C can result in the misexpression of genes typically only found in the testis
	* Misexpression of tissue-specific genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
	* Systematic characterization of ectopic germline genes hasn't been done
		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.

Outline of figures:
**Figure 1: Misexpression of tissue-specific genes in the _Kdm5c_-KO brain**
	* MA-plot and bar graphs of tissue-enriched genes
	* Should we show an example bigwigs/NCBI of another tissue-specific gene or just focus on testis?
	* Example testis-specific genes (NCBI and bigwigs)
	* Overlap of testis DEGs between amygdala and hippocampus
	* Gene ontology of amygdala and hippocampus - ontologies are germline ontologies

**Figure 2: The male _Kdm5c_-KO brain expresses male and female germline-enriched genes**
	* Expression of testis DEGs in germline-depleted testis (this is adult testis data)
	* scRNAseq of testis - # of testis DEGs that are germline-specific markers
	* Although far fewer, 5CKO brain also expresses ovary-enriched genes(NCBI and bigwigs of Zar1) 
	* These ovary enriched genes are also germline specific (NCBI/Li tissues are in adult ovary, so it would be best to show they're oocyte-specific in adult ovary. But I don't think there's a published adult female W/Wv dataset. Could try looking at scRNAseq or just do TPM in embryonic W/Wv data since oocytes are developed at this point? Or both?)
	* Defining what is/isn't a germline gene, and which are male/female biased using embryonic W/Wv data

**Figure 3: Kdm5c-KO epiblast-like cells express master regulators of germline identity**
	* Left - Morphology is unchanged, Right - 5CKO EpiLCs express EpiLC differentiation genes
	* Male EpiLCs express germline genes (example Cyct again)
	* Overlap between brain and EpiLC germline genes - show they're mostly unique
	* GO of Brain and EpiLC germline genes (meiotic genes should be enriched) 
	* Bigwigs or TPM of master regulators
		* Show that while some are also 2-cell genes (Dazl), 2-cell specific genes aren't dysregulated (Zscan4). Important point because published KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I can tell only genes shared between germline and 2-cell are dysregulated.  
	* Staining of Dazl (+ Stra8 if I can get it to work)

**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
	* Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
	* Increase in H3K4me2 in EpiLCs at germline genes 
	* Kdm5c binding in EpiLCs vs PNCs to show that germline repression is happening in early embryo
	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). KDM5C RNA and protein ESC --> EpiLC (increasing then decreasing)
	* RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
	* DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)

**Figure 5: Ectopic, germline-like phenotypes in Kdm5c-KO ESCs/EpiLCs**
	* Sycp3 staining
	* DDX4 staining and repression of retrotransposons
	* Cilia??

### 2024-04-17 submission/manuscript.Rmd
outline results for figure 1 and 2 and render manuscript

@@
-Title: Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
+---
+title:  Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
+output: pdf_document
+geometry: margin = 0.5in
+font-size: 11pt
+header-includes:
+   - \usepackage{caption}
+   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
+   - \usepackage{sectsty} 
+   - \usepackage{helvet}
+   - \renewcommand{\familydefault}{\sfdefault}
+   - \usepackage{lineno}
+   - \linenumbers
+   - \renewcommand\linenumberfont{\normalfont\small}
+   - \usepackage{setspace}
+   - \spacing{1.5}
+   - \pagenumbering{arabic}
+
+bibliography: references.bib
+csl: cell-reports.csl
+indent: true
+editor_options:
+  chunk_output_type: console
+---
 
@@
 
-Outline of figures:
+# Results
+
+## Tissue-specific genes, including testis genes, are enriched within the _Kdm5c_-KO brain
+
+*note: the 5CKO DEGs table we should make a column for if the gene is tissue-specific and which tissue it's in  
+	* see if any liver or muscle genes are dysregulated in both tissues, could add as a bigwig
+
+```{r, DESeq2_params, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+#get the parameters used repeatedly, like the log2fold change cut off
+source("../code/utilities/parameters.R")
+
+```
+Previous RNA sequencing studies performed in the adult male _Kdm5c_ knockout (-KO) hippocampus identified aberrant transcription of genes that are typically only expressed in the testis[@scandagliaLossKdm5cCauses2017].
+Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflect widespread impairment of tissue identity.
+To distinguish these possibilities, we globally assessed if tissue-specific genes were significantly enriched in our previous mRNA-seq dataset of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020].
+We classified differentially expressed genes (DEGs) as tissue-specific if they were significantly upregulated in the _Kdm5c_-KO brain compared to wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) and were previously identified as uniquely enriched within a given mouse tissue[@liComprehensiveMouseTranscriptomic2017].
+
+```{r, tissue_genes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+#get the number and stats for tissue-specific gene expression
+tissue_res <- read.csv("../data/processed/TissueSpecific_AMYHIP_numberofgenesDEGs.csv", header = TRUE)
+
+#get the value you want to print 
+#col is which data column, tiss is which tissue
+gettissval <- function(col, tiss){
+	val <- subset(tissue_res, tissue == tiss)
+	val <- val[,col]
+	return(val)
+}
+
+```
+
+* Based on these criteria, we observed widespread misexpression of tissue-enriched genes within the _Kdm5c_-KO amygdala and hippocampus, with the majority of genes belonging to the testis (Figure 1A-C).
+
+* Even though the testis has the largest number of tissue-specific genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis genes were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact).
+
+* This dysregulation of tissue identity is not unique to the testis, as we also observed significant misexpression of ovary, liver, and muscle genes within the _Kdm5c_-KO brain (Figure 1A-E).
+
+Importantly, these tissue-specific genes show little to no expression in the wild-type brain and have no known brain functions, yet our mRNA-seq data indicates they are polyadenylated and spliced into mature transcripts (Figure 1C,G).
+
+* Of note, we did not observe enrichment of brain-specific genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), even though the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+
+* Together, these results suggest misexpression of testis and other tissue-specific genes within the _Kdm5c_-KO brain is not due to random de-repression of all genes, but rather due to a dysregulation of tissue identity.
+
+### Male and female germline genes are aberrantly expressed in the male _Kdm5c_-KO brain
+
+Intriguingly, many _Kdm5c_-KO testis-enriched DEGs have functions specific to male germ cells, such as _Cytochrome C, testis (Cyct)_, a (Figure 1C), a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002].
+We therefore wondered if the high enrichment of testis genes within the _Kdm5c_-KO brain reflected a failure to demarcate the soma and germline, as the testis contains both germ cells (e.g.
+spermatogonia) as well as somatic cells (e.g.
+Leydig cells) that support hormone production and germline functions.
+Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including cellular process involved in reproduction in multicellular organism (GO:0022412), spermatid development (GO:0007286), and sperm axoneme assembly (GO:0007288) (Figure 1A).
+
+To validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013].
+We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 1B).
+We additionally assessed testis DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
+We found that while some testis-enriched DEGs were classified as specific markers for multiple stages of germ cell development - including spermatogonia, spermatocytes, round spermatids, and elongating spermatids, none marked somatic cells (Figure 1C).
+Altogether, these data suggest testis genes expressed within the _Kdm5c_-KO brain are actually male germ cell genes.
+
+Interestingly, we also observed significant enrichment of ovary genes within the male _Kdm5c_-KO brain (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact).
+Although there are relatively few ovary-enriched DEGs compared to testis-DEGs (Amygdala: `r gettissval("Amygdala_DEGnumb", "Testis")` testis DEGs, `r gettissval("Amygdala_DEGnumb", "Ovary")` ovary DEGs; Hippocampus: `r gettissval("Hippocampus_DEGnumb", "Testis")` testis DEGs, `r gettissval("Hippocampus_DEGnumb", "Ovary")` ovary DEGs), they also had known germline-specific functions.
+For example, _Zygotic arrest 1 (Zar1)_ was recently shown to regulate oocyte translation by sequestering mRNAs[@chengMammalianOocytesStore2022] (Figure 1D).
+
+* To assess if ovary-enriched DEGs were also germline specific, we assessed their expression in wild-type and germline-depleted (Kit^W/Wv^) female embryos, when female germ cells undergo the initial stages of development and meiosis.
+	* We found XYZ.
+	* Altogether, these results indicate male _Kdm5c_-KO mice express both male and female germline genes within the amygdala and hippocampus.
+
+```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+germgenes <- read.csv("../data/processed/germGENES20.csv")
+
+```
+* One difficulty in globally characterizing the misexpression of germline genes is a comprehensive list of mouse germline-specific genes is currently lacking.
+
+* To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating expression in wild-type and germline-depleted (Kit^W/Wv^) male and female mouse embryos from embryonic day 12, 14, and 16.
+	* **note: need to figure out why these time points. We have P6 and adult for males but not females so it would be biased towards males if we included the later ages**
+
+We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
+These criteria yielded `r nrow(germgenes)` germline-enriched genes.
+To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female (XX) or male germline.
+We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` XX-biased, `r nrow(subset(germgenes, sexBias == "XY"))` XY-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
+
+# Discussion
+
+* Papers to read/reference:
+	* Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
+	* two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
+
+# References
+<div id="refs"></div>
+
+\newpage
+# Figures and Tables
+
+\begin{figure}
+  \centering 
+  \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
+  \caption[Misexpression of tissue-specific genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Misexpression of tissue-specific genes in the \textit{Kdm5c}-KO in the brain} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala.
+Left - MA plot of mRNA-sequencing.
+Right - The number of tissue-enriched differentially expressed genes (DEGs).
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{B.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO hippocampus.
+\textbf{C.} Left - Bigwigs of an example aberrantly expressed testis-enriched gene, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
+Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
+\textbf{D.} Left - Bigwigs of an example ovary-enriched germline DEG, \textit{Zygotic arrest 1 (Zar1)}.
+Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.}
+  \label{figurelabel}
+\end{figure}
+
 **Figure 1: Misexpression of tissue-specific genes in the _Kdm5c_-KO brain**
 	* MA-plot and bar graphs of tissue-enriched genes
-	* Should we show an example bigwigs/NCBI of another tissue-specific gene or just focus on testis?
 	* Example testis-specific genes (NCBI and bigwigs)
-	* Overlap of testis DEGs between amygdala and hippocampus
-	* Gene ontology of amygdala and hippocampus - ontologies are germline ontologies
+	* An example ovary tissue-specific gene 
+	* An example muscle/liver tissue-specific gene (NCBI and bigwigs)
 
 **Figure 2: The male _Kdm5c_-KO brain expresses male and female germline-enriched genes**
+	* Gene ontology of testis DEGs in the amygdala and hippocampus - ontologies are germline ontologies
 	* Expression of testis DEGs in germline-depleted testis (this is adult testis data)

### 2024-04-17 code/TissueGenes_AmyHip.R
make excel file of tissue-specific DEGs

@@ -52,3 +52,9 @@ p <- ggbarplot(countingdf, "tissue", "count", fill = "tissue", color = "tissue",
 
 pl <- ggpar(p, legend = "none", ylab = "# of genes", xlab = FALSE, title = paste0("Total Tissue Gene #") )
 ggsave(snakemake@output[[9]], plot = pl, width = 5, height = 5)
+
+### save tissue-specific DEGs in a dataframe
+for(i in 1:length(titles)){
+	tissDEG <- tissue_genes(i)
+	write.table(tissDEG, snakemake@output[[9+i]], sep = ",", row.names = FALSE)
+	}

### 2024-04-18 submission/manuscript.Rmd
edit fig panel order

@@
 **Figure 3: Kdm5c-KO epiblast-like cells express master regulators of germline identity**
-	* Left - Morphology is unchanged, Right - 5CKO EpiLCs express EpiLC differentiation genes
-	* Male EpiLCs express germline genes (example Cyct again)
+	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 
+	* B) 5CKO EpiLCs express EpiLC differentiation genes similar to WT lvls
+	* C) Male EpiLCs express germline genes (example Cyct again)
 	* Overlap between brain and EpiLC germline genes - show they're mostly unique
-	* GO of Brain and EpiLC germline genes (meiotic genes should be enriched) 
+	* GO of Brain and EpiLC germline genes (meiotic enriched) 
 	* Bigwigs or TPM of master regulators

### 2024-04-18 code/GO_brain_testisDEGs.R
gene ontology of testis DEGs from amy and hip

@@ -1,0 +1,22 @@
+#24.04.18 Gene ontology of 5CKO testis DEGs from amygdala and hippocampus
+library(enrichplot)
+library(org.Mm.eg.db)
+library(clusterProfiler)
+library(ggplot2)
+
+
+#read in the testis DEGs
+testisDEGs <- read.csv(snakemake@input[[1]], sep = ",", header = TRUE)
+	#first column is the ENSEMBL IDs
+
+#choose what organism and what type of ontology (biological process)
+ego <- enrichGO(testisDEGs[,1], keyType = 'ENSEMBL', OrgDb = "org.Mm.eg.db", ont="BP", readable=TRUE)
+print(ego[, c("ID", "Description", "p.adjust")])
+
+q <- dotplot(ego, showCategory=15)
+
+
+ggsave(snakemake@output[[1]], plot = q, width = 5.5, height = 6)
+
+
+#ego <- enrichGO(de, keyType = 'ENSEMBL', OrgDb = "org.Mm.eg.db", ont="BP", readable=TRUE)

### 2024-04-18 submission/manuscript.Rmd
add ovary/germline gene section

@@
 Interestingly, we also observed significant enrichment of ovary genes within the male _Kdm5c_-KO brain (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact).
-Although there are relatively few ovary-enriched DEGs compared to testis-DEGs (Amygdala: `r gettissval("Amygdala_DEGnumb", "Testis")` testis DEGs, `r gettissval("Amygdala_DEGnumb", "Ovary")` ovary DEGs; Hippocampus: `r gettissval("Hippocampus_DEGnumb", "Testis")` testis DEGs, `r gettissval("Hippocampus_DEGnumb", "Ovary")` ovary DEGs), they also had known germline-specific functions.
+Although there are relatively few ovary-enriched DEGs compared to testis-DEGs (Amygdala: `r gettissval("Amygdala_DEGnumb", "Testis")` testis DEGs, `r gettissval("Amygdala_DEGnumb", "Ovary")` ovary DEGs; Hippocampus: `r gettissval("Hippocampus_DEGnumb", "Testis")` testis DEGs, `r gettissval("Hippocampus_DEGnumb", "Ovary")` ovary DEGs), they also have known germline-specific functions.
 For example, _Zygotic arrest 1 (Zar1)_ was recently shown to regulate oocyte translation by sequestering mRNAs[@chengMammalianOocytesStore2022] (Figure 1D).

### 2024-04-18 submission/manuscript.Rmd
add updated figs to manuscript

@@
 
+\begin{figure}
+  \centering 
+  \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
+  \caption[Misexpression of tissue-specific genes, including germline genes, in the \textit{Kdm5c}-KO in the brain]{\textbf{Ovary and testis germline genes are expressed in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis.
+Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM).
+\textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis.
+Germline cell types are highlighted in green, somatic cell types in black.
+\textbf{D.} Expression of ovary-enriched DEGs in a wild-type versus germ cell-depleted testis }
+  \label{figurelabel}
+\end{figure}
+
+\newpage
+## Figure outline:
+
 **Figure 1: Misexpression of tissue-specific genes in the _Kdm5c_-KO brain**

### 2024-04-18 submission/manuscript.Rmd
update germline gene text

@@
 Leydig cells) that support hormone production and germline functions.
-Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including cellular process involved in reproduction in multicellular organism (GO:0022412), spermatid development (GO:0007286), and sperm axoneme assembly (GO:0007288) (Figure 1A).
+Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including spermatid development (GO:0007286) and sperm axoneme assembly (GO:0007288) (Figure 1A).
 
@@
 We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 1B).
-We additionally assessed testis DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
+The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
+We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018].
 We found that while some testis-enriched DEGs were classified as specific markers for multiple stages of germ cell development - including spermatogonia, spermatocytes, round spermatids, and elongating spermatids, none marked somatic cells (Figure 1C).

### 2024-04-18 submission/manuscript.Rmd
add EpiLC draft text and figure

@@
 
+## _Kdm5c_-KO cells express master regulators of germline identity in an _in vitro_ model of early embryogenesis
+
+Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity.
+Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020].
+This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 2A).
+Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
+Therefore, to elucidate the mechanism underlying aberrant germline genes expression in the mature _Kdm5c_-KO brain, we first assessed their expression in male (XY) and female (XX) _Kdm5c_ mutant EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].
+Because complete loss of KDM5C in females is not lethal at this time point, we were able to characterize germline transcripts in both _Kdm5c_ heterozygous (XX _Kdm5c_-HET) and homozygous null (XX _Kdm5c_-KO) female EpiLCs.
+
+We found both XX and XY _Kdm5c_ mutant EpiLCs aberrantly express many germline transcripts in comparison to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1)(Figure 2B).
+While many germline genes were dysregulated in all _Kdm5c_ mutants, female heterozygous and homozygous mutants expressed about twice as many germline genes than _Kdm5c_-KO males (XX _Kdm5c_-HET: 212  XX _Kdm5c_-KO: 186, XY _Kdm5c_-KO: 90) (Figure 2C).
+We then wondered if the abundance of germline transcripts in _Kdm5c_ mutant females could be due to defects in X chromosome inactivation caused by KDM5C loss[@samantaActivationXistEvolutionarily2022], since the X chromosome is enriched with many spermatogenesis genes[@wangAbundanceXlinkedGenes2001].
+However, the majority of germline DEGs for both males and females are autosomal genes (Figure 2C), indicating increased germline gene misexpression in females is independent of any potential X chromosome inactivation defects.
+We then asked if the identity of ectopic germline transcripts was influenced by _Kdm5c_ mutant sex, as some germline genes show biased transcription in the wild-type XX or XY gonads.
+We found XX _Kdm5c_-HET, XX _Kdm5c_-KO, and XY _Kdm5c_-KO EpiLCs all expressed similar proportions of XX-biased, XY-biased, and unbiased germline genes that roughly reflected the proportion of genes shared and unique within the male and female wild-type gonads (In WT gonads: 6% XX-biased, 79% XY-biased, 15% unbiased) (Figure 2D).
+*Note: test if unbiased genes are over-represented, I think % wise they are*
+
+To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_ mutant brain.
+The majority of EpiLC germline DEGs were unique to EpiLCs, with only a few shared across brain sequencing datasets (Figure 2E).
+Only two germline genes were consistently misexpressed in all male and female datasets, _DNA segment, Chr 1, Pasteur Institute 1 (D1Pas1)_ and _Testis expressed gene 14 (Tex14)_.
+Surprisingly, germline DEGs unique to _Kdm5c_ mutant EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 2F).
+These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but they are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+_Stra8_'s expression is typically sexually dimorphic, as female germ cells initiate meiosis during embryogenesis while males begin meiosis after birth [@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+Yet we found both XX and XY _Kdm5c_ mutant EpiLCs ectopically express _Stra8_ RNA.
+XX and XY _Kdm5c_ mutants also aberrantly express _Dazl_, a translational regulator essential for germ cell function[@liDAZLMasterTranslational2019].
+Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_ mutant EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (XY _Kdm5c_-KO EpiLCs p = 1.698e-07, Fisher’s Exact Test).
+This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function.
+Therefore, we then stained XY _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 2G).
+We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test, XY KO versus WT), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
+
 # Discussion
@@
 \textbf{D.} Expression of ovary-enriched DEGs in a wild-type versus germ cell-depleted testis }
+  \label{figurelabel}
+\end{figure}
+
+\begin{figure}
+  \centering 
+  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
+  \caption[Male and female \textit{Kdm5c} mutant epiblast-like cells express master regulators of germline identity ]{\textbf{Male and female \textit{Kdm5c} mutant epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of emrbyonic stem cells (ESCs) of the inner cell mass into epiblast stem cells and primordial germ cells.
+Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
+\textbf{B.} Left - Bigwigs of an example germline gene, \textit{Cyct} that is dysregulated in male and female \textit{Kdm5c} mutant EpiLCs.
+Right - Overlap of all germline DEGs between male and female EpiLCs.
+\textbf{C.} Chromosome location of germline genes misexpressed in male and female EpiLCs.
+\textbf{D.} Number of EpiLC germline DEGs that show sex-biased expression in the wild-type gonads.
+\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in male and female \textit{Kdm5c} mutant EpiLC and brain RNA-seq datasets.
+\textbf{F.} Expression of genes key for early germ cell development in \textit{Kdm5c} mutant EpiLCs in transcrips per million (TPM) with p-values for Welch's t-test compared to wild type of the same sex.
+\textbf{G.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.}
   \label{figurelabel}

### 2024-04-18 submission/manuscript.Rmd
bullet point draft of EpiLC data

--- 2024-04-18 submission/manuscript.Rmd
@@
 
+## _Kdm5c_-KO cells express master regulators of germline identity in an _in vitro_ model of early embryogenesis
+
+Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity.
+Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020].
+This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top).
+Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
+Therefore, to elucidate the mechanism underlying aberrant germline genes expression in the mature _Kdm5c_-KO brain, we first assessed their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].
+
+* We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom).
+Additionally, _Kdm5c_-KO EpiLCs downregulated ESC pluripotency genes and upregulated EpiLC differentiation genes similar to wild-type (Figure 3B).
+
+* We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis (Figure 3C).
+
+* Of these testis genes, we also observed misexpression of germline-enriched genes, including _Cyct_ (Figure 3D)
+
+* To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+The majority of EpiLC germline DEGs were unique to EpiLCs, with only a few shared across brain sequencing datasets (Figure 3E).
+Only two germline genes were consistently misexpressed in all datasets, _DNA segment, Chr 1, Pasteur Institute 1 (D1Pas1)_ and _Testis expressed gene 14 (Tex14)_.
+
+* We then compared the typical function of brain and EpiLC germline DEGs and found EpiLC germline DEGs displayed strong enrichment of meiosis-related gene ontologies, including XXXXX (Figure 3F).
+
+* Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
+These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but they are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+
+* _Dazl_ is a translational regulator essential for germ cell function[@liDAZLMasterTranslational2019].
+Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (XY _Kdm5c_-KO EpiLCs p = 1.698e-07, Fisher’s Exact Test).
+This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function.
+
+* Therefore, we then stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H).
+We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test, XY KO versus WT), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
+
 # Discussion

### 2024-04-18 code/utilities/mouseGermlineGenes.R

@@ -221,3 +221,8 @@ write.table(germGENES20, file = snakemake@output[[1]], sep =",", row.names = FAL
 # checking <- subset(XXvsXY, select = c(ENSEMBL, SYMBOL.x, XX_avg, XY_avg, XXoverXY, sexBias))
 
 # XXvsXY <- subset(XXvsXY, select = c(ENSEMBL, sexBias))
+
+
+## make a snakey graph of filtering for germline-enriched genes
+#https://github.com/davidsjoberg/ggsankey
+#devtools::install_github("davidsjoberg/ggsankey")

### 2024-04-18 code/DESeq2_EpiLC.R
EpiLC DESeq2 just males

@@ -1,0 +1,110 @@
+#############################################################################################
+#                                                                                           #
+# 2022 12 14 DESeq2 using snakemake of Kdm5c mutant EpiLCs                                  #
+# Inputs:
+#       cts = counts file (.txt)                                                            #
+#       sampleinfo = sample information sheet #MAKE SURE THE SAMPLES MATCH THE CTS FILE NAMES
+# Libraries: DESeq2, ggplot2                                                                #
+# Outputs: PCA plot of samples and DESeq2 results tables (.csv)                             #
+#                                                                                           #
+#############################################################################################
+#load the color palette utilities first
+source('code/utilities/colorpalettes.R')
+
+###1) format the raw counts table so that all values are integers and there are no gene variant numbers
+cts <- read.csv(snakemake@input[["cts"]], sep ="\t", row.names = 1)
+    #convert counts to integers
+cts[1:ncol(cts)] <- lapply(cts[1:ncol(cts)], as.integer)
+
+###2) Make the sample sheet (a csv file that has the genotype and sex for each sample)
+#get the column (sample) names
+samples <-  colnames(cts)
+
+#get the sex of each sample
+sex <- c()
+ct <- 1
+for(r in samples){
+		#get the first two characters of the string. If they're XX, they're female
+	if (substr(r, 1, 2) == "XX"){
+		rei <- "XX"
+	    } else if (substr(r, 1, 2) == "XY"){ #if they're male
+		rei <- "XY"
+	    } else {
+		rei <- "ERROR"
+	}
+	sex[ct] <- rei
+	ct <- ct + 1
+	
+}
+
+#get the genotype of each sample
+genotype <- c()
+cnt <- 1
+for(g in samples){
+		#get the end fo the string
+	if (substr(g, 10, 11) == "WT"){
+		geno <- "WT"
+        } else if (substr(g, 10, 12) == "HET") {
+        geno <- "5cHET"
+        } else if (substr(g, 10, 11) == "KO"){
+        geno <- "5cKO"
+        } else { #the rest of the samples have 7 characters
+		geno <- "ERROR"
+	}
+	genotype[cnt] <- geno
+	cnt <- cnt + 1
+}
+
+
[…]

### 2024-04-18 code/DESeq2_EpiLC.R
update header

@@ -5,7 +5,7 @@
 #       cts = counts file (.txt)                                                            #
 #       sampleinfo = sample information sheet #MAKE SURE THE SAMPLES MATCH THE CTS FILE NAMES
 # Libraries: DESeq2, ggplot2                                                                #
-# Outputs: PCA plot of samples and DESeq2 results tables (.csv)                             #
+# Outputs: DESeq2 results tables (.csv)                             #
 #                                                                                           #
 #############################################################################################
 #load the color palette utilities first

### 2024-04-18 code/Upset_Brain_EpiLC.R
make and add upset plot to fig

@@ -1,0 +1,39 @@
+### 24.03.06 adapted from dissertation code 23.10.20 upset plot comparing XX and XY EpiLCs and brain germline DEGs
+
+##1) Read in the germline genes list
+germ <- read.csv(snakemake@input[[1]], sep = ",")
+
+##2) Read in the EpiLC and brain XX and XY germline DEGs, make sure files are in same order
+source("code/utilities/parameters.R") #for the log2fc cut off (l2fcco)
+samples <- c("EpiLC XY 5CKO", "AMY 5CKO", "HIP 5CKO")
+DEGs <- list()
+for (s in 1:length(samples)){
+	#read in the dataframe
+	df <- read.csv(snakemake@input[[s+1]], sep = ",")
+	
+	gdf <- subset(df, ENSEMBL %in% germ$ENSEMBL & log2FoldChange > l2fcco)
+	germGENES <- gdf$ENSEMBL
+	#subset for the number of germline genes
+	DEGs[[s]] <- germGENES
+}
+names(DEGs) <- samples
+
+##3) Calculate the overlap between groups
+
+#make an upset plot for the overlap
+library("UpSetR")
+
+modifiedupset <- function(samplelist){
+	upset(fromList(samplelist), order.by = "freq",  sets.x.label = "# Germline DEGs", mainbar.y.label = "# of Overlapping Germline DEGs", empty.intersections = "on")
+}
+
+
+
+#just male samples
+pdf(file = snakemake@output[[1]], width = 8, height = 5)
+
+upset(fromList(DEGs), order.by = "freq",  sets.x.label = "# Germline DEGs", mainbar.y.label = "Overlapping Germline DEGs", empty.intersections = "on", text.scale = 2)
+
+dev.off()
+
+

### 2024-04-18 code/GO_Compare_BrainEpiLC.R
gene ontology of EpiLC vs brain germline DEGs

@@ -1,0 +1,32 @@
+#24.04.09 Gene ontology comparison between male Brain DEGs and EpiLC germline genes to demonstrate the different types of genes dysregulated
+library(enrichplot)
+library(org.Mm.eg.db)
+library(clusterProfiler)
+library(ggplot2)
+
+# make a list with all the germline genes
+
+#sample names, make sure order matches snakefile input
+samples <- c("EpiLC", "Amygdala", "Hippocampus")
+germDEGs <- list()
+
+for(i in 1:length(samples)){
+	DEGs <- read.csv(snakemake@input[[i]], sep = ",", header = TRUE)
+	#get the ensembl names, put in the position of the list
+	germDEGs[[i]] <- DEGs[,1]
+}
+
+names(germDEGs) <- samples
+print(head(germDEGs))
+
+#now run the gene ontology comparison
+ck <- compareCluster(geneCluster = germDEGs, fun = enrichGO,  OrgDb = "org.Mm.eg.db", keyType="ENSEMBL", ont="BP")
+#ck <- setReadable(ck, OrgDb = "org.Mm.eg.db", keyType="ENSEMBL")
+head(ck) 
+
+write.table(data.frame(ck), snakemake@output[[1]], row.names = FALSE)
+
+ggsave(snakemake@output[[2]], plot = dotplot(ck, size = "Count"), width = 9, height = 5.5)
+
+
+    #ego <- enrichGO(de, keyType = 'ENSEMBL', OrgDb = "org.Mm.eg.db", ont="BP", readable=TRUE)

### 2024-04-19 code/GO_Compare_BrainEpiLC.R
fix GO csv file

@@ -24,7 +24,7 @@ ck <- compareCluster(geneCluster = germDEGs, fun = enrichGO,  OrgDb = "org.Mm.eg
 #ck <- setReadable(ck, OrgDb = "org.Mm.eg.db", keyType="ENSEMBL")
 head(ck) 
 
-write.table(data.frame(ck), snakemake@output[[1]], row.names = FALSE)
+write.table(ck, snakemake@output[[1]], row.names = FALSE, sep = ",")
 
 ggsave(snakemake@output[[2]], plot = dotplot(ck, size = "Count"), width = 9, height = 5.5)
 

### 2024-04-19 code/utilities/colorpalettes.R
add captial c to 5cKO EpiLC colors

@@ -38,7 +38,7 @@ tissue_palette2 <-  c("Adrenal Gland" = "#f5372bff", "Brain" = "#f75c52ff","Fore
 #all XX and XY EpiLC genos
 EpiLCpalette <- c("XYWT" = EpiLC_XY_WT, "XY5cKO" = EpiLC_XY_KO, "XXWT" = EpiLC_XX_WT, "XX5cHET" = EpiLC_XX_HET, "XX5cKO" = EpiLC_XX_KO, "Both" = "deepskyblue2")
 #male eplic palette
-EpiLC_XY_palette <- c("WT" = EpiLC_XY_WT, "5cKO" = EpiLC_XY_KO)
+EpiLC_XY_palette <- c("WT" = EpiLC_XY_WT, "5cKO" = EpiLC_XY_KO, "5CKO" = EpiLC_XY_KO)
 
 
 #EpiLC with RA

### 2024-04-19 code/EpiLC_markers.R
plot pluripotency marker tpm

@@ -1,0 +1,153 @@
+### 24.04.19 Assess expression of ESC and EpiLC genes in WT and 5CKO to ensure differentiation occurred properly
+	# use genes in this paper as a starting point: https://pubmed.ncbi.nlm.nih.gov/38177678/
+	#make heatmap of TPM in WT and 5CKO
+
+###1) list of naive and primed pluripotency genes
+
+#should be low in WT EpiLCs
+naivegenes <- data.frame(Symbol = c("Esrrb", "Fgf4", "Klf2", "Klf4", "Klf5", "Nr0b1", "Nr5a2", "Prdm14", "Sox13", "Tbx3", "Tcl1", "Tet2", "Tfcp2l1", "Zfp42"), ENSEMBL = c("ENSMUSG00000021255", "ENSMUSG00000050917", "ENSMUSG00000055148", "ENSMUSG00000003032", "ENSMUSG00000005148", "ENSMUSG00000025056", "ENSMUSG00000026398", "ENSMUSG00000042414", "ENSMUSG00000070643", "ENSMUSG00000018604", "ENSMUSG00000041359", "ENSMUSG00000040943", "ENSMUSG00000026380", "ENSMUSG00000051176"))
+
+#should be high in WT EpiLC
+primedgenes <- data.frame(Symbol = c("Cdh2", "Dnmt3a", "Dnmt3b", "Etv4", "Etv5", "Foxd3", "Lef1", "Otx2", "Pou3f1", "Sox3", "Tcf15"), ENSEMBL = c("ENSMUSG00000024304", "ENSMUSG00000020661", "ENSMUSG00000027478", "ENSMUSG00000017724", "ENSMUSG00000013089", "ENSMUSG00000067261", "ENSMUSG00000027985", "ENSMUSG00000021848", "ENSMUSG00000090125", "ENSMUSG00000045179", "ENSMUSG00000068079"))
+
+#genes for core pluripotency, not expected to change with ESC to EpiLCs
+coregenes <- data.frame(Symbol = c("Lin28a", "Nanog", "Pou5f1", "Sall4", "Sox2", "Utf1", "Zfp281"), ENSEMBL = c("ENSMUSG00000050966", "ENSMUSG00000012396", "ENSMUSG00000024406", "ENSMUSG00000027547", "ENSMUSG00000074637", "ENSMUSG00000047751", "ENSMUSG00000041483"))
+
+
+###2) Read in the TPM file
+
+#format the TPM files
[…]

### 2024-04-19 code/TissueGenes_EpiLC.R
tissue-enriched gene analysis epilcs

@@ -1,0 +1,38 @@
+#24.03.06 - generate graphs for tissue specific genes in EpiLCs and P6 brain
+
+source("code/utilities/tissueSPECIFICgenes.R") #read in the functions
+
+
+### run for the samples, make sure order matches snakefile
+titles <- c("EpiLC XY 5CKO")
+fisherdfs <- list()
+
+for(i in 1:length(titles)){
+	tissDEG <- tissue_genes(i)
+
+	#n = position in snakefile for input DEG table and output volcano plot
+	#title = name of the dataset for the graph
+	#td = tissue-specific DEG 
+	#offset - number of samples to shift the input and output numbers to align correctly
+	#bary - y-axis of bar graph
+	bary <- ifelse(i <= 3, 100, 20) #if it's less than or equal to 3 it's the epilc samples which need a higher y-axis
+	tissue_plots(i, titles[i], tissDEG, length(titles), bary)
+
+	fisherdfs[[i]] <- tissue_fisher(tissDEG, i, titles[i])
+
+}
+
+#merge together the tissue count with the fisher results
+fisherdfs[[length(titles)+1]] <- countingdf
+library(tidyverse)
+fisherdfs <- fisherdfs %>% reduce(full_join, by = "tissue")
+print(fisherdfs)
+
+write.table(fisherdfs, snakemake@output[[4]], sep = ",", row.names = FALSE)
+
+
+### save tissue-specific DEGs in a dataframe
+for(i in 1:length(titles)){
+	tissDEG <- tissue_genes(i)
+	write.table(tissDEG, snakemake@output[[i+4]], sep = ",", row.names = FALSE)
+	}

### 2024-04-19 code/EpiLC_PGCgenes.R
tpm of pgc genes

@@ -1,0 +1,139 @@
+# To plot the expression of primordial germ cell genes in WT and 5CKO EpiLCs #
+
+###1) Read in the TPM file
+
+#function to format the TPM files
+#name is the name in snakemake file
+EpiLC_TPM <- read.csv(file = snakemake@input[[1]], sep ="\t", row.names = 1)
+genes <-(row.names(EpiLC_TPM))
+#remove variant numbers from ensembl gene notation
+ENSEMBL<- read.csv(text = genes, sep=".", header = FALSE, col.names = c("ENSEMBL", "var"))
+   #change the rownames of the counts file to the ensembl genes without the variant numbers
+row.names(EpiLC_TPM) <- ENSEMBL$ENSEMBL
+
+#format the column names by removing extraneous info
+colnames(EpiLC_TPM) <- gsub(".TPM", "", colnames(EpiLC_TPM))
+head(EpiLC_TPM)
+
+
+
+
+###2) read in the sample info
+EpiLC_SI <- read.csv(snakemake@input[[2]], sep = ",")
+#get just the male data
+#XYEpiLC <- subset(EpiLC_TPM, select =  grep("XY", colnames(EpiLC_TPM), value = TRUE))
+#XYEpiLC_SI <- subset(EpiLC_SI, Sex == "XY")
+EpiLC_SI$genosex <- paste0(EpiLC_SI$Sex, EpiLC_SI$Genotype)
+EpiLC_SI$Sample <- colnames(EpiLC_TPM)
+
+
+
+###3) Make a dataframe with the gene info and the ENSEMBL ID
+PGCgenes <- data.frame(ENSEMBL = c("ENSMUSG00000010592","ENSMUSG00000029848", "ENSMUSG00000021758", "ENSMUSG00000046323","ENSMUSG00000025492"), Symbol = c("Dazl", "Stra8",  "Mvh (Ddx4)", "Stella (Dppa3)", "Fragilis (Ifitm3)"))
+#ENSMUSG00000038151 - "Prdm1 (Blimp1)" (not a DEG)
+#ENSMUSG00000005672 - "c-Kit" (only XX DEG)
+
+###4) Function to make boxplot of expression in WT and 5cKO EpiLCs
+library(ggplot2)
+library("ggpubr")
+source('code/utilities/colorpalettes.R') #load the colorpalette
+
+# tpmboxplot <- function(df, gene){
[…]

### 2024-04-19 submission/manuscript.Rmd
editing fig 1 results section

@@
 Gaps in knowledge addressed: 
-* Are other tissue-specific genes dysregulated, or only testis, germline genes?
+* Are other tissue-enriched genes dysregulated, or only testis, germline genes?
 * Curating a robust list of male and female germline genes
@@
 		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
-	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-specific genes
+	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
 		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
@@
 * Loss of KDM5C can result in the misexpression of genes typically only found in the testis
-	* Misexpression of tissue-specific genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
+	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
 	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
@@
 
-## Tissue-specific genes, including testis genes, are enriched within the _Kdm5c_-KO brain
-
-*note: the 5CKO DEGs table we should make a column for if the gene is tissue-specific and which tissue it's in  
-	* see if any liver or muscle genes are dysregulated in both tissues, could add as a bigwig
+## Tissue-enriched genes, including testis genes, are aberrantly expressed in the _Kdm5c_-KO brain
 
@@
 ```
-Previous RNA sequencing studies performed in the adult male _Kdm5c_ knockout (-KO) hippocampus identified aberrant transcription of genes that are typically only expressed in the testis[@scandagliaLossKdm5cCauses2017].
-Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflect widespread impairment of tissue identity.
-To distinguish these possibilities, we globally assessed if tissue-specific genes were significantly enriched in our previous mRNA-seq dataset of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020].
-We classified differentially expressed genes (DEGs) as tissue-specific if they were significantly upregulated in the _Kdm5c_-KO brain compared to wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) and were previously identified as uniquely enriched within a given mouse tissue[@liComprehensiveMouseTranscriptomic2017].
+RNA sequencing (RNA-seq) studies surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017].
+Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity.
+To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]
 
 ```{r, tissue_genes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-#get the number and stats for tissue-specific gene expression
+#get the number and stats for tissue-enriched gene expression
 tissue_res <- read.csv("../data/processed/TissueSpecific_AMYHIP_numberofgenesDEGs.csv", header = TRUE)
@@
 
-* Based on these criteria, we observed widespread misexpression of tissue-enriched genes within the _Kdm5c_-KO amygdala and hippocampus, with the majority of genes belonging to the testis (Figure 1A-C).
+We found many genes significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues, with the majority of differentially expressed genes (DEGs) displaying biased expression towards the testis (Figure 1A-C).
+Even though the testis has the largest number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact).
 
-* Even though the testis has the largest number of tissue-specific genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis genes were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact).
+Even though the _Kdm5c_-KO mice are male, we also observed significant enrichment of ovary-biased DEGs within both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+Intriguingly, many ovary and testis-biased DEGs have functions specific to germ cells and have no known role in the brain.
+For example, the testis-biased DEG _Cytochrome C, testis (Cyct)_ is a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002] (Figure 1C).
+The ovary-enriched DEG _Zygotic arrest 1 (Zar1)_ was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
+Although not consistent across brain regions, we also found enrichment of two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Fisher's Exact) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Fisher's Exact).
+One example of a liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).
+Overall, these tissue-enriched genes show little to no expression in the wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
 
-* This dysregulation of tissue identity is not unique to the testis, as we also observed significant misexpression of ovary, liver, and muscle genes within the _Kdm5c_-KO brain (Figure 1A-E).
-
-Importantly, these tissue-specific genes show little to no expression in the wild-type brain and have no known brain functions, yet our mRNA-seq data indicates they are polyadenylated and spliced into mature transcripts (Figure 1C,G).
-
-* Of note, we did not observe enrichment of brain-specific genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), even though the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
-
-* Together, these results suggest misexpression of testis and other tissue-specific genes within the _Kdm5c_-KO brain is not due to random de-repression of all genes, but rather due to a dysregulation of tissue identity.
+Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), even though the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+Altogether, these results suggest misexpression of testis and other tissue-enriched genes within the _Kdm5c_-KO brain is not due to random de-repression of all genes, but rather due to a dysregulation of tissue identity.
 
@@
 ```
-* One difficulty in globally characterizing the misexpression of germline genes is a comprehensive list of mouse germline-specific genes is currently lacking.
+* One difficulty in globally characterizing the misexpression of germline genes is a comprehensive list of mouse germline-enriched genes is currently lacking.
 
@@
 # Discussion
+* tissue-biased gene expression:
+	* However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function.
+For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
 
@@
   \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
-  \caption[Misexpression of tissue-specific genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Misexpression of tissue-specific genes in the \textit{Kdm5c}-KO in the brain} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala.
+  \caption[Misexpression of tissue-enriched genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Misexpression of tissue-enriched genes in the \textit{Kdm5c}-KO in the brain} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala.
 Left - MA plot of mRNA-sequencing.
@@
   \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
-  \caption[Misexpression of tissue-specific genes, including germline genes, in the \textit{Kdm5c}-KO in the brain]{\textbf{Ovary and testis germline genes are expressed in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis.
+  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis.
 Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM).

### 2024-04-19 code/GO_brain_testisDEGs.R
save GO results as csv

@@ -17,6 +17,7 @@ q <- dotplot(ego, showCategory=15)
 
 
 ggsave(snakemake@output[[1]], plot = q, width = 5.5, height = 6)
+write.table(ego, snakemake@output[[2]], row.names = FALSE, sep = ",")
 
 
 #ego <- enrichGO(de, keyType = 'ENSEMBL', OrgDb = "org.Mm.eg.db", ont="BP", readable=TRUE)

### 2024-04-19 submission/manuscript.Rmd
germline gene section text and GO function

@@
 
-### Male and female germline genes are aberrantly expressed in the male _Kdm5c_-KO brain
-
-Intriguingly, many _Kdm5c_-KO testis-enriched DEGs have functions specific to male germ cells, such as _Cytochrome C, testis (Cyct)_, a (Figure 1C), a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002].
-We therefore wondered if the high enrichment of testis genes within the _Kdm5c_-KO brain reflected a failure to demarcate the soma and germline, as the testis contains both germ cells (e.g.
-spermatogonia) as well as somatic cells (e.g.
-Leydig cells) that support hormone production and germline functions.
-Gene ontology analysis of testis DEGs revealed enrichment of germline-relevant ontologies in the _Kdm5c_-KO amygdala and hippocampus, including spermatid development (GO:0007286) and sperm axoneme assembly (GO:0007288) (Figure 1A).
-
-To validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013].
-We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 1B).
-The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
-We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018].
-We found that while some testis-enriched DEGs were classified as specific markers for multiple stages of germ cell development - including spermatogonia, spermatocytes, round spermatids, and elongating spermatids, none marked somatic cells (Figure 1C).
-Altogether, these data suggest testis genes expressed within the _Kdm5c_-KO brain are actually male germ cell genes.
-
-Interestingly, we also observed significant enrichment of ovary genes within the male _Kdm5c_-KO brain (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact).
-Although there are relatively few ovary-enriched DEGs compared to testis-DEGs (Amygdala: `r gettissval("Amygdala_DEGnumb", "Testis")` testis DEGs, `r gettissval("Amygdala_DEGnumb", "Ovary")` ovary DEGs; Hippocampus: `r gettissval("Hippocampus_DEGnumb", "Testis")` testis DEGs, `r gettissval("Hippocampus_DEGnumb", "Ovary")` ovary DEGs), they also have known germline-specific functions.
-For example, _Zygotic arrest 1 (Zar1)_ was recently shown to regulate oocyte translation by sequestering mRNAs[@chengMammalianOocytesStore2022] (Figure 1D).
-
-* To assess if ovary-enriched DEGs were also germline specific, we assessed their expression in wild-type and germline-depleted (Kit^W/Wv^) female embryos, when female germ cells undergo the initial stages of development and meiosis.
-	* We found XYZ.
-	* Altogether, these results indicate male _Kdm5c_-KO mice express both male and female germline genes within the amygdala and hippocampus.
-
+### Germline genes are aberrantly expressed in the male _Kdm5c_-KO brain
 ```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
@@
 
+#testis gene ontology results
+testisGO <- read.csv("../results/GO_Brain_TestisDEGs.csv")
+
+#function to get the GO ID
+getGO <- function(x, df, column){
+	df_test <- df
+	out <- df_test[df_test$Description == x,][column][,1]
+	return(out)
+}
+
 ```
-* One difficulty in globally characterizing the misexpression of germline genes is a comprehensive list of mouse germline-enriched genes is currently lacking.
 
-* To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating expression in wild-type and germline-depleted (Kit^W/Wv^) male and female mouse embryos from embryonic day 12, 14, and 16.
-	* **note: need to figure out why these time points. We have P6 and adult for males but not females so it would be biased towards males if we included the later ages**
+The testis contains both germ cells (e.g.
+spermatogonia) and somatic cells (e.g.
+Leydig cells) that support hormone production and germline functions.
+We then wondered if the high enrichment of testis genes in the _Kdm5c_-KO brain reflected a failure to demarcate between the soma and germline.
+Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID')`, q = `r signif(getGO("spermatid development", testisGO, 'p.adjust'), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID')`, q = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust'), 3)`) (Figure 2A).
+
+To further validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013].
+We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B).
+The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
+We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018].
+We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
+spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C).
+Together, these data suggest the _Kdm5c_-KO brain expresses germline genes.
+
+We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes.
+To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice.
+We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes.
 
@@
 These criteria yielded `r nrow(germgenes)` germline-enriched genes.
-To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female (XX) or male germline.
+To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female or male germline.
 We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` XX-biased, `r nrow(subset(germgenes, sexBias == "XY"))` XY-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
 

### 2024-04-20 submission/manuscript.Rmd
draft of figure 3 results

@@
 #function to get the GO ID
-getGO <- function(x, df, column){
+getGO <- function(x, df, column, cluster){
 	df_test <- df
+
+	if (cluster == "no") {
 		out <- df_test[df_test$Description == x,][column][,1]
+	} else {
+		out <- df_test[df_test$Description == x & df_test$Cluster == cluster,][column][,1]
+	}
 	return(out)
@@
 We then wondered if the high enrichment of testis genes in the _Kdm5c_-KO brain reflected a failure to demarcate between the soma and germline.
-Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID')`, q = `r signif(getGO("spermatid development", testisGO, 'p.adjust'), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID')`, q = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust'), 3)`) (Figure 2A).
+Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
@@
 We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes.
-
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
 These criteria yielded `r nrow(germgenes)` germline-enriched genes.
-To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female or male germline.
-We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
 
-## _Kdm5c_-KO cells express master regulators of germline identity in an _in vitro_ model of early embryogenesis
+## _Kdm5c_-KO epiblast-like cells aberrantly express master regulators of germline identity
+
+```{r, epilcs, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+
+#testis gene ontology results
+epigerm <- read.csv("../results/DESeq2/germDEGs/germDEGs_EpiLC_XY5cKO.csv")
+GO_EpiBrain <- read.csv("../results/GO_EpiLC_vs_Brain.csv")
+
+epi_restab <- read.csv("../data/processed/restable_EpiLC_XY5cKO.csv", row.names = 1)
+epi_restab$ENSEMBL <- rownames(epi_restab)
+
+zscan4d <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]
+
+```
 
@@
 Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-Therefore, to elucidate the mechanism underlying aberrant germline genes expression in the mature _Kdm5c_-KO brain, we first assessed their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].
+Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by assessing their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].
 
-* We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom).
-Additionally, _Kdm5c_-KO EpiLCs downregulated ESC pluripotency genes and upregulated EpiLC differentiation genes similar to wild-type (Figure 3B).
+We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B).
+We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).
 
-* We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis (Figure 3C).
+Using our curated list, we found `r nrow(epigerm)` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D).
+To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E).
 
-* Of these testis genes, we also observed misexpression of germline-enriched genes, including _Cyct_ (Figure 3D)
+We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
+These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo.
+However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`).
 
-* To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
-The majority of EpiLC germline DEGs were unique to EpiLCs, with only a few shared across brain sequencing datasets (Figure 3E).
-Only two germline genes were consistently misexpressed in all datasets, _DNA segment, Chr 1, Pasteur Institute 1 (D1Pas1)_ and _Testis expressed gene 14 (Tex14)_.
-
-* We then compared the typical function of brain and EpiLC germline DEGs and found EpiLC germline DEGs displayed strong enrichment of meiosis-related gene ontologies, including XXXXX (Figure 3F).
-
-* Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
-These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but they are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
-
-* _Dazl_ is a translational regulator essential for germ cell function[@liDAZLMasterTranslational2019].
-Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (XY _Kdm5c_-KO EpiLCs p = 1.698e-07, Fisher’s Exact Test).
+_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019].
+Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
 This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function.
-
-* Therefore, we then stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H).
-We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test, XY KO versus WT), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
+We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H).
+We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
+Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.
 

### 2024-04-22 submission/manuscript.Rmd
update figure legends

@@
 
-We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
+We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
 These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
@@
 Left - MA plot of mRNA-sequencing.
-Right - The number of tissue-enriched differentially expressed genes (DEGs).
+Right - Number of tissue-enriched differentially expressed genes (DEGs).
 * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{B.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO hippocampus.
-\textbf{C.} Left - Bigwigs of an example aberrantly expressed testis-enriched gene, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
+\textbf{C.} Left - Bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
 Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
-\textbf{D.} Left - Bigwigs of an example ovary-enriched germline DEG, \textit{Zygotic arrest 1 (Zar1)}.
-Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.}
+\textbf{D.} Left - Bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
+Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
+\textbf{E.} Left - Bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
+Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
   \label{figurelabel}
 \end{figure}
+
+__
 
@@
 Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM).
-\textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis.
+\textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis (Green et al 2018).
 Germline cell types are highlighted in green, somatic cell types in black.
-\textbf{D.} Expression of ovary-enriched DEGs in a wild-type versus germ cell-depleted testis }
+\textbf{D.} Snakey diagram of genes filtered for germline enrichment.}
   \label{figurelabel}
@@
   \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
-  \caption[Male and female \textit{Kdm5c} mutant epiblast-like cells express master regulators of germline identity ]{\textbf{Male and female \textit{Kdm5c} mutant epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of emrbyonic stem cells (ESCs) of the inner cell mass into epiblast stem cells and primordial germ cells.
-Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
-\textbf{B.} Left - Bigwigs of an example germline gene, \textit{Cyct} that is dysregulated in male and female \textit{Kdm5c} mutant EpiLCs.
-Right - Overlap of all germline DEGs between male and female EpiLCs.
-\textbf{C.} Chromosome location of germline genes misexpressed in male and female EpiLCs.
-\textbf{D.} Number of EpiLC germline DEGs that show sex-biased expression in the wild-type gonads.
-\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in male and female \textit{Kdm5c} mutant EpiLC and brain RNA-seq datasets.
-\textbf{F.} Expression of genes key for early germ cell development in \textit{Kdm5c} mutant EpiLCs in transcrips per million (TPM) with p-values for Welch's t-test compared to wild type of the same sex.
-\textbf{G.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.}
+  \caption[\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of emrbyonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
+Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
+Bottom - representative images of wild-type (WT) and \textit{Kdm5c}-KO ESC to EpiLC differentiation.
+Brightfield images taken at 20X.
+\textbf{B.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs.
+Welch's t-test, expression in transcripts per million (TPM).
+\textbf{C.} Number of tissue-enriched differentially expressed genes (DEGs).
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
+\textbf{D.} Bigwigs of an example germline gene, \textit{Cyct} that is dysregulated \textit{Kdm5c}-KO EpiLCs.
+\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO mutant EpiLC, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
+\textbf{F.} enrichPlot comparing enriched gene ontoogies for \textit{Kdm5c}-KO EpiLCs, amygdala, and hippocampus germline DEGs.
+\textbf{G.} Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test.
+\textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
+Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test}
   \label{figurelabel}

### 2024-04-22 code/utilities/mouseGermlineGenes.R
draft filtering list, split off germ genes better

@@ -57,7 +57,7 @@ mueller_WT <- merge(mueller_WT, adultFPKM_WT, by = "SYMBOL")
 mueller_WT <- mueller_WT[,c("SYMBOL", "ENSEMBL", "E12XXWT" ,"E12XYWT", "E14XXWT" , "E14XYWT" , "E16XXWT" , "E16XYWT", "P6XYWT", 'AdultXYWT')]
 mueller_WT <- unique(mueller_WT)
 
-#print out number of genes with all timepoints that we can use for analysis 
+#print out number of genes with all time points that we can use for analysis 
     #19160 genes
 cat("Number of genes for analysis: ", nrow(mueller_WT), "\n")
 
@@ -226,3 +226,49 @@ write.table(germGENES20, file = snakemake@output[[1]], sep =",", row.names = FAL
 ## make a snakey graph of filtering for germline-enriched genes
 #https://github.com/davidsjoberg/ggsankey
 #devtools::install_github("davidsjoberg/ggsankey")
+#tutorial: https://rpubs.com/techanswers88/sankey-with-own-data-in-ggplot
+
+library(ggsankey)
+library(ggplot2)
+library(dplyr)
+
+
+
+#1) for every gene, put which category it falls into for each filtering step
+	#categories: mouse genes (all mouse genes), w/wv, tissue
+	#wwv - if it's in the mueller ratio20 list, it lost 80% of its expression in wwv
+	#tissue - if it's expressed in the non-ovary and testis tissues, then it's eliminated. 
+
+uniqueENSEMBL <- unique(mueller_WT$ENSEMBL) 
+
+sankeydf <- data.frame(ENSEMBL = uniqueENSEMBL, genes = rep("mouse gene", length(uniqueENSEMBL)), wwv = ifelse(uniqueENSEMBL %in% muellerRATIO20$ENSEMBL, "yes", "no"), tissue = ifelse(uniqueENSEMBL %in% liRATIO20$ENSEMBL, "gonadal", "non-gonadal"))
+
+row.names(sankeydf) <- sankeydf$ENSEMBL
+sankeydf <- subset(sankeydf, select = -ENSEMBL)
+
+print(head(sankeydf))
+
+##2) make the dataframe for plotting, including how many observations there are for each category
+df <- sankeydf %>%
+  make_long(genes, wwv, tissue)
+
+# count how many there are in each group and merge back with the long df
+dagg <- df%>%
+  dplyr::group_by(node)%>%
[…]

### 2024-04-22 code/utilities/mouseGermlineGenes.R
add sankey diagram of filtering germline genes

@@ -239,9 +239,11 @@ library(dplyr)
 	#wwv - if it's in the mueller ratio20 list, it lost 80% of its expression in wwv
 	#tissue - if it's expressed in the non-ovary and testis tissues, then it's eliminated. 
 
+print(head(germGENES20))
+
 uniqueENSEMBL <- unique(mueller_WT$ENSEMBL) 
 
-sankeydf <- data.frame(ENSEMBL = uniqueENSEMBL, genes = rep("mouse gene", length(uniqueENSEMBL)), wwv = ifelse(uniqueENSEMBL %in% muellerRATIO20$ENSEMBL, "yes", "no"), tissue = ifelse(uniqueENSEMBL %in% liRATIO20$ENSEMBL, "gonadal", "non-gonadal"))
+sankeydf <- data.frame(ENSEMBL = uniqueENSEMBL, Total = rep("mouse genes", length(uniqueENSEMBL)), Wwv = ifelse(uniqueENSEMBL %in% muellerRATIO20$ENSEMBL, "yes", "no"), Tissue = ifelse(uniqueENSEMBL %in% germGENES20$ENSEMBL, "germline-enriched", "somatic"))
 
 row.names(sankeydf) <- sankeydf$ENSEMBL
 sankeydf <- subset(sankeydf, select = -ENSEMBL)
@@ -250,7 +252,7 @@ print(head(sankeydf))
 
 ##2) make the dataframe for plotting, including how many observations there are for each category
 df <- sankeydf %>%
-  make_long(genes, wwv, tissue)
+  make_long(Total, Wwv, Tissue)
 
 # count how many there are in each group and merge back with the long df
 dagg <- df%>%
@@ -260,15 +262,28 @@ dagg <- df%>%
 df2 <- merge(df, dagg, by.x = 'node', by.y = 'node', all.x = TRUE)
 
 # plot the dataframe
-pl <- ggplot(df2, aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = factor(node), label = paste0(node," n=", n)))
 
-pl <- pl +geom_sankey(flow.alpha = 0.5,  color = "gray40", show.legend = TRUE)
-pl <- pl +geom_sankey_label(size = 3, color = "black", fill= "white", hjust = -0.2)
+pl <- ggplot(df2, aes(x = x, next_x = next_x, node = node, next_node = next_node, fill = factor(node), label = paste0(node," n=", n))) +
+  geom_sankey(flow.alpha = .6,
+              node.color = "gray30") +
[…]

### 2024-04-22 submission/manuscript.Rmd
update figure legends

@@
 * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{B.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO hippocampus.
-\textbf{C.} Left - Bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
+\textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
 Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
-\textbf{D.} Left - Bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
+\textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
 Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
-\textbf{E.} Left - Bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
+\textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
 Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
@@
 Germline cell types are highlighted in green, somatic cell types in black.
-\textbf{D.} Snakey diagram of genes filtered for germline enrichment.}
+\textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and germline-depleted mice and in adult mouse non-gonadal tissues (Li et al 2017).}
   \label{figurelabel}
@@
 * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
-\textbf{D.} Bigwigs of an example germline gene, \textit{Cyct} that is dysregulated \textit{Kdm5c}-KO EpiLCs.
+\textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct} that is dysregulated \textit{Kdm5c}-KO EpiLCs.
 \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO mutant EpiLC, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
 \textbf{F.} enrichPlot comparing enriched gene ontoogies for \textit{Kdm5c}-KO EpiLCs, amygdala, and hippocampus germline DEGs.
-\textbf{G.} Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test.
-\textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
+\textbf{G.} Top - Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test.
+Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
 Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test}

### 2024-04-23 submission/manuscript.Rmd
fix figure legend typo

@@
   \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
-  \caption[\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of emrbyonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
+  \caption[\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
 Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).

### 2024-04-23 submission/manuscript.Rmd
update figure 1 title

@@
   \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
-  \caption[Misexpression of tissue-enriched genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Misexpression of tissue-enriched genes in the \textit{Kdm5c}-KO in the brain} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala.
+  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala.
 Left - MA plot of mRNA-sequencing.
@@
 * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
-\textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct} that is dysregulated \textit{Kdm5c}-KO EpiLCs.
+\textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs.
 \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO mutant EpiLC, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.

### 2024-04-24 submission/manuscript.Rmd
draft of fig 4 results

@@
 Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
-This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function.
+This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_-KO cellular function.
 We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H).
@@
 Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.
+
+### KDM5C-mediated removal of H3K4me3 during early embryogenesis promotes long-term repression of germline genes
+
+**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
+	A) KDM5C binding in EpiLCs vs PNCs - number of peaks and percentage of bound genomic regions
+	B) KDM5C binding in Epilcs vs pncs germline genes - GO left, bigwigs right
+	C) Bigwigs of Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
+	D) Increase in H3K4me2 in EpiLCs at germline genes (left - bigwigs, right average signal at germline gene TSS? Just KDM5C-bound genes?)
+	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). 
+	E) KDM5C expression ESC --> EpiLC (increasing then decreasing) Left - RNA right - protein 
+	F) RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
+	G) DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)
+
+Previous work has suggested KDM5C represses germline genes during early development, as re-expression of KDM5C in neuronal cultures fails to suppress their transcription[@scandagliaLossKdm5cCauses2017].
+While KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+However, it is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes.
+
+* To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
+We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: 2,437, PNCs: 1,610, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ).
+
+* Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130).
+Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes are enriched for non-germline-specific functions such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).
+
+* In EpiLCs but not PNCs, KDM5C is bound to the TSS of germline genes that are mixexpressed in both _Kdm5c_-KO EpiLCs and the mature brain, such as _Cyct_ and _D1Pas1_.
+
+* Interestingly, KDM5C is also bound to the germline master regulator _Dazl_ in EpiLCs, but is not bound to other germline regulators like _Stra8_.
+
+* Note: examples of brain germline DEGs that kdm5c is not bound to? That also shows their dysregualtion can be caused by secondary effects.
+
+* Altogether, this indicates that while the misexpression of some germline genes is due to loss of KDM5C binding during emryogenesis, germline genes can also be aberrantly expressed by secondary mechanisms downstream of KDM5C.
+
+* Altogether, this indicates the misexpression of germline genes in _Kdm5c_-KO cells is partly due to direct loss of KDM5C binding during emryogenesis, but also caused by secondary mechanisms activation by secondary mechanisms.
+
+----
+* This indicates long-term misexpression of germline genes is due to developmental loss of KDM5C suppression.
+
+* In EpiLCs, KDM5C was bound to germline genes transcribed in _Kdm5c_-KO EpiLCs as well as those dysregulated in the brain, such as _D1Pas1_ (Figure 4C).
+
+ To further delineate KDM5C's role in suppressing germline identity, we assessed the function of all genes with KDM5C bound to their promoter in EpiLCs and PNCs through gene ontology (GO) analysis (Figure 4D).
+
+Peaks were called using MACS2 software (v 2.1.0.20140616) using input BAM files for normalization, with filters for a q-value
+< 0.1 and a fold enrichment >1. Common peak sets were obtained via DiffBind, and count
+tables for the common peaks were generated with the Bedtools multicov command.
+
+ our previously generated chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) data of KDM5C in
+
+* To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed our previously generated chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) data of KDM5C's substrates, H3K4me2/3[@vallianatosMutuallySuppressiveRoles2020;@samantaActivationXistEvolutionarily2022].
+
+* In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed ectopic H3K4me3 peaks in the _Kdm5c_-KO amygdala at the transcription start site (TSS) of germline genes (Figure 4A).
+* We additionally observed a marked increase in H3K4me2 at germline gene promoters in _Kdm5c_-KO EpiLCs.
 
@@
 \textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs.
-\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO mutant EpiLC, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
-\textbf{F.} enrichPlot comparing enriched gene ontoogies for \textit{Kdm5c}-KO EpiLCs, amygdala, and hippocampus germline DEGs.
+\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
+\textbf{F.} enrichPlot comparing enriched gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
 \textbf{G.} Top - Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test.

### 2024-04-24 submission/manuscript.Rmd
drafting fig 4 results - working on chipseq set up

@@
 
-Previous work has suggested KDM5C represses germline genes during early development, as re-expression of KDM5C in neuronal cultures fails to suppress their transcription[@scandagliaLossKdm5cCauses2017].
-While KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
-However, it is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes.
+Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription.
+
+* However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including master regulators of meiosis and germline identity.
+
+* KDM5C was very recently shown to bind to Dazl in ESCs, unclear if it's bound to other master regulators
+
+* Unclear if KDM5C's binding is maintained throughout development for these master regulators or also lost with differentiation, as KDM5C's binding at germline gene promoters has not been systematically compared.
+
+**check ESC chipseq, did they describe germline genes?**
+
+* Since we found unique genes are dysregulated in EpiLCs
+
+* However, it is currently unclear if KDM5C
+
+ * previous studies only focused on binding at germline DEGs in the brain or focused on Dazl expression, no systematic characterization of KDM5C binding
 
@@
 
-* Note: examples of brain germline DEGs that kdm5c is not bound to? That also shows their dysregualtion can be caused by secondary effects.
+* Altogether, these data indicate that while some germline genes are dysregulated due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
 
-* Altogether, this indicates that while the misexpression of some germline genes is due to loss of KDM5C binding during emryogenesis, germline genes can also be aberrantly expressed by secondary mechanisms downstream of KDM5C.
-
-* Altogether, this indicates the misexpression of germline genes in _Kdm5c_-KO cells is partly due to direct loss of KDM5C binding during emryogenesis, but also caused by secondary mechanisms activation by secondary mechanisms.
+* **Note: examples of brain germline DEGs that kdm5c is not bound to? That also shows their dysregualtion can be caused by secondary effects.**
 
 ----
-* This indicates long-term misexpression of germline genes is due to developmental loss of KDM5C suppression.
 
-* In EpiLCs, KDM5C was bound to germline genes transcribed in _Kdm5c_-KO EpiLCs as well as those dysregulated in the brain, such as _D1Pas1_ (Figure 4C).
-
- To further delineate KDM5C's role in suppressing germline identity, we assessed the function of all genes with KDM5C bound to their promoter in EpiLCs and PNCs through gene ontology (GO) analysis (Figure 4D).
-
-Peaks were called using MACS2 software (v 2.1.0.20140616) using input BAM files for normalization, with filters for a q-value
-< 0.1 and a fold enrichment >1. Common peak sets were obtained via DiffBind, and count
-tables for the common peaks were generated with the Bedtools multicov command.
-
- our previously generated chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) data of KDM5C in
-
-* To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed our previously generated chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) data of KDM5C's substrates, H3K4me2/3[@vallianatosMutuallySuppressiveRoles2020;@samantaActivationXistEvolutionarily2022].
+Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+It is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes, such as promoting the placement of DNA methylation that occurs in EpiLCs.
 

### 2024-04-24 submission/manuscript.Rmd
drafting fig 4 results

--- 2024-04-23 submission/manuscript.Rmd
@@
 Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.
+
+### KDM5C-mediated removal of H3K4me3 during early embryogenesis promotes long-term repression of germline genes
+
+**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
+	A) KDM5C binding in EpiLCs vs PNCs - number of peaks and percentage of bound genomic regions
+	B) KDM5C binding in Epilcs vs pncs germline genes - GO 
+	C) Left - bar graph of germline DEGs KDM5C is bound to in EpiLCs Right - Example bigwigs
+		* Brain and EpiLC shared, EpiLC-specific, Brain-specific, 
+	D) Bigwigs of Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
+	E) Increase in H3K4me2 in EpiLCs at germline genes (left - bigwigs, right average signal at germline gene TSS? Just KDM5C-bound genes?)
+	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). 
+	F) KDM5C expression ESC --> EpiLC (increasing then decreasing) Left - RNA right - protein 
+	G) RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
+	H) DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)
+
+Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription.
+However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including master regulators of meiosis and germline identity.
+While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other master regulators that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.
+
+To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
+We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: 2,437, PNCs: 1,610, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A).
+Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B).
+Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).
+
+We then assessed KDM5C binding at germline genes expressed _Kdm5c_-KO cells to assess if their dysreuglation is due to direct loss of KDM5C binding during embryogenesis.
+
+	* KDM5C was bound to the promoter of many genes dysregulated in both EpiLCs and brain datasets, including _D1Pas1_ and _Cyct_.
+	* KDM5C was not bound to many of the brain-specific germline DEGs, such as XXX.
+	* While KDM5C binds to many germline DEGs in EpiLCs, including _Dazl_, many were also unbound. This included the germline transcription factor _Stra8_.
+	* All genes lacked binding in PNCs.
+	* Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
+
+----
+
+Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+It is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes, such as enabling the placement of DNA methylation.
+
+* In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed ectopic H3K4me3 peaks in the _Kdm5c_-KO amygdala at the transcription start site (TSS) of germline genes (Figure 4).
+* We additionally observed a marked increase in H3K4me2 at germline gene promoters in _Kdm5c_-KO EpiLCs.
+	* This increase was greatest at germline genes directly bound by KDM5C (?)
 

### 2024-05-20 submission/manuscript.Rmd
update figure outlines

@@
 	B) KDM5C binding in Epilcs vs pncs germline genes - GO 
-	C) Left - bar graph of germline DEGs KDM5C is bound to in EpiLCs Right - Example bigwigs
+	C) Left - bar graph of germline DEGs KDM5C is bound to in EpiLCs Right 
+	D) Example bigwigs
 		* Brain and EpiLC shared, EpiLC-specific, Brain-specific, 
+	E) motif analysis of KDM5C-bound vs unbound genes
+
+	--- separate into two figures
 	D) Bigwigs of Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
@@
 	H) DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)
+
+	Direct vs indirect binding motif analysis
 
@@
 For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
+* Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
 

### 2024-05-20 submission/manuscript.Rmd
add commas to large numbers

@@
 ---
+```{r, knitrcommas, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+library(huxtable)
+
+knitr::knit_hooks$set(inline = function(x) {
+  prettyNum(x, big.mark=",")
+})
+
+```
 
@@
 
-Using our curated list, we found `r nrow(epigerm)` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D).
+Using our curated list, we found `r format(nrow(epigerm))` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D).
 To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.

### 2024-05-20 submission/manuscript.Rmd
add commas to large numbers

@@
 ```{r, knitrcommas, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+#adds commas to large numbers
 library(huxtable)

### 2024-05-30 submission/manuscript.Rmd
update text

@@
 source("../code/utilities/parameters.R")
+```
 
-```
 RNA sequencing (RNA-seq) studies surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017].
@@
 To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice.
-We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes.
+We assessed gene expression in male and female mice at embryonic day 12, 14, and 16, as well as male postnatal day 6 and adult testes.
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.

### 2024-06-29 submission/manuscript.Rmd
drafting intro

@@
 ---
-title:  Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
+title:  Somatic misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
 output: pdf_document
@@
    - \pagenumbering{arabic}
+   - \usepackage{indentfirst}
 
@@
 		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
+
+# Introduction
+
+To form a complete organism, embryonic stem cells must differentiate into a myriad of discrete cellular identities.
+This is in part accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications.
+Although initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing technologies unexpectedly found many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators.
+This relationship is partially explained by their regulation of brain-specific genes or chromatin states, such as modulating genes involved in synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuronal and glial developmental programs[@hirabayashiPolycombLimitsNeurogenic2009].
+However, loss of some chromatin regulators can also lead to the misexpression of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
+Currently, very few studies have explored the misexpression of non-neuronal, tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
+To elucidate their contribution to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression.
+Characterizing these features will provide the basis for identifying molecular footholds that can then be exploited for future studies and potential therapeutics.
+
+In this study, we characterized the misexpression of tissue-specific with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
+Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
+Previous work has demonstrated _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016].
+Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017].
+However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to male germline genes, as other tissue-specific genes have not been systematically evaluated.
+
+Interestingly, some of the ectopic testis genes identified in the _Kdm5c_-KO brain are typically expressed in germ cells, such as _Cytochrome C, Testis-specific_, a sperm mitochondrial gene[cit].
+
+- Germ cells are meiotic cells that pass on their genetic material to the next generation and are typically distinguished from somatic cells very early on in embryogenesis. 
+- The demarcation of the germ vs soma is a key feature of multicellularity
+
+* Germ cells, i.e. cells that undergo meiosis, are typically distinguished from somatic cells very early on in embryogenesis and this distinction is a key feature of multicellularity.
+
+* However, the testis also contains somatic cells that support germline function, and it is currently unclear if  It is currently unclear if all testis- (maybe too specific)
+
+	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
+		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
+		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
+		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
+
+Furthermore, the testis contains both somatic and germline
+
+It is currently unclear if these germline genes
+
+	* Systematic characterization of ectopic germline genes hasn't been done
+		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
+		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
+
+However, it is
+
+Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
+
+	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
+
+	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
+	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
+	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
+		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
+		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
+		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
+
+Answering these questions will help identify molecular footholds that can be exploited to test their overall contribution to neuronal impairments that will lay the ground work for therapeutic intervention.
+
+Loss of KDM5C can result in the misexpression of genes typically only found in the testis
+	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
+	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
+	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
+	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
+		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
+		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
+		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
+	* Systematic characterization of ectopic germline genes hasn't been done
+		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
+		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
+
+To determine the
+
+- To ultimately determine their contribution, it is essential to characterize the types of genes expressed and the mechanism behind their de-repression.
+
+	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention
+
+	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
+		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
+	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention
+
+- For example, histone 3 lysine 4 methylation (H3K4me)
+
+- [@lugerCrystalStructureNucleosome1997].
+
+- This relationship is partially 
+It is currently unc
+
+- There is a strong link between H3K4me regulation and NDDs, as mutation of 11 out of 15 genes that read, write, or erase H3K4 mono-, di-, and tri- methylation result in varying neurodevelopmental impairments and syndromic NDDs[@vallianatosDisruptedIntricacyHistone2015].
+Yet it is still unclear why regulation of H3K4 methylation is so crucial for neurodevelopment, as most of H3K4 methylation regulators are ubiquitously expressed across tissues, cell types, and lifespan.
+
+With the recent advancement of next generation sequencing technologies came the unexpected finding that many neurodevelopmental disorders are caused by or linked to mutations in genes that regulate chromatin[@derubeisSynapticTranscriptionalChromatin2014; @gabrieleChromatinBasisNeurodevelopmental2018; @iwaseEpigeneticEtiologyIntellectual2017a].
+
+Chromatin regulators, including those that  DNA and histone modification, is an integral component of cellular identity
+
+Chromatin regulators, including those that  DNA and histone modification, is an integral component of cellular identity
+
+Cellular identity 
+Some of the first chromatin regulators , such as
+
+With the recent advancement of next generation sequencing technologies came the unexpected finding that many neurodevelopmental disorders are caused by or linked to mutations in genes that regulate chromatin[@derubeisSynapticTranscriptionalChromatin2014; @gabrieleChromatinBasisNeurodevelopmental2018; @iwaseEpigeneticEtiologyIntellectual2017a].
+
+The basic subunit of chromatin is the nucleosome and is comprised of DNA wrapped around octamers of histone proteins[@lugerCrystalStructureNucleosome1997].
+
+Histone proteins can undergo covalent chemical modifications on their N-terminal tails that can in turn influence cofactor recruitment, DNA accessibility, and gene expression[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
 

### 2024-06-29 submission/manuscript.Rmd
intro draft

--- 2024-05-30 submission/manuscript.Rmd
@@
 ---
-title:  Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C
+title:  Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
 output: pdf_document
@@
 
+# Introduction
+
+To form a complete organism, embryonic stem cells must differentiate into a myriad of discrete cellular identities.
+This is in part accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
+Although initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing technologies unexpectedly found many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators.
+This relationship is partially explained by their regulation of brain-specific genes or chromatin states, such as modulating genes involved in synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuronal and glial developmental programs[@hirabayashiPolycombLimitsNeurogenic2009].
+However, loss of some chromatin regulators can also lead to the misexpression of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
+Currently, very few studies have explored the misexpression of non-neuronal, tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
+To elucidate their contribution to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression.
+Characterizing these features will enable us to identify molecular footholds common between NDDs that can then be exploited for potential therapeutics.
+
+In this study, we characterized the misexpression of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
+Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
+Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016].
+Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017].
+However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.
+
+Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017].
+Unlike somatic cells, germ cells (e.g.
+sperm and eggs) undergo meiosis and pass on their genetic material to the next generation.
+The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline.
+Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017].
+In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development.
+However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.
+
+To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+We identified widespread dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes.
+The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells.
+To better characterize germline gene misexpression, we then generated a dataset of germline-enriched genes by comparing gene expression in gonads with germ cell depletion.
+We found _Kdm5c_-KO EpiLCs primarily expressed unique germline genes compared to the mature _Kdm5c_-KO brain, including _Dazl_ and _Stra8_, key drivers of germline identity and meiosis.
+While KDM5C is directly bound to some germline gene promoters in EpiLCs, it is not directly bound to many germline-enriched mRNAs expressed with _Kdm5c_-KO cells, indicating germline genes can be aberrantly transcribed through indirect mechanisms.
+Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs.
+Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.
+
+**note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**
+
 # Results
@@
 ## Tissue-enriched genes, including testis genes, are aberrantly expressed in the _Kdm5c_-KO brain
+
+- **note: should I compare amygdala and hippocampus? scandaglia only looked at hippocampus**
 
@@
 
-RNA sequencing (RNA-seq) studies surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017].
+A previous RNA sequencing (RNA-seq) study surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017].
 Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity.
@@
 # Discussion
-* tissue-biased gene expression:
-	* However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function.
+- The demarcation of the germ vs soma is a key feature of multicellularity
+- Other H3K4me regulators, anything known about tissue-biased gene expression?
+- tissue-biased gene expression:
+	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function.
 For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
-* Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
+- Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
 

### 2024-06-29 code/KDM5C_ESCtoEpiLC.R
KDM5C western

@@ -1,0 +1,45 @@
+
+WesternQuant <- function(csvfile, expprotein, housekeep){
+    data <- read.csv(file = csvfile, sep = ",", header = T)
+    colnames(data) <-c("Channel","Signal","Genotype","Time","Area","Mean","Min","Max")
+    head(data)
+
+    #subtract background from signal for normalized value
+    data_green_sig <- subset(data, Channel == "green" & Signal == expprotein)
+    data_green_bg <- subset(data, Channel == "green" & Signal == "background")
+    data_green <- data.frame(data_green_sig, Normalized = data_green_sig$Mean - data_green_bg$Mean)
+
+    data_red_sig <- subset(data, Channel == "red" & Signal == housekeep)
+    data_red_bg <- subset(data, Channel == "red" & Signal == "background")
+    data_red <- data.frame(data_red_sig, Normalized = data_red_sig$Mean - data_red_bg$Mean)
+
+    Loading <- data_green$Normalized/data_red$Normalized
+	data_green[data_green == "5cKO"] <- "5CKO"
+    df <- data.frame(data_green, Loading, GenoTime = paste(data_green$Genotype, data_green$Time) )
+
+    #set the order
+    df$GenoTime <- factor(df$GenoTime, levels = c("WT 0", "WT 24", "WT 48","5CKO 0", "5CKO 24", "5CKO 48"))
+
+    library(ggpubr)
+    #check the dataframe looks correct
+    print(df)
+	nrow(df)
+
+    my_comparisons <- list(c("WT 0", "WT 24"), c("WT 24", "WT 48"))
+    pl <- ggboxplot(df, x = 'GenoTime', y = 'Loading', color = "black", fill="GenoTime", 
+        title = "KDM5C protein expression", 
+        #subtitle = "ESC to EpiLC Differentiation",
+        add = "dotplot", add.params = list(size = 1.5),
+        xlab = " ", ylab = "Normalized KDM5C/DAXX Intensity [au]",
+		ylim = c(0, 2),
+        palette = c("WT 0" = "#f2b722ff", "WT 24" = "#ef8e3eff", "WT 48" = "#b26c93ff","5CKO 0" = "#f2b722ff", "5CKO 24" = "#ef8e3eff", "5CKO 48" = "#b26c93ff")) +
+        stat_compare_means(comparisons = my_comparisons, method="t.test", label = "p.signif", size = 6) +
[…]

### 2024-06-29 code/KDM5C_ESCtoEpiLC.R
KDM5C mRNA expression esc to epiLC qpcr

@@ -1,3 +1,4 @@
+#24.06.24 KDM5C western ESC to EpiLC expression
 
 WesternQuant <- function(csvfile, expprotein, housekeep){
     data <- read.csv(file = csvfile, sep = ",", header = T)
@@ -43,3 +44,158 @@ WesternQuant <- function(csvfile, expprotein, housekeep){
 }
 
 WesternQuant(snakemake@input[[1]], snakemake@params[["expprotein"]], snakemake@params[["housekeep"]])
+
+
+#24.06.29 KDM5C RNA ESC to EpiLC expression
+
+Kdm5c_qpcr <- read.csv(snakemake@input[[2]], sep =",", header = T)
+colnames(Kdm5c_qpcr) <- c("Sample", "Genotype", "Timepoint", "Primer","CT", "Replicate")
+head(Kdm5c_qpcr)
+
+
+#get the average CT value for each technical replicate
+avg_CT <- function(df, samp, primey){
+  reps <- subset(df, Sample == samp & Primer == primey)
+  avg = (reps[1,5]+reps[2,5])/2
+  return(avg)
+}
+
+#difference between CT values for replicates to see if there are any major deviations
+diff_CT <- function(df, samp, primey){
+  reps <- subset(df, Sample == samp & Primer == primey)
+  diff = abs(reps[1,5]-reps[2,5])
+  return(diff)
+}
+
+#get the information that goes with the average (sample, timepoint, primer etc)
+info_CT <- function(df, samp, primey){
+  reps <- subset(df, Sample == samp & Primer == primey)
+  rowsamp = reps[1,1:4]
+  return(rowsamp)
+}
+
+#checking to see if it works
+info_CT(Kdm5c_qpcr, "0-2a", "TBP")
+
+
+#list of all primers and samples
+primerlist <- unique(Kdm5c_qpcr$Primer)
+samplelist <- unique(Kdm5c_qpcr$Sample)
+
+cnt <- 1
+
+#go through each primer and sample and average the replicates together. Store them in a list and dataframe
[…]

### 2024-06-30 code/H3K4_GermlineTSS.R
Fix order to bed files for - strand genes

@@ -54,6 +54,12 @@ library(dplyr)
 library(tidyr)
 print(head(gtf_df))
 gtf_df <- subset(gtf_df, type == "gene")
+
+#need to switch the start and end if the gene is on the minus strand
+gtf_df$TSS <- ifelse(gtf_df$strand == "+", gtf_df$start, gtf_df$end)
+gtf_df$TES <- ifelse(gtf_df$strand == "+", gtf_df$end, gtf_df$start)
+
+
 gtf_df <- pivot_longer(gtf_df, cols =gene_id) %>%
  	separate(value, into = c('ENSEMBL', 'version'), sep = "\\.") %>%
   	select(-name)
@@ -69,7 +75,7 @@ geneTSSandTES <- function(goi, n){
 
 	#regions of interest
 	#goi <- genedf$ENSEMBL
-	roi <- subset(gtf_df, ENSEMBL %in% goi)[,1:3]
+	roi <- subset(gtf_df, ENSEMBL %in% goi, select = c(seqnames, TSS, TES))
 	print('regions of interest')
 	print(tail(roi))
 

### 2024-07-01 submission/manuscript.Rmd
shigeki results edits and ChIPseq line count

@@
 Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016].
-Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017].
+Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017].
 However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.
@@
 To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We identified widespread dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes.
+We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes.
 The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells.
@@
 
-A previous RNA sequencing (RNA-seq) study surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017].
-Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity.
-To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]
+Previous RNA sequencing (RNA-seq) in the adult _Kdm5c_ knockout (-KO) hippocampus identified a few transcripts that are typically expressed in the testis[@scandagliaLossKdm5cCauses2017].
+However, the expression of tissue-specific genes within the _Kdm5c_-KO brain has not systematically characterized.
+Currently, it is unclear if other tissue-specific transcripts are also expressed within the _Kdm5c_-KO brain and if misexpression of testis-enriched genes are a major consequence of KDM5C loss, given the high abundance of testis-enriched genes within the mouse transcriptome.
+Therefore, we first globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]
+
+**note: the idea is there but need to work on wording. The extent to which brain tissue identity is altered is unknown and if testis genes are a major effect**
 
@@
 
-To further validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013].
+To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in germ cell-depleted testes via  _c-Kit_ mutation (Kit^W/Wv^)[@muellerIndependentSpecializationHuman2013].
+Heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ prevents the maturation of germ cells, resulting in overall germ cell depletion[@handelSertoliCellDifferentiation1979].
 We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B).
@@
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes.
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes (Supplementary table 1).
 
@@
 We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B).
-We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).
+We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).
 
@@
 The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E).
+We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+Contrastingly, the _Kdm5c_-KO hippocampus instead expressed genes involved in late-stage sperm development, those involved in the sperm axoneme.
+This suggests the germline developmental program may be occuring ectopically as _Kdm5c_-KO cells progresses through somatic tissue development **(note: placement is awkward.
+Even put here or talk about in dicussion because ChIPseq could help support this?
+- KDM5C not bound to a lot of brain DEGs)**.
 
-We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
+While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
 These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo.
-However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`).
+However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs does not revert cells back to a 2-cell state (Figure 3G).
 
@@
 We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
-Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.
+Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including key drivers of germline identity that can be translated into protein.
 
@@
 Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription.
-However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including master regulators of meiosis and germline identity.
-While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other master regulators that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.
+However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including key regulators of meiosis and germline identity.
+While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other germline drivers that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.
+
+```{r, KDM5C_ChIPseq, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+KDM5C_ChIP_EpiLC <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_EpiLC.csv", header = T)
+
+KDM5C_ChIP_PNC <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_PNC.csv", header = T)
+
+```
 
 To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: 2,437, PNCs: 1,610, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A).
+We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A).
 Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B).
@@
 - Other H3K4me regulators, anything known about tissue-biased gene expression?
+- This suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
 - tissue-biased gene expression:
@@
   \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
-  \caption[\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
+  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
 Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
@@
 
-**Figure 3: Kdm5c-KO epiblast-like cells express master regulators of germline identity**
+**Figure 3: Kdm5c-KO epiblast-like cells express key drivers regulators of germline identity**
 	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 

### 2024-07-01 submission/manuscript.Rmd
update chipseq in text counts

@@
 ```{r, KDM5C_ChIPseq, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-KDM5C_ChIP_EpiLC <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_EpiLC.csv", header = T)
+KDM5C_ChIP_EpiLC_promo <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_EpiLC.csv")
 
-KDM5C_ChIP_PNC <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_PNC.csv", header = T)
+KDM5C_ChIP_PNC_promo <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_PNC.csv")
+
+KDM5C_ChIP_EpiLC <- read.csv("../data/raw/ChIPseq_mm10_EpiLC_WTnoKO_consensus_peaks.bed", sep = "\t")
+
+KDM5C_ChIP_PNC <- read.csv("../data/raw/ChIPseq_mm10_PNC_WTnoKO_consensus_peaks.bed", sep = "\t")
 
@@
 To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A).
+We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`kb from TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A).
 Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B).

### 2024-07-02 code/utilities/mouseGermlineGenes.R
update germline gene embryonic data

@@ -51,10 +51,10 @@ adultFPKM <- merge(avgFPKM, genedf, by = "ENSEMBL")
 
 # generate a dataframe with WT fpkms across all gonad timepoints (add in adult)
 adultFPKM_WT <- subset(adultFPKM, select = c(-AdultXYWWv))
-mueller_WT <- subset(mueller_genes, select = c(SYMBOL, P6XYWT, E12XXWT ,E12XYWT, E14XXWT , E14XYWT , E16XXWT , E16XYWT))
+mueller_WT <- subset(mueller_genes, select = c(SYMBOL, E12XXWT ,E12XYWT, E14XXWT , E14XYWT , E16XXWT , E16XYWT))
 mueller_WT <- merge(mueller_WT, adultFPKM_WT, by = "SYMBOL")
 #change the column order so the numbers are all together and in chronological order
-mueller_WT <- mueller_WT[,c("SYMBOL", "ENSEMBL", "E12XXWT" ,"E12XYWT", "E14XXWT" , "E14XYWT" , "E16XXWT" , "E16XYWT", "P6XYWT", 'AdultXYWT')]
+mueller_WT <- mueller_WT[,c("SYMBOL", "ENSEMBL", "E12XXWT" ,"E12XYWT", "E14XXWT" , "E14XYWT" , "E16XXWT" , "E16XYWT", 'AdultXYWT')]
 mueller_WT <- unique(mueller_WT)
 
 #print out number of genes with all time points that we can use for analysis 
@@ -63,10 +63,10 @@ cat("Number of genes for analysis: ", nrow(mueller_WT), "\n")
 
 #####1C) Generate dataframe with germ-cell depleted WWv FPKMs for all gonad sequencing
 adultFPKM_WWv <- subset(adultFPKM, select = c(-AdultXYWT))
-mueller_WWv <- subset(mueller_genes, select = c(SYMBOL, P6XYWWv, E12XXWWv ,E12XYWWv, E14XXWWv , E14XYWWv , E16XXWWv , E16XYWWv))
+mueller_WWv <- subset(mueller_genes, select = c(SYMBOL, E12XXWWv ,E12XYWWv, E14XXWWv , E14XYWWv , E16XXWWv , E16XYWWv))
 mueller_WWv <- merge(mueller_WWv, adultFPKM_WWv, by = "SYMBOL")
[…]

### 2024-07-02 submission/manuscript.Rmd
editing fig1 resutls text

@@
 
-Previous RNA sequencing (RNA-seq) in the adult _Kdm5c_ knockout (-KO) hippocampus identified a few transcripts that are typically expressed in the testis[@scandagliaLossKdm5cCauses2017].
-However, the expression of tissue-specific genes within the _Kdm5c_-KO brain has not systematically characterized.
-Currently, it is unclear if other tissue-specific transcripts are also expressed within the _Kdm5c_-KO brain and if misexpression of testis-enriched genes are a major consequence of KDM5C loss, given the high abundance of testis-enriched genes within the mouse transcriptome.
-Therefore, we first globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]
-
-**note: the idea is there but need to work on wording. The extent to which brain tissue identity is altered is unknown and if testis genes are a major effect**
+Previous RNA sequencing (RNA-seq) in the adult hippocampus ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
+Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, it is currently unclear if this erosion of brain tissue identity is a major consequence of _Kdm5c_ loss and if it is unique to testis-enriched genes.
+Therefore, we first globally assessed the expression of previously characterized mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the amygdala and hippocampus in adult mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 
-We found many genes significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues, with the majority of differentially expressed genes (DEGs) displaying biased expression towards the testis (Figure 1A-C).
-Even though the testis has the largest number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact).
+We found a large proportion of genes significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C).
+Testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact Test), even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes).
+Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
 
-Even though the _Kdm5c_-KO mice are male, we also observed significant enrichment of ovary-biased DEGs within both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain.
+Although the _Kdm5c_-KO mice were male, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
 Intriguingly, many ovary and testis-biased DEGs have functions specific to germ cells and have no known role in the brain.
-For example, the testis-biased DEG _Cytochrome C, testis (Cyct)_ is a testis-specific cytochrome c isoform found in spermatozoa mitochondria and is important for flagellar beating[@goldbergCytochromeImmunofluorescentLocalization1977; @narisawaTestisSpecificCytochromeNull2002] (Figure 1C).
-The ovary-enriched DEG _Zygotic arrest 1 (Zar1)_ was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Although not consistent across brain regions, we also found enrichment of two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Fisher's Exact) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Fisher's Exact).
-One example of a liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).
-Overall, these tissue-enriched genes show little to no expression in the wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
-
-Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), even though the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
-Altogether, these results suggest misexpression of testis and other tissue-enriched genes within the _Kdm5c_-KO brain is not due to random de-repression of all genes, but rather due to a dysregulation of tissue identity.
+For example, the testis-biased DEG _FK506 binding protein 6 (Fkbp6)_ is a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C) while the ovary-enriched DEG _Zygotic arrest 1 (Zar1)_ was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
+Although not consistent across brain regions, we also found enrichment of two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Fisher's Exact Test).
+An example of a liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).
+Testis, ovary, and liver-enriched DEGs showed little to no expression in the wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
+Together, these results suggest misexpression of testis and other tissue-enriched genes within the _Kdm5c_-KO brain is a major effect of KDM5C loss.
 
@@
 We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B).
-The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
+The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, the aforementioned testis gene that regulates piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
 We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018].

### 2024-07-02 submission/manuscript.Rmd
update germline definition text and notes on figure

@@
 Testis, ovary, and liver-enriched DEGs showed little to no expression in the wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
-Together, these results suggest misexpression of testis and other tissue-enriched genes within the _Kdm5c_-KO brain is a major effect of KDM5C loss.
+Together, these results suggest misexpression of testis and other tissue-enriched genes within the brain is a major effect of KDM5C loss.
 
@@
 
-The testis contains both germ cells (e.g.
-spermatogonia) and somatic cells (e.g.
+The testis contains both germ cells (meiotic cells, e.g.
+spermatogonia) and somatic cells (non-meiotic, e.g.
 Leydig cells) that support hormone production and germline functions.
-We then wondered if the high enrichment of testis genes in the _Kdm5c_-KO brain reflected a failure to demarcate between the soma and germline.
-Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
+Select testis-enriched DEGs that were previously characterized had germline-specific functions[], suggesting _Kdm5c_-KO cells fail to demarcate between the soma and germline.
+To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assed their function through gene onotology.
+We found high enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
-To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in germ cell-depleted testes via  _c-Kit_ mutation (Kit^W/Wv^)[@muellerIndependentSpecializationHuman2013].
-Heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ prevents the maturation of germ cells, resulting in overall germ cell depletion[@handelSertoliCellDifferentiation1979].
-We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B).
-The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, the aforementioned testis gene that regulates piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
-We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018].
+To further validate if these testis DEGs are truly germline genes, we then compared their expression in somatic versus germ cells within the testis.
+We first compared their expression within wild-type versus germ cell-depleted testes[@muellerIndependentSpecializationHuman2013].
+In this study, germ cell depletion was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_
+(Kit^W/Wv^), which prevents the maturation of germ cells and results in overall germline loss[@handelSertoliCellDifferentiation1979].
+Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
+The only testis-enriched DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, the aforementioned testis gene that regulates piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
+We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
 We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
 spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C).
-Together, these data suggest the _Kdm5c_-KO brain expresses germline genes.
+Together, these data suggest the somatic _Kdm5c_-KO brain aberrantly expresses germline genes.
 
-We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes.
-To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice.
-We assessed gene expression in male and female mice at embryonic day 12, 14, and 16, as well as male postnatal day 6 and adult testes.
-We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes (Supplementary table 1).
+We wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes.
+To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in germline-depleted (Kit^W/Wv^) mice.
+Current available Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
+
+We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes with knockout of _Kdm5c_ (Supplementary table 1).
 

### 2024-07-02 submission/manuscript.Rmd
update EpiLC text and figure

@@
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes with knockout of _Kdm5c_ (Supplementary table 1).
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes with loss of _Kdm5c_ (Supplementary table 1).
 
@@
 
-Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity.
-Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020].
+Misexpression of germline genes in the adult _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity.
+Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
 This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top).
 Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by assessing their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].
+Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
 
-We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B).
-We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).
+We first identified _Kdm5c_-KO EpiLC DEGs through DESeq2 analysis of our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (log2 fold change > `r l2fcco`, q < 0.1).
+Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
+Using our curated list, we found _Kdm5c_-KO EpiLCs aberrantly expressed `r format(nrow(epigerm))`germline-enriched genes, including the previously characterized hippocampal DEG[@scandagliaLossKdm5cCauses2017] _Cytochrome C, testis-specific (Cyct)_ (Figure 3C).
+Although we observed aberrant expression of many tissue-enriched genes, we did not observe any significant difference in primed pluripotency genes (Figure 3D) or gross changes in _Kdm5c_-KO cell morpholgy during differentiation (Figure 3E), indicating KDM5C loss does not impair EpiLC formation.
 
-Using our curated list, we found `r format(nrow(epigerm))` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D).
-To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
-The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E).
-We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-Contrastingly, the _Kdm5c_-KO hippocampus instead expressed genes involved in late-stage sperm development, those involved in the sperm axoneme.
-This suggests the germline developmental program may be occuring ectopically as _Kdm5c_-KO cells progresses through somatic tissue development **(note: placement is awkward.
-Even put here or talk about in dicussion because ChIPseq could help support this?
-- KDM5C not bound to a lot of brain DEGs)**.
+To evaluate if all germline DEGs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across all sequencing datasets (Figure 3F).
+We then compared the function of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus showed strong enrichment of late-stage sperm genes, such those involved in the sperm axoneme.
+This shift from meiotic genes to later spermatogenesis genes in the hippocampus suggests the germline developmental program could occur ectopically as _Kdm5c_-KO cells progresses through somatic tissue development.
+**note: this is strengthened by the ChIP-seq data since KDM5C is not directly bound to many brain/flagellar DEGs.This point might be stronger in the ChIPseq figure**
 
-While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G).
+Interestingly, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H).
 These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo.
-However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs does not revert cells back to a 2-cell state (Figure 3G).
+However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).
 
 _Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019].
-Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
-This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_-KO cellular function.
-We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H).
-We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
-Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including key drivers of germline identity that can be translated into protein.
+A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
+This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function.
+We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3H).
+We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020].
+Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
 

### 2024-07-02 submission/manuscript.Rmd
remmoved commas from GO IDs by typing by hand like a peasant

@@
 	}
-	return(out)
+	outformat <- ifelse(column == "ID", toString(out), out) #an attempt to remove the commas from GO ID numbers, didn't work
+	return(outformat) 
 }
@@
 Leydig cells) that support hormone production and germline functions.
-Select testis-enriched DEGs that were previously characterized had germline-specific functions[], suggesting _Kdm5c_-KO cells fail to demarcate between the soma and germline.
+Select testis-enriched DEGs that were characterized previously had germline-specific functions[@scandagliaLossKdm5cCauses2017], suggesting _Kdm5c_-KO cells fail to demarcate between the soma and germline.
 To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assed their function through gene onotology.
-We found high enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
+We found high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
@@
 The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across all sequencing datasets (Figure 3F).
-We then compared the function of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+We then compared the function of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle (GO: 0051321, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
 While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus showed strong enrichment of late-stage sperm genes, such those involved in the sperm axoneme.

### 2024-07-03 submission/manuscript.Rmd
begin editing chipseq section text

@@
 We first compared their expression within wild-type versus germ cell-depleted testes[@muellerIndependentSpecializationHuman2013].
-In this study, germ cell depletion was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_
-(Kit^W/Wv^), which prevents the maturation of germ cells and results in overall germline loss[@handelSertoliCellDifferentiation1979].
+In this study, germ cell depletion was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^), which prevents the maturation of germ cells and results in overall germline loss[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
@@
 
-### KDM5C-mediated removal of H3K4me3 during early embryogenesis promotes long-term repression of germline genes
+### KDM5C binds to a subset of germline gene promoters during early embryogenesis
 
-**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
-	A) KDM5C binding in EpiLCs vs PNCs - number of peaks and percentage of bound genomic regions
-	B) KDM5C binding in Epilcs vs pncs germline genes - GO 
-	C) Left - bar graph of germline DEGs KDM5C is bound to in EpiLCs Right 
-	D) Example bigwigs
-		* Brain and EpiLC shared, EpiLC-specific, Brain-specific, 
-	E) motif analysis of KDM5C-bound vs unbound genes
+- **note: do Direct vs indirect DEGs motif analysis**
+- However, it is currently unclear if KDM5C binds to all germline DEGs and if its binding is maintained at any germline genes in neurons.
 
-	--- separate into two figures
-	D) Bigwigs of Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
-	E) Increase in H3K4me2 in EpiLCs at germline genes (left - bigwigs, right average signal at germline gene TSS? Just KDM5C-bound genes?)
-	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). 
-	F) KDM5C expression ESC --> EpiLC (increasing then decreasing) Left - RNA right - protein 
-	G) RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
-	H) DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)
-
-	Direct vs indirect binding motif analysis
-
-Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription.
-However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including key regulators of meiosis and germline identity.
-While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other germline drivers that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.
+Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress hippocampal germline DEGs and KDM5C is not bound to their promoters in neurons[@scandagliaLossKdm5cCauses2017].
+There is some evidence KDM5C binds to select germline gene promoters in ESCs[@scandagliaLossKdm5cCauses2017], including two recent independent screens that found KDM5C binds to Dazl's promoter[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+As KDM5C's binding at germline gene promoters has not been systematically characterized, it is currently unclear what types of germline genes KDM5C regulates and if its binding is maintained at any germline genes in neurons.
 
@@
 
+GO_KDM5C_ChIP <- read.csv("../results/ChIPseq_KDM5C_GO_EpiLC_vs_PNC_unique.csv")
+
 ```
 
-To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`kb from TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A).
-Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B).
-Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).
+To further characterize the mechanism behind _Kdm5c_-KO germline gene misexpression, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016].
+EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`kb from TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).
+
+- PNC EpiLC shared unique be come explicit (Figure 4B)
+
+While the majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs, many genes were not
+
+a subset of promoter-bound genes were shared between EpiLCs and PNCs
+
+- We then compared the known functions of genes with KDM5C bound to their promoter in EpiLCs vs PNCs.
+
+When we compared the functions of genes with KDM5C bound to their promoters, germline-specific ontologies were only enriched in EpiLCs (Figure 4B).
+
+Ggermline-specific ontologies were only enriched in EpiLCs (Figure 4B).   
+When we compared the functions of genes with KDM5C bound to their promoters,
+
+Ontologies enriched EpiLCs included including meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)` and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3).
+
+Contrastingly, germline-specific ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).
 
@@
 
-----
+### KDM5C erases H3K4me3 to promote long-term repression of germline genes via DNA methylation
 

### 2024-07-03 code/ChIPseq_KDM5C_EpiLC_PNC.R
Update ChIPseq - swap columns and shorten GO height

@@ -113,7 +113,7 @@ plots <- list()
 for (k in 1:length(samples)){
 	plotdf <- subset(qlotdf, ChIP == samples[k])
 	#set the plotting order
-	plotdf$GeneCategory <- factor(plotdf$GeneCategory, levels = germCategory)
+	plotdf$GeneCategory <- factor(plotdf$GeneCategory, levels = c("EpiLC Only", "Common", "Brain Only"))
 
 	library("ggpubr")
 	q <- ggbarplot(plotdf, "GeneCategory", "Count",

### 2024-07-03 code/ChIPseq_KDM5C_GO_Compare_BrainEpiLC.R

@@ -25,6 +25,7 @@ ck <- compareCluster(geneCluster = promogenes, fun = enrichGO,  OrgDb = "org.Mm.
 #ck <- setReadable(ck, OrgDb = "org.Mm.eg.db", keyType="ENSEMBL")
 head(ck) 
 
+ck <- simplify(ck)
 write.table(ck, snakemake@output[[1]], row.names = FALSE, sep = ",")
 
 ggsave(snakemake@output[[2]], plot = dotplot(ck, size = "Count"), width = 9, height = 5.5)
@@ -48,7 +49,7 @@ ck2 <- compareCluster(geneCluster = promo_compare, fun = enrichGO,  OrgDb = "org
 #ck <- setReadable(ck, OrgDb = "org.Mm.eg.db", keyType="ENSEMBL")
 head(ck2) 
 write.table(ck2, snakemake@output[[3]], row.names = FALSE, sep = ",")
-ggsave(snakemake@output[[4]], plot = dotplot(ck2, size = "Count"), width = 5.5, height = 5.5)
+ggsave(snakemake@output[[4]], plot = dotplot(ck2, size = "Count"), width = 5.5, height = 5)
 
 
 #get the overlap between promoters (euler plot)

### 2024-07-03 submission/manuscript.Rmd
draft of KDM5C ChIPseq results

@@
 
-### KDM5C binds to a subset of germline gene promoters during early embryogenesis
+## KDM5C binds to a subset of germline gene promoters during early embryogenesis
 
@@
 
-- PNC EpiLC shared unique be come explicit (Figure 4B)
+The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 4B).
+We then performed gene ontology analysis to compare the functions of genes bound by KDM5C at the promoter in different cell types.
+While there were no significant ontologies for genes with KDM5C only bound in PNCs, gene ontologies for peaks shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C).
+Germline-specific ontologies were only enriched in peaks unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C).
+When comparing KDM5C binding at all germline gene promoters, KDM5C was only bound to a subset of germline gene promoters in EpiLCs and was not bound to any in PNCs (Figure 4D).
+Together, this suggests KDM5C is recruited to a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not regulate germline genes in neurons.
 
-While the majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs, many genes were not
+To determine if the germline mRNAs expressed in the _Kdm5c_-KO brain and EpiLCs are direct targets of KDM5C, we then compared KDM5C binding at RNA-seq DEG promoters (Figure 4E).
+About one third of EpiLC specific and brain-specific (hippocampus, amygdala, or both) germline DEGs were bound by KDM5C in EpiLCs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
+Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above.
+Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F).
+In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs bound by KDM5C (Figure 4G).
+Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H).
+Altogether, this suggests the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters during embryogenesis.
 
-a subset of promoter-bound genes were shared between EpiLCs and PNCs
+## KDM5C erases H3K4me3 to promote long-term repression of germline genes via DNA methylation
 
-- We then compared the known functions of genes with KDM5C bound to their promoter in EpiLCs vs PNCs.
+Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+Somatic repression of germline genes is typically established during the transition between naive and primed pluripotency, which modeled in vitro as ESCs to EpiLC differentiation.
+In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters (**note: check**), including H2AK119ub1 and H3K9me3.
+Germline genes are then silenced long-term by CpG DNA methylation in EpiLCs germline genes.
+As KDM5C's embryonic role in germline gene repression has only been characterized in ESCs, it is currently unclear if KDM5C is involved in the transition between ESCs and EpiLCs.
 
-When we compared the functions of genes with KDM5C bound to their promoters, germline-specific ontologies were only enriched in EpiLCs (Figure 4B).
-
-Ggermline-specific ontologies were only enriched in EpiLCs (Figure 4B).   
-When we compared the functions of genes with KDM5C bound to their promoters,
-
-Ontologies enriched EpiLCs included including meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)` and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3).
-
-Contrastingly, germline-specific ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).
-
-We then assessed KDM5C binding at germline genes expressed _Kdm5c_-KO cells to assess if their dysreuglation is due to direct loss of KDM5C binding during embryogenesis.
-
-	* KDM5C was bound to the promoter of many genes dysregulated in both EpiLCs and brain datasets, including _D1Pas1_ and _Cyct_.
-	* KDM5C was not bound to many of the brain-specific germline DEGs, such as XXX.
-	* While KDM5C binds to many germline DEGs in EpiLCs, including _Dazl_, many were also unbound. This included the germline transcription factor _Stra8_.
-	* All genes lacked binding in PNCs.
-	* Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
-
-### KDM5C erases H3K4me3 to promote long-term repression of germline genes via DNA methylation
-
-Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
 It is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes, such as enabling the placement of DNA methylation.
@@
 For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
-- Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
+- Otx2 is properly expressed in _Kdm5c_-KO EpiLCs and is known to prevent pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
+
+- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
 

### 2024-07-03 submission/manuscript.Rmd
drafting KDM5C mechanism section

@@
 Somatic repression of germline genes is typically established during the transition between naive and primed pluripotency, which modeled in vitro as ESCs to EpiLC differentiation.
-In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters (**note: check**), including H2AK119ub1 and H3K9me3.
-Germline genes are then silenced long-term by CpG DNA methylation in EpiLCs germline genes.
-As KDM5C's embryonic role in germline gene repression has only been characterized in ESCs, it is currently unclear if KDM5C is involved in the transition between ESCs and EpiLCs.
+In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation[@mochizukiRepressionGermlineGenes2021].
+It has been proposed KDM5C may promote germline gene silencing via H3K4me3 removal since DNA methylation is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017] and H3K4me3 can impair DNA methylation placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+However, KDM5C was recently shown to repress _Dazl_ in ESCs indepedent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.
 
-It is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes, such as enabling the placement of DNA methylation.
+To elucidate KDM5C's role in germline gene silencing, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets in wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
+In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 5A).
+We additionally found a marked increase in H3K4me2 germline gene TSSs in _Kdm5c_-KO EpiLCs (Figure 5B).
+Incerase in H3K4me2 and H3K4me3 was highest in _Kdm5c_-KO cells that were highest genes that are bound by KDM5C at their promoter in EpiLCs (**note: do analysis to check if true**).
 
-* In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed ectopic H3K4me3 peaks in the _Kdm5c_-KO amygdala at the transcription start site (TSS) of germline genes (Figure 4).
-* We additionally observed a marked increase in H3K4me2 at germline gene promoters in _Kdm5c_-KO EpiLCs.
-	* This increase was greatest at germline genes directly bound by KDM5C (?)
+To assess KDM5C's embryonic role in germline gene silencing, we first characterized KDM5C's expression in ESCs and EpiLCs by harvesting RNA and protein at 0 hours (ESCs), 24 hours, and 48 hours (EpiLCs).
+While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation, KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5C).
+
+We then determined the role of KDM5C in the initial placement of DNA methylation at germline gene promoters by performing whole genome bislufite sequencing in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) (**note: check**).
+- While wild-type cells accumulate high levels of DNA methylation over the course of ESCs to exEpiLC differentiation, DNA methylation is markedly reduced in _Kdm5c_-KO
 
@@
 - Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
+
+- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
 

### 2024-07-03 submission/manuscript.Rmd
update figure captions but one is too long to render I think

@@
 
-__
-
 \begin{figure}
@@
 Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
-Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test}
+Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test.}
+  \label{figurelabel}
+\end{figure}
+
+\begin{figure}
+  \centering 
+  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
+  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom).
+\textbf{B.} Overlap of genes with KDM5C bound to their promoters in EpiLCs (purple) and PNCs (blue).
+\textbf{C.} Gene ontology comparision of genes with KDM5C bound to their promoter.
+Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared).
+\textbf{D.} Average KDM5C binding at all germline-enriched genes in EpiLCs (left) and PNCs (right).\textbf{E.} KDM5C binding at the promoters of RNA-seq germline DEGs.
+Genes were classified as either only dysregulated in EpiLCs (EpiLC only), genes dyresgulated in the hippocampus or amygdala but not EpiLCs (brain only), or genes dysregulated in both EpiLCs and the brain (common).
+\textbf{F.} Example ChIP-seq bigwigs of DEGs unqiue to EpiLCs.
+Although both are expressed in EpiLCs, KDM5C is bound to the \textit{Dazl} promoter but not the \textit{Stra8} promoter} in EpiLCs.
+\textbf{G.} Example ChIP-seq bigwigs of DEGs common between brain and EpiLCs.
+KDM5C is bound to the \textit{D1Pas1} promoter but not the \textit{XXX} promoter} in EpiLCs.
+\textbf{H.} Example ChIP-seq bigwigs of DEGs unique to the brain.
+KDM5C is bound to the \textit{XXX} promoter but not the \textit{Meig1} promoter.}
+  \label{figurelabel}
+\end{figure}
+
+\addtocounter{figure}{-1}
+\begin{figure} [t!]
+  \caption[]{\textbf{Retinoic acid promotes the prolonged misexpression of germline genes in \textit{Kdm5c}-KO epiblast-like cells.} (Legend continued.) \textbf{D.} Heatmap of germline DEG expression in \textit{Kdm5c}-KO (5CKO) cells at different differentiation time points and culture conditions, grouped by hierarchical clustering.
+Colors indicate the scaled transcripts per million (TPM) expression of germline DEGs, with red being highest expression and blue being lowest.
+\textbf{E.} Line plots of example germline genes in each cluster, displaying their TPM at the different time points.
+Black - wild-type cells differentiated without retinoic acid, dark pink - wild-type cells differentiated with retinoic acid, gray - \textit{Kdm5c}-KO cells differentiated without retionic acid, pale pink - \textit{Kdm5c}-KO cells differentiated with retinoic acid.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
+  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative H3K4me3 ChIP-seq peaks in the wild-type and Kdm5c-KO adult amygdala at two germline gene promoters.
+\textbf{B.} Left - Bigwigs of representative H3K4me2 ChIP-seq peaks in wild-type and Kdm5c-KO EpiLC.
+\textbf{C.} RNA and protein expression of KDM5C across ESC to EpiLC differentiation.
+Top left - diagram of differentiation protocol and collection time points.
+Bottom left - representative lanes of KDM5C western blot with DAXX loading control.
+Middle - KDM5C protein expression normalized to DAXX.
+Quantified intensity using ImageJ (artificial units - au).
+Right - RT-qPCR of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT} \textbf{D.} XXX {E.} XXX.}
   \label{figurelabel}

### 2024-07-03 submission/manuscript.tex
update figure captions but one is too long to render I think

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin = 0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{caption}
\captionsetup[figure]{font=footnotesize, labelfont=bf}
\usepackage{sectsty}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{lineno}
\linenumbers
\renewcommand\linenumberfont{\normalfont\small}
\usepackage{setspace}
\spacing{1.5}
\pagenumbering{arabic}
\usepackage{indentfirst}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Erosion of somatic tissue identity with loss of the X-linked
intellectual disability factor KDM5C}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

Gaps in knowledge addressed: * Are other tissue-enriched genes
dysregulated, or only testis, germline genes? * Curating a robust list
of male and female germline genes * Should talk about 2-cell genes vs
germline genes - way to systematically categorize? * Mechanism behind
long-term germline gene misexpression * Recent evidence suggests loss of
KDM5C in ESCs express some germline genes * Unclear if catalytic
activity is required for long-term silencing * Unclear if their
dysregulation lasts throughout life or the same between brain or not *
When in development does it begin? - Recent evidence suggests some
germline genes expressed in 5CKO ESCs but unclear if their dysregulation
lasts throughout differentiation and if the identity of germline genes
are different compared to the brain * Are there functional consequences
to germline gene misexpression?

Introduction: * Chromatin regulators are important for cellular identity
* H3K4me1-3 linked to active gene promoters and enhancers *
Surprisingly, mutations in Chromatin regulators lead to many NDDs
(including many H3K4 regulators) * Recent studies have shown some
chromatin regulators are important for regulating neuron-specific gene
expression/chromatin stat\_compare\_means * However, loss of some
chromatin regulators can also lead to ectopic expression of
tissue-enriched genes * Very few studies have looked at these genes and
it's unclear if these genes contribute to NDD impairments. * Necessary
to first characterize the mechanism behind their derepression to
identify molecular footholds into testing their contribution to neuronal
impairments and potential for therapeutic intervention

\begin{itemize}
\tightlist
\item
  Loss of KDM5C can result in the misexpression of genes typically only
  found in the testis

  \begin{itemize}
  \tightlist
  \item
    Misexpression of tissue-enriched genes hasn't been systematically
    characterized - Unclear if these genes are exceptions or if other
    tissue-specific genes are dysregulated
  \item
    Interestingly, these genes (Cyct, D1pas1) typically function in the
    germline
  \item
    Germ cells (meiotic cells) are typically distinguished from somatic
    cells very early on in embyrogenesis and is a key feature of
    multicellularity
  \item
    Chromatin regulators are very important for decommissioning germline
    genes and act successively the embryo implants into the uterine wall

    \begin{itemize}
    \tightlist
    \item
      Most studies have focused on ESCs, which have a similar
      transcriptome to germ cells / 2-cells
    \item
      recently, KDM5C was shown to repress DAZL in ESCs, independent of
      its catalytic activity
    \item
      However, DNA methylation is lost in the mature 5CKO brain, DNA
      methylation is placed later and it's Unclear if it's required for
      long-term repression (maybe too specific, just trying to go into
      the fact that the mechanism is partially understood but unclear)
    \end{itemize}
  \item
    Systematic characterization of ectopic germline genes hasn't been
    done

    \begin{itemize}
    \tightlist
    \item
      unknown if other germline-enriched genes are dysregulated,
      including oocyte-specific genes
    \item
      Crucially, it's unknown if misexpression of the germline program
      leads to functional consequences in 5CKO cells.
    \end{itemize}
  \end{itemize}
\end{itemize}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

To form a complete organism, embryonic stem cells must differentiate
into a myriad of discrete cellular identities. This is in part
accomplished by chromatin regulators that can either promote or impede
lineage-specific gene expression through histone and DNA
modifications\textsuperscript{1,2}. Although initially identified for
their roles in cellular identity\textsuperscript{3,4}, recent
advancements in next generation sequencing technologies unexpectedly
found many neurodevelopmental disorders (NDDs) are caused by or linked
to mutations in chromatin regulators. This relationship is partially
explained by their regulation of brain-specific genes or chromatin
states, such as modulating genes involved in synaptic
maturation\textsuperscript{5} or the transition between neuronal and
glial developmental programs\textsuperscript{6}. However, loss of some
chromatin regulators can also lead to the misexpression of
tissue-specific genes outside of their intended
environment\textsuperscript{3,4,7}. Currently, very few studies have
explored the misexpression of non-neuronal, tissue-specific genes in
chromatin-linked neurodevelopmental disorders\textsuperscript{8,9} and
it is unclear if this partial loss of brain identity contributes to
neurodevelopmental impairments. To elucidate their contribution to
neurodevelopmental impairments, it is essential to first characterize
the types of genes misexpressed, the developmental time point the
dysregulation begins, and the molecular mechanism underlying their
de-repression. Characterizing these features will enable us to identify
molecular footholds common between NDDs that can then be exploited for
potential therapeutics.

In this study, we characterized the misexpression of tissue-enriched
genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C,
also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
Pathogenic mutations in \emph{KDM5C} cause Intellectual Developmental
Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534),
whose features include short stature, intellectual disability, seizures,
aberrant aggression, and autistic behaviors\textsuperscript{10--12}.
Previous work has demonstrated constitutive \emph{Kdm5c} knockout (-KO)
mice recapitulate key MRXSCJ patient phenotypes, including
hyperaggression and learning impairments\textsuperscript{13}. Next
generation RNA sequencing (RNA-seq) in the \emph{Kdm5c}-KO hippocampus
unexpectedly revealed ectopic expression of testis-enriched genes within
the brain\textsuperscript{9}. However, it is currently unclear if
misexpression in the \emph{Kdm5c}-KO brain is unique to testis genes, as
other tissue-enriched genes have not been systematically evaluated.

Interestingly, some of the ectopic testis transcripts identified in the
\emph{Kdm5c}-KO brain are typically expressed in germ
cells\textsuperscript{9}. Unlike somatic cells, germ cells (e.g.~sperm
and eggs) undergo meiosis and pass on their genetic material to the next
generation. The germline and the soma are typically distinguished during
early embryogenesis, when germline genes are silenced in epiblast stem
cells soon after implantation and only reactivated in a subset to form
the germline. Chromatin regulators play a key role in decommissioning
germline genes as the embryo transitions from naive to primed
pluripotency by placing repressive histone H2A lysine 119
monoubiquitination (H2AK119ub1)\textsuperscript{14}, histone 3 lysine 9
trimethylation (H3K9me3)\textsuperscript{14,15}, and DNA CpG
methylation\textsuperscript{15--17} at germline gene promoters. KDM5C
may also be involved in this early decommissioning of germline genes, as
re-expression of KDM5C in neurons fails to surpress their
dysregulation\textsuperscript{9}. In support of this, KDM5C was very
recently shown to repress \emph{Deleted in azoospermia like (Dazl)}, a
key regulator of germline development, in mouse embryonic stem cells
(ESCs)\textsuperscript{18,19}. In support of this, two independent
screens in mouse embryonic stem cells (ESCs) recently identified KDM5C
as a repressor of \emph{Deleted in azoospermia like (Dazl)}, a key
regulator of germline development. However, KDM5C's role in embryonic
germline gene repression is currently unclear, given that Dazl is also
expressed in ESCs and in the 2-cell stage and germline gene
misexpression has yet to be globally characterized during
\emph{Kdm5c}-KO embryogenesis.

To elucidate KDM5C's role in tissue identity, here we characterized the
aberrant transcription of tissue-enriched genes within the
\emph{Kdm5c}-KO brain and epiblast-like cells (EpiLCs), an \emph{in
vitro} model of the post-implantation embryo. We identified general
dysregulation of tissue-enriched genes in both the adult \emph{Kdm5c}-KO
brain and EpiLCs, including misexpression of liver, muscle, and ovary
genes. The \emph{Kdm5c}-KO amygdala and hippocampus had significant
enrichment of testis-biased genes that are unique to germ cells. To
better characterize germline gene misexpression, we then generated a
dataset of germline-enriched genes by comparing gene expression in
gonads with germ cell depletion. We found \emph{Kdm5c}-KO EpiLCs
primarily expressed unique germline genes compared to the mature
\emph{Kdm5c}-KO brain, including \emph{Dazl} and \emph{Stra8}, key
drivers of germline identity and meiosis. While KDM5C is directly bound
to some germline gene promoters in EpiLCs, it is not directly bound to
many germline-enriched mRNAs expressed with \emph{Kdm5c}-KO cells,
indicating germline genes can be aberrantly transcribed through indirect
mechanisms. Finally, we found KDM5C loss impairs the placement of DNA
methylation at germline gene promoters as ESCs differentiate into
EpiLCs. Therefore, we propose KDM5C plays a crucial role in the
development of tissue identity during early embryogenesis, including
establishment of the soma-germline boundary.

\textbf{note: need a better conclusion sentence - work on when we know
what's happening with last figure/functional consequences}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{tissue-enriched-genes-including-testis-genes-are-aberrantly-expressed-in-the-kdm5c-ko-brain}{%
\subsection{\texorpdfstring{Tissue-enriched genes, including testis
genes, are aberrantly expressed in the \emph{Kdm5c}-KO
brain}{Tissue-enriched genes, including testis genes, are aberrantly expressed in the Kdm5c-KO brain}}\label{tissue-enriched-genes-including-testis-genes-are-aberrantly-expressed-in-the-kdm5c-ko-brain}}

\begin{itemize}
\tightlist
\item
  \textbf{note: should I compare amygdala and hippocampus? scandaglia
  only looked at hippocampus}
\end{itemize}

Previous RNA sequencing (RNA-seq) in the adult hippocampus ectopic
expression of some testis genes within the \emph{Kdm5c} knockout (-KO)
brain\textsuperscript{9}. Since tissue-enriched genes have not been
systematically characterized in the \emph{Kdm5c}-KO brain, it is
currently unclear if this erosion of brain tissue identity is a major
consequence of \emph{Kdm5c} loss and if it is unique to testis-enriched
genes. Therefore, we first globally assessed the expression of
previously characterized mouse tissue-enriched genes\textsuperscript{20}
in our published mRNA-seq datasets of the amygdala and hippocampus in
adult mice with constitutive knockout of
\emph{Kdm5c}\textsuperscript{21}.

We found a large proportion of genes significantly upregulated within
the \emph{Kdm5c}-KO brain (DESeq2\textsuperscript{22}, log2 fold change
\textgreater{} 0.5, q \textless{} 0.1) are typically enriched in
non-brain tissues (Amygdala: 35\%, Hippocampus: 24\%) (Figure 1A-B). The
majority of tissue-enriched differentially expressed genes
(tissue-enriched DEGs) were testis genes (Figure 1A-C). Testis-biased
DEGs were significantly enriched for both brain regions (Amygdala p =
1.83e-05; Hippocampus p = 4.26e-11, Fisher's Exact Test), even though
the testis has the largest total number of tissue-biased genes compared
to any other tissue (2,496 genes). Of note, we did not observe
enrichment of brain-enriched genes (Amygdala p = 1; Hippocampus p =
0.74, Fisher's Exact), despite the fact these are brain samples and the
brain has the second highest total number of tissue-enriched genes (708
genes).

In addition to the high enrichment of testis genes, we also identified
aberrant expression of other tissue-enriched genes within the
\emph{Kdm5c}-KO brain. Although the \emph{Kdm5c}-KO mice were male, we
also observed significant enrichment of ovary-biased DEGs in both the
amygdala and hippocampus (Amygdala p = 0.00574; Hippocampus p = 0.048,
Fisher's Exact) (Figure 1D). Intriguingly, many ovary and testis-biased
DEGs have functions specific to germ cells and have no known role in the
brain. For example, the testis-biased DEG \emph{FK506 binding protein 6
(Fkbp6)} is a known regulator of piRNA expression and meiosis in germ
cells\textsuperscript{23,24} (Figure 1C) while the ovary-enriched DEG
\emph{Zygotic arrest 1 (Zar1)} was recently shown to sequester mRNAs in
oocytes for meiotic maturation and early zygote
development\textsuperscript{25} (Figure 1D). Although not consistent
across brain regions, we also found enrichment of two non-gonadal
tissues - the liver (Amygdala p = 0.0398, Fisher's Exact Test) and the
muscle (Hippocampus p = 0.0104, Fisher's Exact Test). An example of a
liver-biased DEG is \emph{Apolipoprotein C-I (Apoc1)}, which is involved
in lipoprotein metabolism (Figure 1E). Testis, ovary, and liver-enriched
DEGs showed little to no expression in the wild-type brain, yet our
mRNA-seq data indicate they are polyadenylated and spliced into mature
transcripts (Figure 1C-E). Together, these results suggest misexpression
of testis and other tissue-enriched genes within the brain is a major
effect of KDM5C loss.

\hypertarget{germline-genes-are-aberrantly-expressed-in-the-male-kdm5c-ko-brain}{%
\subsubsection{\texorpdfstring{Germline genes are aberrantly expressed
in the male \emph{Kdm5c}-KO
brain}{Germline genes are aberrantly expressed in the male Kdm5c-KO brain}}\label{germline-genes-are-aberrantly-expressed-in-the-male-kdm5c-ko-brain}}

The testis contains both germ cells (meiotic cells, e.g.~spermatogonia)
and somatic cells (non-meiotic, e.g.~Leydig cells) that support hormone
production and germline functions. Select testis-enriched DEGs that were
characterized previously had germline-specific
functions\textsuperscript{9}, suggesting \emph{Kdm5c}-KO cells fail to
demarcate between the soma and germline. To test if this holds true for
all \emph{Kdm5c}-KO testis-biased DEGs, we first assed their function
through gene onotology. We found high enrichment of germline-relevant
ontologies, including spermatid development (GO: 0007286, p.adjust =
6.2e-12) and sperm axoneme assembly (GO: 0007288, p.adjust = 2.45e-14)
(Figure 2A).

To further validate if these testis DEGs are truly germline genes, we
then compared their expression in somatic versus germ cells within the
testis. We first compared their expression within wild-type versus germ
cell-depleted testes\textsuperscript{26}. In this study, germ cell
depletion was accomplished by heterozygous \emph{W} and \emph{Wv}
mutations in the enzymatic domain of \emph{c-Kit}
(Kit\textsuperscript{W/Wv}), which prevents the maturation of germ cells
and results in overall germline loss\textsuperscript{27}. Almost all
\emph{Kdm5c}-KO testis-enriched DEGs lost expression with germ cell
depletion (Figure 2B). The only testis-enriched DEG that did not show
considerable downregulation with germline depletion was \emph{FK506
binding protein 6 (Fkbp6)}, the aforementioned testis gene that
regulates piRNA expression and meiosis in germ
cells\textsuperscript{23,24}. We then assessed testis-enriched DEG
expression in a published single cell RNA-seq dataset that identified
cell type-specific markers within the testis\textsuperscript{28}. We
found that while some testis-enriched DEGs were classified as specific
markers for different germ cell types (e.g.~spermatogonia,
spermatocytes, round spermatids, and elongating spermatids) none marked
somatic cells (Figure 2C). Together, these data suggest the somatic
\emph{Kdm5c}-KO brain aberrantly expresses germline genes.

We wanted to more deeply characterize germline gene misexpression with
\emph{Kdm5c} loss, but lacked a comprehensive list of mouse
germline-enriched genes. To facilitate downstream analyses, we generated
a list of male and female germline-enriched genes by evaluating gene
expression in germline-depleted (Kit\textsuperscript{W/Wv}) mice.
Current available Kit\textsuperscript{W/Wv} datasets included males and
females at embryonic day 12, 14, and 16\textsuperscript{29}, as well as
adult male testes\textsuperscript{26}.

We defined genes as germline-enriched if their expression met the
following criteria: 1) their expression is greater than 1 FPKM in
wild-type gonads 2) their expression in any wild-type, non-gonadal
tissue\textsuperscript{20} does not exceed 20\% of their maximum
expression in the wild-type germline, and 3) their expression in the
germ cell-depleted germline, for any sex or time point, does not exceed
20\% of their maximum expression in the wild-type germline. These
criteria yielded 1,288 germline-enriched genes (Figure 2D), which was
hereafter used as a resource for assessing misexpression of germline
genes with loss of \emph{Kdm5c} (Supplementary table 1).

\hypertarget{kdm5c-ko-epiblast-like-cells-aberrantly-express-master-regulators-of-germline-identity}{%
\subsection{\texorpdfstring{\emph{Kdm5c}-KO epiblast-like cells
aberrantly express master regulators of germline
identity}{Kdm5c-KO epiblast-like cells aberrantly express master regulators of germline identity}}\label{kdm5c-ko-epiblast-like-cells-aberrantly-express-master-regulators-of-germline-identity}}

Misexpression of germline genes in the adult \emph{Kdm5c}-KO brain
suggests mutants fail to demarcate germline and somatic cellular
identity. Germ cells are typically distinguished from somatic cells soon
after the embryo implants into the uterine wall\textsuperscript{30,31}
when a subset of epiblast stem cells become the primordial germ cells
(PGCs) while the remainder differentiate into the ectoderm, mesoderm,
and endoderm to form the somatic tissues\textsuperscript{32}. This
developmental time point can be modeled \emph{in vitro} through
differentiation of embryonic stem cells (ESCs) into post-implantation
epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies
have demonstrated that while some germline-enriched genes are also
expressed in embryonic stem cells (ESCs), they are silenced as they
differentiate into EpiLCs\textsuperscript{15}. Therefore, we assessed if
KDM5C was necessary for initial germline gene silencing by evaluating
the impact of \emph{Kdm5c} loss in male EpiLCs.

We first identified \emph{Kdm5c}-KO EpiLC DEGs through DESeq2 analysis
of our previously published RNA-seq dataset\textsuperscript{33} (log2
fold change \textgreater{} 0.5, q \textless{} 0.1). Similar to the
\emph{Kdm5c}-KO brain, we observed general dysregulation of
tissue-enriched genes, with the largest number of genes belonging to the
brain and testis, although they were not significantly enriched (Figure
3B). Using our curated list, we found \emph{Kdm5c}-KO EpiLCs aberrantly
expressed 54germline-enriched genes, including the previously
characterized hippocampal DEG\textsuperscript{9} \emph{Cytochrome C,
testis-specific (Cyct)} (Figure 3C). Although we observed aberrant
expression of many tissue-enriched genes, we did not observe any
significant difference in primed pluripotency genes (Figure 3D) or gross
changes in \emph{Kdm5c}-KO cell morpholgy during differentiation (Figure
3E), indicating KDM5C loss does not impair EpiLC formation.

To evaluate if all germline DEGs are constitutively dysregulated or
change over the course of development, we then compared EpiLC germline
DEGs to those expressed in the \emph{Kdm5c}-KO brain. The majority of
EpiLC germline DEGs were unique to EpiLCs, with only \emph{Cyct} shared
across all sequencing datasets (Figure 3F). We then compared the
function of EpiLC and brain germline DEGs and found particularly high
enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G),
such as meiotic cell cycle (GO: 0051321, p.adjust = 4.44e-07). While
there was modest enrichment of meiotic gene ontologies in in both brain
regions, the \emph{Kdm5c}-KO hippocampus showed strong enrichment of
late-stage sperm genes, such those involved in the sperm axoneme. This
shift from meiotic genes to later spermatogenesis genes in the
hippocampus suggests the germline developmental program could occur
ectopically as \emph{Kdm5c}-KO cells progresses through somatic tissue
development. \textbf{note: this is strengthened by the ChIP-seq data
since KDM5C is not directly bound to many brain/flagellar DEGs.This
point might be stronger in the ChIPseq figure}

Interestingly, DEGs unique to \emph{Kdm5c}-KO EpiLCs included key
drivers of germline identity, such as \emph{Stimulated by retinoic acid
8 (Stra8)} and \emph{Deleted in azoospermia like (Dazl)} (Figure 3H).
These genes are typically expressed during embryonic germ cell
development to commit PGCs to the germline fate, but are also expressed
later in life to trigger meiotic gene expression
programs\textsuperscript{34--36}. Of note, some germline genes,
including \emph{Dazl}, are also expressed in the two-cell embryo.
However, we did not see misexpression of two-cell embryo-specific genes,
like \emph{Zscan4d} (q = 0.381), indicating \emph{Kdm5c}-KO in EpiLCs do
not revert back to a 2-cell state (Figure 3H).

\emph{Dazl} is an RNA-binding protein that regulates the translation of
germline mRNAs and is essential for germ cell
function\textsuperscript{37}. A significant portion of germline
transcripts misexpressed in \emph{Kdm5c}-KO EpiLCs are known binding
targets of DAZL, including \emph{Stra8}\textsuperscript{38} (p =
1.698e-07, Fisher's Exact Test). This suggests expression of DAZL
protein could promote the translation of other aberrant germline
transcripts, influencing their ability to impact \emph{Kdm5c}-KO
cellular function. We thus tested DAZL protein expression in
\emph{Kdm5c}-KO EpiLCs through immunocytochemistry (Figure 3H). We
observed about 25\% of \emph{Kdm5c}-KO EpiLCs expressed DAZL protein and
it was localized to the cytoplasm (p = 0.0015, Welch's t-test),
consistent with the pattern of DAZL expression in
spermatogonia\textsuperscript{38}. Altogether these results suggest
tissue-specific genes are misexpressed during \emph{Kdm5c}-KO
embryogenesis, including key drivers of germline identity that can be
translated into protein.

\hypertarget{kdm5c-binds-to-a-subset-of-germline-gene-promoters-during-early-embryogenesis}{%
\subsection{KDM5C binds to a subset of germline gene promoters during
early
embryogenesis}\label{kdm5c-binds-to-a-subset-of-germline-gene-promoters-during-early-embryogenesis}}

\begin{itemize}
\tightlist
\item
  \textbf{note: do Direct vs indirect DEGs motif analysis}
\item
  However, it is currently unclear if KDM5C binds to all germline DEGs
  and if its binding is maintained at any germline genes in neurons.
\end{itemize}

Previous work suggests KDM5C represses germline genes during early
development, as re-expression of KDM5C in knockout neuronal cultures
fails to suppress hippocampal germline DEGs and KDM5C is not bound to
their promoters in neurons\textsuperscript{9}. There is some evidence
KDM5C binds to select germline gene promoters in
ESCs\textsuperscript{9}, including two recent independent screens that
found KDM5C binds to Dazl's promoter\textsuperscript{18,19}. As KDM5C's
binding at germline gene promoters has not been systematically
characterized, it is currently unclear what types of germline genes
KDM5C regulates and if its binding is maintained at any germline genes
in neurons.

To further characterize the mechanism behind \emph{Kdm5c}-KO germline
gene misexpression, we analyzed KDM5C chromatin immunoprecipitation
followed by DNA sequencing (ChIP-seq) datasets in
EpiLCs\textsuperscript{33} and primary neuron cultures (PNCs) from the
cortex and hippocampus\textsuperscript{13}. EpiLCs had a higher total
number of KDM5C peaks than PNCs (EpiLCs: 5,808, PNCs: 1,276, MACS2 q
\textless{} 0.1 and fold enrichment \textgreater{} 1) and KDM5C was
primarily localized to gene promoters in both cell types (EpiLCs: 4,190,
PNCs: 745 +/- 500kb from TSS), although PNCs showed increased
localization to non-promoter regions (Figure 4A).

The majority of promoters bound by KDM5C in PNCs were also bound in
EpiLCs (513 shared promoters), however a large portion of gene promoters
were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure
4B). We then performed gene ontology analysis to compare the functions
of genes bound by KDM5C at the promoter in different cell types. While
there were no significant ontologies for genes with KDM5C only bound in
PNCs, gene ontologies for peaks shared between PNCs and EpiLCs were
enriched for functions involving nucleic acid turnover, such as
deoxyribonucleotide metabolic process (\url{GO:0009262}, p.adjust =
8.28e-05) (Figure 4C). Germline-specific ontologies were only enriched
in peaks unique to EpiLCs, such as meiotic nuclear division (GO: 0007127
p.adjust = 6.77e-16) and meiotic cell cycle process (\url{GO:1903046},
p.adjust = 5.05e-16) (Figure 3C). When comparing KDM5C binding at all
germline gene promoters, KDM5C was only bound to a subset of germline
gene promoters in EpiLCs and was not bound to any in PNCs (Figure 4D).
Together, this suggests KDM5C is recruited to a subset of germline gene
promoters in EpiLCs, including meiotic genes, but does not regulate
germline genes in neurons.

To determine if the germline mRNAs expressed in the \emph{Kdm5c}-KO
brain and EpiLCs are direct targets of KDM5C, we then compared KDM5C
binding at RNA-seq DEG promoters (Figure 4E). About one third of EpiLC
specific and brain-specific (hippocampus, amygdala, or both) germline
DEGs were bound by KDM5C in EpiLCs (EpiLC only: 36\%, Brain only:
33.3\%). Some notable differences in KDM5C binding for EpiLC-specific
DEGs included \emph{Dazl} and \emph{Stra8}, two key drivers of germline
identity discussed above. Although both mRNAs are expressed in
\emph{Kdm5c}-KO EpiLCs, KDM5C is only bound to \emph{Dazl}'s promoter
and not \emph{Stra8}'s in EpiLCs (Figure 4F). In contrast to the unique
DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs
bound by KDM5C (Figure 4G). Again, we did not observe any KDM5C binding
at germline gene promoters in PNCs, even for brain-specific DEGs (Figure
4H). Altogether, this suggests the majority of germline mRNAs expressed
in \emph{Kdm5c}-KO cells are dysregulated independent of direct KDM5C
recruitment to their promoters during embryogenesis.

\hypertarget{kdm5c-erases-h3k4me3-to-promote-long-term-repression-of-germline-genes-via-dna-methylation}{%
\subsection{KDM5C erases H3K4me3 to promote long-term repression of
germline genes via DNA
methylation}\label{kdm5c-erases-h3k4me3-to-promote-long-term-repression-of-germline-genes-via-dna-methylation}}

Although KDM5C is generally thought to supress transcription through
erasure of histone 3 lysine 4 di- and trimethylation
(H3K4me2/3)\textsuperscript{39}, recent studies in ESCs have suggested
KDM5C's repression \emph{Dazl} is independent of its catalytic
activity\textsuperscript{18}. Somatic repression of germline genes is
typically established during the transition between naive and primed
pluripotency, which modeled in vitro as ESCs to EpiLC differentiation.
In ESCs, chromatin regulators place repressive histone modifications at
germline gene promoters, including histone 2A 119 monoubiquitination
(H2AK119ub1) and histone 3 lysine 9 trimethylation
(H3K9me3)\textsuperscript{14,15,40}. Germline genes are then silenced
long-term in EpiLCs by \emph{de novo} placement of DNA CpG
methylation\textsuperscript{15}. It has been proposed KDM5C may promote
germline gene silencing via H3K4me3 removal since DNA methylation is
lost at select germline gene promoters in the
hippocampus\textsuperscript{9} and H3K4me3 can impair DNA methylation
placement\textsuperscript{41,42}. However, KDM5C was recently shown to
repress \emph{Dazl} in ESCs indepedent of its catalytic
activity\textsuperscript{18}. Because KDM5C's role in germline gene
repression has only been characterized in ESCs and in the mature brain,
it is currently unclear to what extent KDM5C is involved during
transition between ESCs and EpiLCs and if its catalytic activity is
required for long-term silencing.

To elucidate KDM5C's role in germline gene silencing, we first
characterized KDM5C's substrates, histone 3 lysine 4 di- and
trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets
in wild type and \emph{Kdm5c}-KO amygdala\textsuperscript{21} and
EpiLCs\textsuperscript{33}. In congruence with previous work in the
\emph{Kdm5c}-KO hippocampus\textsuperscript{9}, we observed aberrant
accumulation of H3K4me3 around the transcription start site (TSS) of
germline genes in the \emph{Kdm5c}-KO amygdala (Figure 5A). We
additionally found a marked increase in H3K4me2 germline gene TSSs in
\emph{Kdm5c}-KO EpiLCs (Figure 5B). Incerase in H3K4me2 and H3K4me3 was
highest in \emph{Kdm5c}-KO cells that were highest genes that are bound
by KDM5C at their promoter in EpiLCs (\textbf{note: do analysis to check
if true}).

To assess KDM5C's embryonic role in germline gene silencing, we first
characterized KDM5C's expression in ESCs and EpiLCs by harvesting RNA
and protein at 0 hours (ESCs), 24 hours, and 48 hours (EpiLCs). While
\emph{Kdm5c} mRNA steadily decreased from 0 to 48 hours of
diferentiation, KDM5C protein initially increased from 0 to 24 hours but
then decreased to near knockout levels by 48 hours (Figure 5C).

We then determined the role of KDM5C in the initial placement of DNA
methylation at germline gene promoters by performing whole genome
bislufite sequencing in wild-type and \emph{Kdm5c}-KO ESCs and 96 hour
extended EpiLCs (exEpiLCs) (\textbf{note: check}). - While wild-type
cells accumulate high levels of DNA methylation over the course of ESCs
to exEpiLC differentiation, DNA methylation is markedly reduced in
\emph{Kdm5c}-KO

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\begin{itemize}
\item
  The demarcation of the germ vs soma is a key feature of
  multicellularity
\item
  Other H3K4me regulators, anything known about tissue-biased gene
  expression?
\item
  This suggests the germline developmental program is occuring
  ectopically as \emph{Kdm5c}-KOs progresses through somatic tissue
  development
\item
  tissue-biased gene expression:

  \begin{itemize}
  \tightlist
  \item
    However unlike the gonadal-biased DEGs, many liver and muscle-biased
    DEGs have a known involvement in brain function. For example, the
    liver-biased DEG is \emph{Apolipoprotein C-I (Apoc1)}, is important
    for lipoprotein metabolism but has also been shown to influence
    learning and memory (Figure 1E).
  \end{itemize}
\item
  Otx2 is properly expressed in \emph{Kdm5c}-KO EpiLCs and is known to
  prevent pgc identity
  \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/}
\item
  Altogether, these data indicate that while some germline genes are
  misexpressed due to direct loss of KDM5C binding during emryogenesis,
  secondary downstream mechanisms can also promote their aberrant
  transcription.
\item
  KDM5C is dynamically regulated during the window of embryonic germline
  gene silencing, our EpiLC ChIP-seq is likely catching the tail end of
  KDM5C's main involvement.
\item
  Papers to read/reference:

  \begin{itemize}
  \tightlist
  \item
    Reconstitution of the Mouse Germ Cell Specification Pathway in
    Culture by Pluripotent Stem Cells:
    \url{https://www.cell.com/fulltext/S0092-8674(11)00771-9}
  \item
    two cell gene list used by Suzuki et al Max paper is based on 2 cell
    sequencing:
    \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/}
  \end{itemize}
\end{itemize}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-strahlLanguageCovalentHistone2000}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Strahl, B.D., and Allis, C.D. (2000). The language of
covalent histone modifications. Nature \emph{403}, 41--45.
\url{https://doi.org/10.1038/47412}.}

\leavevmode\vadjust pre{\hypertarget{ref-jenuweinTranslatingHistoneCode2001}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Jenuwein, T., and Allis, C.D. (2001). Translating the
{Histone Code}. Science \emph{293}, 1074--1080.
\url{https://doi.org/10.1126/science.1063127}.}

\leavevmode\vadjust pre{\hypertarget{ref-lewisGeneComplexControlling1978a}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Lewis, E.B. (1978). A gene complex controlling
segmentation in {Drosophila}. Nature \emph{276}, 565--570.
\url{https://doi.org/10.1038/276565a0}.}

\leavevmode\vadjust pre{\hypertarget{ref-kennisonDosagedependentModifiersPolycomb1988a}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Kennison, J.A., and Tamkun, J.W. (1988).
Dosage-dependent modifiers of polycomb and antennapedia mutations in
{Drosophila}. Proc. Natl. Acad. Sci. U.S.A. \emph{85}, 8136--8140.
\url{https://doi.org/10.1073/pnas.85.21.8136}.}

\leavevmode\vadjust pre{\hypertarget{ref-zhouBrainspecificPhosphorylationMeCP22006}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Zhou, Z., Hong, E.J., Cohen, S., Zhao, W.-N., Ho,
H.-Y.H., Schmidt, L., Chen, W.G., Lin, Y., Savner, E., Griffith, E.C.,
et al. (2006). Brain-specific phosphorylation of {MeCP2} regulates
activity-dependent {Bdnf} transcription, dendritic growth, and spine
maturation. Neuron \emph{52}, 255--269.
\url{https://doi.org/10.1016/j.neuron.2006.09.037}.}

\leavevmode\vadjust pre{\hypertarget{ref-hirabayashiPolycombLimitsNeurogenic2009}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Hirabayashi, Y., Suzki, N., Tsuboi, M., Endo, T.A.,
Toyoda, T., Shinga, J., Koseki, H., Vidal, M., and Gotoh, Y. (2009).
Polycomb {Limits} the {Neurogenic Competence} of {Neural Precursor
Cells} to {Promote Astrogenic Fate Transition}. Neuron \emph{63},
600--613. \url{https://doi.org/10.1016/j.neuron.2009.08.021}.}

\leavevmode\vadjust pre{\hypertarget{ref-kassisPolycombTrithoraxGroup2017}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Kassis, J.A., Kennison, J.A., and Tamkun, J.W. (2017).
Polycomb and {Trithorax Group Genes} in {\emph{Drosophila}}. Genetics
\emph{206}, 1699--1725.
\url{https://doi.org/10.1534/genetics.115.185116}.}

\leavevmode\vadjust pre{\hypertarget{ref-schaeferControlCognitionAdaptive2009}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Schaefer, A., Sampath, S.C., Intrator, A., Min, A.,
Gertler, T.S., Surmeier, D.J., Tarakhovsky, A., and Greengard, P.
(2009). Control of cognition and adaptive behavior by the {GLP}/{G9a}
epigenetic suppressor complex. Neuron \emph{64}, 678--691.
\url{https://doi.org/10.1016/j.neuron.2009.11.019}.}

\leavevmode\vadjust pre{\hypertarget{ref-scandagliaLossKdm5cCauses2017}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Scandaglia, M., Lopez-Atalaya, J.P., Medrano-Fernandez,
A., Lopez-Cascales, M.T., Del Blanco, B., Lipinski, M., Benito, E.,
Olivares, R., Iwase, S., Shi, Y., et al. (2017). Loss of {Kdm5c Causes
Spurious Transcription} and {Prevents} the {Fine-Tuning} of
{Activity-Regulated Enhancers} in {Neurons}. Cell Rep \emph{21}, 47--59.
\url{https://doi.org/10.1016/j.celrep.2017.09.014}.}

\leavevmode\vadjust pre{\hypertarget{ref-claesNovelSyndromicForm2000}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Claes, S., Devriendt, K., Van Goethem, G., Roelen, L.,
Meireleire, J., Raeymaekers, P., Cassiman, J.J., and Fryns, J.P. (2000).
\href{https://www.ncbi.nlm.nih.gov/pubmed/10982473}{Novel syndromic form
of {X-linked} complicated spastic paraplegia}. Am J Med Genet \emph{94},
1--4.}

\leavevmode\vadjust pre{\hypertarget{ref-jensenMutationsJARID1CGene2005}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Jensen, L.R., Amende, M., Gurok, U., Moser, B., Gimmel,
V., Tzschach, A., Janecke, A.R., Tariverdian, G., Chelly, J., Fryns,
J.P., et al. (2005). Mutations in the {JARID1C} gene, which is involved
in transcriptional regulation and chromatin remodeling, cause {X-linked}
mental retardation. Am J Hum Genet \emph{76}, 227--236.
\url{https://doi.org/10.1086/427563}.}

\leavevmode\vadjust pre{\hypertarget{ref-carmignacFurtherDelineationFemale2020}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Carmignac, V., Nambot, S., Lehalle, D., Callier, P.,
Moortgat, S., Benoit, V., Ghoumid, J., Delobel, B., Smol, T., Thuillier,
C., et al. (2020). Further delineation of the female phenotype with
{KDM5C} disease causing variants: 19 new individuals and review of the
literature. Clin Genet \emph{98}, 43--55.
\url{https://doi.org/10.1111/cge.13755}.}

\leavevmode\vadjust pre{\hypertarget{ref-iwaseMouseModelXlinked2016}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Iwase, S., Brookes, E., Agarwal, S., Badeaux, A.I., Ito,
H., Vallianatos, C.N., Tomassy, G.S., Kasza, T., Lin, G., Thompson, A.,
et al. (2016). A {Mouse Model} of {X-linked Intellectual Disability
Associated} with {Impaired Removal} of {Histone Methylation}. Cell
Reports \emph{14}, 1000--1009.
\url{https://doi.org/10.1016/j.celrep.2015.12.091}.}

\leavevmode\vadjust pre{\hypertarget{ref-endohPCGF6PRC1SuppressesPremature2017}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Endoh, M., Endo, T.A., Shinga, J., Hayashi, K., Farcas,
A., Ma, K.W., Ito, S., Sharif, J., Endoh, T., Onaga, N., et al. (2017).
{PCGF6-PRC1} suppresses premature differentiation of mouse embryonic
stem cells by regulating germ cell-related genes. Elife \emph{6}.
\url{https://doi.org/10.7554/eLife.21064}.}

\leavevmode\vadjust pre{\hypertarget{ref-mochizukiRepressionGermlineGenes2021}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Mochizuki, K., Sharif, J., Shirane, K., Uranishi, K.,
Bogutz, A.B., Janssen, S.M., Suzuki, A., Okuda, A., Koseki, H., and
Lorincz, M.C. (2021). Repression of germline genes by {PRC1}.6 and
{SETDB1} in the early embryo precedes {DNA} methylation-mediated
silencing. Nat Commun \emph{12}, 7020.
\url{https://doi.org/10.1038/s41467-021-27345-x}.}

\leavevmode\vadjust pre{\hypertarget{ref-velascoDnmt3bRecruitmentE2F62010}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Velasco, G., Hubé, F., Rollin, J., Neuillet, D.,
Philippe, C., Bouzinba-Segard, H., Galvani, A., Viegas-Péquignot, E.,
and Francastel, C. (2010). Dnmt3b recruitment through {E2F6}
transcriptional repressor mediates germ-line gene silencing in murine
somatic tissues. Proc Natl Acad Sci U S A \emph{107}, 9281--9286.
\url{https://doi.org/10.1073/pnas.1000473107}.}

\leavevmode\vadjust pre{\hypertarget{ref-hackettPromoterDNAMethylation2012}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Hackett, J.A., Reddington, J.P., Nestor, C.E., Dunican,
D.S., Branco, M.R., Reichmann, J., Reik, W., Surani, M.A., Adams, I.R.,
and Meehan, R.R. (2012). Promoter {DNA} methylation couples
genome-defence mechanisms to epigenetic reprogramming in the mouse
germline. Development \emph{139}, 3623--3632.
\url{https://doi.org/10.1242/dev.081661}.}

\leavevmode\vadjust pre{\hypertarget{ref-guptaGenomewideScreenReveals2023}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Gupta, N., Yakhou, L., Albert, J.R., Azogui, A., Ferry,
L., Kirsh, O., Miura, F., Battault, S., Yamaguchi, K., Laisné, M., et
al. (2023). A genome-wide screen reveals new regulators of the
2-cell-like cell state. Nat Struct Mol Biol.
\url{https://doi.org/10.1038/s41594-023-01038-z}.}

\leavevmode\vadjust pre{\hypertarget{ref-aladhamiSystematicIdentificationFactors2023}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Al Adhami, H., Vallet, J., Schaal, C., Schumacher, P.,
Bardet, A.F., Dumas, M., Chicher, J., Hammann, P., Daujat, S., and
Weber, M. (2023). Systematic identification of factors involved in the
silencing of germline genes in mouse embryonic stem cells. Nucleic Acids
Research \emph{51}, 3130--3149.
\url{https://doi.org/10.1093/nar/gkad071}.}

\leavevmode\vadjust pre{\hypertarget{ref-liComprehensiveMouseTranscriptomic2017}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Li, B., Qing, T., Zhu, J., Wen, Z., Yu, Y., Fukumura,
R., Zheng, Y., Gondo, Y., and Shi, L. (2017). A {Comprehensive Mouse
Transcriptomic BodyMap} across 17 {Tissues} by {RNA-seq}. Sci Rep
\emph{7}, 4200. \url{https://doi.org/10.1038/s41598-017-04520-z}.}

\leavevmode\vadjust pre{\hypertarget{ref-vallianatosMutuallySuppressiveRoles2020}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Vallianatos, C.N., Raines, B., Porter, R.S., Bonefas,
K.M., Wu, M.C., Garay, P.M., Collette, K.M., Seo, Y.A., Dou, Y., Keegan,
C.E., et al. (2020). Mutually suppressive roles of {KMT2A} and {KDM5C}
in behaviour, neuronal structure, and histone {H3K4} methylation. Commun
Biol \emph{3}, 278. \url{https://doi.org/10.1038/s42003-020-1001-6}.}

\leavevmode\vadjust pre{\hypertarget{ref-loveModeratedEstimationFold2014a}{}}%
\CSLLeftMargin{22. }%
\CSLRightInline{Love, M.I., Huber, W., and Anders, S. (2014). Moderated
estimation of fold change and dispersion for {RNA-seq} data with
{DESeq2}. Genome Biol \emph{15}, 550.
\url{https://doi.org/10.1186/s13059-014-0550-8}.}

\leavevmode\vadjust pre{\hypertarget{ref-crackowerEssentialRoleFkbp62003}{}}%
\CSLLeftMargin{23. }%
\CSLRightInline{Crackower, M.A., Kolas, N.K., Noguchi, J., Sarao, R.,
Kikuchi, K., Kaneko, H., Kobayashi, E., Kawai, Y., Kozieradzki, I.,
Landers, R., et al. (2003). Essential {Role} of {Fkbp6} in {Male
Fertility} and {Homologous Chromosome Pairing} in {Meiosis}. Science
\emph{300}, 1291--1295. \url{https://doi.org/10.1126/science.1083022}.}

\leavevmode\vadjust pre{\hypertarget{ref-xiolRoleFkbp6Chaperone2012}{}}%
\CSLLeftMargin{24. }%
\CSLRightInline{Xiol, J., Cora, E., Koglgruber, R., Chuma, S.,
Subramanian, S., Hosokawa, M., Reuter, M., Yang, Z., Berninger, P.,
Palencia, A., et al. (2012). A {Role} for {Fkbp6} and the {Chaperone
Machinery} in {piRNA Amplification} and {Transposon Silencing}.
Molecular Cell \emph{47}, 970--979.
\url{https://doi.org/10.1016/j.molcel.2012.07.019}.}

\leavevmode\vadjust pre{\hypertarget{ref-chengMammalianOocytesStore2022}{}}%
\CSLLeftMargin{25. }%
\CSLRightInline{Cheng, S., Altmeppen, G., So, C., Welp, L.M., Penir, S.,
Ruhwedel, T., Menelaou, K., Harasimov, K., Stützer, A., Blayney, M., et
al. (2022). Mammalian oocytes store {mRNAs} in a mitochondria-associated
membraneless compartment. Science \emph{378}, eabq4835.
\url{https://doi.org/10.1126/science.abq4835}.}

\leavevmode\vadjust pre{\hypertarget{ref-muellerIndependentSpecializationHuman2013}{}}%
\CSLLeftMargin{26. }%
\CSLRightInline{Mueller, J.L., Skaletsky, H., Brown, L.G., Zaghlul, S.,
Rock, S., Graves, T., Auger, K., Warren, W.C., Wilson, R.K., and Page,
D.C. (2013). Independent specialization of the human and mouse {X}
chromosomes for the male germ line. Nat Genet \emph{45}, 1083--1087.
\url{https://doi.org/10.1038/ng.2705}.}

\leavevmode\vadjust pre{\hypertarget{ref-handelSertoliCellDifferentiation1979}{}}%
\CSLLeftMargin{27. }%
\CSLRightInline{Handel, M.A., and Eppig, J.J. (1979). Sertoli {Cell
Differentiation} in the {Testes} of {Mice Genetically Deficient} in
{Germ Cells}. Biology of Reproduction \emph{20}, 1031--1038.
\url{https://doi.org/10.1095/biolreprod20.5.1031}.}

\leavevmode\vadjust pre{\hypertarget{ref-greenComprehensiveRoadmapMurine2018}{}}%
\CSLLeftMargin{28. }%
\CSLRightInline{Green, C.D., Ma, Q., Manske, G.L., Shami, A.N., Zheng,
X., Marini, S., Moritz, L., Sultan, C., Gurczynski, S.J., Moore, B.B.,
et al. (2018). A {Comprehensive Roadmap} of {Murine Spermatogenesis
Defined} by {Single-Cell RNA-Seq}. Dev Cell \emph{46}, 651--667.e10.
\url{https://doi.org/10.1016/j.devcel.2018.07.025}.}

\leavevmode\vadjust pre{\hypertarget{ref-sohGeneRegulatoryProgram2015}{}}%
\CSLLeftMargin{29. }%
\CSLRightInline{Soh, Y.Q., Junker, J.P., Gill, M.E., Mueller, J.L., van
Oudenaarden, A., and Page, D.C. (2015). A {Gene Regulatory Program} for
{Meiotic Prophase} in the {Fetal Ovary}. PLoS Genet \emph{11}, e1005531.
\url{https://doi.org/10.1371/journal.pgen.1005531}.}

\leavevmode\vadjust pre{\hypertarget{ref-magnusdottirHowMakePrimordial2014}{}}%
\CSLLeftMargin{30. }%
\CSLRightInline{Magnúsdóttir, E., and Surani, M.A. (2014). How to make a
primordial germ cell. Development \emph{141}, 245--252.
\url{https://doi.org/10.1242/dev.098269}.}

\leavevmode\vadjust pre{\hypertarget{ref-gunesdoganPrimordialGermCell2014}{}}%
\CSLLeftMargin{31. }%
\CSLRightInline{Günesdogan, U., Magnúsdóttir, E., and Surani, M.A.
(2014). Primordial germ cell specification: A context-dependent cellular
differentiation event {[}corrected{]}. Philos Trans R Soc Lond B Biol
Sci \emph{369}. \url{https://doi.org/10.1098/rstb.2013.0543}.}

\leavevmode\vadjust pre{\hypertarget{ref-bardotMouseGastrulationCoordination2020}{}}%
\CSLLeftMargin{32. }%
\CSLRightInline{Bardot, E.S., and Hadjantonakis, A.-K. (2020). Mouse
gastrulation: {Coordination} of tissue patterning, specification and
diversification of cell fate. Mechanisms of Development \emph{163},
103617. \url{https://doi.org/10.1016/j.mod.2020.103617}.}

\leavevmode\vadjust pre{\hypertarget{ref-samantaActivationXistEvolutionarily2022}{}}%
\CSLLeftMargin{33. }%
\CSLRightInline{Samanta, M.K., Gayen, S., Harris, C., Maclary, E.,
Murata-Nakamura, Y., Malcore, R.M., Porter, R.S., Garay, P.M.,
Vallianatos, C.N., Samollow, P.B., et al. (2022). Activation of {Xist}
by an evolutionarily conserved function of {KDM5C} demethylase. Nat
Commun \emph{13}, 2602.
\url{https://doi.org/10.1038/s41467-022-30352-1}.}

\leavevmode\vadjust pre{\hypertarget{ref-koubovaRetinoicAcidRegulates2006}{}}%
\CSLLeftMargin{34. }%
\CSLRightInline{Koubova, J., Menke, D.B., Zhou, Q., Capel, B., Griswold,
M.D., and Page, D.C. (2006). Retinoic acid regulates sex-specific timing
of meiotic initiation in mice. Proc. Natl. Acad. Sci. U.S.A. \emph{103},
2474--2479. \url{https://doi.org/10.1073/pnas.0510813103}.}

\leavevmode\vadjust pre{\hypertarget{ref-linGermCellIntrinsicExtrinsic2008}{}}%
\CSLLeftMargin{35. }%
\CSLRightInline{Lin, Y., Gill, M.E., Koubova, J., and Page, D.C. (2008).
Germ {Cell-Intrinsic} and -{Extrinsic Factors Govern Meiotic Initiation}
in {Mouse Embryos}. Science \emph{322}, 1685--1687.
\url{https://doi.org/10.1126/science.1166340}.}

\leavevmode\vadjust pre{\hypertarget{ref-endoRetinoicAcidGerm2019}{}}%
\CSLLeftMargin{36. }%
\CSLRightInline{Endo, T., Mikedis, M.M., Nicholls, P.K., Page, D.C., and
De Rooij, D.G. (2019). Retinoic {Acid} and {Germ Cell Development} in
the {Ovary} and {Testis}. Biomolecules \emph{9}, 775.
\url{https://doi.org/10.3390/biom9120775}.}

\leavevmode\vadjust pre{\hypertarget{ref-liDAZLMasterTranslational2019}{}}%
\CSLLeftMargin{37. }%
\CSLRightInline{Li, H., Liang, Z., Yang, J., Wang, D., Wang, H., Zhu,
M., Geng, B., and Xu, E.Y. (2019). {DAZL} is a master translational
regulator of murine spermatogenesis. Natl Sci Rev \emph{6}, 455--468.
\url{https://doi.org/10.1093/nsr/nwy163}.}

\leavevmode\vadjust pre{\hypertarget{ref-mikedisDAZLMediatesBroad2020}{}}%
\CSLLeftMargin{38. }%
\CSLRightInline{Mikedis, M.M., Fan, Y., Nicholls, P.K., Endo, T.,
Jackson, E.K., Cobb, S.A., De Rooij, D.G., and Page, D.C. (2020). {DAZL}
mediates a broad translational program regulating expansion and
differentiation of spermatogonial progenitors. eLife \emph{9}, e56523.
\url{https://doi.org/10.7554/eLife.56523}.}

\leavevmode\vadjust pre{\hypertarget{ref-iwaseXLinkedMentalRetardation2007}{}}%
\CSLLeftMargin{39. }%
\CSLRightInline{Iwase, S., Lan, F., Bayliss, P., De La Torre-Ubieta, L.,
Huarte, M., Qi, H.H., Whetstine, J.R., Bonni, A., Roberts, T.M., and
Shi, Y. (2007). The {X-Linked Mental Retardation Gene SMCX}/{JARID1C
Defines} a {Family} of {Histone H3 Lysine} 4 {Demethylases}. Cell
\emph{128}, 1077--1088.
\url{https://doi.org/10.1016/j.cell.2007.02.017}.}

\leavevmode\vadjust pre{\hypertarget{ref-liuPolycombGroupProtein2020}{}}%
\CSLLeftMargin{40. }%
\CSLRightInline{Liu, M., Zhu, Y., Xing, F., Liu, S., Xia, Y., Jiang, Q.,
and Qin, J. (2020). The polycomb group protein {PCGF6} mediates germline
gene silencing by recruiting histone-modifying proteins to target gene
promoters. J Biol Chem \emph{295}, 9712--9724.
\url{https://doi.org/10.1074/jbc.RA119.012121}.}

\leavevmode\vadjust pre{\hypertarget{ref-otaniStructuralBasisRecognition2009}{}}%
\CSLLeftMargin{41. }%
\CSLRightInline{Otani, J., Nankumo, T., Arita, K., Inamoto, S.,
Ariyoshi, M., and Shirakawa, M. (2009). Structural basis for recognition
of {H3K4} methylation status by the {DNA} methyltransferase {3A
ATRX}--{DNMT3}--{DNMT3L} domain. EMBO Reports \emph{10}, 1235--1241.
\url{https://doi.org/10.1038/embor.2009.218}.}

\leavevmode\vadjust pre{\hypertarget{ref-guoStructuralInsightAutoinhibition2015}{}}%
\CSLLeftMargin{42. }%
\CSLRightInline{Guo, X., Wang, L., Li, J., Ding, Z., Xiao, J., Yin, X.,
He, S., Shi, P., Dong, L., Li, G., et al. (2015). Structural insight
into autoinhibition and histone {H3-induced} activation of {DNMT3A}.
Nature \emph{517}, 640--644. \url{https://doi.org/10.1038/nature13899}.}

\end{CSLReferences}

\newpage

\hypertarget{figures-and-tables}{%
\section{Figures and Tables}\label{figures-and-tables}}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala. Left - MA plot of mRNA-sequencing. Right - Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{B.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO hippocampus. \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal). Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red. \textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}. Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red. \textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}. Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
  \label{figurelabel}
\end{figure}

\_\_

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis. Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM). \textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis (Green et al 2018). Germline cell types are highlighted in green, somatic cell types in black. \textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and germline-depleted mice and in adult mouse non-gonadal tissues (Li et al 2017).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). Bottom - representative images of wild-type (WT) and \textit{Kdm5c}-KO ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{B.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{C.} Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} enrichPlot comparing enriched gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{G.} Top - Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test. Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs. Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test.}
  \label{figurelabel}
\end{figure}

\textbackslash begin\{figure\} \centering 
\includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}

\caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology comparision of genes with KDM5C bound to their promoter. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding at all germline-enriched genes in EpiLCs (left) and PNCs (right).\textbf{E.} KDM5C binding at the promoters of RNA-seq germline DEGs. Genes were classified as either only dysregulated in EpiLCs (EpiLC only), genes dyresgulated in the hippocampus or amygdala but not EpiLCs (brain only), or genes dysregulated in both EpiLCs and the brain (common). \textbf{F.} Example ChIP-seq bigwigs of DEGs unqiue to EpiLCs. Although both are expressed in EpiLCs, KDM5C is bound to the \textit{Dazl} promoter but not the \textit{Stra8} promoter}

in EpiLCs. \textbf{G.} Example ChIP-seq bigwigs of DEGs common between
brain and EpiLCs. KDM5C is bound to the \textit{D1Pas1} promoter but not
the \textit{XXX} promoter\} in EpiLCs. \textbf{H.} Example ChIP-seq
bigwigs of DEGs unique to the brain. KDM5C is bound to the \textit{XXX}
promoter but not the \textit{Meig1} promoter.\} \label{figurelabel}
\textbackslash end\{figure\}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative H3K4me3 ChIP-seq peaks in the wild-type and Kdm5c-KO adult amygdala at two germline gene promoters. \textbf{B.} Left - Bigwigs of representative H3K4me2 ChIP-seq peaks in wild-type and Kdm5c-KO EpiLC. \textbf{C.} RNA and protein expression of KDM5C across ESC to EpiLC differentiation. Top left - diagram of differentiation protocol and collection time points. Bottom left - representative lanes of KDM5C western blot with DAXX loading control. Middle - KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - RT-qPCR of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2-deltaCT) \textbf{D.} XXX {E.} XXX.}
  \label{figurelabel}
\end{figure}

\newpage

\hypertarget{figure-outline}{%
\subsection{Figure outline:}\label{figure-outline}}

\textbf{Figure 1: Misexpression of tissue-specific genes in the
\emph{Kdm5c}-KO brain} * MA-plot and bar graphs of tissue-enriched genes
* Example testis-specific genes (NCBI and bigwigs) * An example ovary
tissue-specific gene * An example muscle/liver tissue-specific gene
(NCBI and bigwigs)

\textbf{Figure 2: The male \emph{Kdm5c}-KO brain expresses male and
female germline-enriched genes} * Gene ontology of testis DEGs in the
amygdala and hippocampus - ontologies are germline ontologies *
Expression of testis DEGs in germline-depleted testis (this is adult
testis data) * scRNAseq of testis - \# of testis DEGs that are
germline-specific markers * Although far fewer, 5CKO brain also
expresses ovary-enriched genes(NCBI and bigwigs of Zar1) * These ovary
enriched genes are also germline specific (NCBI/Li tissues are in adult
ovary, so it would be best to show they're oocyte-specific in adult
ovary. But I don't think there's a published adult female W/Wv dataset.
Could try looking at scRNAseq or just do TPM in embryonic W/Wv data
since oocytes are developed at this point? Or both?) * Defining what
is/isn't a germline gene, and which are male/female biased using
embryonic W/Wv data

\textbf{Figure 3: Kdm5c-KO epiblast-like cells express key drivers
regulators of germline identity} * A) ESC to EpiLC differentiation Left
- Morphology is unchanged, * B) 5CKO EpiLCs express EpiLC
differentiation genes similar to WT lvls * C) Male EpiLCs express
germline genes (example Cyct again) * Overlap between brain and EpiLC
germline genes - show they're mostly unique * GO of Brain and EpiLC
germline genes (meiotic enriched) * Bigwigs or TPM of master regulators
* Show that while some are also 2-cell genes (Dazl), 2-cell specific
genes aren't dysregulated (Zscan4). Important point because published
KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I
can tell only genes shared between germline and 2-cell are
dysregulated.\\
* Staining of Dazl (+ Stra8 if I can get it to work)

\textbf{Figure 4: Loss of KDM5C's catalytic activity impairs DNAme
placement and long-term silencing of germline genes} * Increase in
H3K4me3 in Kdm5c-KO amygdala at germline genes * Increase in H3K4me2 in
EpiLCs at germline genes * Kdm5c binding in EpiLCs vs PNCs to show that
germline repression is happening in early embryo * Previous studies only
looked at ESCs, unknown if catalytic activity is required for long-term
repression, especially since DNA methylation is placed later). KDM5C RNA
and protein ESC --\textgreater{} EpiLC (increasing then decreasing) *
RNA expression of germline genes with catalytic dead rescue (Ilakkiya) *
DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)

\textbf{Figure 5: Ectopic, germline-like phenotypes in Kdm5c-KO
ESCs/EpiLCs} * Sycp3 staining * DDX4 staining and repression of
retrotransposons * Cilia??

\end{document}

### 2024-07-04 code/GeneTSS.R
tss and tes genes

@@ -1,0 +1,44 @@
+
+## 2024.06.30 Get the transcript start and end site of each gene in the clusters for bedtools plotting of KDM5C binding
+source("code/utilities/GeneTSSandTES.R")
+
+#tss window
+WIND <- 1000
+
+#germline genes
+germ <- read.csv(snakemake@input[[1]], sep = ",")
+geneTSSandTES(germ$ENSEMBL, 1)
+geneTSSwindow(germ$ENSEMBL, 2, WIND)
+
+#test genes 
+testgenes <- c("ENSMUSG00000010592", "ENSMUSG00000020059", "ENSMUSG00000009628")
+geneTSSandTES(testgenes, 3)
+geneTSSwindow(testgenes, 4, WIND)
+
+
+#just the brain or EpiLC DEGs
+samples <- c("EpiLC", "Amygdala", "Hippocampus")
+allgermDEGs <- c()
+allupDEGs <- c()
+for (i in 1:length(samples)){
+	DEGs <- read.csv(snakemake@input[[i+1]], sep = ",")
+	#non-germline up DEGs
+	upDEGs <-subset(DEGs, Direction == "Up" & !(ENSEMBL %in% germ$ENSEMBL))
+	allupDEGs <- c(allupDEGs, upDEGs$ENSEMBL)
+
+	#germline up DEGs
+	germDEGs <-subset(DEGs, Direction == "Up" & ENSEMBL %in% germ$ENSEMBL)
+	allgermDEGs <- c(allgermDEGs, germDEGs$ENSEMBL)
+
+}
+
+allgermDEGs <- unique(allgermDEGs)
+print(length(allgermDEGs))
+geneTSSandTES(allgermDEGs, 5)
+geneTSSwindow(allgermDEGs, 6, WIND)
+
+
+allupDEGs <- unique(allupDEGs)
+print(length(allupDEGs))
+geneTSSandTES(allupDEGs, 7)
+geneTSSwindow(allupDEGs, 8, WIND)

### 2024-07-04 code/H3K4_GermlineTSS.R

@@ -1,105 +1,0 @@
-## 2024.06.27 Get the transcript start and end site of each gene in the clusters for bedtools plotting of KDM5C binding
-# #biomart doesn't work, returns the incorrect cooridnates
-# library(TxDb.Mmusculus.UCSC.mm10.knownGene)
-# txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
-# library(biomaRt)
-
-# genes <- genes(txdb)
-# mart <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
-
-# #get the list of germline genes
-# germ <- read.csv(snakemake@input[[1]], sep = ",") 
-
-# #genelist - list of ensembl gene IDs you want the window for
-# #window - # of bp upstream and downstream
-# #out - output location in snakemake for bedfile
-# TSSwindowBED <- function(genelist, window, out){
-# 	#listAttributes(mart, page = "structure")
-# 	#get the TSS for list of genes
-# 	tss <- getBM(attributes = c("chromosome_name", "transcription_start_site", "transcript_end",
-#                             "ensembl_gene_id", "start_position", "end_position", "ensembl_transcript_id", "external_gene_name"), filters = "ensembl_gene_id", values = genelist, mart = mart)
-
-# 	#add and subtract bp window and format in bed style
-# 	print(tss)
-# 	bed <- unique(data.frame(chr = paste0("chr", tss$chromosome_name), start = tss$transcription_start_site - window, end =  tss$transcription_start_site + window))
-# 	print(bed)
-# 	write.table(bed, snakemake@output[[out]], sep = "\t", row.names = FALSE, col.names=FALSE, quote = FALSE)
-
-# 	bed2 <- unique(data.frame(chr = paste0("chr", tss$chromosome_name), start = tss$start_position, end =  tss$end_position))
-# 	print(bed2)
-# 	write.table(bed2, snakemake@output[[out+1]], sep = "\t", row.names = FALSE, col.names=FALSE, quote = FALSE)
-
-# }
-
-	
-# WIND <- 1000
-# #TSSwindowBED(germ$ENSEMBL, WIND, 1)
-# testgenes <- c("ENSMUSG00000010592", "ENSMUSG00000056436", "ENSMUSG00000021758")
-
[…]

### 2024-07-04 code/utilities/GeneTSSandTES.R

@@ -1,0 +1,54 @@
+#script for getting bed files of the TSS and TES for genes of interest - input is the gtf file used in RNA/ChIPseq quantification
+
+library("rtracklayer")
+gtf <- rtracklayer::import('../../gencode.vM21.annotation.gtf') #gtf file used in ChIPseq quantification
+gtf_df <- as.data.frame(gtf)
+# print(head(gtf_df))
+
+library(dplyr)
+library(tidyr)
+print(head(gtf_df))
+gtf_df <- subset(gtf_df, type == "gene")
+
+#need to switch the start and end if the gene is on the minus strand instead of +
+gtf_df$TSS <- ifelse(gtf_df$strand == "+", gtf_df$start, gtf_df$end)
+gtf_df$TES <- ifelse(gtf_df$strand == "+", gtf_df$end, gtf_df$start)
+
+#removing the ensembl gene IDs
+gtf_df <- pivot_longer(gtf_df, cols =gene_id) %>%
+ 	separate(value, into = c('ENSEMBL', 'version'), sep = "\\.") %>%
+  	select(-name)
+
+	# gtf_df[c('ENSEMBL', 'version')] <- str_split(gtf_df$gene_id, '.')
+print('formatted gtf')
+print(head(gtf_df))
+
+#get TSS and TES for genes of interest
+#goi - genes of interest - vector of ENSEMBL gene IDs 
+#n - location of output table in snakefile
+geneTSSandTES <- function(goi, n){
+	#split the gene_id column into ensembl and variant names
+
+	#regions of interest
+	roi <- subset(gtf_df, ENSEMBL %in% goi, select = c(seqnames, TSS, TES))
+	print('regions of interest')
+	print(tail(roi))
+
+	write.table(roi, snakemake@output[[n]], sep = "\t", row.names = FALSE, col.names=FALSE, quote = FALSE)
+
+}
+
+#goi - genes of interest - vector of ENSEMBL gene IDs 
+#window - window of bp around the tss
+#n - location of output table in snakefile
+geneTSSwindow <- function(goi, n, window){
+	#regions of interest
+	roi <- subset(gtf_df, ENSEMBL %in% goi, select = c(seqnames, TSS))
+	
+	#Add and subtract the window from the TSS
+	roi$TSS_up <- roi$TSS - window
+	roi$TSS_down <- roi$TSS + window
+
[…]

### 2024-07-05 submission/manuscript.Rmd
figure legends and H3k4me23 avg binding

--- 2024-07-03 submission/manuscript.Rmd
@@
 Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
-Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test}
+Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test.}
   \label{figurelabel}
@@
 
+\begin{figure}
+  \centering 
+  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
+  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom).
+\textbf{B.} Overlap of genes with KDM5C bound to their promoters in EpiLCs (purple) and PNCs (blue).
+\textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter.
+Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared).
+\textbf{D.} Average KDM5C binding at all germline-enriched genes in EpiLCs (left) and PNCs (right).
+\textbf{E.} KDM5C binding at the promoters of RNA-seq germline DEGs.
+Genes were classified as either only dysregulated in EpiLCs (EpiLC only), genes dyresgulated in the hippocampus or amygdala but not EpiLCs (brain only), or genes dysregulated in both EpiLCs and the brain (common).
+(Legend continued on next page.)}
+  \label{figurelabel}
+\end{figure}
+\addtocounter{figure}{-1}
+\begin{figure} [t!]
+  \caption[]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} (Legend continued.) \textbf{F.} Example ChIP-seq bigwigs of DEGs unqiue to EpiLCs.
+Although both are expressed in EpiLCs, KDM5C is bound to the \textit{Dazl} promoter but not the \textit{Stra8} promoter in EpiLCs.
+\textbf{F.} Bigwigs of the upregulated imprinted gene \textit{Dlk1} that shows rescue in the double mutant brain.
+\textbf{G.} Example ChIP-seq bigwigs of DEGs common between brain and EpiLCs.
+KDM5C is bound to the \textit{D1Pas1} promoter but not the \textit{XXX} promoter in EpiLCs.
+\textbf{H.} Example ChIP-seq bigwigs of DEGs unique to the brain.
+KDM5C is bound to the \textit{XXX} promoter but not the \textit{Meig1} promoter}
+\end{figure}
+
 \newpage
+\begin{figure}
+  \centering
+  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
+  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
+Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO amygdala.
+\textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs.
+\textbf{C.} RNA and protein expression of KDM5C across ESC to EpiLC differentiation.
+Top left - diagram of differentiation protocol and collection time points.
+Bottom left - representative lanes of Western blot for KDM5C and DAXX.
+Middle - KDM5C protein expression normalized to DAXX.
+Quantified intensity using ImageJ (artificial units - au).
+Right - RT-qPCR of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}) \textbf{D.} XXX {E.} XXX.}
+  \label{figurelabel}
+\end{figure}
+
+\clearpage
 ## Figure outline:

### 2024-07-05 submission/manuscript.Rmd
begin editing intro

@@
 ```
-
-Gaps in knowledge addressed: 
-* Are other tissue-enriched genes dysregulated, or only testis, germline genes?
-* Curating a robust list of male and female germline genes
-	* Should talk about 2-cell genes vs germline genes - way to systematically categorize? 
-* Mechanism behind long-term germline gene misexpression
-	* Recent evidence suggests loss of KDM5C in ESCs express some germline genes
-		* Unclear if catalytic activity is required for long-term silencing
-		* Unclear if their dysregulation lasts throughout life or  the same between brain or not
-	* When in development does it begin?
-- Recent evidence suggests some germline genes expressed in 5CKO ESCs but unclear if their dysregulation lasts throughout differentiation and if the identity of germline genes are different compared to the brain
-* Are there functional consequences to germline gene misexpression?
-
-Introduction:
-* Chromatin regulators are important for cellular identity
-	* H3K4me1-3 linked to active gene promoters and enhancers
-	* Surprisingly, mutations in Chromatin regulators lead to many NDDs (including many H3K4 regulators)
-		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
-	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
-		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
-	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention
-
-* Loss of KDM5C can result in the misexpression of genes typically only found in the testis
-	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
-	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
-	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
-	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
-		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
-		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
-		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
-	* Systematic characterization of ectopic germline genes hasn't been done
-		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
-		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
+# Abstract
 
@@
 
-To form a complete organism, embryonic stem cells must differentiate into a myriad of discrete cellular identities.
-This is in part accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
-Although initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing technologies unexpectedly found many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators.
-This relationship is partially explained by their regulation of brain-specific genes or chromatin states, such as modulating genes involved in synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuronal and glial developmental programs[@hirabayashiPolycombLimitsNeurogenic2009].
-However, loss of some chromatin regulators can also lead to the misexpression of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Currently, very few studies have explored the misexpression of non-neuronal, tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
-To elucidate their contribution to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression.
+Embryonic stem cells employ a single genome to form a myriad of discrete cellular identities.
+This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
+Although many chromatin regulators were initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators[CIT].
+The connection between neurodevelopment and chromatin regulation can be partially explained by their regulation of brain-specific genes or chromatin states, such as orchestrating synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuron and glia during cortical development[@hirabayashiPolycombLimitsNeurogenic2009].
+However, loss of some chromatin regulators can also result in aberrant transcription of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
+Very few studies have explored the ectopic expression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
+
+To elucidate the contribution of tissue identity dysregulation to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression.
 Characterizing these features will enable us to identify molecular footholds common between NDDs that can then be exploited for potential therapeutics.
 
-In this study, we characterized the misexpression of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
+In this study, we characterized the dysregulation of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
@@
 	* Cilia??
+
+# Notes
+
+Gaps in knowledge addressed: 
+* Are other tissue-enriched genes dysregulated, or only testis, germline genes?
+* Curating a robust list of male and female germline genes
+	* Should talk about 2-cell genes vs germline genes - way to systematically categorize? 
+* Mechanism behind long-term germline gene misexpression
+	* Recent evidence suggests loss of KDM5C in ESCs express some germline genes
+		* Unclear if catalytic activity is required for long-term silencing
+		* Unclear if their dysregulation lasts throughout life or  the same between brain or not
+	* When in development does it begin?
+- Recent evidence suggests some germline genes expressed in 5CKO ESCs but unclear if their dysregulation lasts throughout differentiation and if the identity of germline genes are different compared to the brain
+* Are there functional consequences to germline gene misexpression?
+
+Introduction:
+* Chromatin regulators are important for cellular identity
+	* H3K4me1-3 linked to active gene promoters and enhancers
+	* Surprisingly, mutations in Chromatin regulators lead to many NDDs (including many H3K4 regulators)
+		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
+	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
+		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
+	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention
+
+* Loss of KDM5C can result in the misexpression of genes typically only found in the testis
+	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
+	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
+	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
+	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
+		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
+		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
+		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
+	* Systematic characterization of ectopic germline genes hasn't been done
+		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
+		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.

### 2024-07-05 submission/manuscript.Rmd
editing intro

@@
 
-Embryonic stem cells employ a single genome to form a myriad of discrete cellular identities.
+Embryonic stem cells employ a single genome to form the myriad of discrete cellular identities that make up a complete organism.
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
@@
 
-To elucidate the contribution of tissue identity dysregulation to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed, the developmental time point the dysregulation begins, and the molecular mechanism underlying their de-repression.
-Characterizing these features will enable us to identify molecular footholds common between NDDs that can then be exploited for potential therapeutics.
-
+To elucidate if dysregulation of brain tissue identity contributes to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed and the molecular mechanisms underlying their de-repression.
 In this study, we characterized the dysregulation of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
@@
 Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017].
-However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.
+However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been characterized.
+Furthermore, some of the ectopic testis transcripts have known functions in germ cells[@scandagliaLossKdm5cCauses2017], potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss.
+While there is emerging evidence that KDM5C regulates select germline genes, it is unclear which germline  _Kdm5c_-KO.
+- KDM5C's molecular mechanisms unclear - direct vs indirect
+- Developmental time point of involvement
+- How the genes change over time.
 
-Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017].
-Unlike somatic cells, germ cells (e.g.
-sperm and eggs) undergo meiosis and pass on their genetic material to the next generation.
-The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline.
-Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
-KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017].
-In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development.
-However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.
+- Systematically characterizing KDM5C's role in repressing germline identity will provide molecular footholds for testing the impact of ectopic germline gene expression on neurodevelopment.
+These can then be developed into novel therapeutics for MRXSCJ treatment and other NDDs that share a similar etiology
+**note: don't want to be too specific about therapeutics**
 
 To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes.
-The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells.
-To better characterize germline gene misexpression, we then generated a dataset of germline-enriched genes by comparing gene expression in gonads with germ cell depletion.
-We found _Kdm5c_-KO EpiLCs primarily expressed unique germline genes compared to the mature _Kdm5c_-KO brain, including _Dazl_ and _Stra8_, key drivers of germline identity and meiosis.
-While KDM5C is directly bound to some germline gene promoters in EpiLCs, it is not directly bound to many germline-enriched mRNAs expressed with _Kdm5c_-KO cells, indicating germline genes can be aberrantly transcribed through indirect mechanisms.
+We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, ovary, and testis genes.
+Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are typically only expressed in germ cells.
+While the _Kdm5c_-KO brain primarily expresses genes important for late germ cell development, _Kdm5c_-KO EpiLCs aberrantly express key drivers of germline identity and meiosis, such as _Dazl_ and _Stra8_.
+We found that KDM5C only binds a subset of germline genes expressed in the _Kdm5c_-KO brain and EpiLCs, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms.
 Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs.
@@
 
-**note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**
+- We then generated a dataset of germline-enriched genes to aid in the characterization of germline gene misexpression
+- **note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**
+- **note: I feel like I'm missing the key points still on the last paragraph**
 
@@
 		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
+
+**Germline gene repression background:**
+
+Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017].
+Unlike somatic cells, germ cells (e.g.
+sperm and eggs) undergo meiosis and pass on their genetic material to the next generation.
+The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline.
+Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017].
+In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development.
+However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.

### 2024-07-07 submission/manuscript.Rmd
cleaning up intro, need to add germline bacgkround

@@
 
-Embryonic stem cells employ a single genome to form the myriad of discrete cellular identities that make up a complete organism.
-This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
-Although many chromatin regulators were initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators[CIT].
-The connection between neurodevelopment and chromatin regulation can be partially explained by their regulation of brain-specific genes or chromatin states, such as orchestrating synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuron and glia during cortical development[@hirabayashiPolycombLimitsNeurogenic2009].
-However, loss of some chromatin regulators can also result in aberrant transcription of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Very few studies have explored the ectopic expression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
+A single genome holds the instructions to generate the myriad of cell types found within the adult organism.
+This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
+Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed their dysregulation underlies many neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
+Many studies have suggested this connection between neurodevelopment and chromatin regulators is due to their regulation of brain-specific genes, such as orchestrating the transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and the transition between neuronal and glial differentiation in neural precursor cells[@hirabayashiPolycombLimitsNeurogenic2009].
+However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
+Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
 
-To elucidate if dysregulation of brain tissue identity contributes to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed and the molecular mechanisms underlying their de-repression.
-In this study, we characterized the dysregulation of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase.
-Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
-Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016].
-Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017].
-However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been characterized.
-Furthermore, some of the ectopic testis transcripts have known functions in germ cells[@scandagliaLossKdm5cCauses2017], potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss.
+To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms underlying their de-repression.
+In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
+KDM5C is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters.
+Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
+_Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO brain unexpectedly revealed ectopic expression of testis genes within the hippocampus[@scandagliaLossKdm5cCauses2017].
+It is currently unclear if these testis genes impair _Kdm5c_-KO neurodevelopment and if this dysregulation in _Kdm5c_-KO brain tissue identity is unique to testis genes.
+
+Intriguingly, some of the ectopic testis transcripts expressed within the _Kdm5c_-KO brain have known functions in germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic and germline identity.
+
+- germ line background
+
+potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss.
 While there is emerging evidence that KDM5C regulates select germline genes, it is unclear which germline  _Kdm5c_-KO.
@@
 
-To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, ovary, and testis genes.
-Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are typically only expressed in germ cells.
-While the _Kdm5c_-KO brain primarily expresses genes important for late germ cell development, _Kdm5c_-KO EpiLCs aberrantly express key drivers of germline identity and meiosis, such as _Dazl_ and _Stra8_.
-We found that KDM5C only binds a subset of germline genes expressed in the _Kdm5c_-KO brain and EpiLCs, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms.
-Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs.
-Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.
+- an _in vitro_ model of the post-implantation embryo.
 
-- We then generated a dataset of germline-enriched genes to aid in the characterization of germline gene misexpression
-- **note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**
-- **note: I feel like I'm missing the key points still on the last paragraph**
+To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the male _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs).
+We observed general dysregulation of tissue-enriched genes the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary.
+Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are typically unique to germ cells.
+While the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_.
+KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
+Finally, we found KDM5C promotes the embryonic silencing of germline genes in somatic cells by aiding the placement of DNA methylation in EpiLCs through H3K4me3 removal.
+Thus, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+
+- **note: last sentence needs work**
 

### 2024-07-08 submission/manuscript.Rmd
discussion formatting

@@
 
-A single genome holds the instructions to generate the myriad of cell types found within the adult organism.
+A single genome holds the instructions to generate all the myriad of cell types found within the adult organism.
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
@@
 # Discussion
+Random thoughts
 - The demarcation of the germ vs soma is a key feature of multicellularity
-- Other H3K4me regulators, anything known about tissue-biased gene expression?
-- This suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
+- Anything known about tissue-biased gene expression in other H3K4me regulators?
+- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
 - tissue-biased gene expression:
 	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function.
-For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
-- Otx2 is properly expressed in _Kdm5c_-KO EpiLCs and is known to prevent pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
+For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory.
+- Otx2 is expressed in EpiLCs and is known to repress PGC identity. 
+	- It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
 
@@
 
-* Papers to read/reference:
-	* Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
-	* two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
+- Papers to read/reference:
+	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
+	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
 

### 2024-07-08 submission/manuscript.Rmd
added intro germ paragraph

@@
 Intriguingly, some of the ectopic testis transcripts expressed within the _Kdm5c_-KO brain have known functions in germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic and germline identity.
+Germ cells produce the gametes (e.g.
+sperm and eggs) that pass on their genetic material to the next generation while the somatic cells perform all the other bodily functions.
+Distinguishing between soma and germline is a key feature of multicellularity and occurs in mammals soon after the embryo is implanted into the uterine wall.
+Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to suppress their dysregulation[@scandagliaLossKdm5cCauses2017].
+In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+However, KDM5C binding and _Kdm5c_-KO germline gene misexpression has yet to be globally characterized during early embryogenesis.
+Given that _Dazl_ and other germline genes can be expressed in ESCs and in the 2-cell stage, it is unclear if KDM5C has a direct role in long-term germline gene silencing in post-implantation epiblast.
+Systematically characterizing KDM5C's role in germline gene repression will unveil key mechanisms underlying of soma versus germline identity and provide molecular footholds for testing the impact of ectopic germline gene expression on neurodevelopment.
 
-- germ line background
-
-potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss.
-While there is emerging evidence that KDM5C regulates select germline genes, it is unclear which germline  _Kdm5c_-KO.
-- KDM5C's molecular mechanisms unclear - direct vs indirect
-- Developmental time point of involvement
-- How the genes change over time.
-
-- Systematically characterizing KDM5C's role in repressing germline identity will provide molecular footholds for testing the impact of ectopic germline gene expression on neurodevelopment.
-These can then be developed into novel therapeutics for MRXSCJ treatment and other NDDs that share a similar etiology
-**note: don't want to be too specific about therapeutics**
-
-- an _in vitro_ model of the post-implantation embryo.
-
-To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the male _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs).
+To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the male _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
 We observed general dysregulation of tissue-enriched genes the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary.
@@
 Thus, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-
-- **note: last sentence needs work**
 
@@
 \clearpage
+# Notes
+
 ## Figure outline:
@@
 	* Cilia??
-
-# Notes
 

### 2024-07-08 submission/manuscript.Rmd
editing results fig 1 and 2

@@
 
-A single genome holds the instructions to generate all the myriad of cell types found within the adult organism.
+A single genome holds the instructions to generate the myriad of cell types found within the adult organism.
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
-Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed their dysregulation underlies many neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
-Many studies have suggested this connection between neurodevelopment and chromatin regulators is due to their regulation of brain-specific genes, such as orchestrating the transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and the transition between neuronal and glial differentiation in neural precursor cells[@hirabayashiPolycombLimitsNeurogenic2009].
-However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.
+Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018].
+Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009].
+However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if this partial loss of brain identity contributes to neurodevelopmental impairments.
 
-To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms underlying their de-repression.
-In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
-KDM5C is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters.
+To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms governing their de-repression.
+In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C).
+KDM5C, , also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters.
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
 _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO brain unexpectedly revealed ectopic expression of testis genes within the hippocampus[@scandagliaLossKdm5cCauses2017].
-It is currently unclear if these testis genes impair _Kdm5c_-KO neurodevelopment and if this dysregulation in _Kdm5c_-KO brain tissue identity is unique to testis genes.
+Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
+It is currently unknown if this dysregulation of brain tissue identity further impairs _Kdm5c_-KO neurodevelopment and if ectopic gene expression within the _Kdm5c_-KO brain is unique to testis genes.
 
-Intriguingly, some of the ectopic testis transcripts expressed within the _Kdm5c_-KO brain have known functions in germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic and germline identity.
-Germ cells produce the gametes (e.g.
-sperm and eggs) that pass on their genetic material to the next generation while the somatic cells perform all the other bodily functions.
-Distinguishing between soma and germline is a key feature of multicellularity and occurs in mammals soon after the embryo is implanted into the uterine wall.
-Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
-KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to suppress their dysregulation[@scandagliaLossKdm5cCauses2017].
+Intriguingly, some of the ectopic testis transcripts identified within the _Kdm5c_-KO hippocampus have known functions unique to germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic versus germline identity.
+Distinguishing between the germline (cells that pass on their genetic material, e.g.
+sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis.
+In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells as the embryo implants into the uterine wall and transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in knockout neurons fails to suppress their dysregulation[@scandagliaLossKdm5cCauses2017].
 In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
 However, KDM5C binding and _Kdm5c_-KO germline gene misexpression has yet to be globally characterized during early embryogenesis.
-Given that _Dazl_ and other germline genes can be expressed in ESCs and in the 2-cell stage, it is unclear if KDM5C has a direct role in long-term germline gene silencing in post-implantation epiblast.
-Systematically characterizing KDM5C's role in germline gene repression will unveil key mechanisms underlying of soma versus germline identity and provide molecular footholds for testing the impact of ectopic germline gene expression on neurodevelopment.
+Given that _Dazl_ and other germline-enriched genes can also be expressed in ESCs and at the 2-cell stage, it is unclear if KDM5C has a direct role in the long-term germline gene silencing that occurs in the post-implantation epiblast.
+Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
 
-To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the male _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We observed general dysregulation of tissue-enriched genes the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary.
-Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are typically unique to germ cells.
+To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary.
+Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are typically uniquely expressed in germ cells.
 While the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_.
 KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
-Finally, we found KDM5C promotes the embryonic silencing of germline genes in somatic cells by aiding the placement of DNA methylation in EpiLCs through H3K4me3 removal.
-Thus, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation in EpiLCs through H3K4me2/3 removal.
+Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
 
@@
 
-Previous RNA sequencing (RNA-seq) in the adult hippocampus ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
+Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
 Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, it is currently unclear if this erosion of brain tissue identity is a major consequence of _Kdm5c_ loss and if it is unique to testis-enriched genes.
-Therefore, we first globally assessed the expression of previously characterized mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the amygdala and hippocampus in adult mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
+Therefore, we first globally assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 
-We found a large proportion of genes significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
 The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C).
-Testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact Test), even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes).
-Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact Test).
+One example of a testis-enriched DEG is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
 
 In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain.
-Although the _Kdm5c_-KO mice were male, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
-Intriguingly, many ovary and testis-biased DEGs have functions specific to germ cells and have no known role in the brain.
-For example, the testis-biased DEG _FK506 binding protein 6 (Fkbp6)_ is a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C) while the ovary-enriched DEG _Zygotic arrest 1 (Zar1)_ was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Although not consistent across brain regions, we also found enrichment of two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Fisher's Exact Test).
-An example of a liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).
-Testis, ovary, and liver-enriched DEGs showed little to no expression in the wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
-Together, these results suggest misexpression of testis and other tissue-enriched genes within the brain is a major effect of KDM5C loss.
+Although the _Kdm5c_-KO mice were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
+Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Fisher's Exact Test).
+An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).
+Testis, ovary, and liver-enriched DEGs showed little to no expression in the developing and adult wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
+Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+Together, these results suggest misexpression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
 
-### Germline genes are aberrantly expressed in the male _Kdm5c_-KO brain
 ```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
@@
 ```
+### Germline genes are misexpressed in the _Kdm5c_-KO brain
 
@@
 Leydig cells) that support hormone production and germline functions.
-Select testis-enriched DEGs that were characterized previously had germline-specific functions[@scandagliaLossKdm5cCauses2017], suggesting _Kdm5c_-KO cells fail to demarcate between the soma and germline.
-To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assed their function through gene onotology.
-We found high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
+Intriugingly, many _Kdm5c_-KO testis and ovary enriched-DEGs have germline-specific functions, suggesting _Kdm5c_-KO cells fail to distinguish between the soma and germline.
+To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assed their known functions through gene onotology analysis.
+We found _Kdm5c_-KO testis-enriched DEGs high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
-To further validate if these testis DEGs are truly germline genes, we then compared their expression in somatic versus germ cells within the testis.
-We first compared their expression within wild-type versus germ cell-depleted testes[@muellerIndependentSpecializationHuman2013].
-In this study, germ cell depletion was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^), which prevents the maturation of germ cells and results in overall germline loss[@handelSertoliCellDifferentiation1979].
+To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in somatic versus germ cells within the testis.
+We first compared their expression in the testis with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^) that prevent the maturation of germ cells[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
@@
 
-We wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes.
-To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in germline-depleted (Kit^W/Wv^) mice.
-Current available Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
-
-We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes with loss of _Kdm5c_ (Supplementary table 1).
+We then wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes.
+To facilitate downstream analyses, we generated a curated list of germline-enriched genes using currently available RNA-seq datasets in Kit^W/Wv^ mice.
+Wild-type and Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
+We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any adult wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
 
@@
 zscan4d <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]
+duxf3 <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['padj'][,1]
 
@@
 
-Misexpression of germline genes in the adult _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity.
+Misexpression of germline genes in the adult brain suggests _Kdm5c_-KO cells fail to demarcate between germline and somatic identity.
 Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
 This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top).
-Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
+Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
+Therefore, we assessed if KDM5C was necessary for embryonic silencing of germline genes by evaluating the impact of _Kdm5c_ loss in EpiLCs.
 
-We first identified _Kdm5c_-KO EpiLC DEGs through DESeq2 analysis of our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (log2 fold change > `r l2fcco`, q < 0.1).
+We first identified _Kdm5c_-KO EpiLC DEGs in of our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression.
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
-Using our curated list, we found _Kdm5c_-KO EpiLCs aberrantly expressed `r format(nrow(epigerm))`germline-enriched genes, including the previously characterized hippocampal DEG[@scandagliaLossKdm5cCauses2017] _Cytochrome C, testis-specific (Cyct)_ (Figure 3C).
+Using the curated list of germline genes generated above, we found _Kdm5c_-KO EpiLCs aberrantly expressed `r format(nrow(epigerm))` germline-enriched genes, including the previously characterized hippocampal DEG[@scandagliaLossKdm5cCauses2017] _Cytochrome C, testis-specific (Cyct)_ (Figure 3C).
 Although we observed aberrant expression of many tissue-enriched genes, we did not observe any significant difference in primed pluripotency genes (Figure 3D) or gross changes in _Kdm5c_-KO cell morpholgy during differentiation (Figure 3E), indicating KDM5C loss does not impair EpiLC formation.
 
-To evaluate if all germline DEGs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
-The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across all sequencing datasets (Figure 3F).
-We then compared the function of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle (GO: 0051321, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus showed strong enrichment of late-stage sperm genes, such those involved in the sperm axoneme.
-This shift from meiotic genes to later spermatogenesis genes in the hippocampus suggests the germline developmental program could occur ectopically as _Kdm5c_-KO cells progresses through somatic tissue development.
-**note: this is strengthened by the ChIP-seq data since KDM5C is not directly bound to many brain/flagellar DEGs.This point might be stronger in the ChIPseq figure**
+We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if all germline DEGs, like _Cyct_, are constitutively dysregulated or if they can change over the course of development.
+We found this was primarily not the case, as the majority of germline DEGs expressed in EpiLCs and the brain were unique, with only _Cyct_ shared across all sequencing datasets (Figure 3F).
+We then compared the known functions of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.
 
@@
 These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
-Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo.
-However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).
+Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
+However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).
 
-_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019].
+We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019].
 A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
 This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function.
-We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3H).
+We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I).
 We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020].
@@
 - Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
+	- This shift from meiotic genes to later spermatogenesis genes in the hippocampus suggests the germline developmental program could occur ectopically as _Kdm5c_-KO cells progresses through somatic tissue development.
+**note: this is strengthened by the ChIP-seq data since KDM5C is not directly bound to many brain/flagellar DEGs.This point might be stronger in the ChIPseq figure**
 
@@
   \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
-  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO amygdala.
+  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
 Left - MA plot of mRNA-sequencing.
 Right - Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{B.} Expression of tissue-enriched genes in the male \textit{Kdm5c}-KO hippocampus.
-\textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{Cytochrome C, testis-specific (Cyct)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
 Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.

### 2024-07-08 submission/manuscript.Rmd
update fig 1-3 figure legends

@@
   \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
-  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis.
+  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis (Mueller et al 2013).
 Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM).
@@
   \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
-  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
-Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
-Bottom - representative images of wild-type (WT) and \textit{Kdm5c}-KO ESC to EpiLC differentiation.
-Brightfield images taken at 20X.
-\textbf{B.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs.
+  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
+Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
+\textbf{B.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs.
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
+\textbf{C.}  Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs.
+\textbf{D.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs.
 Welch's t-test, expression in transcripts per million (TPM).
-\textbf{C.} Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
-\textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs.
-\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
-\textbf{F.} enrichPlot comparing enriched gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
-\textbf{G.} Top - Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test.
-Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
-Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test.}
+\textbf{E.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation.
+Brightfield images taken at 20X.\textbf{F.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
+\textbf{G.} enrichPlot comparing enriched biological process gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
+\textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs.
+Top right - Example 2-cell genes that are not dysregulated in \textit{Kdm5c}-KO EpiLCs.
+p-values for Welch's t-test.
+Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{I.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
+Percentage of DAZL-positive cells normalized to number of DAPI-positive cells, p-value for Welch's t-test.}
   \label{figurelabel}

### 2024-07-09 submission/manuscript.Rmd
editing results/figs 4-5

@@
 
-- **note: do Direct vs indirect DEGs motif analysis**
-- However, it is currently unclear if KDM5C binds to all germline DEGs and if its binding is maintained at any germline genes in neurons.
-
 Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress hippocampal germline DEGs and KDM5C is not bound to their promoters in neurons[@scandagliaLossKdm5cCauses2017].
-There is some evidence KDM5C binds to select germline gene promoters in ESCs[@scandagliaLossKdm5cCauses2017], including two recent independent screens that found KDM5C binds to Dazl's promoter[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-As KDM5C's binding at germline gene promoters has not been systematically characterized, it is currently unclear what types of germline genes KDM5C regulates and if its binding is maintained at any germline genes in neurons.
+There is some evidence KDM5C binds to select germline gene promoters in ESCs[@scandagliaLossKdm5cCauses2017], including _Dazl_[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+As KDM5C's binding at germline gene promoters has not been systematically characterized, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
 
@@
 To further characterize the mechanism behind _Kdm5c_-KO germline gene misexpression, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016].
-EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`kb from TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).
+EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).
 
 The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 4B).
-We then performed gene ontology analysis to compare the functions of genes bound by KDM5C at the promoter in different cell types.
-While there were no significant ontologies for genes with KDM5C only bound in PNCs, gene ontologies for peaks shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C).
-Germline-specific ontologies were only enriched in peaks unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C).
-When comparing KDM5C binding at all germline gene promoters, KDM5C was only bound to a subset of germline gene promoters in EpiLCs and was not bound to any in PNCs (Figure 4D).
-Together, this suggests KDM5C is recruited to a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not regulate germline genes in neurons.
+We then performed gene ontology analysis to compare the known functions of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
+While there were no ontologies significantly enriched for genes only bound by KDM5C in PNCs, gene ontologies for genes shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C).
+Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C).
+We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
+While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D).
+Together, this suggests KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not regulate germline genes in neurons.
 
-To determine if the germline mRNAs expressed in the _Kdm5c_-KO brain and EpiLCs are direct targets of KDM5C, we then compared KDM5C binding at RNA-seq DEG promoters (Figure 4E).
-About one third of EpiLC specific and brain-specific (hippocampus, amygdala, or both) germline DEGs were bound by KDM5C in EpiLCs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
+We then compared KDM5C binding at RNA-seq DEG promoters to determine if the germline mRNAs expressed in the _Kdm5c_-KO brain and EpiLCs are direct targets of KDM5C (Figure 4E).
+About one third of EpiLC-specific and brain-specific (hippocampus, amygdala, or both) germline DEGs were bound by KDM5C in EpiLCs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
 Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above.
 Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F).
-In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs bound by KDM5C (Figure 4G).
+In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G).
 Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H).
 Altogether, this suggests the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters during embryogenesis.
+
+**notes:**
+- do Direct vs indirect DEGs motif analysis
 
@@
 Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
-Somatic repression of germline genes is typically established during the transition between naive and primed pluripotency, which modeled in vitro as ESCs to EpiLC differentiation.
+Somatic repression of germline genes is typically established during the transition between naive and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.
 In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation[@mochizukiRepressionGermlineGenes2021].
-It has been proposed KDM5C may promote germline gene silencing via H3K4me3 removal since DNA methylation is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017] and H3K4me3 can impair DNA methylation placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
-However, KDM5C was recently shown to repress _Dazl_ in ESCs indepedent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021].
+Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017].
 Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.
@@
 In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 5A).
-We additionally found a marked increase in H3K4me2 germline gene TSSs in _Kdm5c_-KO EpiLCs (Figure 5B).
-Incerase in H3K4me2 and H3K4me3 was highest in _Kdm5c_-KO cells that were highest genes that are bound by KDM5C at their promoter in EpiLCs (**note: do analysis to check if true**).
+We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 5B).
 
-To assess KDM5C's embryonic role in germline gene silencing, we first characterized KDM5C's expression in ESCs and EpiLCs by harvesting RNA and protein at 0 hours (ESCs), 24 hours, and 48 hours (EpiLCs).
-While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation, KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5C).
+We then evaluated KDM5C's embryonic role in germline gene silencing during ESC to EpiLC differentiation.
+We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C).
+While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E).
+We then perfomed whole genome bislufite sequencing in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters.
+While wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).
 
-We then determined the role of KDM5C in the initial placement of DNA methylation at germline gene promoters by performing whole genome bislufite sequencing in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) (**note: check**).
-- While wild-type cells accumulate high levels of DNA methylation over the course of ESCs to exEpiLC differentiation, DNA methylation is markedly reduced in _Kdm5c_-KO
+- Catalytic activity
+- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via removal of H3K4me2/3 during early embryogenesis.
+
+- **note:** should I test if incerase in H3K4me2 and H3K4me3 was highest in _Kdm5c_-KO cells at genes bound by KDM5C at their promoter in EpiLCs? Don't think it's that impactful
 
 # Discussion
-Random thoughts
+
 - The demarcation of the germ vs soma is a key feature of multicellularity
@@
 \textbf{B.} Overlap of genes with KDM5C bound to their promoters in EpiLCs (purple) and PNCs (blue).
-\textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter.
+\textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
 Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared).
-\textbf{D.} Average KDM5C binding at all germline-enriched genes in EpiLCs (left) and PNCs (right).
+\textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right).
 \textbf{E.} KDM5C binding at the promoters of RNA-seq germline DEGs.
@@
 \begin{figure} [t!]
-  \caption[]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} (Legend continued.) \textbf{F.} Example ChIP-seq bigwigs of DEGs unqiue to EpiLCs.
-Although both are expressed in EpiLCs, KDM5C is bound to the \textit{Dazl} promoter but not the \textit{Stra8} promoter in EpiLCs.
-\textbf{F.} Bigwigs of the upregulated imprinted gene \textit{Dlk1} that shows rescue in the double mutant brain.
-\textbf{G.} Example ChIP-seq bigwigs of DEGs common between brain and EpiLCs.
+  \caption[]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} (Legend continued.) \textbf{F.} Example KDM5C ChIP-seq bigwigs of DEGs unqiue to EpiLCs.
+Although both are expressed in \textit{Kdm5c}-KO EpiLCs, KDM5C is only bound to the \textit{Dazl} promoter and not the \textit{Stra8} promoter in EpiLCs.
+\textbf{G.} Example KDM5C ChIP-seq bigwigs of RNA-seq DEGs common between the brain and EpiLCs.
 KDM5C is bound to the \textit{D1Pas1} promoter but not the \textit{XXX} promoter in EpiLCs.
-\textbf{H.} Example ChIP-seq bigwigs of DEGs unique to the brain.
+\textbf{H.} Example KDM5C ChIP-seq bigwigs of RNA-seq DEGs unique to the brain.
 KDM5C is bound to the \textit{XXX} promoter but not the \textit{Meig1} promoter}
@@
   \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
-Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO amygdala.
+Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala.
 \textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs.
-\textbf{C.} RNA and protein expression of KDM5C across ESC to EpiLC differentiation.
-Top left - diagram of differentiation protocol and collection time points.
-Bottom left - representative lanes of Western blot for KDM5C and DAXX.
-Middle - KDM5C protein expression normalized to DAXX.
+Right - Average H3K4me2 at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs.
+\textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein.
+\textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}).
+* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test.
+\textbf{E.} KDM5C protein expression normalized to DAXX.
 Quantified intensity using ImageJ (artificial units - au).
-Right - RT-qPCR of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}) \textbf{D.} XXX {E.} XXX.}
+Right - representative lanes of Western blot for KDM5C and DAXX.
+* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} XXX {G.} XXX.}
   \label{figurelabel}

### 2024-07-11 code/TissueGenes_AmyHip.R
fixed fisher so odds ratio is correct

@@ -44,7 +44,6 @@ print(head(testisDEGs))
 write.table(testisDEGs, snakemake@output[[8]], sep = ",", row.names = FALSE)
 
 
-
 ### make a histogram of the total number of tissue specific genes
 p <- ggbarplot(countingdf, "tissue", "count", fill = "tissue", color = "tissue", palette = tissue_palette, 
 		label = TRUE, lab.pos = "out", lab.col = "black") + rotate_x_text(45) 

### 2024-07-11 code/utilities/tissueSPECIFICgenes.R

@@ -244,6 +244,7 @@ tissue_fisher <- function(td, n, title){
 	#for every tissue, count how many are DEGs
 	pvalues <- c()
 	DEGnumb <- c()
+	oddsratio <- c()
 	cnt <- 1
 	for (t in unique(Tissues$tissue)){
 		Tissues_genes <- subset(Tissues, tissue == t)
@@ -256,7 +257,7 @@ tissue_fisher <- function(td, n, title){
 
 		tissue_yes_DEG_no <- nrow(subset(Tissues_genes, !(ENSEMBL %in% DEGs$ENSEMBL)))
 
-		fisherdf <- data.frame("tissue_no" = c(tissue_no_DEG_no, tissue_no_DEG_yes), "tissue_yes" = c(tissue_yes_DEG_no, tissue_yes_DEG_yes), row.names = c("DEG_no", "DEG_yes"))
+		fisherdf <- data.frame("tissue_yes" = c(tissue_yes_DEG_yes, tissue_yes_DEG_no), "tissue_no" = c(tissue_no_DEG_yes, tissue_no_DEG_no), row.names = c("DEG_yes", "DEG_no"))
 		
 		fishtest <- fisher.test(fisherdf)
 		# print(paste0(title, " fisher results for ", t))
@@ -265,13 +266,16 @@ tissue_fisher <- function(td, n, title){
 
 		pvalues[cnt] <- fishtest$p.value
 		DEGnumb[cnt] <- tissue_yes_DEG_yes
+		oddsratio[cnt] <- fishtest$estimate[[1]]
+
 		cnt <- cnt + 1
 
 	}
 		print(paste0(title, " fisher results"))
-		fishp <- data.frame(tissue = unique(Tissues$tissue), pvalues = pvalues, DEGnumb = DEGnumb)
+		fishp <- data.frame(tissue = unique(Tissues$tissue), pvalues = pvalues, DEGnumb = DEGnumb, oddsratio = oddsratio)
 		names(fishp)[names(fishp) == 'pvalues'] <- paste0(title, '_pvalues')
 		names(fishp)[names(fishp) == 'DEGnumb'] <- paste0(title, '_DEGnumb')
+		names(fishp)[names(fishp) == 'oddsratio'] <- paste0(title, '_oddsratio')
 		print(fishp)
 		return(fishp)
 

### 2024-07-11 submission/manuscript.Rmd
add odds ratio to manuscript new fisher

@@
 The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C).
-Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Fisher's Exact Test).
+Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
 One example of a testis-enriched DEG is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
@@
 In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain.
-Although the _Kdm5c_-KO mice were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+Although the _Kdm5c_-KO mice were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Fisher's Exact Test).
+Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)` Fisher's Exact Test).
 An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).

### 2024-07-11 submission/manuscript.Rmd
add odds ratio to manuscript new fisher

@@
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)` Fisher's Exact Test).
+Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test).
 An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).

### 2024-07-19 submission/manuscript.Rmd
outlining results section

@@
 
-Intriguingly, some of the ectopic testis transcripts identified within the _Kdm5c_-KO hippocampus have known functions unique to germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic versus germline identity.
+- one key process chrom reg are involved in are is distinguishing bw germ + soma cell identity
+
 Distinguishing between the germline (cells that pass on their genetic material, e.g.
 sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis.
-In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells as the embryo implants into the uterine wall and transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
-KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in knockout neurons fails to suppress their dysregulation[@scandagliaLossKdm5cCauses2017].
-In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-However, KDM5C binding and _Kdm5c_-KO germline gene misexpression has yet to be globally characterized during early embryogenesis.
-Given that _Dazl_ and other germline-enriched genes can also be expressed in ESCs and at the 2-cell stage, it is unclear if KDM5C has a direct role in the long-term germline gene silencing that occurs in the post-implantation epiblast.
+In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
 Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
@@
 ### Germline genes are misexpressed in the _Kdm5c_-KO brain
+- Intriguingly, some of the ectopic testis transcripts identified within the _Kdm5c_-KO hippocampus have known functions unique to germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic versus germline identity.
 
@@
 Intriugingly, many _Kdm5c_-KO testis and ovary enriched-DEGs have germline-specific functions, suggesting _Kdm5c_-KO cells fail to distinguish between the soma and germline.
-To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assed their known functions through gene onotology analysis.
+To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assessed their known functions through gene onotology analysis.
 We found _Kdm5c_-KO testis-enriched DEGs high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
@@
 ## KDM5C binds to a subset of germline gene promoters during early embryogenesis
+
+- KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in knockout neurons fails to suppress their dysregulation[@scandagliaLossKdm5cCauses2017].
 
@@
 
-- The demarcation of the germ vs soma is a key feature of multicellularity
+** overall outline **
+- Tissue identity
+	- We identified multiple tissue-specific genes in the brain
+	- some non-brain, tissue-enriched genes are lowly expressed do have brain-specific functions.
+	- KDM5C might be important for fine-tuning their expression to match the level important in the tissue. 
+	- this could be an avenue to impact _Kdm5c_-KO brain function.
+
+- testis genes are germline genes, germline gene list
+	- While a select few tissue-enriched DEGs had known roles in the brain, the majority were testis genes with no known function outside the germline. 
+	- Studies on chromatin regulators in germline gene repression has focused on specific genes 
+	- Curated a list of germline-enriched genes to robustly assess the types of genes dysregulated and KDM5C's role
+
+- KDM5C represses germline genes during early embryogenesis
+	- Found this dysregulation in cellular identity begins during _Kdm5c_-KO embryogenesis
+	- EpiLC express early regualtors, Brain late sperm - program seems to be continuing across development
+	- Importantly, even though germline genes are misexpressed, markers of EpiLC differentiation and EpiLC morphology is unimpaired. 
+		- Epiblast can become primoridal germ cells
+		- Otx2 is expressed in EpiLCs and is known to repress PGC identity.
+It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
+		- Germline Program happening in the background of typical development.
+
+- One of the genes misexpressed in _Kdm5c_-KO EpiLCs is Dazl, _Deleted in azoospermia like (Dazl)_, a key regulator of germline development that promotes the translation of germline-specific transcripts[XXX].
+	- 2 other studies of dazl regulators in ESCs also found KDM5C represses dazl. 
+	- While crucial for germ cell development, Dazl is also expressed in ESCs and at the 2-cell stage. 
+	- Some genes important for germ cell, ESC, and 2-cell development given they are important for self-renewal.  
+	- We did not observe dysregulation of 2-cell-specific regulators like Dux. 
+	- We found Dazl was expressed in _Kdm5c_-KO EpiLCs, when _Dazl_ it typically decommissioned. This indicates KDM5C loss results in long-term ectopic expression.
+	- We additionally found DAZL protein is ectopically expressed and localized to the cytoplasm.
+Given that a significant number of KDM5C-KO gemrline DEGs are DAZL targets, it may promote the ectopic germline program.
+
+	- Very recently two other studies identified KDM5C represses Dazl in ESCs. 
+	- - KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+However, KDM5C binding and _Kdm5c_-KO germline gene misexpression has yet to be globally characterized during early embryogenesis.
+Given that _Dazl_ and other germline-enriched genes can also be expressed in ESCs and at the 2-cell stage, it is unclear if KDM5C has a direct role in the long-term germline gene silencing that occurs in the post-implantation epiblast.
+
+- We globally characterized KDM5C's repression of germline genes during early embryogenesis and in mature neurons. 
+	- In line with previous work, we found KDM5C did not regulate germline genes in neurons. 
+	- KDM5C-bound genes in EpiLCs were enriched for germline ontologies, suggesting a major role of KDM5c during embryogenesis
+	- While KDM5C directly binds some germline genes, including _Dazl_, many of the genes dysregulated in _Kdm5c_-KO were not direct targets. 
+		- Stra8 notable exception
+		- Brain genes, including late-stage spermatogenesis genes
+		- Suggests germline genes can be dysregulated direct and indirect of KDM5C resgulation
+		- Further supports germline programs can be ectopically activated
+
+- Germline genes typcially decomissioned during ESCs and EpiLC differentiation. 
+	- Initially repressive histone marks are placed in ESCs, then DNA methylation is placed in EpiLCs.
+	- Dysregulation of germline genes later in life and in EpiLCs indicates KDM5C is involved in this processes, but unclear when. 
+	- Recently in ESCs KDM5C's catalytic acivity was found to be unnecessary for dazl suppression.  
+	- Some evidence that KDM5C is important for DNAme, as DNAme is reduced at two germ gene promoters in the adult 5CKO hippocampus.
+	- We found KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing.
+
+- Altogether, this suggests KDM5C plays a pivotal role in the development of tissue identity, including the soma-germline distinction.
+
+- In this study we demonstrated KDM5C play a pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the _Kdm5c_-KO hippocampus and amygdala.
+	- In addition to testis genes identified previously[XXX], we also observed significant enrichment of muscle, liver, and even ovary-biased genes aberrantly transcribed within the male brain.
+
+	- Many liver and muscle-biased DEGs are weakly expressed in the brain and have known roles in brain function, such as _Apolipoprotein C-I (Apoc1)_, a liver-biased DEG that important for lipoprotein metabolism but has also been shown to influence learning and memory[XXX].
+
+	- This suggests tissue-enriched genes that are lowly expressed in the brain might be more sensitive to dysregulation with KDM5C loss.
+	- increased propensity for dysregulation 
+	- However, the majority of tissue-enriched DEGs are testis and ovary genes that have known functions in the germline and no known function in the brain.
+
+- Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
+- Studies on other chromatin regulators have focused on key marker genes to identify germline gene misexpression, such as _Dazl_. 
+- Using a germline-depleted mouse model, we curated a list of mouse germline-enriched genes to globally assess germline gene dysregulation. 
+- Many of the genes dysregulated in the mature _Kdm5c_-KO brain were important for late stages of spermatogenesis, such as those important for sperm flagellar structure.
+Contrastingly, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germ cell identity, including _Dazl_ and _Stra8_.
+
+- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
+- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
+-
+
+While Directly binds to a subset of germline genes during early embryogenesis 
+- KDM5C binds to a subset of germline genes during early embryogenesis
+
+- The including the demarcation between soma and germline fates.
+
+	the _Kdm5c_-KO brain also 
+	Previous work identified testis-enriched genes in _Kdm5c_ genes
+	- 
+	- However unlike the gonadal-biased DEGs,
+
+- The demarcation of the germline versus soma is a key feature of multicellularity and sexually dimorphic reproduction
+
 - Anything known about tissue-biased gene expression in other H3K4me regulators?
@@
 - tissue-biased gene expression:
-	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function.
-For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory.
-- Otx2 is expressed in EpiLCs and is known to repress PGC identity. 
-	- It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
 

### 2024-07-19 code/ChIPseq_KDM5C_EpiLC_PNC.R
list of germline genes bound by KDM5C vs not

@@ -116,20 +116,6 @@ for (k in samples){
 
 }
 
-germ <- read.csv(snakemake@input[[6]], sep = ",")
-print("germ genes")
-print(head(germ))
-
-print("genestatus")
-genestatus <- merge(genestatus, germ, by = "ENSEMBL")
-print(head(genestatus))
-
-#save all of the bound and unbound DEGs in EpiLCs
-write.table(genestatus, snakemake@output[[5]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
-
-#save list of genes for HOMER
-write.table(subset(genestatus, KDM5C_binding == "Bound")[,"SYMBOL"], snakemake@output[[6]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
-write.table(subset(genestatus, KDM5C_binding == "Unbound")[,"SYMBOL"], snakemake@output[[7]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
 
 #plotting data frame
 print(qlotdf)
@@ -152,3 +138,33 @@ for (k in 1:length(samples)){
 
 ggsave(snakemake@output[[4]], grid.arrange(grobs = plots, ncol = 2), width = 7.5, height = 3.5)
 
+
+### Which germline genes are bound by KDM5C
+germ <- read.csv(snakemake@input[[6]], sep = ",")
+# print("germ genes")
+# # print(head(germ))
+
+# print("genestatus")
+genestatus <- merge(genestatus, germ, by = "ENSEMBL")
+# print(head(genestatus))
+
+#save all of the bound and unbound DEGs in EpiLCs
+write.table(genestatus, snakemake@output[[5]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
+#save list of genes for HOMER
+write.table(subset(genestatus, KDM5C_binding == "Bound")[,"SYMBOL"], snakemake@output[[6]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
+write.table(subset(genestatus, KDM5C_binding == "Unbound")[,"SYMBOL"], snakemake@output[[7]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
+
+
+#all germline genes, which ones are kdm5c bound to in EpiLCs
+germ$KDM5C_binding <- ifelse(germ$ENSEMBL %in% peakENSEMBL[["EpiLC"]], "Bound", "Unbound")
[…]

### 2024-07-20 code/ChIPseq_KDM5C_EpiLC_PNC.R
euler plot for germline gene binding

@@ -168,3 +168,17 @@ write.table(subset(germ, KDM5C_binding == "Unbound")[,"SYMBOL"], snakemake@outpu
 
 
 #make eulerr of kdm5c binding at all germline genes
+library("eulerr")
+source("code/utilities/colorpalettes.R")
+
+#genes bound in epilc that are not germline genes
+allepi <- peakENSEMBL[["EpiLC"]]
+epilc_nogerm <- allepi[!allepi %in% germ$ENSEMBL]
+
+
+together <- euler(c("Bound_EpiLCs" = length(epilc_nogerm), "Germline" = nrow(subset(germ, KDM5C_binding == "Unbound")), "Bound_EpiLCs&Germline" = nrow(subset(germ, KDM5C_binding == "Bound"))))
+
+p <- plot(together, quantities = TRUE, labels = list(font = 4), fills = c(EpiLC_XY_KO, germcolor))
+
+library("ggplot2")
+ggsave(snakemake@output[[11]], plot = p, width = 4, height = 3)

### 2024-07-20 submission/manuscript.Rmd
table notes
drafting discussion

@@
 	- While a select few tissue-enriched DEGs had known roles in the brain, the majority were testis genes with no known function outside the germline. 
+	- distinguishing between germline and soma a key feature of multicellularity, chromatin regulators are important for this process
 	- Studies on chromatin regulators in germline gene repression has focused on specific genes 
@@
 - KDM5C represses germline genes during early embryogenesis
-	- Found this dysregulation in cellular identity begins during _Kdm5c_-KO embryogenesis
+	- Found this dysregulation in tissue identity begins during _Kdm5c_-KO early embryogenesis
 	- EpiLC express early regualtors, Brain late sperm - program seems to be continuing across development
@@
 		- Further supports germline programs can be ectopically activated
+		- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
+			- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
 
-- Germline genes typcially decomissioned during ESCs and EpiLC differentiation. 
-	- Initially repressive histone marks are placed in ESCs, then DNA methylation is placed in EpiLCs.
-	- Dysregulation of germline genes later in life and in EpiLCs indicates KDM5C is involved in this processes, but unclear when. 
-	- Recently in ESCs KDM5C's catalytic acivity was found to be unnecessary for dazl suppression.  
-	- Some evidence that KDM5C is important for DNAme, as DNAme is reduced at two germ gene promoters in the adult 5CKO hippocampus.
-	- We found KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing.
+- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from naive to primed pluripotency.
+	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
+	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.
+		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
+	- However, Recently KDM5C's catalytic acivity was found to be unnecessary for dazl suppression in ESCs.  
+	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes.
 
+- In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain.
+	- In EpiLCs, KDM5C directly represses key drivers of germline identity like Dazl, likely through PRC1.6 recruitment and promoting CpG methylation.
+However, _Kdm5c_-KO also ectopically expresses germline genes activated indirectly, including _Stra8_.
+	- The germline developmental program to some extent continues ectopically during _Kdm5c_-KO development, resulting in aberrant transcription of late stage spermatogenesis genes later in life. 
+	- These results define KDM5C's role in the demarcation between soma and gemrline identity and offer a window into potential targets to assess the deletirous ecfects these ectopic genes on neurodevelopment.
+
+## Discussion notes
 - Altogether, this suggests KDM5C plays a pivotal role in the development of tissue identity, including the soma-germline distinction.
@@
 # Figures and Tables
+- Supplementary table 1: list of all germline genes. 
+	- Columns to include: 
+		- KDM5C bound vs not
+		- DEG in EpiLC, brain, both, neither (separate columns?)
 

### 2024-07-22 submission/manuscript.Rmd
fleshing out discussion

@@
 
+# Experimental Procedures
+
 # Discussion
 
-** overall outline **
-- Tissue identity
-	- We identified multiple tissue-specific genes in the brain
-	- some non-brain, tissue-enriched genes are lowly expressed do have brain-specific functions.
-	- KDM5C might be important for fine-tuning their expression to match the level important in the tissue. 
-	- this could be an avenue to impact _Kdm5c_-KO brain function.
+In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala.
+We defined tissue-enriched genes based on the criteria used in a previously published dataset of 17 C57/Bl6J mouse tissues[@liComprehensiveMouseTranscriptomic2017], which defined genes as tissue-enriched if they had more than 4-fold higher expression when compared to any other tissue.
+In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
+It is currently unclear what the impact of these tissue-enriched genes are upon _Kdm5c_-KO brain function and if they ultimately contribute to MRXSCJ-related phenotypes.
+While the majority of tissue-enriched DEGs were testis and ovary genes with no known brain functions, select liver and muscle-biased DEGs do have known roles in the wild-type brain.
+For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver[XXX], but is lowly expressed in the wild-type brain.
+Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
+Our results suggest KDM5C fine-tunes the expression of tissue-enriched genes like _Apoc1_ to match the level requried for proper brain function.
+However, futher studies are required to detrmine if these genes contribute to _Kdm5c_-KO neuronal impairments and MRXSCJ pathology.
 
@@
 ## Discussion notes
-- Altogether, this suggests KDM5C plays a pivotal role in the development of tissue identity, including the soma-germline distinction.
-
-- In this study we demonstrated KDM5C play a pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the _Kdm5c_-KO hippocampus and amygdala.
-	- In addition to testis genes identified previously[XXX], we also observed significant enrichment of muscle, liver, and even ovary-biased genes aberrantly transcribed within the male brain.
-
-	- Many liver and muscle-biased DEGs are weakly expressed in the brain and have known roles in brain function, such as _Apolipoprotein C-I (Apoc1)_, a liver-biased DEG that important for lipoprotein metabolism but has also been shown to influence learning and memory[XXX].
-
-	- This suggests tissue-enriched genes that are lowly expressed in the brain might be more sensitive to dysregulation with KDM5C loss.
-	- increased propensity for dysregulation 
-	- However, the majority of tissue-enriched DEGs are testis and ovary genes that have known functions in the germline and no known function in the brain.
 

### 2024-07-22 submission/manuscript.Rmd
discussion drafting

@@
 sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis.
-In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na\"{i}ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
 Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
@@
 Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
-Somatic repression of germline genes is typically established during the transition between naive and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.
+Somatic repression of germline genes is typically established during the transition between na\"{i}ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.
 In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
@@
 
-- testis genes are germline genes, germline gene list
-	- While a select few tissue-enriched DEGs had known roles in the brain, the majority were testis genes with no known function outside the germline. 
-	- distinguishing between germline and soma a key feature of multicellularity, chromatin regulators are important for this process
-	- Studies on chromatin regulators in germline gene repression has focused on specific genes 
-	- Curated a list of germline-enriched genes to robustly assess the types of genes dysregulated and KDM5C's role
+A substantial portion of genes upregulated in _Kdm5c_-KO brain are testis genes that have no known function within the brain.
+Through the use of publically available RNA-seq datasets, we demonstrated these testis-enriched DEGs are typically unique to germ cells.
+Misexpression of germline genes in the brain suggests _Kdm5c_-KOs fail to demarcate between somatic and germline lineages, a key feature of multicellularity and sexual reproduction.
+Previous studies have demonstrated germline genes are decommissioned in somatic cells by multiple chromatin-modifying enzymes, but have primarily focused on the repression of key marker genes.
+To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets.
+This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which types of germline genes are directly or indirectly regulated by KDM5C.
+
+We found misexpression of tissue-enriched genes, including germline genes, also occurs in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+
+- _In vivo_, a subset of epiblast stem cells will reactivate germline gene expression and form primordial germ cells (PGCs) that will eventually become the mature germline.
+This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]
+
+- Importantly, we did not observe any significant changes in the expression of EpiLC marker genes and cellular morphology, indicating _Kdm5c_-KO EpiLCs are not simply becoming PGCLCs. 
+- This is further supported proper expression of _Otx2_ in _Kdm5c_-KO EpiLCs, an epiblast stem cell marker that is known to repress differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018].
+This suggests aberrant germline gene expression is occuring ectopically in conjunction with typical developmental programs, in conjunction with _Kdm5c_-KO mice having overall unimpaired brain morphology and very few brain-specific DEGs.
+
+- EpiLCs are an _in vitro_ model of the post-implantation embryo, when cells transition from na\"{i}ve to primed pluripotency.
+
+- Only a small number of germline genes were shared between _Kdm5c_-KO EpiLCs and in the brain.
+
+dysregulation of tissue identity
+
+- In line with previous work on other chromatin regulators, we found dysregulation of _Kdm5c_-KO tissue identity began during early embryogenesis. While _Kdm5c_-KO epiblast-like cells (EpiLCs)
 
@@
 
-- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from naive to primed pluripotency.
+- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na\"{i}ve to primed pluripotency.
 	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
 	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.
+
 		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
@@
 	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes. 
+	- This would be the first (?) example in which removal of an active mark is required for germline gene repression.
 

### 2024-07-23 code/KDM5C_chip_HOMER.R
new unique motifs

@@ -54,3 +54,8 @@ makemotifs <- function(motif){
 for(m in motifs){
 	makemotifs(m)
 }
+
+
+### 240723 read in the instances file from homer
+# e2finst <- read.csv(snakemake@input[[9]], sep = "\t")
+# print(head(e2finst))

### 2024-07-24 submission/manuscript.Rmd
add accent for naive

@@
 sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis.
-In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na\"{i}ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
 Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
@@
 Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
-Somatic repression of germline genes is typically established during the transition between na\"{i}ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.
+Somatic repression of germline genes is typically established during the transition between na&#239;ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.
 In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
@@
 
-We found misexpression of tissue-enriched genes, including germline genes, also occurs in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-
-- _In vivo_, a subset of epiblast stem cells will reactivate germline gene expression and form primordial germ cells (PGCs) that will eventually become the mature germline.
-This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]
-
-- Importantly, we did not observe any significant changes in the expression of EpiLC marker genes and cellular morphology, indicating _Kdm5c_-KO EpiLCs are not simply becoming PGCLCs. 
-- This is further supported proper expression of _Otx2_ in _Kdm5c_-KO EpiLCs, an epiblast stem cell marker that is known to repress differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018].
-This suggests aberrant germline gene expression is occuring ectopically in conjunction with typical developmental programs, in conjunction with _Kdm5c_-KO mice having overall unimpaired brain morphology and very few brain-specific DEGs.
-
-- EpiLCs are an _in vitro_ model of the post-implantation embryo, when cells transition from na\"{i}ve to primed pluripotency.
+We found this dysregulation in tissue identity likely begins during early emrbyogenesis, as germline genes and other tissue-enriched genes are aberrantly expressed in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+_In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[].
+However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[].
+This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX].
+Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs.
+Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation.
+Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018].
+This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.
 
@@
 
-- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na\"{i}ve to primed pluripotency.
+- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na&#239;ve to primed pluripotency.
 	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
@@
 The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline.
-Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
 KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017].

### 2024-07-24 submission/manuscript.Rmd
make margins larger

@@
 output: pdf_document
-geometry: margin = 0.5in
+geometry: margin = 1in
 font-size: 11pt

### 2024-07-24 submission/manuscript.Rmd
discussion draft

@@
 
-We found this dysregulation in tissue identity likely begins during early emrbyogenesis, as germline genes and other tissue-enriched genes are aberrantly expressed in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs).
 _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[].
@@
 
-- Only a small number of germline genes were shared between _Kdm5c_-KO EpiLCs and in the brain.
+- change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C
+We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).
+While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs.
+This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress germline transcripts[@scandagliaLossKdm5cCauses2017].
+Although KDM5C is enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of EpiLC germline-enriched DEGs.
+One notable DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells that promotes meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
+Retinoic acid can only activate _Stra8_ expression when DAZL is present, which is a direct target of KDM5C and is aberrantly transcribed and translated in _Kdm5c_-KO EpiLCs.
+This indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C regulation through activation by other ectopic germline programs.
+Consistent with this idea, many _Kdm5c_-KO EpiLC germline DEGs are important for early germ cell development and meiosis while those expressed in the mature _Kdm5c_-KO brain are involved in late sperm development.
+Altogether, this indicates ectopic germline programs are, to some extent, progressing through germ cell developmental stages over the course of _Kdm5c_-KO development.
 
-dysregulation of tissue identity
-
-- In line with previous work on other chromatin regulators, we found dysregulation of _Kdm5c_-KO tissue identity began during early embryogenesis. While _Kdm5c_-KO epiblast-like cells (EpiLCs)
-
-- KDM5C represses germline genes during early embryogenesis
-	- Found this dysregulation in tissue identity begins during _Kdm5c_-KO early embryogenesis
-	- EpiLC express early regualtors, Brain late sperm - program seems to be continuing across development
-	- Importantly, even though germline genes are misexpressed, markers of EpiLC differentiation and EpiLC morphology is unimpaired. 
-		- Epiblast can become primoridal germ cells
-		- Otx2 is expressed in EpiLCs and is known to repress PGC identity.
-It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/
-		- Germline Program happening in the background of typical development.
-
-- One of the genes misexpressed in _Kdm5c_-KO EpiLCs is Dazl, _Deleted in azoospermia like (Dazl)_, a key regulator of germline development that promotes the translation of germline-specific transcripts[XXX].
-	- 2 other studies of dazl regulators in ESCs also found KDM5C represses dazl. 
-	- While crucial for germ cell development, Dazl is also expressed in ESCs and at the 2-cell stage. 
-	- Some genes important for germ cell, ESC, and 2-cell development given they are important for self-renewal.  
-	- We did not observe dysregulation of 2-cell-specific regulators like Dux. 
-	- We found Dazl was expressed in _Kdm5c_-KO EpiLCs, when _Dazl_ it typically decommissioned. This indicates KDM5C loss results in long-term ectopic expression.
-	- We additionally found DAZL protein is ectopically expressed and localized to the cytoplasm.
-Given that a significant number of KDM5C-KO gemrline DEGs are DAZL targets, it may promote the ectopic germline program.
-
-	- Very recently two other studies identified KDM5C represses Dazl in ESCs. 
-	- - KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-However, KDM5C binding and _Kdm5c_-KO germline gene misexpression has yet to be globally characterized during early embryogenesis.
-Given that _Dazl_ and other germline-enriched genes can also be expressed in ESCs and at the 2-cell stage, it is unclear if KDM5C has a direct role in the long-term germline gene silencing that occurs in the post-implantation epiblast.
-
-- We globally characterized KDM5C's repression of germline genes during early embryogenesis and in mature neurons. 
-	- In line with previous work, we found KDM5C did not regulate germline genes in neurons. 
-	- KDM5C-bound genes in EpiLCs were enriched for germline ontologies, suggesting a major role of KDM5c during embryogenesis
-	- While KDM5C directly binds some germline genes, including _Dazl_, many of the genes dysregulated in _Kdm5c_-KO were not direct targets. 
-		- Stra8 notable exception
-		- Brain genes, including late-stage spermatogenesis genes
-		- Suggests germline genes can be dysregulated direct and indirect of KDM5C resgulation
-		- Further supports germline programs can be ectopically activated
 - Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
 	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
+- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
+	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
+- Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654
+
+It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal.
+For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021].
+Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+Interestingly, one screen found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency in ESCs[@guptaGenomewideScreenReveals2023].
+We found that while KDM5C also repressess _Dazl_ expression in EpiLCs, _Kdm5c_-KO EpiLCs do not express 2-cell specific genes like _Dux_ and _Zscan4c_.
+Out of the four 2-cell regulators characterized in ESCs, KDM5C was the only factor whose repression of _Dazl_ was independent of _Dux_ expression[@guptaGenomewideScreenReveals2023].
+Together, shis suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression, including germline genes that are also involved in pluripotency and self-renewal.
 
@@
 
+- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
+
+- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
+	- then transition into the long term silencing mechanism paragraph
+
+- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
+
 - Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
@@
 - One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
--
-
-While Directly binds to a subset of germline genes during early embryogenesis 
-- KDM5C binds to a subset of germline genes during early embryogenesis
 
@@
 - Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
-	- This shift from meiotic genes to later spermatogenesis genes in the hippocampus suggests the germline developmental program could occur ectopically as _Kdm5c_-KO cells progresses through somatic tissue development.
-**note: this is strengthened by the ChIP-seq data since KDM5C is not directly bound to many brain/flagellar DEGs.This point might be stronger in the ChIPseq figure**
-
-- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
 

### 2024-07-24 submission/manuscript.Rmd
fix formatting discussion

@@
 - change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C
+
 We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).
@@
 It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal.
-For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021].
+For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239;ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021].
 Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
@@
 - We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
+**note: maybe just put in results**.
+Could move around depending upon if I get pheno working.
 

### 2024-07-25 submission/manuscript.Rmd
add methods

@@
 
-# Experimental Procedures
-
 # Discussion
@@
 While the majority of tissue-enriched DEGs were testis and ovary genes with no known brain functions, select liver and muscle-biased DEGs do have known roles in the wild-type brain.
-For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver[XXX], but is lowly expressed in the wild-type brain.
+For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver, but is also lowly expressed in the wild-type brain.
 Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
@@
 Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs).
-_In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[].
-However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[].
-This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX].
+_In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020].
+However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
+This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
 Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs.
@@
 	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
+
+# Materials and Methods
+
+## Classifying tissue-enriched and germline-enriched genes
+Tissue-enriched differentially expressd genes were determined by their classification in a previously published list of genes enriched in 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017].
+This study defined expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.
+
+We curated a list of germline-enriched genes using an RNA-seq dataset from wild-type and germline-depleted (Kit^W/Wv^) male and female mouse embryos from embryonic day 12, 14, and 16 [@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
+Germline-enriched genes met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type somatic tissues[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted (Kit^W/Wv^) germline, for any sex or time point, does not exceed 20% of maximum expression in wild-type germline.
+
+## Cell culture
+We utilized our previously established cultures of male wild-type and _Kdm5c_ mutant embryonic stem cells[@samantaActivationXistEvolutionarily2022].
+Sex was confirmed by genotyping _Uba1_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'.
+Deletion of _Kdm5c_ was confirmed through the primers 5'-ATGCCCATATTAAGAGTCCCTG-3', 5'-TCTGCCTTGATGGGACTGTT-3', and 5'-GGTTCTCAACACTCACATAGTG-3'.
+
+Emrbyonic stem cells (ESCs) were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
+They were then transitioned into ground-state "naive" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
+Both pESC and nESC media were supplemented with the GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), the MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and leukemia inhibitory factor (LIF, Millipore#ESG1107).
+
+nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF).
+
+Immunocytochemistry of DAZL in EpiLCs was performed by washing cells thrice with phosphobuffered saline (PBS), fixing cells in 4% paraformaldehyde, washing thrice in PBS, blocking in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour, washing thrice with PBS, and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4C.
+The next day cells were rinsed thrice with PBS, incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1000) in blocking buffer, rinsed thrice in PBS, and then imaged.
+Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and then quantified via ImageJ.
+
+## RNA sequencing 
+### Published datasets
+All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo.
+Previously published RNA sequencing datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GSE: GSE96797).
+
+### Alignment and analysis
+After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads.
+Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser.
+RStudio (v3.6.0) was then used to analyze counts files by DESeq2 (v1.26.0)[@loveModeratedEstimationFold2014a] to identify differentially expressed genes (DEGs) with a q-value (p-adjusted via FDR/Benjamini–Hochberg correction) less than 0.1 and a log2 fold change greater than 0.5.
+For all DESeq2 analyses, log2 fold changes were calculated with lfcShrink using the ashr package[@stephensFalseDiscoveryRates2016].
+MA-plots were generated by ggpubr (v0.4.0), and Eulerr diagrams were generated by eulerr (v6.1.1).
+Boxplots and scatterplots were generated by ggpubr (v0.6.0) and ggplot2 (v3.3.2).
+Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
+The Upset plot was generated via the package UpSetR (v1.4.0)[@conwayUpSetRPackageVisualization2017].
+Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting.
+
+## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
+We analyzed our previously published KDM5C ChIP-seq experiments in primary neuron cultures from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036) and EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797).
+ChIP-seq of histone 3 lysine 4 dimethylation in male EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
+ChIP-seq of histone 3 lysine 4 trimethylation in the male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
+
+ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches.
+Only uniquely mapped reads were used for analysis.
+Peaks were called using MACS2 software (v2.1.0.20140616) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1.
+Common peak sets were obtained in R via DiffBind (v3.6.5), and count tables for the common peaks were generated with the Bedtools multicov command.
+We removed “black-listed” genomic regions that often give aberrant signals.
+Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1).
+Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010].
+Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting.
+Average binding across the genome was visualized using deeptools (v3.1.3).
+Bigwigs were visualized using the UCSC genome browser.
+
+## Whole genome bisulfite sequencing (WGBS)
+
+## Data availability
+
+# Acknowledgements
 

### 2024-07-25 submission/manuscript.Rmd
move icc to its own section

@@
 ## Cell culture
-We utilized our previously established cultures of male wild-type and _Kdm5c_ mutant embryonic stem cells[@samantaActivationXistEvolutionarily2022].
+We utilized our previously established cultures of male wild-type and _Kdm5c_ knockout (-KO) embryonic stem cells[@samantaActivationXistEvolutionarily2022].
 Sex was confirmed by genotyping _Uba1_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'.
@@
 
-Immunocytochemistry of DAZL in EpiLCs was performed by washing cells thrice with phosphobuffered saline (PBS), fixing cells in 4% paraformaldehyde, washing thrice in PBS, blocking in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour, washing thrice with PBS, and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4C.
+## Immunocytochemistry (ICC)
+ICC of DAZL in EpiLCs was performed by washing cells thrice with phosphobuffered saline (PBS), fixing cells in 4% paraformaldehyde, washing thrice in PBS, blocking in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour, washing thrice with PBS, and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4C.
 The next day cells were rinsed thrice with PBS, incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1000) in blocking buffer, rinsed thrice in PBS, and then imaged.

### 2024-07-25 submission/manuscript.Rmd
finish discussion draft

@@
 
-- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na&#239;ve to primed pluripotency.
-	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
-	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.
+It is hypothesized distinct repressive chromatin marks are placed at germline gene promoters as the embryo transitions from na&#239;ve to primed pluripotency.
+Initially, germline genes are repressed by placement of hisone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3) in ESCs and then gain _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+KDM5C may be instead required to remove an active mark, histone 3 lysine 4 trimethylation (H3K4me3), since H3K4me3 can impede _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found an increase in H3K4me3 around the transcription start site (TSS) of germline DEGs and loss of CpGme at at least two germline promoters[@scandagliaLossKdm5cCauses2017].
+However, KDM5C's role in embryonic germline gene silencing is still unclear, ask KDM5C's catalytic activity was recently found to be dispensible for silencing DAZL in ESCs[@guptaGenomewideScreenReveals2023].
+In this study, we observed a global increase in H3K4me3 around the TSS of germline-enriched gene in the _Kdm5c_-KO amygdala and an increase in H3K4me2 in _Kdm5c_-KO EpiLCs.
+We found KDM5C's expression is dynamically regulated during ESC to EpiLC differentiation and loss of KDM5C leads to impaired placement of CpGme in extended EpiLCs.
+Altogether, this suggests KDM5C is necessary during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.
 
-		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
-	- However, Recently KDM5C's catalytic acivity was found to be unnecessary for dazl suppression in ESCs.  
-	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes. 
-	- This would be the first (?) example in which removal of an active mark is required for germline gene repression.
+In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain.
+In EpiLCs, KDM5C directly represses key drivers of germline identity like _Dazl_, by promoting intial CpG methylation placement in the post-implantation embryo.
+However, germline genes can also become ectopically expressesed in _Kdm5c_-KO cells independent of direct KDM5C regulation, including the meiotic transcription factor _Stra8_.
+These ectopic germline developmental programs can, to some extent, mimic typical germ cell development, resulting in aberrant transcription early developmental and meiotic genes in _Kdm5c_-KO EpiLCs and late-stage spermatogenesis genes in the _Kdm5c_-KO brain.
+Altogether, these results define KDM5C's role in the demarcation between soma and gemrline identity and offers novel insight into how this dysregulation of tissue identity changes over the course of development.
+Additionally, this study provides the mechanistic foundation required to ultimately investigate the impact of aberrant germline identity upon neurodevelopment.
 
-- In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain.
-	- In EpiLCs, KDM5C directly represses key drivers of germline identity like Dazl, likely through PRC1.6 recruitment and promoting CpG methylation.
-However, _Kdm5c_-KO also ectopically expresses germline genes activated indirectly, including _Stra8_.
-	- The germline developmental program to some extent continues ectopically during _Kdm5c_-KO development, resulting in aberrant transcription of late stage spermatogenesis genes later in life. 
-	- These results define KDM5C's role in the demarcation between soma and gemrline identity and offer a window into potential targets to assess the deletirous ecfects these ectopic genes on neurodevelopment.
+- include cancer in there somewhere (Somatic misexpression of germline genes has been implicated in many cancers.)
 

### 2024-07-25 submission/manuscript.Rmd
abstract draft

@@
 # Abstract
+
+It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
+While some chromatin regulators are important for neuron-specific gene expression, loss of repressive chromatin regulators can also lead to the aberrant transcription of tissue-specific genes outside of their intended context.
+Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown.
+Here, we explored this crisis in cellular identity in NDDs by investigating the ectopic expression of germline (e.g.
+sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+Male _Kdm5c_ knockout (-KO) mice recapitulate key phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534), including heightened aggression and learning impairments.
+We found male _Kdm5c_ knockout (-KO) mice that recapitulate key phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534) aberrantly express liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+The majority of tissue-enriched genes were typically unique to germ cells, indicating the erosion of the soma-germline boundary is a major effect of KDM5C loss.
+Using RNA sequencing datasets of germ cell-depleted mice, we curated a list of germline-enriched genes and found KDM5C regulates many key drivers of meiosis and germline identity in an _in vitro_ model of the post-implantation epiblast.
+While KDM5C promotes the initial placement of DNA methylation at a subset of germline gene promoters during early embryogenesis, germline genes can also be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C binding.
+Intriguingly, many of the germline genes expressed in the _Kdm5c_-KO brain are late-stage spermatogenesis genes that are not directly bound by KDM5C.
+This suggests germline developmental programs can occur ectopically in the background of typical _Kdm5c_KO development.
+Ultimately, this work provides insight into the demarcation of somatic and germline identity in mammals while also providing the foundational mechanisms by which to test the impact of germline gene dysregulation in neurodevelopmental disorders.
 

### 2024-07-25 submission/manuscript.Rmd
move discussion notes to end
edit abstract

--- 2024-07-25 submission/manuscript.Rmd
@@
 # Abstract
+
+It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
+While some chromatin regulators are important for neuron-specific gene expression, loss of repressive chromatin regulators can also lead to the aberrant transcription of tissue-specific genes outside of their intended context.
+Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown.
+Here, we explored this crisis in cellular identity in NDDs by investigating the ectopic expression of germline (e.g.
+sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534), aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+Many genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
+Using germ cell-depleted RNA sequencing datasets, we curated a a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression.
+Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
+We found that while KDM5C promotes the initial placement of DNA methylation at a subset of germline gene promoters in EpiLCs, germline genes can also become activated independent of direct KDM5C regulation.
+Intriguingly, many of the germline transcripts expressed in the _Kdm5c_-KO brain were late-stage spermatogenesis genes that are not directly bound by KDM5C.
+This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development.
+Ultimately, this work provides novel insight into the demarcation of somatic and germline lineages in mammals while also characterizing the cellular identity crisis within a chromatin-linked neurodevelopmental disorder.
 
@@
 
-## Discussion notes
-
-- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
-**note: maybe just put in results**.
-Could move around depending upon if I get pheno working.
-
-- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
-	- then transition into the long term silencing mechanism paragraph
-
-- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
-
-- Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
-- Studies on other chromatin regulators have focused on key marker genes to identify germline gene misexpression, such as _Dazl_. 
-- Using a germline-depleted mouse model, we curated a list of mouse germline-enriched genes to globally assess germline gene dysregulation. 
-- Many of the genes dysregulated in the mature _Kdm5c_-KO brain were important for late stages of spermatogenesis, such as those important for sperm flagellar structure.
-Contrastingly, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germ cell identity, including _Dazl_ and _Stra8_.
-
-- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
-- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
-
-- The including the demarcation between soma and germline fates.
-
-	the _Kdm5c_-KO brain also 
-	Previous work identified testis-enriched genes in _Kdm5c_ genes
-	- 
-	- However unlike the gonadal-biased DEGs,
-
-- The demarcation of the germline versus soma is a key feature of multicellularity and sexually dimorphic reproduction
-
-- Anything known about tissue-biased gene expression in other H3K4me regulators?
-- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
-- tissue-biased gene expression:
-
-- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
-
-- Papers to read/reference:
-	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
-	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
-
 # Materials and Methods
@@
 # Notes
+## Discussion notes
+
+- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
+**note: maybe just put in results**.
+Could move around depending upon if I get pheno working.
+
+- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
+	- then transition into the long term silencing mechanism paragraph
+
+- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
+
+- Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
+- Studies on other chromatin regulators have focused on key marker genes to identify germline gene misexpression, such as _Dazl_. 
+- Using a germline-depleted mouse model, we curated a list of mouse germline-enriched genes to globally assess germline gene dysregulation. 
+- Many of the genes dysregulated in the mature _Kdm5c_-KO brain were important for late stages of spermatogenesis, such as those important for sperm flagellar structure.
+Contrastingly, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germ cell identity, including _Dazl_ and _Stra8_.
+
+- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
+- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
+
+- The including the demarcation between soma and germline fates.
+
+	the _Kdm5c_-KO brain also 
+	Previous work identified testis-enriched genes in _Kdm5c_ genes
+	- 
+	- However unlike the gonadal-biased DEGs,
+
+- The demarcation of the germline versus soma is a key feature of multicellularity and sexually dimorphic reproduction
+
+- Anything known about tissue-biased gene expression in other H3K4me regulators?
+- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
+- tissue-biased gene expression:
+
+- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
+
+- Papers to read/reference:
+	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
+	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
 

### 2024-07-25 submission/manuscript.Rmd
discussion notes

@@
 
+- Talk about impact - Other germline gene chromatin regulators, NDDs, cancer. Maybe talk about myt1l
+- include cancer in there somewhere (Somatic misexpression of germline genes has been implicated in many cancers.)
+
 In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain.
@@
 Additionally, this study provides the mechanistic foundation required to ultimately investigate the impact of aberrant germline identity upon neurodevelopment.
-
-- include cancer in there somewhere (Somatic misexpression of germline genes has been implicated in many cancers.)
 

### 2024-07-26 submission/manuscript.Rmd
intro shigeki edits

@@
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
-Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018].
+While many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018].
 Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009].
-However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
-Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if this partial loss of brain identity contributes to neurodevelopmental impairments.
+However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if partial loss of brain identity contributes to neurodevelopmental impairments.
 
@@
 Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
-It is currently unknown if this dysregulation of brain tissue identity further impairs _Kdm5c_-KO neurodevelopment and if ectopic gene expression within the _Kdm5c_-KO brain is unique to testis genes.
+It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.
 
-- one key process chrom reg are involved in are is distinguishing bw germ + soma cell identity
-
-Distinguishing between the germline (cells that pass on their genetic material, e.g.
-sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis.
-In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
+The testis contains both germ cells (cells that pass on their genetic material, e.g.
+sperm) and somatic cells (cells that perform all other bodily functions).
+Distinguishing between the germline and the soma is a key feature of multicellularity and that typically occurs during early embryogenesis.
+In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
+Initially, repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], are placed at germline gene promoters in embryonic stem cells and are then decorated with DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] in the post-implantation embryo.
+How KDM5C interacts with known germline gene silencers during this time point is currently unknown.
+Additionally, most studies have focused on select genes important for early germ cell development rather than germline-enriched genes as a whole given the lack of a curated list of germline-enriched genes.
+Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g.
+meiotic genes versus spermatid differentiation genes.
 Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
@@
 We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary.
-Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are typically uniquely expressed in germ cells.
-While the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_.
-KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
-Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation in EpiLCs through H3K4me2/3 removal.
+Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes.
+We demonstrated testis and ovary-enriched genes genes are germline genes by generating a list of germline enriched-genes using germ cell-depleted RNA sequencing datasets.
+Using this curated list, we found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
+Additionally, KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
+Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs.
 Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
@@
 ## Tissue-enriched genes, including testis genes, are aberrantly expressed in the _Kdm5c_-KO brain
-
-- **note: should I compare amygdala and hippocampus? scandaglia only looked at hippocampus**
 

### 2024-07-26 submission/manuscript.Rmd
shigeki results 1 edits

@@
 Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
-Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, it is currently unclear if this erosion of brain tissue identity is a major consequence of _Kdm5c_ loss and if it is unique to testis-enriched genes.
-Therefore, we first globally assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
+Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed and if loss of tissue identity a major consequence of KDM5C loss.
+Therefore, we globally assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 
-We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are enriched in non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
 The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C).
@@
 In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain.
-Although the _Kdm5c_-KO mice were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
-Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which was recently shown to sequester mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
+Although the _Kdm5c_-KO mice we examined were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
 Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test).
-An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, which is involved in lipoprotein metabolism (Figure 1E).
+An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] (Figure 1E).
 Testis, ovary, and liver-enriched DEGs showed little to no expression in the developing and adult wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).

### 2024-07-26 submission/manuscript.Rmd
shigeki results 2 edits

@@
 Leydig cells) that support hormone production and germline functions.
-Intriugingly, many _Kdm5c_-KO testis and ovary enriched-DEGs have germline-specific functions, suggesting _Kdm5c_-KO cells fail to distinguish between the soma and germline.
-To test if this holds true for all _Kdm5c_-KO testis-biased DEGs, we first assessed their known functions through gene onotology analysis.
+To determine if _Kdm5c_-KO testis-enriched DEGs are somatic or germline genes, we first assessed their known functions through gene onotology analysis.
 We found _Kdm5c_-KO testis-enriched DEGs high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
@@
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
-The only testis-enriched DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, the aforementioned testis gene that regulates piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012].
 We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
@@
 spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C).
-Together, these data suggest the somatic _Kdm5c_-KO brain aberrantly expresses germline genes.
+Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, reflecting an erosion between somatic versus germline identity.
 
 We then wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes.
-To facilitate downstream analyses, we generated a curated list of germline-enriched genes using currently available RNA-seq datasets in Kit^W/Wv^ mice.
+To facilitate downstream analyses, we generated a list of germline-enriched genes using currently available RNA-seq datasets in Kit^W/Wv^ mice.
 Wild-type and Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any adult wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally assess germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
 

### 2024-07-26 submission/manuscript.Rmd
shigeki updates fig 2

@@
 
-Misexpression of germline genes in the adult brain suggests _Kdm5c_-KO cells fail to demarcate between germline and somatic identity.
 Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
 This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top).
-Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-Therefore, we assessed if KDM5C was necessary for embryonic silencing of germline genes by evaluating the impact of _Kdm5c_ loss in EpiLCs.
+While some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
+Therefore, we tested if KDM5C was necessary for silencing germline genes in the post-implation embryo by evaluating the impact of _Kdm5c_ loss in EpiLCs.
 
-We first identified _Kdm5c_-KO EpiLC DEGs in of our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression.
+We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression.
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
-Using the curated list of germline genes generated above, we found _Kdm5c_-KO EpiLCs aberrantly expressed `r format(nrow(epigerm))` germline-enriched genes, including the previously characterized hippocampal DEG[@scandagliaLossKdm5cCauses2017] _Cytochrome C, testis-specific (Cyct)_ (Figure 3C).
-Although we observed aberrant expression of many tissue-enriched genes, we did not observe any significant difference in primed pluripotency genes (Figure 3D) or gross changes in _Kdm5c_-KO cell morpholgy during differentiation (Figure 3E), indicating KDM5C loss does not impair EpiLC formation.
+Although we observed aberrant expression of many tissue-enriched genes, including  `r format(nrow(epigerm))` germline-enriched genes, we did not find any significant difference in primed pluripotency genes, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
+We also did not observe any gross changes in _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D), altogether indicating KDM5C loss does not impair EpiLC formation.
 
-We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if all germline DEGs, like _Cyct_, are constitutively dysregulated or if they can change over the course of development.
-We found this was primarily not the case, as the majority of germline DEGs expressed in EpiLCs and the brain were unique, with only _Cyct_ shared across all sequencing datasets (Figure 3F).
-We then compared the known functions of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we compared compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+We found the majority of germline DEGs expressed in EpiLCs and the brain were unique, with only _Cyct_ shared across all sequencing datasets (Figure 3E-F).
+We then evaluated the known functions of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
 While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.

### 2024-07-26 submission/manuscript.Rmd
remove dazl protein from fig 2, put in functional section

@@
 However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).
-
-We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019].
-A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
-This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function.
-We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I).
-We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020].
-Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
 
@@
 # Notes
+## Dazl
+We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019].
+A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
+This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function.
+We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I).
+We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020].
+Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
+
 ## Discussion notes

### 2024-07-26 submission/manuscript.Rmd
shigeki updates fig 3-4

@@
 
-- KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in knockout neurons fails to suppress their dysregulation[@scandagliaLossKdm5cCauses2017].
-
-Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress hippocampal germline DEGs and KDM5C is not bound to their promoters in neurons[@scandagliaLossKdm5cCauses2017].
-There is some evidence KDM5C binds to select germline gene promoters in ESCs[@scandagliaLossKdm5cCauses2017], including _Dazl_[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-As KDM5C's binding at germline gene promoters has not been systematically characterized, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
+Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017].
+There is some evidence KDM5C binds to select germline gene promoters in ESCs but is lost in neurons[@scandagliaLossKdm5cCauses2017].
+However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters.
+Thus, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
 
@@
 
-To further characterize the mechanism behind _Kdm5c_-KO germline gene misexpression, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016].
+To further characterize KDM5C's role in germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016].
 EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).
@@
 We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D).
-Together, this suggests KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not regulate germline genes in neurons.
+While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D-E).
+Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not regulate germline genes in neurons.
 
-We then compared KDM5C binding at RNA-seq DEG promoters to determine if the germline mRNAs expressed in the _Kdm5c_-KO brain and EpiLCs are direct targets of KDM5C (Figure 4E).
+We then compared KDM5C binding at RNA-seq DEG promoters to determine if the germline mRNAs expressed in the _Kdm5c_-KO brain and EpiLCs are direct targets of KDM5C (Figure 4F).
 About one third of EpiLC-specific and brain-specific (hippocampus, amygdala, or both) germline DEGs were bound by KDM5C in EpiLCs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
+Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H).
 Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above.
@@
 In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G).
-Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H).
-Altogether, this suggests the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters during embryogenesis.
+Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons.
+Furthermore, majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters in EpiLCs.
 
-**notes:**
-- do Direct vs indirect DEGs motif analysis
+## KDM5C promotes _de novo_ DNA methylation at germline genes
 
-## KDM5C erases H3K4me3 to promote long-term repression of germline genes via DNA methylation
+KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], however it is unclear how KDM5C interacts with other suppr.
 
-Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
+
  Somatic repression of germline genes is typically established during the transition between na&#239;ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.

### 2024-07-26 submission/manuscript.Rmd
shigeki updates fig 5

@@
 Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017].
-There is some evidence KDM5C binds to select germline gene promoters in ESCs but is lost in neurons[@scandagliaLossKdm5cCauses2017].
+There is some evidence KDM5C binds to select germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017].
 However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters.
@@
 
-KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], however it is unclear how KDM5C interacts with other suppr.
-
-recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].
-
- Somatic repression of germline genes is typically established during the transition between na&#239;ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation.
-In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021].
-Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017].
-Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.
+In the early embryo, germline gene promoters are initially decorated with repressive histone modifications before long-term silencing by _de novo_ DNA CpG methylation (CpGme)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear.
+KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), marks enriched at active gene promoters[@iwaseXLinkedMentalRetardation2007].
+However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023].
+Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for long-term germline gene silencing.
+In support of this, CpGme is markedly reduced at two germline gene promoters in the adult hippocampus[@scandagliaLossKdm5cCauses2017].
+However, because KDM5C's role in germline gene repression has only been characterized at select germline genes in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during initial CpGme placement.
 
@@
 
-We then evaluated KDM5C's embryonic role in germline gene silencing during ESC to EpiLC differentiation.
+We then evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation.
 We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C).
 While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E).
-We then perfomed whole genome bislufite sequencing in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters.
-While wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).
+To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs).
+Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).
 
-- Catalytic activity
 - Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via removal of H3K4me2/3 during early embryogenesis.
-
-- **note:** should I test if incerase in H3K4me2 and H3K4me3 was highest in _Kdm5c_-KO cells at genes bound by KDM5C at their promoter in EpiLCs? Don't think it's that impactful
 

### 2024-07-26 submission/manuscript.Rmd
edit methods

@@
 ## Classifying tissue-enriched and germline-enriched genes
-Tissue-enriched differentially expressd genes were determined by their classification in a previously published list of genes enriched in 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017].
-This study defined expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.
+Tissue-enriched differentially expressd genes (DEGs) were determined by their classification in a previously published dataset from 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017].
+This study defined tissue expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.
 
@@
 Emrbyonic stem cells (ESCs) were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
-They were then transitioned into ground-state "naive" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
+They were then transitioned into ground-state "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
 Both pESC and nESC media were supplemented with the GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), the MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and leukemia inhibitory factor (LIF, Millipore#ESG1107).
@@
 ## Immunocytochemistry (ICC)
-ICC of DAZL in EpiLCs was performed by washing cells thrice with phosphobuffered saline (PBS), fixing cells in 4% paraformaldehyde, washing thrice in PBS, blocking in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour, washing thrice with PBS, and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4C.
-The next day cells were rinsed thrice with PBS, incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1000) in blocking buffer, rinsed thrice in PBS, and then imaged.
-Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and then quantified via ImageJ.
+ICC of DAZL in EpiLCs was performed by first growing cells on coverslips.
+Cells were then washed thrice with phosphobuffered saline (PBS), fixed in 4% paraformaldehyde, washed thrice in PBS, and blacked in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour.
+They were then washed thrice with PBS and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4°C with gentle rocking.
+The next day, cells were rinsed thrice with PBS, and incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1,000) with DAPI (1:1,000) in blocking buffer for 1 hour at room temperature.
+Coverslips were then rinsed thrice in PBS, and mounted onto slides using Prolong Gold (Invitrogen #P36930).
+Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and quantified via ImageJ before unblinding.
 
-## RNA sequencing 
-### Published datasets
-All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo.
-Previously published RNA sequencing datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GSE: GSE96797).
-
-### Alignment and analysis
+## RNA sequencing (RNA-seq)
 After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads.
@@
 For all DESeq2 analyses, log2 fold changes were calculated with lfcShrink using the ashr package[@stephensFalseDiscoveryRates2016].
-MA-plots were generated by ggpubr (v0.4.0), and Eulerr diagrams were generated by eulerr (v6.1.1).
+MA-plots were generated by ggpubr (v0.6.0), and Eulerr diagrams were generated by eulerr (v6.1.1).
 Boxplots and scatterplots were generated by ggpubr (v0.6.0) and ggplot2 (v3.3.2).
-Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
 The Upset plot was generated via the package UpSetR (v1.4.0)[@conwayUpSetRPackageVisualization2017].
-Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting.
+Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
 
 ## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
-We analyzed our previously published KDM5C ChIP-seq experiments in primary neuron cultures from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036) and EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797).
-ChIP-seq of histone 3 lysine 4 dimethylation in male EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
-ChIP-seq of histone 3 lysine 4 trimethylation in the male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
-
 ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches.
@@
 Peaks were called using MACS2 software (v2.1.0.20140616) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1.
-Common peak sets were obtained in R via DiffBind (v3.6.5), and count tables for the common peaks were generated with the Bedtools multicov command.
 We removed “black-listed” genomic regions that often give aberrant signals.
+Common peak sets were obtained in R via DiffBind (v3.6.5).
+In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (XXX).
 Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1).
+Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
 Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010].
-Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting.
 Average binding across the genome was visualized using deeptools (v3.1.3).
@@
 ## Data availability
+### Published datasets
+All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo.
+Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797).
+
+Previously published ChIP-seq experiments included KDM5C in wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797) and mouse primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036).
+ChIP-seq of histone 3 lysine 4 dimethylation in male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
+ChIP-seq of histone 3 lysine 4 trimethylation in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
 
@@
 
+- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
 ## Discussion notes

### 2024-07-26 submission/manuscript.Rmd
discussion edit

@@
 In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala.
-We defined tissue-enriched genes based on the criteria used in a previously published dataset of 17 C57/Bl6J mouse tissues[@liComprehensiveMouseTranscriptomic2017], which defined genes as tissue-enriched if they had more than 4-fold higher expression when compared to any other tissue.
 In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
-It is currently unclear what the impact of these tissue-enriched genes are upon _Kdm5c_-KO brain function and if they ultimately contribute to MRXSCJ-related phenotypes.
-While the majority of tissue-enriched DEGs were testis and ovary genes with no known brain functions, select liver and muscle-biased DEGs do have known roles in the wild-type brain.
-For example, _Apolipoprotein C-I (Apoc1)_ is a lipid transport gene highly enriched within the liver, but is also lowly expressed in the wild-type brain.
+Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
+However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
 Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
-Our results suggest KDM5C fine-tunes the expression of tissue-enriched genes like _Apoc1_ to match the level requried for proper brain function.
-However, futher studies are required to detrmine if these genes contribute to _Kdm5c_-KO neuronal impairments and MRXSCJ pathology.
+Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.
 

### 2024-07-26 submission/manuscript.Rmd
discussion update 2

@@
 
-A substantial portion of genes upregulated in _Kdm5c_-KO brain are testis genes that have no known function within the brain.
-Through the use of publically available RNA-seq datasets, we demonstrated these testis-enriched DEGs are typically unique to germ cells.
-Misexpression of germline genes in the brain suggests _Kdm5c_-KOs fail to demarcate between somatic and germline lineages, a key feature of multicellularity and sexual reproduction.
-Previous studies have demonstrated germline genes are decommissioned in somatic cells by multiple chromatin-modifying enzymes, but have primarily focused on the repression of key marker genes.
-To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets.
-This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which types of germline genes are directly or indirectly regulated by KDM5C.
+We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain.
+Distinguishing the germline (i.e.
+cells that pass on their genetic material to the next generation) from the soma (i.e.
+all other cells of the body) is a key feature of multicellularity and sexual reproduction[XXX].
+Previous work identifying chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes, such as _Dazl_ and _Ddx4_[XXX].
+To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[].
+This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.
 
@@
 # Notes
+## Things to do
+- Move dazl to new figure if other staining works
+- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
+- Add cell culture concentrations to methods
+
 ## Dazl

### 2024-07-26 submission/manuscript.Rmd
edit discussion 3

@@
 
-Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs).
+RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
 _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020].
@@
 This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
-Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs.
-Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation.
-Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018].
-This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.
-
-- change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C
+Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs.
+Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018].
+Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation.
+Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
 

### 2024-07-26 submission/manuscript.Rmd
edit discussion 3

@@
 
-We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).
+We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).
 While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs.
-This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress germline transcripts[@scandagliaLossKdm5cCauses2017].
-Although KDM5C is enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of EpiLC germline-enriched DEGs.
-One notable DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells that promotes meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
-Retinoic acid can only activate _Stra8_ expression when DAZL is present, which is a direct target of KDM5C and is aberrantly transcribed and translated in _Kdm5c_-KO EpiLCs.
-This indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C regulation through activation by other ectopic germline programs.
-Consistent with this idea, many _Kdm5c_-KO EpiLC germline DEGs are important for early germ cell development and meiosis while those expressed in the mature _Kdm5c_-KO brain are involved in late sperm development.
-Altogether, this indicates ectopic germline programs are, to some extent, progressing through germ cell developmental stages over the course of _Kdm5c_-KO development.
-
-- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
-	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
-- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
-	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
-- Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654
+This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017].
+Although enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of the germline genes aberrantly transcribed in _Kdm5c_-KO EpiLCs.
+One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
+However, retinoic acid can only activate _Stra8_ expression when DAZL is present.
+Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiCs and is both transcribed and translated in _Kdm5c_-KO EpiLCs.
+Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs.
+These ectopic germline programs appear to loosely mimic the trajectory of typical germline development, as germline genes important for early germ cell development and meiosis are expressed in _Kdm5c_-KO EpiLC while late spermatogenesis genes are expressed in the mature _Kdm5c_-KO brain.
+Altogether, this indicates once a subset of germline genes are activated due to loss of direct KDM5C repression in the early embryo, these ectopic germline programs can continue to progress in the background of _Kdm5c_-KO somatic development.
 
@@
 - Add cell culture concentrations to methods
+- Motif analysis
+	- Discussion - talk about motifs
 
@@
 
-- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
-## Discussion notes
-
 - We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
@@
 Could move around depending upon if I get pheno working.
+
+## Discussion notes
+
+- For other paper:
+	- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
+	- - Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654
+
+- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
+	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
+- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
+	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
 

### 2024-07-27 submission/manuscript.Rmd
discussion edits and final paragraph

@@
 ```
+Katherine M. Bonefas, Ilakkiya Venkatachalam, and Shigeki Iwase. 
 # Abstract
@@
 Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009].
-However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of heart[@mallMyt1lSafeguardsNeuronal2017;@weigelMYT1LHaploinsufficiencyHuman2023] and liver-specific[@schaeferControlCognitionAdaptive2009] genes within adult neurons.
 Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if partial loss of brain identity contributes to neurodevelopmental impairments.
@@
 cells that pass on their genetic material to the next generation) from the soma (i.e.
-all other cells of the body) is a key feature of multicellularity and sexual reproduction[XXX].
-Previous work identifying chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes, such as _Dazl_ and _Ddx4_[XXX].
-To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[].
+all other cells of the body) is a key feature of multicellularity and sexual reproduction.
+Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+To globally characterize KDM5C's role in germline gene repression throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
 This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.
@@
 It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal.
-For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239;ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021].
-Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-Interestingly, one screen found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency in ESCs[@guptaGenomewideScreenReveals2023].
-We found that while KDM5C also repressess _Dazl_ expression in EpiLCs, _Kdm5c_-KO EpiLCs do not express 2-cell specific genes like _Dux_ and _Zscan4c_.
-Out of the four 2-cell regulators characterized in ESCs, KDM5C was the only factor whose repression of _Dazl_ was independent of _Dux_ expression[@guptaGenomewideScreenReveals2023].
-Together, shis suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression, including germline genes that are also involved in pluripotency and self-renewal.
+For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
+Very recently, two screens of _Dazl_-repressors identified KDM5C as a direct regulator in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
+Interestingly, one study found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency[@guptaGenomewideScreenReveals2023].
+However, out of the four regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023].
+Although expressed at the 2-cell stage and in ESCs, _Dazl_ is typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
+We found KDM5C also repressess _Dazl_ in EpiLCs, but _Kdm5c_-KO EpiLCs did not express 2-cell specific genes like _Dux_ and _Zscan4c_.
+Together, this suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
 
-It is hypothesized distinct repressive chromatin marks are placed at germline gene promoters as the embryo transitions from na&#239;ve to primed pluripotency.
-Initially, germline genes are repressed by placement of hisone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3) in ESCs and then gain _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-KDM5C may be instead required to remove an active mark, histone 3 lysine 4 trimethylation (H3K4me3), since H3K4me3 can impede _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
-This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found an increase in H3K4me3 around the transcription start site (TSS) of germline DEGs and loss of CpGme at at least two germline promoters[@scandagliaLossKdm5cCauses2017].
-However, KDM5C's role in embryonic germline gene silencing is still unclear, ask KDM5C's catalytic activity was recently found to be dispensible for silencing DAZL in ESCs[@guptaGenomewideScreenReveals2023].
-In this study, we observed a global increase in H3K4me3 around the TSS of germline-enriched gene in the _Kdm5c_-KO amygdala and an increase in H3K4me2 in _Kdm5c_-KO EpiLCs.
-We found KDM5C's expression is dynamically regulated during ESC to EpiLC differentiation and loss of KDM5C leads to impaired placement of CpGme in extended EpiLCs.
-Altogether, this suggests KDM5C is necessary during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.
+_Dazl_ and other germline-enriched genes are initially decommissioned in ESCs by repressive histone modifications and then _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+Unlike the previously characterized germline gene suppressors, KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
+In this study, we observed a global increase in H3K4me2/3 in _Kdm5c_-KO EpiLCs and amygdala around the transcription start site (TSS) of germline-enriched genes.
+However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+Although not necessary in ESCs, KDM5C's catalytic activity be required for CpGme in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found CpGme is significantly eroded at at least two germline promoters[@scandagliaLossKdm5cCauses2017].
+To elucidate the mechanism behind KDM5C-mediated silencing of germline genes, we characterized KDM5C in ESCs and EpiLCs and found its expression is dynamically regulated during differenitation.
+Furthermore, KDM5C acts during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.
 
-- Talk about impact - Other germline gene chromatin regulators, NDDs, cancer. Maybe talk about myt1l
-- include cancer in there somewhere (Somatic misexpression of germline genes has been implicated in many cancers.)
-
-In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain.
-In EpiLCs, KDM5C directly represses key drivers of germline identity like _Dazl_, by promoting intial CpG methylation placement in the post-implantation embryo.
-However, germline genes can also become ectopically expressesed in _Kdm5c_-KO cells independent of direct KDM5C regulation, including the meiotic transcription factor _Stra8_.
-These ectopic germline developmental programs can, to some extent, mimic typical germ cell development, resulting in aberrant transcription early developmental and meiotic genes in _Kdm5c_-KO EpiLCs and late-stage spermatogenesis genes in the _Kdm5c_-KO brain.
-Altogether, these results define KDM5C's role in the demarcation between soma and gemrline identity and offers novel insight into how this dysregulation of tissue identity changes over the course of development.
-Additionally, this study provides the mechanistic foundation required to ultimately investigate the impact of aberrant germline identity upon neurodevelopment.
+The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development.
+Although their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012].
+Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression.
+Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that may have a similar underlying cause of germline versus soma dysregulation.
+However, further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.
 
@@
 # Acknowledgements
+- Jacob Mueller for providing insight in germline gene regulation. 
+- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast like cells
+- Funding sources
 

### 2024-07-27 submission/manuscript.Rmd
add micro units

@@
    - \usepackage{setspace}
+   - \usepackage{siunitx}
    - \spacing{1.5}
@@
 They were then transitioned into ground-state "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
-Both pESC and nESC media were supplemented with the GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), the MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and leukemia inhibitory factor (LIF, Millipore#ESG1107).
+Both pESC and nESC media were supplemented with 3 $\mu$M GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), 1 $\mu$M MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and 1,000 units/mL leukemia inhibitory factor (LIF, Millipore#ESG1107).
 
-nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF).
+nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF), as previously described [@samantaGeneratingPrimedPluripotent2020].
 
@@
 Only uniquely mapped reads were used for analysis.
-Peaks were called using MACS2 software (v2.1.0.20140616) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1.
+Peaks were called using MACS2 software (v2.2.9.1) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1.
 We removed “black-listed” genomic regions that often give aberrant signals.
 Common peak sets were obtained in R via DiffBind (v3.6.5).
-In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (XXX).
+In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (v2.25.0).
 Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1).
@@
 - Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
-- Add cell culture concentrations to methods
 - Motif analysis

### 2024-07-27 submission/manuscript.Rmd
fix formatting

@@
 Katherine M. Bonefas, Ilakkiya Venkatachalam, and Shigeki Iwase.
+
 # Abstract
@@
 ### Germline genes are misexpressed in the _Kdm5c_-KO brain
-- Intriguingly, some of the ectopic testis transcripts identified within the _Kdm5c_-KO hippocampus have known functions unique to germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic versus germline identity.
 

### 2024-08-01 submission/manuscript.Rmd
shigeki edits abstract, intro, results

@@
 Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown.
-Here, we explored this crisis in cellular identity in NDDs by investigating the ectopic expression of germline (e.g.
+Here, we explored this cellular identity crisis in NDDs by investigating the ectopic expression of germline (e.g.
 sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
@@
 This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development.
-Ultimately, this work provides novel insight into the demarcation of somatic and germline lineages in mammals while also characterizing the cellular identity crisis within a chromatin-linked neurodevelopmental disorder.
+Ultimately, this work links soma-vs-germline demarcation to a chromatin-linked neurodevelopmental disorder.
 
@@
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
-While many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018].
+Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
+Recent advancements in next generation sequencing unexpectedly revealed mutations in many chromatin regulators cause neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
 Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009].
-However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of heart[@mallMyt1lSafeguardsNeuronal2017;@weigelMYT1LHaploinsufficiencyHuman2023] and liver-specific[@schaeferControlCognitionAdaptive2009] genes within adult neurons.
-Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if partial loss of brain identity contributes to neurodevelopmental impairments.
+However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if ectopic gene expression contributes to neurodevelopmental impairments.
 
-To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms governing their de-repression.
+To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of genes dysregulated and the molecular mechanisms governing their de-repression.
 In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C).
@@
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
-_Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
+_Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
 Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
@@
 sperm) and somatic cells (cells that perform all other bodily functions).
-Distinguishing between the germline and the soma is a key feature of multicellularity and that typically occurs during early embryogenesis.
+Distinguishing between the germline and the soma is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] and that typically occurs during early embryogenesis[@lehmannGermlineStemCells2012].
 In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
@@
 How KDM5C interacts with known germline gene silencers during this time point is currently unknown.
-Additionally, most studies have focused on select genes important for early germ cell development rather than germline-enriched genes as a whole given the lack of a curated list of germline-enriched genes.
+Additionally, most studies have focused on select genes important for early germ cell development rather than germline genes as a whole given the lack of a curated list of germline-enriched genes.
 Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g.
@@
 To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary.
+We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes.
 Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes.
-We demonstrated testis and ovary-enriched genes genes are germline genes by generating a list of germline enriched-genes using germ cell-depleted RNA sequencing datasets.
-Using this curated list, we found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
-Additionally, KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
+We demonstrated testis-enriched genes genes are germline genes and not somatic testis genes by.
+We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
+Additionally, KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
 Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs.
@@
 Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
-Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed and if loss of tissue identity a major consequence of KDM5C loss.
-Therefore, we globally assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
+Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed.
+Therefore, we systematically assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 
-In addition to the high enrichment of testis genes, we also identified aberrant expression of other tissue-enriched genes within the _Kdm5c_-KO brain.
-Although the _Kdm5c_-KO mice we examined were male, we observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Although not consistent across brain regions, we also found significant enrichment of DEGS biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test).
+Given the _Kdm5c_-KO mice we analyzed are male, this demonstrates ectopic expression of tissue-enriched genes is independent of organismal sex.
+Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test).
 An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] (Figure 1E).
-Testis, ovary, and liver-enriched DEGs showed little to no expression in the developing and adult wild-type brain, yet our mRNA-seq data indicate they are polyadenylated and spliced into mature transcripts (Figure 1C-E).
+These aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
 Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
@@
 ```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-germgenes <- read.csv("../data/processed/germGENES20.csv")
+germgenes <- read.csv("../data/processed/germGENES20.csv", header = TRUE)
 
@@
 
-To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in somatic versus germ cells within the testis.
-We first compared their expression in the testis with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^) that prevent the maturation of germ cells[@handelSertoliCellDifferentiation1979].
+To further validate if these testis DEGs are truly germline genes, we evaluated their expression in somatic versus germ cells within the testis.
+We first compared their expression in the testis without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
 We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
-We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
-spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C).
-Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, reflecting an erosion between somatic versus germline identity.
+We found some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
+spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C).
+Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion between somatic versus germline identity.
 
-We then wanted to more deeply characterize germline gene misexpression with _Kdm5c_ loss, but lacked a comprehensive list of mouse germline-enriched genes.
-To facilitate downstream analyses, we generated a list of germline-enriched genes using currently available RNA-seq datasets in Kit^W/Wv^ mice.
-Wild-type and Kit^W/Wv^ datasets included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
-We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any adult wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
+We then wanted to characterize germline gene misexpression with _Kdm5c_ loss more deeply, but lacked a comprehensive list of mouse germline-enriched genes.
+We thus generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that include males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
+We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue for adult wild type [@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
 These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally assess germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
@@
 
-Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
-This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top).
+Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
+A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
+This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A, top)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
 While some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-Therefore, we tested if KDM5C was necessary for silencing germline genes in the post-implation embryo by evaluating the impact of _Kdm5c_ loss in EpiLCs.
+Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in EpiLCs.
 
@@
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
-Although we observed aberrant expression of many tissue-enriched genes, including  `r format(nrow(epigerm))` germline-enriched genes, we did not find any significant difference in primed pluripotency genes, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
-We also did not observe any gross changes in _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D), altogether indicating KDM5C loss does not impair EpiLC formation.
+Primed pluripotency genes, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs(Figure 3C).
+Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) was normal, indicating KDM5C loss does not impair EpiLC formation.
 
-To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we compared compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
-We found the majority of germline DEGs expressed in EpiLCs and the brain were unique, with only _Cyct_ shared across all sequencing datasets (Figure 3E-F).
-We then evaluated the known functions of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.
+To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
+We found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.
 
-Interestingly, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H).
+Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H).
 These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
@@
 Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017].
-There is some evidence KDM5C binds to select germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017].
+KDM5C binds to several germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017].
 However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters.
@@
 
-To further characterize KDM5C's role in germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016].
+To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
 EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).
@@
 While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D-E).
-Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not regulate germline genes in neurons.
+Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
 
-We then compared KDM5C binding at RNA-seq DEG promoters to determine if the germline mRNAs expressed in the _Kdm5c_-KO brain and EpiLCs are direct targets of KDM5C (Figure 4F).
-About one third of EpiLC-specific and brain-specific (hippocampus, amygdala, or both) germline DEGs were bound by KDM5C in EpiLCs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
+We then compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F).
+In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
 Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H).
@@
 Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons.
-Furthermore, majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters in EpiLCs.
+Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.
 
@@
 
-In the early embryo, germline gene promoters are initially decorated with repressive histone modifications before long-term silencing by _de novo_ DNA CpG methylation (CpGme)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+In the early embryo, germline gene promoters are initially silenced by repressive histone modifications, including histone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3), and are then silenced long-term via DNA CpG methylation (CpGme)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
 Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear.
-KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), marks enriched at active gene promoters[@iwaseXLinkedMentalRetardation2007].
-However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023].
-Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for long-term germline gene silencing.
-In support of this, CpGme is markedly reduced at two germline gene promoters in the adult hippocampus[@scandagliaLossKdm5cCauses2017].
-However, because KDM5C's role in germline gene repression has only been characterized at select germline genes in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during initial CpGme placement.
+KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
+However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
+Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for CpGme-mediated, long-term silencing of germline genes.
+In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
+Based on these observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
 
-To elucidate KDM5C's role in germline gene silencing, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets in wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
+To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets of the wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
 In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 5A).

### 2024-08-02 submission/manuscript.Rmd
update methods

@@
 We utilized our previously established cultures of male wild-type and _Kdm5c_ knockout (-KO) embryonic stem cells[@samantaActivationXistEvolutionarily2022].
-Sex was confirmed by genotyping _Uba1_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'.
+Sex was confirmed by genotyping _Uba1_/_Uba1y_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'.
 Deletion of _Kdm5c_ was confirmed through the primers 5'-ATGCCCATATTAAGAGTCCCTG-3', 5'-TCTGCCTTGATGGGACTGTT-3', and 5'-GGTTCTCAACACTCACATAGTG-3'.
 
-Emrbyonic stem cells (ESCs) were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
-They were then transitioned into ground-state "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
+Emrbyonic stem cells (ESCs) and epiblast-like cells were cultured using previously established methods[@samantaGeneratingPrimedPluripotent2020].
+Briefly, ESCs were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
+They were then transitioned into ground-state, "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
 Both pESC and nESC media were supplemented with 3 $\mu$M GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), 1 $\mu$M MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and 1,000 units/mL leukemia inhibitory factor (LIF, Millipore#ESG1107).
@@
 ## Immunocytochemistry (ICC)
-ICC of DAZL in EpiLCs was performed by first growing cells on coverslips.
+ICC of DAZL in EpiLCs was performed by first growing cells on fibronectin-coated coverslips.
 Cells were then washed thrice with phosphobuffered saline (PBS), fixed in 4% paraformaldehyde, washed thrice in PBS, and blacked in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour.
@@
 ## Whole genome bisulfite sequencing (WGBS)
+Genomic DNA (gDNA) from na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells.
+gDNA was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
 
@@
 
+### Data analysis
+Scripts used to generate the results, tables, and figures of this study are available via a GitHub repository: XXX
+
 # Acknowledgements
 - Jacob Mueller for providing insight in germline gene regulation. 
-- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast like cells
+- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast-like cells
 - Funding sources

### 2024-08-04 code/serverscripts/WGBS/bismark_align.py
add WGBS server scripts

@@ -1,0 +1,23 @@
+#24.07.24 - bismark BAM alignment of trimmed reads. 
+
+import sys
+import os
+import glob
+
+
+#fastq file input location (trimmed)
+FASTQ_DIR = "../data/processed/trimmed_reads/"
+
+#output directory
+OUTDIR = "../data/processed/BAM"
+
+#get the fastq files for alignment (just the first read)
+firstREAD = glob.glob("{fastq_dir}*val_1.fq.gz".format(fastq_dir = FASTQ_DIR))
+
+for i in firstREAD:
+	#clean up the read name for getting the second read
+	clean = i.replace(FASTQ_DIR, "")
+	clean = clean.replace("1_val_1.fq.gz", "")
+
+	#run bismark
+	os.system("nice -10 bismark --non_directional --parallel 6 -o {outdir} /nfs/value/siwase2/GENOMES/MM10/bismark_genome/ -1 {READ_1} -2 {fastq_dir}{READ_2}2_val_2.fq.gz".format(outdir = OUTDIR, READ_1 = i, fastq_dir = FASTQ_DIR, READ_2 = clean))

### 2024-08-04 code/serverscripts/WGBS/bismark_genome_prep.py

@@ -1,0 +1,8 @@
+#24.07.24 - Prepare the mm10 genome for bisulfite analysis. Only need to do once
+
+import sys
+import os
+
+
+
+os.system("bismark_genome_preparation --parallel 4 --path_to_aligner ~/mambaforge/envs/bismark/bin/ --verbose /nfs/value/siwase2/GENOMES/MM10/bismark_genome/")

### 2024-08-04 code/serverscripts/WGBS/methylKit_diff.R

@@ -1,0 +1,12 @@
+# 24.08.24 DNA methylation analysis of bismark bam files
+
+#read in the bam files
+bamfiles <- list.files(path = "../data/processed/BAM_dedup/", pattern = "*.sorted.dedup.bam",
+           full.names = TRUE)
+
+print(bamfiles)
+
+# my.methRaw=processBismarkAln(location = system.file("extdata", "test.fastq_bismark.sorted.min.sam", 
+#                              package = "methylKit"),
+# 		                         sample.id="test1", assembly="mm10", 
+# 		                         read.context="CpG", save.folder=getwd())

### 2024-08-06 code/serverscripts/WGBS/methylKit_diff.R
script for differentially methylated cpgs:

@@ -1,12 +1,81 @@
-# 24.08.24 DNA methylation analysis of bismark bam files
+# 24.08.24 DNA methylation analysis of bismark bam files using methylKit
+    #https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#run from command line by 'Rscript [nameofscript].R'
+
+library(methylKit)
+library(ggplot2)
+# library(stringr)
 
 #read in the bam files
-bamfiles <- list.files(path = "../data/processed/BAM_dedup/", pattern = "*.sorted.dedup.bam",
+bamfiles <- list.files(path = "../data/processed/BAM_dedup", pattern = "*.sorted.dedup.bam",
            full.names = TRUE)
 
-print(bamfiles)
+#remove the bam.bai files
+bamfiles <- bamfiles[!endsWith(bamfiles, '.bai')]
 
-# my.methRaw=processBismarkAln(location = system.file("extdata", "test.fastq_bismark.sorted.min.sam", 
-#                              package = "methylKit"),
-# 		                         sample.id="test1", assembly="mm10", 
-# 		                         read.context="CpG", save.folder=getwd())
+
+#we want to compare WT and 5CKO for ESCs and EpiLCs
+cells <- c("esc", "EpiLC")
+#empty lists to store the output
+
+mylists <- list(list(), list())
+names(mylists) <- cells
+
+for (i in cells){
+    #subset for samples of that cell type
+    sampCell <- grep(i, bamfiles, value=TRUE)
+
+    #for every sample in sampcell, get the treatment (WT or KO)
+    #get the sample ID
+    ID <- gsub("../data/processed/BAM_dedup/", "", sampCell)
+    ID <- gsub(".sorted.dedup.bam", "", ID)
+    
+    #format as a list
+    ID_list <- list()
+    for (d in 1:length(ID)){
+        ID_list[[d]] <- ID[d]
+    }
+    # print("IDs as list")
+    # print(ID_list)
+    
+    #get the sample treatment (first two characters) - vector format
+    treat <- substr(ID, start = 1, stop = 2)
+    # print("treatments")
+    # print(treat)
+
+    #samples in list format
+    samp_list <- list()
+    for (s in 1:length(sampCell)){
+        samp_list[[s]] <- sampCell[s]
+    }
+
+
+    #save the alignment into the correct list
[…]

### 2024-08-06 code/serverscripts/WGBS/methylKit_diff.R
fix treatment, change to integers

@@ -38,10 +38,16 @@ for (i in cells){
     # print("IDs as list")
     # print(ID_list)
     
-    #get the sample treatment (first two characters) - vector format
-    treat <- substr(ID, start = 1, stop = 2)
+    #get the sample genotype (first two characters) - vector format
+    geno <- substr(ID, start = 1, stop = 2)
     # print("treatments")
     # print(treat)
+    # treatment is 1 or 0, 1 for KO, 0 for WT
+    treat <- c()
+    for (g in 1:length(geno)) {
+        treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
+    }
+    print(treat)
 
     #samples in list format
     samp_list <- list()

### 2024-08-06 submission/manuscript.Rmd
checking if new vsc works
update kdm5c-bound promoters results

@@
 We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D-E).
+In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 4D).
+Based on our ChIP-seq peak criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 4E).
+In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 4D).
 Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
+
+- read in the homer results txt file above here
+
+- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) binding.
+	- We asessed genomic features enriched at KDM5C-bound versus unbound gemrline gene promoters using HOMER.
+		- window I analyzed
 
@@
 ## KDM5C promotes _de novo_ DNA methylation at germline genes
+- if there are differences, say at germline gene CpG islands
 

### 2024-08-06 submission/manuscript.Rmd
add motif to text

@@
 - read in the homer results txt file above here
+- Add in euler KDM5C-bound vs KDM5C-unbound coloring
+		- Make sure instances window is 500 bp, change if not
 
-- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) binding.
-	- We asessed genomic features enriched at KDM5C-bound versus unbound gemrline gene promoters using HOMER.
-		- window I analyzed
+- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) and CpG islands (CGIs).
+We asessed transcription factor motifs enriched at germline gene promoters that are either bound or unbound by KDM5C in EpiLCs using HOMER[XXX] (TSS +/- 500 bp, q-value < 0.1).
+MAX and E2F6 binding sites were significantly enriched at KDM5C-bound germline genes, with about one third of KDM5C-bound promoters contained one or both of these motifs (Figure 4).
+Contrastingly, KDM5C-unbound germline genes were enriched for RFX transcription factor motifs, a family of transcription factors important for ciliogenesis.
+The RFX X-box motif is present at XXX% of KDM5C-unbound promoters (Figure 4).
 
-We then compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F).
+	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
+	 - https://www.nature.com/articles/srep20435
+	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
+	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
+
+Finally, we compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F).
 In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
@@
 While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E).
+
+- Germline genes are known to accumulate CpGme at (CGIs).
 To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs).
 Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).
+
+- discussion 
+ - 	- However, CPG islands greatly detrmine KDM5C recruitment - KDM5C is known to be enriched at CGIs.)
 
@@
 # Discussion
+
+Motif Discussion
+- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C (discussion?)
 

### 2024-08-06 submission/manuscript.Rmd
began discussion edits

--- 2024-08-02 submission/manuscript.Rmd
@@
 In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
+Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes, but germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary.
 Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
@@
 Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
-Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.
 
 We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain.
-Distinguishing the germline (i.e.
-cells that pass on their genetic material to the next generation) from the soma (i.e.
-all other cells of the body) is a key feature of multicellularity and sexual reproduction.
+Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023].
 Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-To globally characterize KDM5C's role in germline gene repression throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
-This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.
+To characterize KDM5C's role in germline gene repression at a genome-wide level throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
+This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which groups of germline genes are directly and indirectly regulated by KDM5C.
+This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.
 
 RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
-_In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020].
+_In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020].
 However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
@@
 Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs.
-Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018].
+Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
 Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation.
@@
 We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).
+
+Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells.
+
 While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs.
@@
 However, further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.
+
+- Last paragraph
+	- Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.
 

### 2024-08-07 code/serverscripts/HOMER/MotifEnrich_KDM5C_boundvsunbound.py
homer results and scripts

@@ -1,0 +1,30 @@
+# 24.07.19 HOMER to find motifs enriched in KDM5C bound vs unbound germline DEGs to see what is promoting their dysregulation independent of KDM5C
+
+import sys
+import os
+import glob
+
+#where HOMER is located
+HOMER = "../../HOMER/"
+
+#output directory
+OUT = "../results"
+
+#window of bp that you want to search aroudn the TSS
+WINDOW = "500"
+
+#input files of gene list
+#input file needs to be a .txt file with the gene IDs in the first column, and no quotes around them.
+GENES = glob.glob("../data/raw/*.txt")
+
+for i in GENES:
+	#clean the file name 
+	clean = i.replace("../data/raw/", "")
+	clean = clean.replace("_HOMER.txt", "")
+
+	#make a directory for the outputs for that sample because for some reason that's what homer does
+	OUTDIR = "{OUT}/{CLEAN}".format(OUT = OUT, CLEAN = clean)
+	os.system("mkdir {OUTDIR}".format(OUTDIR = OUTDIR))
+
+	#run the HOMER command for that gene list
+	os.system("nohup perl {HOMER}bin/findMotifs.pl {GENES} mouse {OUTDIR} -start -{WINDOW} -end {WINDOW} &".format(HOMER = HOMER, OUT = OUT, GENES = i, OUTDIR = OUTDIR, WINDOW = WINDOW))

### 2024-08-07 code/serverscripts/HOMER/motif_instances.py

@@ -1,0 +1,81 @@
+#2024.07.19 - find specific instances of motifs using HOMER
+#activate blueberry conda env to get python 3
+#have to run without nohup to avoid the running output being put on the top of the file
+
+#24.07.23 - using my custom e2f and ebox motifs
+
+import sys
+import os
+import glob
+
+#where HOMER is located
+HOMER = "../../HOMER/"
+
+#output directory
+OUT = "../results"
+
+#make a dictionary of the custom motifs
+motifs = {"E2F": "TCCCGC", "Ebox": "CACGTG", "xbox" : "usingHOMERs"}
+
+
+#make a directory for the outputs for that sample because for some reason that's what homer does
+outdir = "{OUT}/MotifsCount".format(OUT = OUT)
+#os.system("mkdir {OUTDIR}".format(OUTDIR = outdir))
+
+
+
+#function for searching for motifs in list of genes
+#MOTIF - motif you're searching for
+#GENES - genes you're searching (txt file)
+#OUTDIR - output directory
+#CLEAN - cleaned file name
+def motifSearch(MOTIF, GENES):
+
+	#set a header so x-box motifs can use the homer provided one
+	if MOTIF == "xbox":
+		header = "{HOMER}motifs/".format(HOMER = HOMER)
+	else:
+		header = ""
+		
+	os.system("perl {HOMER}bin/findMotifs.pl {GENES} mouse {OUTDIR}/ -find {header}{MOTIF}.motif > {OUTDIR}/{MOTIF}_instances_findMotifs_{CLEAN}.txt".format(HOMER = HOMER, MOTIF = MOTIF, header = header, GENES = GENES, OUTDIR = outdir, CLEAN = clean))
+
+	#os.system("perl {HOMER}bin/annotatePeaks.pl tss mm10 -list {GENES} -m {header}{MOTIF}.motif -size -500,500 > {OUTDIR}/{MOTIF}_instances_annoPeaks_300_300_{CLEAN}.txt".format(HOMER = HOMER, MOTIF = MOTIF, GENES = GENES, header = header, OUTDIR = outdir, CLEAN = clean))
+	
+	#histogram of motif instances
[…]

### 2024-08-07 submission/manuscript.Rmd

@@
 We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D-E).
+In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 4D).
+Based on our ChIP-seq peak criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 4E).
+In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 4D).
 Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
 
-We then compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F).
+- read in the homer results txt file above here
+- Add in euler KDM5C-bound vs KDM5C-unbound coloring
+		- Make sure instances window is 500 bp, change if not
+
+- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) and CpG islands (CGIs).
+We asessed transcription factor motifs enriched at germline gene promoters that are either bound or unbound by KDM5C in EpiLCs using HOMER[XXX] (TSS +/- 500 bp, q-value < 0.1).
+MAX and E2F6 binding sites were significantly enriched at KDM5C-bound germline genes, with about one third of KDM5C-bound promoters contained one or both of these motifs (Figure 4).
+Contrastingly, KDM5C-unbound germline genes were enriched for RFX transcription factor motifs, a family of transcription factors important for ciliogenesis.
+The RFX X-box motif is present at XXX% of KDM5C-unbound promoters (Figure 4).
+
+	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
+	 - https://www.nature.com/articles/srep20435
+	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
+	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
+
+Finally, we compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F).
 In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).

### 2024-08-07 code/ChIPseq_KDM5C_HOMER.R
add x box motifs and remove degs

@@ -1,61 +1,4 @@
-# 24.07.20 - plot the e2f and ebox motifs for kdm5c-bound and unbound germline genes
-
-#read in everything
-#make a dataframe that is bound and unbound together
-#plot the data
-
-
-# #read in the HOMER output files for the histogram
-
-# motifs <- c("E2F", "E-box")
-# genes <- c("all germline genes", "germline DEGs")
-
-
-# #function to format the HOMER output
-# formatdf <- function(df){
-# 	colnames(df) <- c("Distance_Center", "Total_Sites", "Pos_Sites", "Neg_Sites", "A_Freq", "C_Freq","G_Freq", "T_Freq")
-# 	phist <- data.frame("TSS_Distance" = rep(df$Distance_Center, 2), "Sites_per_bp" = c(df$Pos_Sites, df$Neg_Sites), "Type" = c(rep("+ Sites", length(df$Pos_Sites)), rep("- Sites", length(df$Neg_Sites))))
-# 	return(phist)
-# }
-
-
-# library('ggpubr')
-# histplot <- function(df, motif, n){
-# 	p <- ggline(phist, "TSS_Distance", "Sites_per_bp", color = "Type", shape = 46, title = paste(motif, "binding sites")) + geom_vline(xintercept = 0, linetype = "dashed", color = "gray")
-# 	ggsave(snakemake@output[[n]], ggpar(p, legend = "bottom"), width = 6, height = 3.5)
-# }
-
-# #graphing bound and unbound on same graph
-# histplot_tog <- function(df, title, n){
-# 	p <- ggline(df, "TSS_Distance", "Sites_per_bp", color = "Bound_Type", shape = 46, title =  title) + geom_vline(xintercept = 0, linetype = "dashed", color = "gray")
-# 	ggsave(snakemake@output[[n]], ggpar(p, legend = "right"), width = 9, height = 3.5)
-# }
-
-
-# makemotifs <- function(motif){
-# 	offset <- ifelse(motif == "E2F", 0, ifelse(motif == "E-box", 4, "whoop"))
-
-# 	for(i in 1:length(genes)){
-# 		bound <- formatdf(read.csv(snakemake@input[[i+offset]], sep = "\t"))
-# 		bound$binding <- "KDM5C bound"
[…]

### 2024-08-07 code/ChIPseq_KDM5C_HOMER.R
percent bargraphs for motifs

@@ -1,4 +1,5 @@
 # 24.07.28 - percentage of KDM5C-bound and unbound promoters with e2f and ebox motifs
+library("ggpubr")
 
 #read in the KDM5C bound genes
 germ_KDM5C <- read.csv(snakemake@input[[1]], sep = ",")
@@ -20,7 +21,7 @@ print(head(germ_KDM5C))
 	#Motifs
 	#Percentage
 
-motifdf <- data.frame(Kdm5c_binding = c(rep("Bound", 4), rep("Unbound", 4)), Motifs = c("E2F", "Ebox", "Both", "Neither"))
+motifdf <- data.frame(Kdm5c_binding = c(rep("Bound", 4), rep("Unbound", 4)), Motifs = c("E2F", "E-box", "Both", "Neither"))
 print(motifdf)
 
 
@@ -72,40 +73,42 @@ library(wesanderson)
 #set the plotting order
 
 motifbar <- function(df, colum, TITLE){
-	df$Motifs <- factor(df$Motifs, levels = c("E2F", "Ebox", "Both", "Neither"))
-	library("ggpubr")
+	df$Motifs <- factor(df$Motifs, levels = c("E2F", "E-box", "Both", "Neither"))
 	q <- ggbarplot(df, "Kdm5c_binding", colum,
-	fill = "Motifs", color = "Motifs", palette = wes_palette("GrandBudapest1", 4),
+	fill = "Motifs", color = "Motifs", palette = wes_palette("GrandBudapest2", 4),
 	title = TITLE, label = TRUE, lab.col = "black", lab.vjust = 1, xlab = "KDM5C Binding at Promoter", ylab = "% of genes with motif", orientation = "vert") 
 
 	return(q)
 }
 
 
-ggsave(snakemake@output[[1]], motifbar(motifdf, "Percent", "E2F and Ebox motifs - All germline genes"), width = 4, height = 4)
+ggsave(snakemake@output[[1]], motifbar(motifdf, "Percent", "All germline genes"), width = 4, height = 4)
 
 
 #X-box motifs
-xbox_df <- data.frame(Kdm5c_binding = rep(c("Bound", "Unbound"), 2), Xbox_status = c(rep("Xbox",2), rep("No", 2)))
[…]

### 2024-08-07 code/ChIPseq_KDM5C_HOMER.R
update xbox colors

@@ -105,7 +105,7 @@ xbox_df
 
 xbox_df$Xbox_status <- factor(xbox_df$Xbox_status, level = c("X-box", "No"))
 
-q <- ggbarplot(xbox_df, "Kdm5c_binding", "Percent", fill = "Xbox_status", color = "Xbox_status", palette = c("royalblue1", "royalblue4"),
+q <- ggbarplot(xbox_df, "Kdm5c_binding", "Percent", fill = "Xbox_status", color = "Xbox_status", palette = c("steelblue1", "steelblue4"),
 title = "All germline genes", label = TRUE, lab.col = "black", lab.vjust = 1, xlab = "KDM5C Binding at Promoter", ylab = "% of genes", orientation = "vert") 
 
 ggsave(snakemake@output[[2]], q, width = 4, height = 4)

### 2024-08-08 submission/manuscript.Rmd
add motif results paragraph

@@
 
-- read in the homer results txt file above here
-- Add in euler KDM5C-bound vs KDM5C-unbound coloring
-		- Make sure instances window is 500 bp, change if not
-
-- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) and CpG islands (CGIs).
-We asessed transcription factor motifs enriched at germline gene promoters that are either bound or unbound by KDM5C in EpiLCs using HOMER[XXX] (TSS +/- 500 bp, q-value < 0.1).
-MAX and E2F6 binding sites were significantly enriched at KDM5C-bound germline genes, with about one third of KDM5C-bound promoters contained one or both of these motifs (Figure 4).
-Contrastingly, KDM5C-unbound germline genes were enriched for RFX transcription factor motifs, a family of transcription factors important for ciliogenesis.
-The RFX X-box motif is present at XXX% of KDM5C-unbound promoters (Figure 4).
-
-	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
-	 - https://www.nature.com/articles/srep20435
-	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
-	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
+Many germline-specific genes, including meiotic genes, are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
+Thus, we identified transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters using HOMER[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1).
+MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: , E2F6 q-value:, E2F q-value: ) (Figure 4).
+One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 4).
+KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites.
+RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
+Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
+Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs, but is also not a direct target of KDM5C (Figure 4).
+Thus, distinct transcription factor programs regulate KDM5C-bound and unbound germline genes.
 
@@
 - Germline genes are known to accumulate CpGme at (CGIs).
+	- What percentage of germline genes have CGIs
+- Do differential methylation analysis for WT ESCs to WT EpiLCs
+	- What percentage of germline genes significanty gain methylation (at CGI or at promoter)
+	- Out of the ones that gain methylation, which are significantly reduced
+
 To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs).
@@
 # Discussion
-
-Motif Discussion
-- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C (discussion?)
 
@@
 Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
+
+- Motif analysis
+	- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C
+	- KDM5C does not contain motif-specific binding
+	- controls cilia in mutiple cell types, initially identified in neurons. Shared axis that could contribute to NDDs
+		- RFX Cilia in NDDs - https://pubmed.ncbi.nlm.nih.gov/33658631/ 
+	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
+	 - https://www.nature.com/articles/srep20435
+	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
+	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
 

### 2024-08-08 submission/manuscript.Rmd
Add female section unedited

@@
 However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).
+
+## Females have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
+```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+#get the parameters used repeatedly, like the log2fold change cut off
+source("../../KDM5C_Germ_Dev_Sex_RA/code/utilities/parameters.R")
+```
+
+- Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes
+
+It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repression have been conducted exclusively in males.
+Sex is particularly pertinent in the case of KDM5C, as KDM5C is more highly expressed in females than males.
+We assessed germline gene misexpression in _Kdm5c_ mutant male and female mouse epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo that reflects when germline genes are typically silenced in somatic cells.
+
+We first identified differentially expressed genes (DEGs) compared to wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs and then filtered for germline-enriched genes using our previously curated list.
+While both male and female _Kdm5c_ mutants misexpressed germline genes, about twice as many were dysregulated in homozygous and heterozygous females compared to homozygous males (XX KO: , XX HET: , XY KO: ) (Figure 1A).
+
+A significant portion of germline genes were dysreguated in both male and female _Kdm5c_ mutants (Figure 1B).
+We then compared the log 2 fold change of shared germline DEGs to assess if sex impacts the degree of germline gene misexpression.
+	* Results of this, are females more sensitive? (Figure 1C)
+
+We then wondered if the increased number of germline DEGs in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes.
+However, both shared and female-specific germline DEGs did not display a significant bias towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 1E).
+Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.
+
+## Germline gene misexpression in _Kdm5c_ mutants is independent of germ cell sex
+
+While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
+Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions.
+It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.
+
+To comprehensively assess the impact of gonadal sex on germline gene misexpression, we filtered our list of germline-enriched genes for egg (XX) or sperm (XY) biased genes.
+We determined male and female biased expression  using RNAseq data from wild-type male and female mice at embryonic day 12, 14, and 16.
+We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased
+germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 2A).
 
@@
   \centering 
+  \includegraphics{../../KDM5C_Germ_Dev_Sex_RA/submission/compiled_figs/EpiLC_XXvsXY.pdf}
+  \caption[Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
+Left - MA plot of mRNA-sequencing.
+Right - Number of tissue-enriched differentially expressed genes (DEGs).
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
+Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
+\textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
+Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
+\textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
+Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
+  \label{figurelabel}
+\end{figure}
+
+\begin{figure}
+  \centering 
   \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}

### 2024-08-08 submission/manuscript.Rmd
female edits

@@
 Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
+
+- Add females - KDM5C is a sexually dimorphic chromatin regulator, embryonic lethality makes it difficult to compare
 
@@
 
-It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repression have been conducted exclusively in males.
-Sex is particularly pertinent in the case of KDM5C, as KDM5C is more highly expressed in females than males.
-We assessed germline gene misexpression in _Kdm5c_ mutant male and female mouse epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo that reflects when germline genes are typically silenced in somatic cells.
+It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males.
+We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs.
+[@samantaActivationXistEvolutionarily2022].
 
-We first identified differentially expressed genes (DEGs) compared to wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs and then filtered for germline-enriched genes using our previously curated list.
-While both male and female _Kdm5c_ mutants misexpressed germline genes, about twice as many were dysregulated in homozygous and heterozygous females compared to homozygous males (XX KO: , XX HET: , XY KO: ) (Figure 1A).
+We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
+While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B).
+We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only).
+
+- Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic cell cycle and cilium movement, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion - )
+- XY only - protein localization to mitochondrion
+- Shared germline DEGs
+	- meiotic and 
+- Shared germline genes included female gamete generation
 

### 2024-08-09 submission/manuscript.Rmd
draft text female specific

@@
 [@samantaActivationXistEvolutionarily2022].
+We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.
 
-We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
+Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
 While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B).
 We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only).
+Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic (meiotic cell cycle) and flagellar (cilium movement) genes, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion).
 
-- Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic cell cycle and cilium movement, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion - )
-- XY only - protein localization to mitochondrion
-- Shared germline DEGs
-	- meiotic and 
-- Shared germline genes included female gamete generation
-
-A significant portion of germline genes were dysreguated in both male and female _Kdm5c_ mutants (Figure 1B).
-We then compared the log 2 fold change of shared germline DEGs to assess if sex impacts the degree of germline gene misexpression.
-	* Results of this, are females more sensitive? (Figure 1C)
-
-We then wondered if the increased number of germline DEGs in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes.
-However, both shared and female-specific germline DEGs did not display a significant bias towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 1E).
+Germline genes dysregulated in both sexes were also enriched for meiotic ontologies (meiotic nuclear division), as well as egg-specific genes (female gamete generation).
+The majority of these shared germline DEGs had a greater log2 fold change from wild-type in females compared to males (Figure 4D-F).
+The increased number of germline genes and degree of dysregulation in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes[XXX].
+However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G).
 Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.
@@
 Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
+
+- XX vs XY
+- While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
+Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions.
+It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.
 
@@
   \centering 
-  \includegraphics{../../KDM5C_Germ_Dev_Sex_RA/submission/compiled_figs/EpiLC_XXvsXY.pdf}
-  \caption[Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
+  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
+  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
 Left - MA plot of mRNA-sequencing.

### 2024-08-09 submission/manuscript.Rmd
egg sperm genes

@@
 
-- Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes
-
 It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males.
@@
 However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G).
-Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.
+Thus, while female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, it is likely independent of potential XCI defects.
 
@@
 While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
-Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions.
-It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.
-
-To comprehensively assess the impact of gonadal sex on germline gene misexpression, we filtered our list of germline-enriched genes for egg (XX) or sperm (XY) biased genes.
-We determined male and female biased expression  using RNAseq data from wild-type male and female mice at embryonic day 12, 14, and 16.
+To comprehensively assess if germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes.
 We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased
-germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 2A).
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 4H).
+We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L).
+_Kdm5c_ mutant male and female-specific germline DEGs were also not biased to the corresponding germ cell sex, indicating differences between male and female germline gene dysregulation not due to sex-specific activation of sperm or egg-specific transcriptional programs.
+These results demonstrate sex influences the degree of gene misexpression with loss of KDM5C, but not the sex of germ cell-enriched genes.
 

### 2024-08-12 code/EpiLC_Rfx2.R
rfx2 mRNA expression

@@ -1,0 +1,85 @@
+#24.08.11 - To plot the expression of primordial germ cell genes in WT and 5CKO EpiLCs #
+
+###1) Read in the TPM file
+
+#function to format the TPM files
+#name is the name in snakemake file
+EpiLC_TPM <- read.csv(file = snakemake@input[[1]], sep ="\t", row.names = 1)
+genes <-(row.names(EpiLC_TPM))
+#remove variant numbers from ensembl gene notation
+ENSEMBL<- read.csv(text = genes, sep=".", header = FALSE, col.names = c("ENSEMBL", "var"))
+   #change the rownames of the counts file to the ensembl genes without the variant numbers
+row.names(EpiLC_TPM) <- ENSEMBL$ENSEMBL
+
+#format the column names by removing extraneous info
+colnames(EpiLC_TPM) <- gsub(".TPM", "", colnames(EpiLC_TPM))
+head(EpiLC_TPM)
+
+
+
+
+###2) read in the sample info
+EpiLC_SI <- read.csv(snakemake@input[[2]], sep = ",")
+#get just the male data
+#XYEpiLC <- subset(EpiLC_TPM, select =  grep("XY", colnames(EpiLC_TPM), value = TRUE))
+#XYEpiLC_SI <- subset(EpiLC_SI, Sex == "XY")
+EpiLC_SI$genosex <- paste0(EpiLC_SI$Sex, EpiLC_SI$Genotype)
+EpiLC_SI$Sample <- colnames(EpiLC_TPM)
+
+
+
+###3) Make a dataframe with the gene info and the ENSEMBL ID
+genelist <- data.frame(ENSEMBL = c("ENSMUSG00000024206"), Symbol = c("Rfx2"))
+
+###4) Function to make boxplot of expression in WT and 5cKO EpiLCs
+library(ggplot2)
+library("ggpubr")
+source('code/utilities/colorpalettes.R') #load the colorpalette
+
+#### plot just the male samples, using facet
+#generates a tpm plot based on a gene dataframe that has the ensembl IDs in the first column
+#ymax - maximum y value
+#get the genes that match by ensembl ID
+EpiLC_pgc_TPM <- subset(EpiLC_TPM, rownames(EpiLC_TPM) %in% genelist[,1])
+#make a new dataframe with  columns being the sample, genotype, and TPM
[…]

### 2024-08-12 submission/manuscript.Rmd
update abstract'

@@
 Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-We found that while KDM5C promotes the initial placement of DNA methylation at a subset of germline gene promoters in EpiLCs, germline genes can also become activated independent of direct KDM5C regulation.
+Germline gene dysregualtion is sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene dysregulation than knockout males.
+We found KDM5C acts at a subset of germline gene promoters in EpiLCs to promote the initial placement of DNA methylation at CpG islands.
+However, germline genes can also become activated in _Kdm5c_-KO cells independent of direct KDM5C regulation.
 Intriguingly, many of the germline transcripts expressed in the _Kdm5c_-KO brain were late-stage spermatogenesis genes that are not directly bound by KDM5C.
 This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development.
-Ultimately, this work links soma-vs-germline demarcation to a chromatin-linked neurodevelopmental disorder.
+Ultimately, this work characterizes a novel suppressor of germline gene transcription and links impaired soma-vs-germline demarcation to a chromatin-linked neurodevelopmental disorder.
 

### 2024-08-12 submission/manuscript.Rmd
edit intro and add females

@@
 Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Recent advancements in next generation sequencing unexpectedly revealed mutations in many chromatin regulators cause neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
+Unexpectedly, next generation sequencing revealed mutations in many chromatin regulators directly cause or are associated with neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
 Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009].
@@
 
-To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of genes dysregulated and the molecular mechanisms governing their de-repression.
+To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
 In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C).
-KDM5C, , also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters.
-Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
-_Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
+KDM5C, also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters.
+Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
+MRXSCJ is more common and often more severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
+Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
 Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
-It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.
+It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, if this phenomenon is sexually dimorphic, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.
 
@@
 sperm) and somatic cells (cells that perform all other bodily functions).
-Distinguishing between the germline and the soma is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] and that typically occurs during early embryogenesis[@lehmannGermlineStemCells2012].
-In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
-Initially, repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], are placed at germline gene promoters in embryonic stem cells and are then decorated with DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] in the post-implantation embryo.
-How KDM5C interacts with known germline gene silencers during this time point is currently unknown.
-Additionally, most studies have focused on select genes important for early germ cell development rather than germline genes as a whole given the lack of a curated list of germline-enriched genes.
+Distinguishing between the germline and the soma is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012].
+In mammals, chromatin regulators are crucial for decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+Currently, most studies have focused on marker genes important for early germ cell development rather than germline genes as a whole given the lack of a curated list of germline-enriched genes.
 Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g.
 meiotic genes versus spermatid differentiation genes.
-Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.
-
-- Add females - KDM5C is a sexually dimorphic chromatin regulator, embryonic lethality makes it difficult to compare
+Furthermore, it is currently unknown if ectopic germline gene expression is influenced by chromosomal sex, as previous studies on germline gene repression have been exclusively in males.
+Sex is particularly pertinent in the case of sexually dimorphic chromatin regulators like KDM5C, which lies on the X chromosome and partially escapes X chromosome inactivation, resulting in 1.2-2-fold higher in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+Thus, systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, will unveil key mechanisms underlying the demarcation between soma and germline identity.
+These results will provide molecular footholds that can then be exploited to test the contribution of ectopic germline genes on neurodevelopment.
 
@@
 We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes.
-Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes.
-We demonstrated testis-enriched genes genes are germline genes and not somatic testis genes by.
+The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells and not somatic cells of the testis.
 We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
-Additionally, KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
+Female _Kdm5c_-KO EpiLCs expressed over double the number of germline genes than _Kdm5c_-KO males and germline genes dysregulated in both sexes were expressed more highly in females, demonstrating females have increased sensitivity to germline gene dysregulation.
+KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
 Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs.
@@
 
-Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
+Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the male _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
 Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed.

### 2024-08-12 submission/manuscript.Rmd
edit results 1

@@
 
-## Tissue-enriched genes, including testis genes, are aberrantly expressed in the _Kdm5c_-KO brain
+## Tissue-enriched genes are aberrantly expressed in the _Kdm5c_-KO brain
 
@@
 Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the male _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
-Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed.
-Therefore, we systematically assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
+It is currently unclear if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
+We thus characterized dysregulation of _Kdm5c_-KO brain tissue identity by systematically assessing the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017], using our published mRNA-seq datasets of the adult amygdala and hippocampus for male mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
-One example of a testis-enriched DEG is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
+One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
 
 Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
-Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
+Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
 Given the _Kdm5c_-KO mice we analyzed are male, this demonstrates ectopic expression of tissue-enriched genes is independent of organismal sex.
+
 Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test).
-An example of a liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] (Figure 1E).
-These aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
-Of note, we did not observe enrichment of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
-Together, these results suggest misexpression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
+A liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] that has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
+For all tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
+Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+Together, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
 

### 2024-08-12 submission/manuscript.Rmd
edit results 2

@@
 Leydig cells) that support hormone production and germline functions.
-To determine if _Kdm5c_-KO testis-enriched DEGs are somatic or germline genes, we first assessed their known functions through gene onotology analysis.
-We found _Kdm5c_-KO testis-enriched DEGs high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
+To determine if _Kdm5c_-KO testis-enriched DEGs are somatic or germline genes, we first evaluated their known functions through gene onotology analysis.
+We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
-To further validate if these testis DEGs are truly germline genes, we evaluated their expression in somatic versus germ cells within the testis.
-We first compared their expression in the testis without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
+To further validate if these testis DEGs are truly germline genes, we then assessed their expression in somatic versus germ cells within the testis.
+We first compared their expression in wild-type testes to those without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
 We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
-We found some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
+Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
 spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C).
@@
 
-We then wanted to characterize germline gene misexpression with _Kdm5c_ loss more deeply, but lacked a comprehensive list of mouse germline-enriched genes.
-We thus generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that include males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
-We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue for adult wild type [@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
+We then wanted to characterize germline gene misexpression with _Kdm5c_ loss more deeply, but lacked a comprehensive list of mouse germline genes.
+We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
+We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
 These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally assess germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

### 2024-08-12 submission/manuscript.Rmd
edit results 3

@@
 
-## _Kdm5c_-KO epiblast-like cells aberrantly express master regulators of germline identity
+## _Kdm5c_-KO epiblast-like cells aberrantly express key regulators of germline identity
 
@@
 This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A, top)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
-While some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in EpiLCs.
+While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
+Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
 
-We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression.
+We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression.
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
-Primed pluripotency genes, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs(Figure 3C).
-Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) was normal, indicating KDM5C loss does not impair EpiLC formation.
+Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C).
+Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) appeared normal, indicating KDM5C loss does not impair EpiLC formation.
 
-To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
 We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
-We found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.
+EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).
 
@@
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell state (Figure 3H).
+However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H).
+Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.
 
-## Females have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
+## Female epiblast-like cells have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
 ```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
@@
 
-In the early embryo, germline gene promoters are initially silenced by repressive histone modifications, including histone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3), and are then silenced long-term via DNA CpG methylation (CpGme)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+In the early embryo, germline gene promoters are initially silenced by repressive histone modifications, including histone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3), and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
 Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear.

### 2024-08-13 submission/manuscript.Rmd
edit results 4

@@
 
-It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males.
+It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have focused on their regulation in males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
 We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs.
@@
 Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
-While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B).
-We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only).
-Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic (meiotic cell cycle) and flagellar (cilium movement) genes, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion).
+While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B).
+We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared).
+Female-specific germline DEGs were enriched for meiotic (GO:XXX meiotic cell cycle) and flagellar (GO:XXX cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:XXX protein localization to mitochondrion).
+Germline transcripts expressed in both sexes were enriched for meiotic (meiotic nuclear division) and egg-specific functions (female gamete generation).
 
-Germline genes dysregulated in both sexes were also enriched for meiotic ontologies (meiotic nuclear division), as well as egg-specific genes (female gamete generation).
-The majority of these shared germline DEGs had a greater log2 fold change from wild-type in females compared to males (Figure 4D-F).
-The increased number of germline genes and degree of dysregulation in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes[XXX].
-However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G).
-Thus, while female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, it is likely independent of potential XCI defects.
+The majority of germline genes expressed in both sexes had a greater log2 fold change from wild-type in females compared to males (Figure 4D).
+This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to the loss of KDM5C-mediated suppression of germline genes.
+Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants, as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes instead lying on autosomes (Figure 4G).
+Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.
 

### 2024-08-13 submission/manuscript.Rmd
edit results 4

@@
 
-While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
-To comprehensively assess if germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes.
+While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
+To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes.
 We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017] (Figure 4H).
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
 We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L).
-_Kdm5c_ mutant male and female-specific germline DEGs were also not biased to the corresponding germ cell sex, indicating differences between male and female germline gene dysregulation not due to sex-specific activation of sperm or egg-specific transcriptional programs.
-These results demonstrate sex influences the degree of gene misexpression with loss of KDM5C, but not the sex of germ cell-enriched genes.
+Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex.
+This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs.
+These results demonstrate that sex influences the degree of gene misexpression with KDM5C loss, but not the germ cell sex in which the germline genes originate.
 

### 2024-08-13 submission/manuscript.Rmd
edit results 4

@@
 
-## Germline gene misexpression in _Kdm5c_ mutants is independent of germ cell sex
+### Germline gene misexpression in _Kdm5c_ mutants is independent of germ cell sex
 
@@
 This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs.
-These results demonstrate that sex influences the degree of gene misexpression with KDM5C loss, but not the germ cell sex in which the germline genes originate.
+These results demonstrate that sex influences the degree of germline gene misexpression with KDM5C loss, but not the germ cell sex from which the germline genes originate.
 

### 2024-08-13 submission/manuscript.Rmd
edit results 5

@@
 This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
-We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-L).
+We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J).
 Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex.
@@
 However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters.
-Thus, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
+It is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
 
@@
 To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 4A).
+EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
 
-The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 4B).
-We then performed gene ontology analysis to compare the known functions of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
-While there were no ontologies significantly enriched for genes only bound by KDM5C in PNCs, gene ontologies for genes shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C).
-Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C).
-We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 4D).
-Based on our ChIP-seq peak criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 4E).
-In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 4D).
+The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B).
+Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
+Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
+There were no ontologies significantly enriched for genes only bound by KDM5C in PNCs.
+We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
+In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 5D).
+Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).
+In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 5D).
 Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
 
-Many germline-specific genes, including meiotic genes, are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
-Thus, we identified transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters using HOMER[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1).
-MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: , E2F6 q-value:, E2F q-value: ) (Figure 4).
-One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 4).
-KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites.
+Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
+To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1).
+MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5F).
+One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5G).
+KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, Rfx2 q-value < 0.0001, Rfx5 q-value < 0.0001) (Figure 5H).
 RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
 Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs, but is also not a direct target of KDM5C (Figure 4).
-Thus, distinct transcription factor programs regulate KDM5C-bound and unbound germline genes.
+Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs (Figure 5I), but is not a direct target of KDM5C.
+Thus, distinct transcription factor programs regulate the expression of KDM5C-bound and unbound germline genes.
 
-Finally, we compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F).
-In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
-Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H).
+Finally, we compared KDM5C binding at the promoters of RNA-seq DEGs to determine if the germline genes transcribed in _Kdm5c_-KO cells are direct targets of KDM5C (Figure 5J-L).
+In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEG promoters (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
+Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs.
 Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above.
-Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F).
-In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G).
-Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons.
+Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 5K).
+In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS.
+Altogether, this suggests KDM5C decommssions a subset of germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons.
 Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

### 2024-08-13 submission/manuscript.Rmd
edit results 6

@@
 
-## KDM5C promotes _de novo_ DNA methylation at germline genes
-- if there are differences, say at germline gene CpG islands
+## KDM5C promotes _de novo_ DNA methylation at germline genes (if there are differences, say at germline gene CpG islands)
 
-In the early embryo, germline gene promoters are initially silenced by repressive histone modifications, including histone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3), and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear.
-KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
-However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
-Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for CpGme-mediated, long-term silencing of germline genes.
+In the early embryo, germline gene promoters are initially decorated with repressive histone modifications and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
+Our results above indicate KDM5C also acts at germline gene promoters during this time period.
+However, how KDM5C interacts with other germline gene silencing mechanisms is currently unclear.
+KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], yet KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
+Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required later in development for long-term silencing of germline genes.
 In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
-Based on these observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
 
+Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
 To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets of the wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
-In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 5A).
-We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 5B).
+In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
+We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
 
 We then evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation.
-We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C).
-While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E).
-
-- Germline genes are known to accumulate CpGme at (CGIs).
-	- What percentage of germline genes have CGIs
-- Do differential methylation analysis for WT ESCs to WT EpiLCs
-	- What percentage of germline genes significanty gain methylation (at CGI or at promoter)
-	- Out of the ones that gain methylation, which are significantly reduced
+We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation into EpiLCs (Figure 6C).
+While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6E).
 
 To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs).
-Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).
 
-- discussion 
- - 	- However, CPG islands greatly detrmine KDM5C recruitment - KDM5C is known to be enriched at CGIs.)
+- Germline genes are known to accumulate CpGme at (CGIs) during the transition from naive to primed pluripotency.
+- We first identified the number of germline genes with CGIs neighboring their promoter.
+	- Found XXX% of germline genes had CGIs, XXX% of which were direct KDM5C targets in EpiLCs (Eulerr).
 
-- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via removal of H3K4me2/3 during early embryogenesis.
+- We then curated a list of germline genes that significantly gained CpGme in wild-type exEpiLCs compared to ESCs.
+	- Example gene bedgraph
+	- Majority of CGI germline genes?
+
+- Out of the CGI genes, which had significantly reduced CpGme in 5CKO (heatmap of % methylation)
+	- Highlight interesting genes affected vs unaffected by KDM5C (especially if same/different from E2F6, PRC1.6, Setdb1 targets)
+	- CGIs that never gain DNAme in WT
+	- Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 6F).
+
+- Non-CGI germline genes, any significant changes at their promoter? (Heatmap again?)
+
+- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via DNA methylation during early embryogenesis.
 
@@
 	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
+
+- However, CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception
 

### 2024-08-13 submission/manuscript.Rmd
add females to discussion

@@
 In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
-Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes, but germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary.
-Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
+Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
 However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
-Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
+Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.
 
-We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain.
+Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes but instead germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary.
 Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023].
-Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-To characterize KDM5C's role in germline gene repression at a genome-wide level throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
-This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which groups of germline genes are directly and indirectly regulated by KDM5C.
+Previous work characterizing repressors of germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+To characterize KDM5C's role in germline gene repression throughout life at a genome-wide level, we curated a list of germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
+This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, 3) the impact of germ cell sex upon germline gene misexpression, and 4) which groups of germline genes are directly and indirectly regulated by KDM5C.
 This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.
 
-RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
-_In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020].
-However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
+Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
+_In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020].
+However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
 This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
-Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs.
-Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
+Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs.
+Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
 Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation.
-Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
+Proper EpiLC differentiation, coupled with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
 
-- XX vs XY
-- While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
-Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions.
-It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.
+While many genes involved in early germline development and meiosis are common between males and females, some germline genes display sex-biased expression or perform functions unique to eggs and sperm.
+We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex.
+However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs.
+Knockout of KDM5C in females is embryonic lethal, likely due to impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022].
+XCI defects could explain why _Kdm5c_-KO females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome.
+Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+Intriguingly, females heterozygous for _Kdm5c_ loss also had over double the number of germline DEGs than males, even though their XCI is largely preserved and their expression of KDM5C should be roughly equivalent to that of wild-type males.
+Altogether, this suggests female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity.
+Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators like KDM5C or a general feature of female cells.
 

### 2024-08-13 submission/manuscript.Rmd

@@
 
-- Motif analysis
-	- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C
-	- KDM5C does not contain motif-specific binding
-	- controls cilia in mutiple cell types, initially identified in neurons. Shared axis that could contribute to NDDs
-		- RFX Cilia in NDDs - https://pubmed.ncbi.nlm.nih.gov/33658631/ 
-	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
-	 - https://www.nature.com/articles/srep20435
-	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
-	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
-
-- However, CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception
-
-We then globally characterized KDM5C binding at germline-enriched gene promoters through KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs).
-
 Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells.
-
 While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs.
 This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017].
-Although enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of the germline genes aberrantly transcribed in _Kdm5c_-KO EpiLCs.
+It is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding.
+In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
+While MAX and E2F6 motifs were enriched at KDM5C-bound promoters in EpiLCs, only about one third of KDM5C-bound germline genes contained their consesus sequence at their promoter.
+Thus, other, unknown factors likely facilitate KDM5C's recruitment to germline genes.
+
+We also found germline genes can become activated in _Kdm5c_-KO cells indepedent of direct loss of KDM5C suppression, given that KDM5C was only bound to about a third of germline-enriched DEGs.
 One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
 However, retinoic acid can only activate _Stra8_ expression when DAZL is present.
-Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiCs and is both transcribed and translated in _Kdm5c_-KO EpiLCs.
+Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiLCs and is transcribed _Kdm5c_-KO EpiLCs.
 Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs.
-These ectopic germline programs appear to loosely mimic the trajectory of typical germline development, as germline genes important for early germ cell development and meiosis are expressed in _Kdm5c_-KO EpiLC while late spermatogenesis genes are expressed in the mature _Kdm5c_-KO brain.
-Altogether, this indicates once a subset of germline genes are activated due to loss of direct KDM5C repression in the early embryo, these ectopic germline programs can continue to progress in the background of _Kdm5c_-KO somatic development.
+In support of this, germline genes unbound by KDM5C were significantly enriched for RFX transcription factors.
+RFX transcription factors bind to X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
+One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs.
+Intruigingly, _Kdm5c_-KO EpiLCs primarily expressed germline genes involved in early germ cell development and meiosis while the mature _Kdm5c_-KO brain expressed late-stage spermatogenesis genes, including known RFX2 targets.
+Altogether, these data suggest once activated due to loss of direct KDM5C repression, ectopic germline programs can loosely mimic germline development and progress in the background of _Kdm5c_-KO somatic development.
 
@@
 
+- CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception. KDM5C promotes this process.
+
 The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development.
-Although their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012].
+However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
+KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes, such as _Apoc1_.
+KDM5C could also ensure tissue-specific timing of transcription factor family members, like RFX genes that are broadly required for cilia and flagella formation, including neuronal cilia[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009; @chungRFX2BroadlyRequired2012; @coschieraPrimaryCiliaPromote2024].
+While their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012].
 Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression.
-Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that may have a similar underlying cause of germline versus soma dysregulation.
-However, further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.
-
-- Last paragraph
-	- Thus, failure to fine-tune the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_ could be one route by which loss of brain tissue identity contributes to _Kdm5c_-KO impairments.
+Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation.
+Further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.
 
@@
 - Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast-like cells
+- Ilakkiya
 - Funding sources

### 2024-08-13 code/WGBS_germ_CGI.R
WGBS CPG island analysis

@@ -1,0 +1,98 @@
+
+### 24.04.26 compare binding locations of KDM5C ChIp-seq peaks and number of germline genes bound
+
+
+# Visualize genomic locations using ChIPseeker
+library(ChIPseeker)
+library(TxDb.Mmusculus.UCSC.mm10.knownGene)
+txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
+library(clusterProfiler)
+library(ggplot2)
+library(dplyr)
+
+
+#get the parameter used for the promoter range
+source("code/utilities/parameters.R")
+
+
+#read in the bedfile
+
+#using WT no KO for both EpiLCs and PNCs
+samples <- c("EpiLC", "PNC") #make sure order matches snakefile
+#a list to save the plots in
+annoplot <- list()
+#a list of the peak gene names
+peakENSEMBL <- list()
+
+#read in the bed file of CGIs
+peak <- readPeakFile(snakemake@input[[1]])
+print(head(peak))
+
+#annotate where the peaks are located and plot
+peakAnno <- annotatePeak(peak, tssRegion=c(-promorange, promorange), TxDb=txdb, annoDb="org.Mm.eg.db")
+
+#subset for just the promoter peaks
+annot <- as.data.frame(peakAnno@anno)
+promo <- subset(annot, annotation == "Promoter")
+
+print("promo")
+print(head(promo))
+	
+promo_genes <- promo %>% distinct(ENSEMBL, .keep_all = TRUE) #keep only distinct annotations (in case 2 CGIs are within 1000bp)
+CGI_ENSEMBL <- unique(promo_genes$ENSEMBL)
+print("promo_genes")
+print(head(promo_genes))
+
+print("CGI_ENSEMBL")
+print(CGI_ENSEMBL[1:5])
+
+
+#read in the germline genes list with KDM5C binding at promoter
+germ <- read.csv(snakemake@input[[2]], sep = ",")
+
+### save a bed file of the CGI regions that are at germline gene promoters
+germ_promo_CGI_bed <- promo[promo$ENSEMBL %in% germ$ENSEMBL,]
+
+write.table(germ_promo_CGI_bed[1:3], snakemake@output[[1]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
+
+
[…]

### 2024-08-13 code/serverscripts/WGBS/CGI_WT_methyl_diff.R
troubleshooting methylkit cgi

@@ -1,0 +1,119 @@
+# 24.08.08 DNA methylation analysis of bismark bam files at CGI using methylKit
+#https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#https://nbis-workshop-epigenomics.readthedocs.io/en/latest/content/tutorials/methylationSeq/Seq_Tutorial.html
+#run from command line by 'Rscript [nameofscript].R'
+
+library(methylKit)
+library(ggplot2)
+library(genomation)
+
+
+#we want to compare WT and 5CKO for ESCs and EpiLCs
+cells <- c("esc", "EpiLC")
+
+
+#read in the methylation differences results from the bam files using methRead
+#   #samples in list format
+#   samp_list <- list()
+#   for (s in 1:length(sampCell)){
+#     samp_list[[s]] <- sampCell[s]
+#   }
+
+
+
+for (i in cells){
+  #get list of files
+  difffiles <- list.files(path = "../results/methylKit", pattern = paste0("*",i,"_CpG.txt"),
+                         full.names = TRUE)
+  #make as a list
+  samp_list <- list()
+  for (s in 1:length(difffiles)){
+    samp_list[[s]] <- difffiles[s]
+  }
+  
+  #get the sample names (IDs)   
+  ID <- gsub("../results/methylKit/", "", difffiles)
+  ID <- gsub("_CpG.txt", "", ID)
+  #format as a list
+  ID_list <- list()
+  for (d in 1:length(ID)){
+    ID_list[[d]] <- ID[d]
+  }
+
+
+  #get the sample genotype (first two characters) - vector format
+  geno <- substr(ID, start = 1, stop = 2)
+
+  # treatment is 1 or 0, 1 for KO, 0 for WT
+  treat <- c()
+  for (g in 1:length(geno)) {
+    treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
+  }
+  
+  #read in the files
+  myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
+                   context="CpG", mincov = 10)
+  #filter based on read coverage
+  myobj <- filterByCoverage(myobj,lo.count=10,lo.perc=NULL,
+                                  hi.count=NULL,hi.perc=99.9)
+  
[…]

### 2024-08-13 code/serverscripts/WGBS/CGI_methyl_diff.R

@@ -1,0 +1,119 @@
+# 24.08.08 DNA methylation analysis of bismark bam files at CGI using methylKit
+#https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#https://nbis-workshop-epigenomics.readthedocs.io/en/latest/content/tutorials/methylationSeq/Seq_Tutorial.html
+#run from command line by 'Rscript [nameofscript].R'
+
+library(methylKit)
+library(ggplot2)
+library(genomation)
+
+
+#we want to compare WT and 5CKO for ESCs and EpiLCs
+cells <- c("esc", "EpiLC")
+
+
+#read in the methylation differences results from the bam files using methRead
+#   #samples in list format
+#   samp_list <- list()
+#   for (s in 1:length(sampCell)){
+#     samp_list[[s]] <- sampCell[s]
+#   }
+
+
+
+for (i in cells){
+  #get list of files
+  difffiles <- list.files(path = "../results/methylKit", pattern = paste0("*",i,"_CpG.txt"),
+                         full.names = TRUE)
+  #make as a list
+  samp_list <- list()
+  for (s in 1:length(difffiles)){
+    samp_list[[s]] <- difffiles[s]
+  }
+  
+  #get the sample names (IDs)   
+  ID <- gsub("../results/methylKit/", "", difffiles)
+  ID <- gsub("_CpG.txt", "", ID)
+  #format as a list
+  ID_list <- list()
+  for (d in 1:length(ID)){
+    ID_list[[d]] <- ID[d]
+  }
+
+
+  #get the sample genotype (first two characters) - vector format
+  geno <- substr(ID, start = 1, stop = 2)
+
+  # treatment is 1 or 0, 1 for KO, 0 for WT
+  treat <- c()
+  for (g in 1:length(geno)) {
+    treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
+  }
+  
+  #read in the files
+  myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
+                   context="CpG", mincov = 5)
+  #filter based on read coverage
+  myobj <- filterByCoverage(myobj,lo.count=5,lo.perc=NULL,
+                                  hi.count=NULL,hi.perc=99.9)
+  
[…]

### 2024-08-13 code/serverscripts/WGBS/CGI_WT_methyl_diff.R
pooling CGI WGBS

@@ -1,6 +1,7 @@
 # 24.08.08 DNA methylation analysis of bismark bam files at CGI using methylKit
 #https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
 #https://nbis-workshop-epigenomics.readthedocs.io/en/latest/content/tutorials/methylationSeq/Seq_Tutorial.html
+#https://compgenomr.github.io/book/extracting-interesting-regions-differential-methylation-and-segmentation.html
 #run from command line by 'Rscript [nameofscript].R'
 
 library(methylKit)
@@ -8,17 +9,23 @@ library(ggplot2)
 library(genomation)
 
 
+
+#regional summary analysis
+library(ChIPseeker)
+library(TxDb.Mmusculus.UCSC.mm10.knownGene)
+txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
+library(ggplot2)
+library(dplyr)
+
+#need granges as input for the bedfile
+#read in the bed file of CGIs
+bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
+print(head(bed))
+
+
+
 #we want to compare WT and 5CKO for ESCs and EpiLCs
 cells <- c("esc", "EpiLC")
-
-
-#read in the methylation differences results from the bam files using methRead
-#   #samples in list format
-#   samp_list <- list()
-#   for (s in 1:length(sampCell)){
-#     samp_list[[s]] <- sampCell[s]
-#   }
-
 
 
 for (i in cells){
@@ -40,6 +47,9 @@ for (i in cells){
     ID_list[[d]] <- ID[d]
   }
 
+  print("sample ID_list")
+  print(ID_list)
+
 
   #get the sample genotype (first two characters) - vector format
   geno <- substr(ID, start = 1, stop = 2)
@@ -52,31 +62,35 @@ for (i in cells){
   
   #read in the files
   myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
-                   context="CpG", mincov = 10)
+                   context="CpG", mincov = 5)
   #filter based on read coverage
-  myobj <- filterByCoverage(myobj,lo.count=10,lo.perc=NULL,
+  myobj <- filterByCoverage(myobj,lo.count=5,lo.perc=NULL,
                                   hi.count=NULL,hi.perc=99.9)
   
[…]

### 2024-08-13 code/serverscripts/WGBS/CGI_WT_methyl_diff.R
pooling CGI WGBS

@@ -89,17 +89,13 @@ for (i in cells){
     #make a bedgraph file of the differences
   bedgraph(myDiff10p, col.name = "meth.diff", file.name = paste0("../results/methylKit/diff_cpg_10p_pooled_", i,".bed"))
   
-  #just hypo methylated
-  myDiff10p.hypo <- getMethylDiff(myDiff,difference=10,qvalue=0.01,type="hypo")
-  
   
   #get the singificantly different cpgs that fall within CGIs
   cpg_anot <- readFeatureFlank("../data/raw/CGI_UCSC.bed", feature.flank.name = c("CpGi", "shores"), flank=2000)
   print(head(cpg_anot))  
+    
   
-  
-  
-  diffCpGann <- annotateWithFeatureFlank(as(myDiff10p.hypo,"GRanges"), feature = cpg_anot$CpGi, flank = cpg_anot$shores, feature.name = "CpGi", flank.name = "shores")
+  diffCpGann <- annotateWithFeatureFlank(as(myDiff10p,"GRanges"), feature = cpg_anot$CpGi, flank = cpg_anot$shores, feature.name = "CpGi", flank.name = "shores")
 
   plotTargetAnnotation(diffCpGann, main = "Differential Methylation Annotation")
   q <- plotTargetAnnotation(diffCpGann, main = "Differential Methylation Annotation")

### 2024-08-13 code/serverscripts/WGBS/CGI_WT_methyl_diff.R
remove unnecessary packages

@@ -12,16 +12,12 @@ library(genomation)
 
 #regional summary analysis
 library(ChIPseeker)
-library(TxDb.Mmusculus.UCSC.mm10.knownGene)
-txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
-library(ggplot2)
 library(dplyr)
 
 #need granges as input for the bedfile
 #read in the bed file of CGIs
 bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
 print(head(bed))
-
 
 
 #we want to compare WT and 5CKO for ESCs and EpiLCs
@@ -100,7 +96,7 @@ for (i in cells){
   plotTargetAnnotation(diffCpGann, main = "Differential Methylation Annotation")
   q <- plotTargetAnnotation(diffCpGann, main = "Differential Methylation Annotation")
   
-  pdf(paste0("../results/methylKit/WGBS_", i, "CGI_hypo_summary.pdf"), p, width = 4, height = 4)
+  pdf(paste0("../results/methylKit/WGBS_", i, "CGI_hypo_summary.pdf"), q, width = 4, height = 4)
   dev.off()
   
 

### 2024-08-13 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
make new file for WT

@@ -1,0 +1,125 @@
+# 24.08.08 DNA methylation analysis of bismark bam files at CGI using methylKit
+#https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#https://nbis-workshop-epigenomics.readthedocs.io/en/latest/content/tutorials/methylationSeq/Seq_Tutorial.html
+#https://compgenomr.github.io/book/extracting-interesting-regions-differential-methylation-and-segmentation.html
+#run from command line by 'Rscript [nameofscript].R'
+
+library(methylKit)
+library(ggplot2)
+library(genomation)
+
+
+
+#regional summary analysis
+library(ChIPseeker)
+library(dplyr)
+
+#need granges as input for the bedfile
+#read in the bed file of CGIs
+bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
+print(head(bed))
+
+
+#we want to compare WT and 5CKO for ESCs and EpiLCs
+cells <- c("esc", "EpiLC")
+
+
+for (i in cells){
+  #get list of files
+  difffiles <- list.files(path = "../results/methylKit", pattern = paste0("*",i,"_CpG.txt"),
+                         full.names = TRUE)
+  #make as a list
+  samp_list <- list()
+  for (s in 1:length(difffiles)){
+    samp_list[[s]] <- difffiles[s]
+  }
+  
+  #get the sample names (IDs)   
+  ID <- gsub("../results/methylKit/", "", difffiles)
+  ID <- gsub("_CpG.txt", "", ID)
+  #format as a list
+  ID_list <- list()
+  for (d in 1:length(ID)){
+    ID_list[[d]] <- ID[d]
+  }
+
+  print("sample ID_list")
+  print(ID_list)
+
+
+  #get the sample genotype (first two characters) - vector format
+  geno <- substr(ID, start = 1, stop = 2)
+
+  # treatment is 1 or 0, 1 for KO, 0 for WT
+  treat <- c()
+  for (g in 1:length(geno)) {
+    treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
+  }
+  
+  #read in the files
+  myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
+                   context="CpG", mincov = 5)
+  #filter based on read coverage
[…]

### 2024-08-14 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
wt only ESc to EpiLc analysis

@@ -15,89 +15,88 @@ library(ChIPseeker)
 library(dplyr)
 
 #need granges as input for the bedfile
-#read in the bed file of CGIs
-bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
+#read in the bed file of CGIs for germline gene promoters
+bed <- readPeakFile("../data/raw/CGI_UCSC_all_germ.bed")
 print(head(bed))
 
 
-#we want to compare WT and 5CKO for ESCs and EpiLCs
-cells <- c("esc", "EpiLC")
-
-
-for (i in cells){
-  #get list of files
-  difffiles <- list.files(path = "../results/methylKit", pattern = paste0("*",i,"_CpG.txt"),
-                         full.names = TRUE)
-  #make as a list
-  samp_list <- list()
-  for (s in 1:length(difffiles)){
+#get list of WT files
+difffiles <- list.files(path = "../results/methylKit", pattern = paste0("WT*_CpG.txt"), full.names = TRUE)
+#make as a list
+samp_list <- list()
+for (s in 1:length(difffiles)){
     samp_list[[s]] <- difffiles[s]
-  }
+}
   
-  #get the sample names (IDs)   
-  ID <- gsub("../results/methylKit/", "", difffiles)
-  ID <- gsub("_CpG.txt", "", ID)
-  #format as a list
-  ID_list <- list()
-  for (d in 1:length(ID)){
-    ID_list[[d]] <- ID[d]
-  }
+#get the sample names (IDs)   
+ID <- gsub("../results/methylKit/", "", difffiles)
+ID <- gsub("_CpG.txt", "", ID)
+#format as a list
+ID_list <- list()
+for (d in 1:length(ID)){
+	ID_list[[d]] <- ID[d]
+}
 
-  print("sample ID_list")
-  print(ID_list)
+print("sample ID_list")
+print(ID_list)
 
 
-  #get the sample genotype (first two characters) - vector format
-  geno <- substr(ID, start = 1, stop = 2)
+#get the sample cell type (first two characters) - vector format
+cell <- rep(c("ESC", "EpiLC"), 2)
 
-  # treatment is 1 or 0, 1 for KO, 0 for WT
-  treat <- c()
-  for (g in 1:length(geno)) {
-    treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
-  }
+# treatment is 1 or 0, 1 for KO, 0 for WT
+treat <- c()
+for (g in 1:length(cell)) {
[…]

### 2024-08-14 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
sample info

@@ -21,7 +21,8 @@ print(head(bed))
 
 
 #get list of WT files
-difffiles <- list.files(path = "../results/methylKit", pattern = paste0("WT*_CpG.txt"), full.names = TRUE)
+difffiles <- list.files(path = "../results/methylKit", pattern = "WT*_CpG.txt", full.names = TRUE)
+difffiles
 #make as a list
 samp_list <- list()
 for (s in 1:length(difffiles)){

### 2024-08-14 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
using sys glob instead of list.files

@@ -21,7 +21,7 @@ print(head(bed))
 
 
 #get list of WT files
-difffiles <- list.files(path = "../results/methylKit", pattern = "WT*_CpG.txt", full.names = TRUE)
+difffiles <- Sys.glob("../results/methylKit/WT*_CpG.txt")
 difffiles
 #make as a list
 samp_list <- list()

### 2024-08-14 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
fix naming

@@ -43,7 +43,7 @@ print(ID_list)
 
 
 #get the sample cell type (first two characters) - vector format
-cell <- rep(c("ESC", "EpiLC"), 2)
+cell <- rep(c("EpiLC", "ESC"), 2)
 
 # treatment is 1 or 0, 1 for KO, 0 for WT
 treat <- c()
@@ -69,7 +69,7 @@ print("united regions")
 print(head(united_regions))
 
 #just saving the one methylation results. Want the average methylation
-write.table(united_regions, paste0("../results/methylKit/regionCounts_germCGI_", i, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
+write.table(united_regions, paste0("../results/methylKit/regionCounts_germCGI_WT.csv"), sep = ',', row.names = FALSE, quote = FALSE)
 
 
 #pool the samples for calculating the methylation differences

### 2024-08-14 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
remove old variable

@@ -101,25 +101,23 @@ dev.off()
   
 
   
-  #get the list of differentially methylated CGIs
-  myobj_islands <- regionCounts(myobj, cpg_anot$CpGi)
-  # Filter the summarized counts by coverage
-  myobj_islands_filt <- filterByCoverage(myobj_islands, lo.count=5, lo.perc=NULL, hi.count=NULL, hi.perc=99.9)
-  # Perform simple normalization 
-  myobj_islands_filt_norm <- normalizeCoverage(myobj_islands_filt, method = "median")
-  # Merge the samples again
-  meth_islands <- unite(myobj_islands_filt_norm, destrand=TRUE)
+#get the list of differentially methylated CGIs
+myobj_islands <- regionCounts(myobj, cpg_anot$CpGi)
+# Filter the summarized counts by coverage
+myobj_islands_filt <- filterByCoverage(myobj_islands, lo.count=5, lo.perc=NULL, hi.count=NULL, hi.perc=99.9)
+# Perform simple normalization 
+myobj_islands_filt_norm <- normalizeCoverage(myobj_islands_filt, method = "median")
+# Merge the samples again
+meth_islands <- unite(myobj_islands_filt_norm, destrand=TRUE)
   
-  #differentially methylation for islands
-  myDiff_islands <- calculateDiffMeth(meth_islands)
+#differentially methylation for islands
+myDiff_islands <- calculateDiffMeth(meth_islands)
   # Rank by significance
-  myDiff_islands <- myDiff_islands[order(myDiff_islands$qvalue),]
+myDiff_islands <- myDiff_islands[order(myDiff_islands$qvalue),]
   # get all differentially methylated CpG Islands
-  myDiff_islands_10p <- getMethylDiff(myDiff_islands, difference=10, qvalue=0.01)
+myDiff_islands_10p <- getMethylDiff(myDiff_islands, difference=10, qvalue=0.01)
   
-  write.table(myDiff_islands_10p, paste0("../results/methylKit/WGBS_CGI_10_", i, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
+write.table(myDiff_islands_10p, paste0("../results/methylKit/WGBS_getmethyldiff_germCGI_10_WT_ESCvsEpiLC.csv"), sep = ',', row.names = FALSE, quote = FALSE)
   
 
-  
-}
 

### 2024-08-14 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
loosen cut off

@@ -4,6 +4,10 @@
 #https://compgenomr.github.io/book/extracting-interesting-regions-differential-methylation-and-segmentation.html
 #run from command line by 'Rscript [nameofscript].R'
 
+#qvalue cut off
+qval <- 0.1
+
+
 library(methylKit)
 library(ggplot2)
 library(genomation)
@@ -80,7 +84,7 @@ dm.pooledf <- calculateDiffMeth(pooled.meth, mc.cores = 8)
 
 
 #all of the bases
-myDiff10p <- getMethylDiff(dm.pooledf, difference=10, qvalue=0.01)
+myDiff10p <- getMethylDiff(dm.pooledf, difference=10, qvalue=qval)
 #make a bedgraph file of the differences
 bedgraph(myDiff10p, col.name = "meth.diff", file.name = paste0("../results/methylKit/bedgraph_diff_WT_ESC_EpiLC_pooled.bed"))
   
@@ -89,7 +93,7 @@ bedgraph(myDiff10p, col.name = "meth.diff", file.name = paste0("../results/methy
 cpg_anot <- readFeatureFlank("../data/raw/CGI_UCSC.bed", feature.flank.name = c("CpGi", "shores"), flank=2000)
 print(head(cpg_anot))  
 
-myDiff10p_hyper <- getMethylDiff(dm.pooledf, difference=10, qvalue=0.01, type = "hyper")
+myDiff10p_hyper <- getMethylDiff(dm.pooledf, difference=10, qvalue=qval, type = "hyper")
 
 diffCpGann <- annotateWithFeatureFlank(as(myDiff10p_hyper,"GRanges"), feature = cpg_anot$CpGi, flank = cpg_anot$shores, feature.name = "CpGi", flank.name = "shores")
 
@@ -115,7 +119,7 @@ myDiff_islands <- calculateDiffMeth(meth_islands)
   # Rank by significance
 myDiff_islands <- myDiff_islands[order(myDiff_islands$qvalue),]
   # get all differentially methylated CpG Islands
-myDiff_islands_10p <- getMethylDiff(myDiff_islands, difference=10, qvalue=0.01)
+myDiff_islands_10p <- getMethylDiff(myDiff_islands, difference=10, qvalue=qval)
   
 write.table(myDiff_islands_10p, paste0("../results/methylKit/WGBS_getmethyldiff_germCGI_10_WT_ESCvsEpiLC.csv"), sep = ',', row.names = FALSE, quote = FALSE)
   

### 2024-08-14 code/serverscripts/WGBS/methylKit_diff.R
loosen min cov to 5

@@ -5,6 +5,14 @@
 library(methylKit)
 library(ggplot2)
 # library(stringr)
+library(ChIPseeker)
+library(dplyr)
+
+#need granges as input for the bedfile
+#read in the bed file of CGIs
+bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
+print(head(bed))
+
 
 #read in the bam files
 bamfiles <- list.files(path = "../data/processed/BAM_dedup", pattern = "*.sorted.dedup.bam",
@@ -58,7 +66,7 @@ for (i in cells){
 
     #save the alignment into the correct list
     #When it's run and if you save the results you can then read the results in directly using methRead
-    mylists[[i]] <- processBismarkAln(location = samp_list, nolap = TRUE,
+    mylists[[i]] <- processBismarkAln(location = samp_list, nolap = TRUE, mincov = 5,
 		                    sample.id = ID_list, assembly="mm10", treatment = treat, 
                           read.context="CpG", save.folder="../results/methylKit")
     
@@ -68,20 +76,31 @@ for (i in cells){
 
 
 for (n in cells){
+	#get the methylation for cgis
+	regions <- regionCounts(mylists[[n]],bed)
+
+	united_regions <- unite(regions, destrand=FALSE)
+	print("united regions")
+	print(head(united_regions))
+
+	write.table(united_regions, paste0("../results/methylKit/regionCounts_CGI_5min_", n, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
+
+
     #combine methylation calls into one
     meth <- unite(mylists[[n]], destrand = TRUE)
     print(head(meth))
 
     #get the PCA plot
-    ggsave(paste0("../results/methylKit/PCA_WGBS_", n, ".pdf"), PCASamples(meth), width = 4, height = 4)
+    PCASamples(meth)
+	 
+	pooled.meth <- pool(meth, sample.ids=c("KO","WT"))
+	dm.pooledf <- calculateDiffMeth(pooled.meth, mc.cores = 8)
 
-    #calculate methylation differences
-    myDiff <- calculateDiffMeth(meth, mc.cores = 8)
-    #write.table(myDiff, paste0("../data/processed/WGBS_methylDiff_", n, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
-
-    # get all differentially methylated bases with greater than 10% difference in methylation
[…]

### 2024-08-15 code/serverscripts/WGBS/methyl_diff_WTvsKO.R
wt vs KO mincov 5 qvalue 0.1

@@ -1,0 +1,132 @@
+# 24.08.15 DNA methylation analysis of bismark bam files at CGI using methylKit
+#https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#https://nbis-workshop-epigenomics.readthedocs.io/en/latest/content/tutorials/methylationSeq/Seq_Tutorial.html
+#https://compgenomr.github.io/book/extracting-interesting-regions-differential-methylation-and-segmentation.html
+#run from command line by 'Rscript [nameofscript].R'
+
+## calculating methylation differences in WT vs Ko ESCs and EpiLCs
+
+library(methylKit)
+library(ggplot2)
+library(genomation)
+
+
+
+#regional summary analysis
+library(ChIPseeker)
+library(dplyr)
+
+#need granges as input for the bedfile
+#read in the bed file of CGIs
+bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
+print(head(bed))
+
+
+#we want to compare WT and 5CKO for ESCs and EpiLCs
+cells <- c("esc", "EpiLC")
+
+
+for (i in cells){
+  #get list of files
+  difffiles <- list.files(path = "../results/methylKit", pattern = paste0("*",i,"_CpG.txt"),
+                         full.names = TRUE)
+  #make as a list
+  samp_list <- list()
+  for (s in 1:length(difffiles)){
+    samp_list[[s]] <- difffiles[s]
+  }
+  
+  #get the sample names (IDs)   
+  ID <- gsub("../results/methylKit/", "", difffiles)
+  ID <- gsub("_CpG.txt", "", ID)
+  #format as a list
+  ID_list <- list()
+  for (d in 1:length(ID)){
+    ID_list[[d]] <- ID[d]
+  }
+
+  print("sample ID_list")
+  print(ID_list)
+
+
+  #get the sample genotype (first two characters) - vector format
+  geno <- substr(ID, start = 1, stop = 2)
+
+  # treatment is 1 or 0, 1 for KO, 0 for WT
+  treat <- c()
+  for (g in 1:length(geno)) {
+    treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
+  }
+  
+  #read in the files
+  myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
[…]

### 2024-08-15 code/serverscripts/WGBS/methylkit_mincov3.R
mincov 3

@@ -1,0 +1,106 @@
+# 24.08.24 DNA methylation analysis of bismark bam files using methylKit
+    #https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#run from command line by 'Rscript [nameofscript].R'
+
+library(methylKit)
+library(ggplot2)
+# library(stringr)
+library(ChIPseeker)
+library(dplyr)
+
+#need granges as input for the bedfile
+#read in the bed file of CGIs
+bed <- readPeakFile("../data/raw/CGI_UCSC.bed")
+print(head(bed))
+
+
+#read in the bam files
+bamfiles <- list.files(path = "../data/processed/BAM_dedup", pattern = "*.sorted.dedup.bam",
+           full.names = TRUE)
+
+#remove the bam.bai files
+bamfiles <- bamfiles[!endsWith(bamfiles, '.bai')]
+
+
+#we want to compare WT and 5CKO for ESCs and EpiLCs
+cells <- c("esc", "EpiLC")
+#empty lists to store the output
+
+mylists <- list(list(), list())
+names(mylists) <- cells
+
+for (i in cells){
+    #subset for samples of that cell type
+    sampCell <- grep(i, bamfiles, value=TRUE)
+
+    #for every sample in sampcell, get the treatment (WT or KO)
+    #get the sample ID
+    ID <- gsub("../data/processed/BAM_dedup/", "", sampCell)
+    ID <- gsub(".sorted.dedup.bam", "", ID)
+    
+    #format as a list
+    ID_list <- list()
+    for (d in 1:length(ID)){
+        ID_list[[d]] <- ID[d]
+    }
+    # print("IDs as list")
+    # print(ID_list)
+    
+    #get the sample genotype (first two characters) - vector format
+    geno <- substr(ID, start = 1, stop = 2)
+    # print("treatments")
+    # print(treat)
+    # treatment is 1 or 0, 1 for KO, 0 for WT
+    treat <- c()
+    for (g in 1:length(geno)) {
+        treat[g] <- ifelse(geno[g] == "KO", 1, ifelse(geno[g] == "WT", 0, NA))
+    }
+    print(treat)
+
+    #samples in list format
+    samp_list <- list()
+    for (s in 1:length(sampCell)){
+        samp_list[[s]] <- sampCell[s]
+    }
+
+
+    #save the alignment into the correct list
[…]

### 2024-08-15 code/serverscripts/WGBS/methylkit_mincov5.R
make new folder for outputs

@@ -68,7 +68,7 @@ for (i in cells){
     #When it's run and if you save the results you can then read the results in directly using methRead
     mylists[[i]] <- processBismarkAln(location = samp_list, nolap = TRUE, mincov = 5,
 		                    sample.id = ID_list, assembly="mm10", treatment = treat, 
-                          read.context="CpG", save.folder="../results/methylKit")
+                          read.context="CpG", save.folder="../results/methylKit_5min")
     
 
 }
@@ -79,11 +79,11 @@ for (n in cells){
 	#get the methylation for cgis
 	regions <- regionCounts(mylists[[n]],bed)
 
-	united_regions <- unite(regions, destrand=FALSE)
+	united_regions <- unite(regions, destrand=TRUE)
 	print("united regions")
 	print(head(united_regions))
 
-	write.table(united_regions, paste0("../results/methylKit/regionCounts_CGI_5min_", n, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
+	write.table(united_regions, paste0("../results/methylKit_5min/regionCounts_CGI_5min_", n, "_WTvsKO.csv"), sep = ',', row.names = FALSE, quote = FALSE)
 
 
     #combine methylation calls into one
@@ -96,11 +96,11 @@ for (n in cells){
 	pooled.meth <- pool(meth, sample.ids=c("KO","WT"))
 	dm.pooledf <- calculateDiffMeth(pooled.meth, mc.cores = 8)
 
-	myDiff10p <- getMethylDiff(dm.pooledf, difference=10, qvalue=0.01)
+	myDiff10p <- getMethylDiff(dm.pooledf, difference=10, qvalue=0.1)
     #make a bedgraph file of the differences
-	bedgraph(myDiff10p, col.name = "meth.diff", file.name = paste0("../results/methylKit/bedgraph_diff_cpg_10p_5min_pooled_", n,".bed"))
+	bedgraph(myDiff10p, col.name = "meth.diff", file.name = paste0("../results/methylKit_5min/bedgraph_diff_cpg_10p_q1_5min_pooled_", n,"_WTvsKO.bed"))
   
-    write.table(myDiff10p, paste0("../results/methylKit/WGBS_CpG_10p_5min_", n, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
[…]

### 2024-08-15 submission/manuscript.Rmd
shigeki abstract edits

@@
 It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
-While some chromatin regulators are important for neuron-specific gene expression, loss of repressive chromatin regulators can also lead to the aberrant transcription of tissue-specific genes outside of their intended context.
-Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown.
-Here, we explored this cellular identity crisis in NDDs by investigating the ectopic expression of germline (e.g.
-sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
-We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534), aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
-Many genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
-Using germ cell-depleted RNA sequencing datasets, we curated a a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression.
+Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms behind their dysregulation and their functional consequences are largely unknown.
+Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
+We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
+We then curated a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression.
 Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-Germline gene dysregualtion is sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene dysregulation than knockout males.
-We found KDM5C acts at a subset of germline gene promoters in EpiLCs to promote the initial placement of DNA methylation at CpG islands.
-However, germline genes can also become activated in _Kdm5c_-KO cells independent of direct KDM5C regulation.
-Intriguingly, many of the germline transcripts expressed in the _Kdm5c_-KO brain were late-stage spermatogenesis genes that are not directly bound by KDM5C.
-This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development.
-Ultimately, this work characterizes a novel suppressor of germline gene transcription and links impaired soma-vs-germline demarcation to a chromatin-linked neurodevelopmental disorder.
+Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males.
+We found KDM5C represses germline-specific transcription by binding to a subset of germline gene promoters in EpiLCs to facilitate DNA methylation of CpG islands.
+However, germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding.
+This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development, due to downstream activation by germline transcription factors.
+These results define KDM5C's role in germline gene suppression and potentially link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.
+
+- Not sure if I like the last sentence, I think we should focus on what we did do versus what we could do in the future
 

### 2024-08-15 submission/manuscript.Rmd
shigeki edits intro

@@
 - Not sure if I like the last sentence, I think we should focus on what we did do versus what we could do in the future
+	- Add something that we deepened the characterization of germline gene classes that use different regulatory mechanisms (CpG islands vs not, meiotic vs late-stage, E2F6/MAX vs no)
 
@@
 Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Unexpectedly, next generation sequencing revealed mutations in many chromatin regulators directly cause or are associated with neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
-Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009].
+Unexpectedly, human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
+Most studies investigating this relationship have explored their regulation of brain-specific genes and chromatin marks.
 However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
-Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if ectopic gene expression contributes to neurodevelopmental impairments.
+Very few studies have investigated this severe crisis in cellular identity in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if these ectopic genes contribute to neurodevelopmental impairments.
 
 To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C).
-KDM5C, also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters.
+We characterized lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3) found at active gene promoters[@iwaseXLinkedMentalRetardation2007].
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
-MRXSCJ is more common and often more severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
+MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
 Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
-It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, if this phenomenon is sexually dimorphic, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017], however it is unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development testis gene dysregulation begins, and if other tissue-specific genes are also aberrantly transcribed with KDM5C loss.
 
-The testis contains both germ cells (cells that pass on their genetic material, e.g.
-sperm) and somatic cells (cells that perform all other bodily functions).
-Distinguishing between the germline and the soma is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012].
+Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012].
 In mammals, chromatin regulators are crucial for decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
 Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
-Currently, most studies have focused on marker genes important for early germ cell development rather than germline genes as a whole given the lack of a curated list of germline-enriched genes.
-Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g.
+The precise roles of KDM5C during this process remains unclear.
+Additionally, studies on germline gene repression have primarily focused on marker genes important for germ cell development rather than germline genes as a whole, given they lacked a curated list of germline-enriched genes.
+Therefore, it is unknown if the mechanism of repression differs for certain classes of germline genes, e.g.
 meiotic genes versus spermatid differentiation genes.
-Furthermore, it is currently unknown if ectopic germline gene expression is influenced by chromosomal sex, as previous studies on germline gene repression have been exclusively in males.
-Sex is particularly pertinent in the case of sexually dimorphic chromatin regulators like KDM5C, which lies on the X chromosome and partially escapes X chromosome inactivation, resulting in 1.2-2-fold higher in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Thus, systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, will unveil key mechanisms underlying the demarcation between soma and germline identity.
-These results will provide molecular footholds that can then be exploited to test the contribution of ectopic germline genes on neurodevelopment.
+
+It is also unknown if ectopic germline gene expression is influenced by chromosomal sex, as previous studies on germline gene repression have been exclusively in males.
+Sex is particularly pertinent in the case of KDM5C, which lies on the X chromosome and partially escapes X chromosome inactivation, resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 
 To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes.
-The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells and not somatic cells of the testis.
-We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
-Female _Kdm5c_-KO EpiLCs expressed over double the number of germline genes than _Kdm5c_-KO males and germline genes dysregulated in both sexes were expressed more highly in females, demonstrating females have increased sensitivity to germline gene dysregulation.
-KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
-Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs.
-Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+We curated list of germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms.
+Based on the data presented below, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
 
@@
 # Discussion
+We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes.
+The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells and not somatic cells of the testis.
+We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
+Female _Kdm5c_-KO EpiLCs expressed over double the number of germline genes than _Kdm5c_-KO males and germline genes dysregulated in both sexes were expressed more highly in females, demonstrating females have increased sensitivity to germline gene dysregulation.
+KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
+Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs.
+- Change: we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+
+- Thus, systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, will unveil key mechanisms underlying the demarcation between soma and germline identity.
+These results will provide molecular footholds that can then be exploited to test the contribution of ectopic germline genes on neurodevelopment.
 

### 2024-08-15 submission/manuscript.Rmd
shigeki edits results

@@
 The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C).
-Even though the testis has the largest total number of tissue-biased genes compared to any other tissue (`r gettissval("count", "Testis")` genes), testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
-One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
+Even though the testis has the largest total number of biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
+One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
 
@@
 
-To further validate if these testis DEGs are truly germline genes, we then assessed their expression in somatic versus germ cells within the testis.
+To further validate if these testis DEGs are truly germline genes, we assessed their expression in somatic versus germ cells within the testis.
 We first compared their expression in wild-type testes to those without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
@@
 spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C).
-Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion between somatic versus germline identity.
+Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.
 
@@
 Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C).
-Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D) appeared normal, indicating KDM5C loss does not impair EpiLC formation.
+Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating KDM5C loss does not impair EpiLC formation.
 
-To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain.
+We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline DEGs are constitutively dysregulated or change over the course of development.
 We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
 EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).
+While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).
 
@@
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see misexpression of two-cell embryo-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H).
+However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H).
 Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.
@@
 
-It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have focused on their regulation in males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs.
+It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
 [@samantaActivationXistEvolutionarily2022].
-We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.
 
 Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
-While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B).
+While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B), such as _Tktl2_ and _Esx1_.
 We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared).
-Female-specific germline DEGs were enriched for meiotic (GO:XXX meiotic cell cycle) and flagellar (GO:XXX cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:XXX protein localization to mitochondrion).
-Germline transcripts expressed in both sexes were enriched for meiotic (meiotic nuclear division) and egg-specific functions (female gamete generation).
+Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:0070585 protein localization to mitochondrion).
+Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).
 
-The majority of germline genes expressed in both sexes had a greater log2 fold change from wild-type in females compared to males (Figure 4D).
-This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to the loss of KDM5C-mediated suppression of germline genes.
-Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants, as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes instead lying on autosomes (Figure 4G).
+The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D).
+This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated suppression of germline genes.
+Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+However, both shared and female-specific germline DEGs were not biased towards the X chromosome and the majority of female DEGs instead lie on autosomes (Figure 4G).
 Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.
@@
 While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
-To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg (XX) and sperm (XY) biased genes.
-We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H), which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].
+Therefore, we wondered if _Kdm5c_ mutant males express sperm genes while mutant females express egg genes.
+To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes.
+We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H).
 We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J).
-Germline genes dysregulated exclusively in males and females were also not biased towards their corresponding germ cell sex.
-This indicates differences between male and female germline gene dysregulation is not due to sex-specific activation of sperm or egg-specific transcriptional programs.
-These results demonstrate that sex influences the degree of germline gene misexpression with KDM5C loss, but not the germ cell sex from which the germline genes originate.
+Germline genes dysregulated exclusively in_Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex.
+This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.
+These results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.
+
+- **note: the edited last sentence ended with "independent of germ cell sex", but I think if it's read out of context that sounds like we tested KDM5C in the four core genotypes.
+But I'm not sure if "germline gene sex" is confusing**
 
@@
 
-Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017].
-KDM5C binds to several germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017].
+KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017].
 However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters.
-It is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
+Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
 
@@
 We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 5D).
+In EpiLCs, we observed modest KDM5C signal at about half of all germline genes (Figure 5D).
 Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).
-In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 5D).
-Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
+One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs (Figure 5K).
+In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%), but 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX).
+In condordance with our gene ontology results, we did not observe KDM5C accumulation at any germline gene promtoers in PNCs (Figure 5D).
+Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
+Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
 
@@
 One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5G).
-KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, Rfx2 q-value < 0.0001, Rfx5 q-value < 0.0001) (Figure 5H).
-RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
-Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs (Figure 5I), but is not a direct target of KDM5C.
-Thus, distinct transcription factor programs regulate the expression of KDM5C-bound and unbound germline genes.
-
-Finally, we compared KDM5C binding at the promoters of RNA-seq DEGs to determine if the germline genes transcribed in _Kdm5c_-KO cells are direct targets of KDM5C (Figure 5J-L).
-In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEG promoters (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%).
-Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs.
-Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above.
-Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 5K).
-In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS.
-Altogether, this suggests KDM5C decommssions a subset of germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons.
-Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.
+KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5H).
+RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
+Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5I), however it is not a direct target of KDM5C (Supplementary figure XXX).
+Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.
 

### 2024-08-16 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
calculate DMRs for WT

@@ -5,7 +5,7 @@
 #run from command line by 'Rscript [nameofscript].R'
 
 #qvalue cut off
-qval <- 0.1
+qval <- 0.01
 
 
 library(methylKit)
@@ -59,69 +59,79 @@ treat[g] <- ifelse(cell[g] == "EpiLC", 1, ifelse(cell[g] == "ESC", 0, NA))
 myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
                    context="CpG", mincov = 5)
 #filter based on read coverage
-myobj <- filterByCoverage(myobj,lo.count=5,lo.perc=NULL,
+myobj <- filterByCoverage(myobj,lo.count=3,lo.perc=NULL,
                     hi.count=NULL,hi.perc=99.9)
   
 
 #get the methylation for cgis
-regions <- regionCounts(myobj,bed)
+regions <- regionCounts(myobj, bed)
 print("all regions")
 print(head(regions))
 
-united_regions <- unite(regions, destrand=FALSE)
-print("united regions")
-print(head(united_regions))
+# united_regions <- unite(regions, destrand=TRUE)
+# print("united regions")
+# print(head(united_regions))
+
+#then do differential methylation for the regions
+dm.regions <- calculateDiffMeth(regions, mc.cores = 8)
+#this should be like a DESEQ2 results table
+head(dm.regions)
 
 #just saving the one methylation results. Want the average methylation
-write.table(united_regions, paste0("../results/methylKit/regionCounts_germCGI_WT.csv"), sep = ',', row.names = FALSE, quote = FALSE)
+write.table(dm.regions, "../results/methylKit/WGBS_restab_regionCounts_germCGI_WT_min3_ESCvsEpiLC.csv", sep = ',', row.names = FALSE, quote = FALSE)
+
+#get just the significant ones - should be like the DESeq2 DEGs
+myDiff_regions <- getMethylDiff(dm.regions, difference=10, qvalue=qval)
+write.table(myDiff_regions, "../results/methylKit/WGBS_getmethyldiff_germCGI_p10_min3_q01_WT_ESCvsEpiLC.csv", sep = ',', row.names = FALSE, quote = FALSE)
+
 
 
 #pool the samples for calculating the methylation differences
 #unite the samples
[…]

### 2024-08-16 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
fixed diffmeth for CGIs to united regions

@@ -65,15 +65,16 @@ myobj <- filterByCoverage(myobj,lo.count=3,lo.perc=NULL,
 
 #get the methylation for cgis
 regions <- regionCounts(myobj, bed)
-print("all regions")
-print(head(regions))
+# print("all regions")
+# print(head(regions))
 
-# united_regions <- unite(regions, destrand=TRUE)
-# print("united regions")
-# print(head(united_regions))
+united_regions <- unite(regions, destrand=FALSE)
+  #can try setting min per group to 1 so bases only have to be covered in one sample/group to be counted
+print("united regions")
+print(head(united_regions))
 
 #then do differential methylation for the regions
-dm.regions <- calculateDiffMeth(regions, mc.cores = 8)
+dm.regions <- calculateDiffMeth(united_regions, mc.cores = 8)
 #this should be like a DESEQ2 results table
 head(dm.regions)
 

### 2024-08-16 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
fixed diffmeth for Cs to united CpGs

@@ -95,8 +95,10 @@ write.table(myDiff_regions, "../results/methylKit/WGBS_getmethyldiff_germCGI_p10
 
 #methylation differences unpooled
 dmC <- calculateDiffMeth(myobj, mc.cores = 8)
+united_dmC <- unite(dmC, destrand=FALSE)
+
 #all of the bases
-myDiff10p <- getMethylDiff(dmC, difference=10, qvalue=qval)
+myDiff10p <- getMethylDiff(united_dmC, difference=10, qvalue=qval)
 #make a bedgraph file of the differences
 bedgraph(myDiff10p, col.name = "WT ESC vs EpiLC min3 q01 unpooled", file.name = "../results/methylKit/bedgraph_diff_WT_ESCvsEpiLC_p10_min3_q01_notpooled.bed")  
   

### 2024-08-16 code/serverscripts/WGBS/methyl_diff_germCGI_WT.R
fix grouping problem with CpG methylation differences

@@ -93,12 +93,13 @@ write.table(myDiff_regions, "../results/methylKit/WGBS_getmethyldiff_germCGI_p10
 # pooled.meth <- pool(meth, sample.ids=c("EpiLC","ESC"))
 # dm.pooledf <- calculateDiffMeth(pooled.meth, mc.cores = 8)
 
-#methylation differences unpooled
-dmC <- calculateDiffMeth(myobj, mc.cores = 8)
-united_dmC <- unite(dmC, destrand=FALSE)
+#methylation differences pooled
+united_dmC <- unite(myobj, destrand=FALSE)
+
+dm.C <- calculateDiffMeth(united_dmC, mc.cores = 8)
 
 #all of the bases
-myDiff10p <- getMethylDiff(united_dmC, difference=10, qvalue=qval)
+myDiff10p <- getMethylDiff(dm.C, difference=10, qvalue=qval)
 #make a bedgraph file of the differences
 bedgraph(myDiff10p, col.name = "WT ESC vs EpiLC min3 q01 unpooled", file.name = "../results/methylKit/bedgraph_diff_WT_ESCvsEpiLC_p10_min3_q01_notpooled.bed")  
   

### 2024-08-16 code/serverscripts/WGBS/deeptools_heatmap.py
begin writing deeptools script

@@ -1,0 +1,1 @@
+#20240816 deeptools heatmap of bedgraph files for differential methylation

### 2024-08-16 submission/manuscript.Rmd
add summary paragraph at beg of discussion

@@
 # Discussion
-We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes.
-The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells and not somatic cells of the testis.
+In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
+We observed global dysregulation of tissue-enriched genes in the _Kdm5c_-KO brain and EpiLCs, including substantial misexpression of testis, liver, muscle, and ovary-enriched genes.
+The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells, rather than somatic cells.
 We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
-Female _Kdm5c_-KO EpiLCs expressed over double the number of germline genes than _Kdm5c_-KO males and germline genes dysregulated in both sexes were expressed more highly in females, demonstrating females have increased sensitivity to germline gene dysregulation.
-KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms.
-Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs.
-- Change: we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+We demonstrated female EpiLCs are more sensitive to germline gene de-repression, as the number germline genes misexpressed and their degree of dysregulation was greater in _Kdm5c_-KO females compared to males.
+KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms.
+Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation at CpG islands for a subset of germline genes in EpiLCs.
+Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+By systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, we unveiled unique regulatory mechansisms governing distinct germline gene classes in somatic lineages.
+Altogether, these results provide molecular footholds that in the future can be exploited to test the contribution of ectopic germline genes upon neurodevelopment.
 
-- Thus, systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, will unveil key mechanisms underlying the demarcation between soma and germline identity.
-These results will provide molecular footholds that can then be exploited to test the contribution of ectopic germline genes on neurodevelopment.
+---
+
+### Combining sections
 

### 2024-08-16 submission/manuscript.Rmd
merging paragraphs

@@
 
----
+It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal.
+For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
+Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].
+However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023].
+We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes.
+This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
 
-### Combining sections
+Although expressed at the 2-cell stage an in ESCs, _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
 
-In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala.
-In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
-Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
-However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
-Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.
+Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
+Proper _Otx2_ expression, coupled normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
 
-Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes but instead germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary.
-Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023].
-Previous work characterizing repressors of germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-To characterize KDM5C's role in germline gene repression throughout life at a genome-wide level, we curated a list of germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
-This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, 3) the impact of germ cell sex upon germline gene misexpression, and 4) which groups of germline genes are directly and indirectly regulated by KDM5C.
-This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.
+misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells
 
-Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
-_In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020].
-However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
-This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
-Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs.
-Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation.
-Proper EpiLC differentiation, coupled with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
+Since epiblast stem cells can reactivate germline genes and differentiate into primoridal germ cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
+
+### males vs females
 
@@
 However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs.
-Knockout of KDM5C in females is embryonic lethal, likely due to impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022].
-XCI defects could explain why _Kdm5c_-KO females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+Knockout of KDM5C in females is embryonic lethal and results in impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022].
+XCI defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
 However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome.
 Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Intriguingly, females heterozygous for _Kdm5c_ loss also had over double the number of germline DEGs than males, even though their XCI is largely preserved and their expression of KDM5C should be roughly equivalent to that of wild-type males.
-Altogether, this suggests female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity.
-Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators like KDM5C or a general feature of female cells.
+Intriguingly, heterozygous loss of _Kdm5c_ in females also had over double the number of germline DEGs than males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males.
+Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity.
+Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.
+
+### ChIPseq motifs and transcription factors
 
@@
 
-It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal.
-For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-Very recently, two screens of _Dazl_-repressors identified KDM5C as a direct regulator in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-Interestingly, one study found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency[@guptaGenomewideScreenReveals2023].
-However, out of the four regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023].
-Although expressed at the 2-cell stage and in ESCs, _Dazl_ is typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-We found KDM5C also repressess _Dazl_ in EpiLCs, but _Kdm5c_-KO EpiLCs did not express 2-cell specific genes like _Dux_ and _Zscan4c_.
-Together, this suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
+### Germline gene decommissioning dna methylation
+- End with comment on catalytic dependent vs independent functions of chromatin regulators
+	- Requirement for catalytic activity may change depending upon the context
 
@@
 
+### Functional consequences 
+In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
+Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
+However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
+Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.
+
 The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development.
@@
 Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation.
-Further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.
+Ultimately, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
+
+---
+
+Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes but instead germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary.
+Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023].
+Previous work characterizing repressors of germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+To characterize KDM5C's role in germline gene repression throughout life at a genome-wide level, we curated a list of germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
+This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, 3) the impact of germ cell sex upon germline gene misexpression, and 4) which groups of germline genes are directly and indirectly regulated by KDM5C.
+This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.
+
+Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
+_In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020].
+However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
+This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
 

### 2024-08-16 submission/manuscript.Rmd
discussion 2-cell and sex

@@
 This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
+_Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
+Yet, _Kdm5c_-KO EpiLCs properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
+Proper _Otx2_ expression, coupled with normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
 
-Although expressed at the 2-cell stage an in ESCs, _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
-
-Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Proper _Otx2_ expression, coupled normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
-
-misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells
-
-Since epiblast stem cells can reactivate germline genes and differentiate into primoridal germ cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
-
-### males vs females
-
-While many genes involved in early germline development and meiosis are common between males and females, some germline genes display sex-biased expression or perform functions unique to eggs and sperm.
+While many genes involved in early germline development are common between males and females, some display sex-biased expression or perform functions unique to eggs and sperm.
 We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex.
@@
 
-Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which germline genes can be misexpressed in _Kdm5c_-KO cells.
-While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs.
-This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress two aberrant germline transcripts[@scandagliaLossKdm5cCauses2017].
-It is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding.
+Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells.
+Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017],
+
+However, it is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding.
+
  In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].

### 2024-08-17 submission/manuscript.Rmd
discussion motifs

@@
 
-### ChIPseq motifs and transcription factors
-
 Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells.
 Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017],
-
-However, it is unclear how KDM5C is recruited to germline genes given that KDM5C itself does not contain domains for sequence-specific binding.
-
+It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding.
 In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
-While MAX and E2F6 motifs were enriched at KDM5C-bound promoters in EpiLCs, only about one third of KDM5C-bound germline genes contained their consesus sequence at their promoter.
-Thus, other, unknown factors likely facilitate KDM5C's recruitment to germline genes.
-
-We also found germline genes can become activated in _Kdm5c_-KO cells indepedent of direct loss of KDM5C suppression, given that KDM5C was only bound to about a third of germline-enriched DEGs.
-One notable EpiLC DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells to promote meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
-However, retinoic acid can only activate _Stra8_ expression when DAZL is present.
-Unlike _Stra8_, _Dazl_ is a direct target of KDM5C in EpiLCs and is transcribed _Kdm5c_-KO EpiLCs.
-Expression of indirect DEGs like _Stra8_ indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells through downstream activation by other ectopic germline programs.
-In support of this, germline genes unbound by KDM5C were significantly enriched for RFX transcription factors.
-RFX transcription factors bind to X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
+While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promtoers contained their consesus sequence.
+Additionally, germline genes like _Stra8_ can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters.
+_Stra8_ is a meiotic transcription factor that is typically turned on in the presence of retinoic acid signaling and DAZL[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006], which is a direct KDM5C target.
+Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors that promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
 One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs.
-Intruigingly, _Kdm5c_-KO EpiLCs primarily expressed germline genes involved in early germ cell development and meiosis while the mature _Kdm5c_-KO brain expressed late-stage spermatogenesis genes, including known RFX2 targets.
-Altogether, these data suggest once activated due to loss of direct KDM5C repression, ectopic germline programs can loosely mimic germline development and progress in the background of _Kdm5c_-KO somatic development.
+Altogether, these data suggest that once germline genes are activated due to loss of direct KDM5C repression, they can activate other ectopic germline programs downstream that loosely mimic germline development.
 

### 2024-08-20 code/serverscripts/WGBS/bedgraph_heat.py
draft code deeptools

@@ -1,45 +1,0 @@
-#2024.08.19 Plotting average WGBS methylation at germline loci using bedgraph, divided by KDM5C bound and unbound genes
-#activate the bedtools environment in conda
-
-import os
-import sys
-import glob
-
-#bigwigs
-EpiLC = ['/scratch/webserver/siwase/Testis_genes/KDM5C_H3K4me2_EpiLCs_mm10/SMCX_WT_Merged_IP_mm10.bw','/scratch/webserver/siwase/Testis_genes/KDM5C_H3K4me2_EpiLCs_mm10/SMCX_KO_Merged_IP_mm10.bw']
-PNC = ['/scratch/webserver/siwase/Testis_genes/Iwase_et_al_2016_Cell_Reports_mm10/KDM5C_ChIP_WT_CN_rep1_mm10.bw','/scratch/webserver/siwase/Testis_genes/Iwase_et_al_2016_Cell_Reports_mm10/KDM5C_ChIP_KO_CN_rep1_mm10.bw']
-
-
-
-#bedfile of the region that you want to plot
-germregions = glob.glob("../data/processed/GermTSSandTES/TSS_window_3000*.bed")
-
-#wrap all the deeptools fucntions together
-#regions - gene coordinates
-#bigwigs - bigwigs you're plotting
-#colors - color for heatmap
-def deeptoolswrap(regions, bigwigs, name):
-	for i in regions:
-
-		#name of the output - ID is the bigwig id, coord is the bed cooridnates
-		coord = i.replace("../data/processed/GermTSSandTES/TSS_window_", "")
-		coord = coord.replace(".bed", "")
-		
-		window = 3000
-		out = "../results/plotProfile_TSS_KDM5C_{NAME}_{COORD}".format(COORD = coord, NAME = name)
-		
-		#make the compute matrix
-		os.system('computeMatrix reference-point -S {BIGWIGS} -R {REGIONS} --referencePoint center -a {WINDOW} -b {WINDOW} -o {OUT}_matrix.mat.gz'.format(BIGWIGS = " ".join(bigwigs), REGIONS = i, WINDOW = window, OUT = out))
-		title = "KDM5C_{COORD}".format(COORD = coord)
-		
-		#make the average plot --perGroup splits by bedfile so all bigwigs are plotted on same graph
[…]

### 2024-08-20 code/serverscripts/WGBS/deeptools_heatmap.py

@@ -1,1 +1,47 @@
-#20240816 deeptools heatmap of bedgraph files for differential methylation
+#2024.08.19 Plotting average WGBS methylation at germline loci using bedgraph, divided by KDM5C bound and unbound genes
+#activate the bedtools environment in conda
+
+#convert bedgraph to bigwig - https://anaconda.org/bioconda/ucsc-bedgraphtobigwig
+
+import os
+import sys
+import glob
+
+#bigwigs
+EpiLC = ['/scratch/webserver/siwase/Testis_genes/KDM5C_H3K4me2_EpiLCs_mm10/SMCX_WT_Merged_IP_mm10.bw','/scratch/webserver/siwase/Testis_genes/KDM5C_H3K4me2_EpiLCs_mm10/SMCX_KO_Merged_IP_mm10.bw']
+PNC = ['/scratch/webserver/siwase/Testis_genes/Iwase_et_al_2016_Cell_Reports_mm10/KDM5C_ChIP_WT_CN_rep1_mm10.bw','/scratch/webserver/siwase/Testis_genes/Iwase_et_al_2016_Cell_Reports_mm10/KDM5C_ChIP_KO_CN_rep1_mm10.bw']
+
+
+
+#bedfile of the region that you want to plot
+germregions = glob.glob("../data/processed/GermTSSandTES/TSS_window_3000*.bed")
+
+#wrap all the deeptools fucntions together
+#regions - gene coordinates
+#bigwigs - bigwigs you're plotting
+#colors - color for heatmap
+def deeptoolswrap(regions, bigwigs, name):
+	for i in regions:
+
+		#name of the output - ID is the bigwig id, coord is the bed cooridnates
+		coord = i.replace("../data/processed/GermTSSandTES/TSS_window_", "")
+		coord = coord.replace(".bed", "")
+		
+		window = 3000
+		out = "../results/plotProfile_TSS_KDM5C_{NAME}_{COORD}".format(COORD = coord, NAME = name)
+		
+		#make the compute matrix
+		os.system('computeMatrix reference-point -S {BIGWIGS} -R {REGIONS} --referencePoint center -a {WINDOW} -b {WINDOW} -o {OUT}_matrix.mat.gz'.format(BIGWIGS = " ".join(bigwigs), REGIONS = i, WINDOW = window, OUT = out))
+		title = "KDM5C_{COORD}".format(COORD = coord)
+		
[…]

### 2024-08-20 code/WGBS_ESCvsEpiLC.R
volcano of WGBS

@@ -1,0 +1,56 @@
+# 24.08.19 - WGBS volcano plots
+
+#read in the KDM5C bound genes
+KDM5C_binding <- read.csv(snakemake@input[[1]], sep = ",")
+
+print(head(KDM5C_binding))
+
+
+samples <- c("all germ promoters", "germ CGI promoters")
+
+library(ChIPseeker)
+library(TxDb.Mmusculus.UCSC.mm10.knownGene)
+txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
+library(ggplot2)
+library(dplyr)
+
+volcano_WGBS <- function(n){
+	df <- read.csv(snakemake@input[[n+1]], sep = ",")
+
+	#rename chr to seqnames to match granges for merging later
+	names(df)[names(df) == "chr"]<- "seqnames"
+	print(paste("rows in", samples[n], "-", nrow(df)))
+	print(head(df))
+	
+
+	#get the germline genes for each coordinate
+		#make granges object with bed coordinates
+	peak <- makeGRangesFromDataFrame(df)
+	print(head(peak))
+
+		#annotate locations
+	peakAnno <- annotatePeak(peak, tssRegion=c(-500, 500),
+                         TxDb=txdb, annoDb="org.Mm.eg.db")
+	
+	annot <- as.data.frame(peakAnno@anno)
+	annot_promo <- subset(annot, annotation == "Promoter")
+	#annot_promo_genes <- annot_promo %>% distinct(ENSEMBL, .keep_all = TRUE)
+
+	print(paste("rows in annotated", samples[n], "-", nrow(annot_promo)))
+	print(head(annot_promo))
+
+	merge <- merge(annot_promo, df)
+	print(paste("rows in merged", samples[n], "-", nrow(merge)))
+	print(head(merge))
+
+	#its working for the CGIs but all of the promoters there are extras. maybe its because multiple genes have TSSs in that location
+		#Do I necessarily need the gene location for the volcano plot? no. Just need a unique name for each region
+
+}
+
+
+for (i in 1:length(samples)){
+	volcano_WGBS(i)
+}
+
+

### 2024-08-20 submission/manuscript.Rmd
combine discussion cpgme and impact

@@
 KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms.
-Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation at CpG islands for a subset of germline genes in EpiLCs.
+Finally, we found KDM5C is dynamically regulated during the differentiation of ESCs in EpiLCs and promotes long-term germline gene silencing by aiding the placement of DNA methylation at CpG islands in EpiLCs.
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
@@
 Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells.
-Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017],
+Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017].
 It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding.
@@
 
-### Germline gene decommissioning dna methylation
-- End with comment on catalytic dependent vs independent functions of chromatin regulators
-	- Requirement for catalytic activity may change depending upon the context
+Unlike the previously characterized germline gene suppressors[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
+However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+Emerging work indicate many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression at least as potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
+While dispensible in ESCs, KDM5C's catalytic activity may be necessary in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
+Indeed, we found _Kdm5c_-KO EpiLCs have impaired CpGme accumulation at germline gene CpG islands, indicating H3K4me3 removal is necessary at this developmental stage.
+Thus, the shift in the mechanism of KDM5C-mediated repression during the transition from na&#239;ve to primed pluripotency demonstrates the cataltyic activity of histone-modifying enzymes can become necessary within different cellular contexts.
 
-_Dazl_ and other germline-enriched genes are initially decommissioned in ESCs by repressive histone modifications and then _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-Unlike the previously characterized germline gene suppressors, KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
-In this study, we observed a global increase in H3K4me2/3 in _Kdm5c_-KO EpiLCs and amygdala around the transcription start site (TSS) of germline-enriched genes.
-However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
-Although not necessary in ESCs, KDM5C's catalytic activity be required for CpGme in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
-This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found CpGme is significantly eroded at at least two germline promoters[@scandagliaLossKdm5cCauses2017].
-To elucidate the mechanism behind KDM5C-mediated silencing of germline genes, we characterized KDM5C in ESCs and EpiLCs and found its expression is dynamically regulated during differenitation.
-Furthermore, KDM5C acts during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.
-
-- CPG islands greatly detrmine KDM5C recruitment - KDM5C is previously known to be enriched at CGIs. CGIs not typcially methylated, germline genes exception. KDM5C promotes this process.
-
-### Functional consequences 
-In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
-Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression.
-However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
-Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.
+- DNAme and CpG islands
+	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
+	- Germline genes are methylated at promoter CGIs, which are typically not methylated in other gene types
+	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
+	- We found not all germline-enriched genes have CGIs, but still gain CpGme around TSS. Unclear what the recruitment mechanism is for these genes
+	- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they also participate in non-CGI TSS CpGme
+	- CGIs highly determine KDM5C recruitment to germline gene promoters and KDM5C germline gene CGI methylation.
+		- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
+	- Different classes of germline genes use distinct silencing mechanisms
 
@@
 However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
-KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes, such as _Apoc1_.
-KDM5C could also ensure tissue-specific timing of transcription factor family members, like RFX genes that are broadly required for cilia and flagella formation, including neuronal cilia[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009; @chungRFX2BroadlyRequired2012; @coschieraPrimaryCiliaPromote2024].
-While their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012].
-Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression.
-Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation.
-Ultimately, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
-
----
-
-Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes but instead germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary.
-Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023].
-Previous work characterizing repressors of germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-To characterize KDM5C's role in germline gene repression throughout life at a genome-wide level, we curated a list of germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015].
-This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, 3) the impact of germ cell sex upon germline gene misexpression, and 4) which groups of germline genes are directly and indirectly regulated by KDM5C.
-This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.
-
-Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis.
-_In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020].
-However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
-This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
+In addition to germline genes, we also found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
+Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
+_Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
+Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.
+While the impact of ectopic germline transcripts upon neurodevelopment is currently unknown, some models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
+Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression.
+Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation, however further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
 

### 2024-08-20 submission/manuscript.Rmd
edit discussion p1

@@
 # Discussion
-In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
-We observed global dysregulation of tissue-enriched genes in the _Kdm5c_-KO brain and EpiLCs, including substantial misexpression of testis, liver, muscle, and ovary-enriched genes.
-The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells, rather than somatic cells.
-We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis.
-We demonstrated female EpiLCs are more sensitive to germline gene de-repression, as the number germline genes misexpressed and their degree of dysregulation was greater in _Kdm5c_-KO females compared to males.
-KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms.
-Finally, we found KDM5C is dynamically regulated during the differentiation of ESCs in EpiLCs and promotes long-term germline gene silencing by aiding the placement of DNA methylation at CpG islands in EpiLCs.
+In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity and characterized substantial misexpression of testis, liver, muscle, and ovary-enriched genes within the _Kdm5c_-KO brain.
+Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not somatic cells.
+_Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis.
+We demonstrated that while _Kdm5c_ mutant sex did not influence the whether sperm or egg genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
+Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as KDM5C is enriched at germline gene promoters in EpiLCs but only bound to a subset of germline-enriched DEGs.
+Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands.
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, we unveiled unique regulatory mechansisms governing distinct germline gene classes in somatic lineages.
-Altogether, these results provide molecular footholds that in the future can be exploited to test the contribution of ectopic germline genes upon neurodevelopment.
+By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages.
+Furthermore, these results provide molecular footholds that can be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.
 
-It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal.
-For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
+It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal.
+For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
 Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].

### 2024-08-20 submission/manuscript.Rmd
edit discussion p2

@@
 # Discussion
-In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity and characterized substantial misexpression of testis, liver, muscle, and ovary-enriched genes within the _Kdm5c_-KO brain.
+In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity and identified substantial misexpression of testis, liver, muscle, and ovary-enriched genes within the _Kdm5c_-KO brain.
 Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not somatic cells.
@@
 For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].
-However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023].
-We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes.
-This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
-_Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
-Yet, _Kdm5c_-KO EpiLCs properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Proper _Otx2_ expression, coupled with normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
+KDM5C may therefore negatively regulate totipotency, since KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].
+However, _Kdm5c_-KO misexpression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter.
+Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into EpiLCs[@mochizukiRepressionGermlineGenes2021].
+We found _Kdm5c_-KO EpiLCs expressed _Dazl_ but did not express 2-cell specific genes, suggesting the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
+Misexpression of germline genes in _Kdm5c_-KO EpiLCs could indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
+Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
+Altogether, these results demonstrate KDM5C repressess germline identity during embryogenesis and _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs.
 

### 2024-08-20 submission/manuscript.Rmd
edit discussion p3

@@
 
-While many genes involved in early germline development are common between males and females, some display sex-biased expression or perform functions unique to eggs and sperm.
-We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex.
-However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs.
-Knockout of KDM5C in females is embryonic lethal and results in impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022].
-XCI defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome.
+Although eggs and sperm intially express very similar sets of germline genes for PGC formation and meiosis, some germline genes are sexually dimorphic.
+We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of repression is independent of a given germline gene's sex.
+However, organismal sex did greatly influence the manifestation of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males.
+X chromosome inactivation (XCI)  defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022].
+However, female germline DEGs were not significantly biased towards the X chromosome.
 Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Intriguingly, heterozygous loss of _Kdm5c_ in females also had over double the number of germline DEGs than males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males.
+Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
 Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity.
-Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.
+Future studies are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.
 

### 2024-08-20 submission/manuscript.Rmd
last 3 discussion paragraphs

@@
 
-Our analysis of KDM5C ChIP-seq data provided novel insights into direct and indirect mechanisms by which different classes of germline genes can be misexpressed in _Kdm5c_-KO cells.
-Transcription of germline genes in the mature _Kdm5c_-KO brain is likely due to loss of KDM5c-mediated repression during embryogenesis, given that KDM5C was bound to their promoters in EpiLCs, is not recruited to germline genes in neurons, and germline transcription is not repressed when KDM5C in reintroduced in neurons[@scandagliaLossKdm5cCauses2017].
-It is unclear how KDM5C is recruited to germline genes, given that KDM5C itself does not contain domains for sequence-specific binding.
+It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding.
 In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
-While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promtoers contained their consesus sequence.
-Additionally, germline genes like _Stra8_ can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters.
-_Stra8_ is a meiotic transcription factor that is typically turned on in the presence of retinoic acid signaling and DAZL[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006], which is a direct KDM5C target.
-Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors that promote the transcription of cilia and flagellar genes[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009].
-One of the top RFX members enriched included RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016] whose mRNA is also expressed in _Kdm5c_-KO EpiLCs.
-Altogether, these data suggest that once germline genes are activated due to loss of direct KDM5C repression, they can activate other ectopic germline programs downstream that loosely mimic germline development.
+While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promoters contained their consesus sequence.
+Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters.
+Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
+Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
+Together, this suggests once germline genes like _Dazl_ and _Rfx2_ are activated in _Kdm5c_-KO cells, they can then turn on other ectopic germline programs downstream and loosely mimic germ cell development.
 
-Unlike the previously characterized germline gene suppressors[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
-However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
-Emerging work indicate many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression at least as potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
-While dispensible in ESCs, KDM5C's catalytic activity may be necessary in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
-Indeed, we found _Kdm5c_-KO EpiLCs have impaired CpGme accumulation at germline gene CpG islands, indicating H3K4me3 removal is necessary at this developmental stage.
-Thus, the shift in the mechanism of KDM5C-mediated repression during the transition from na&#239;ve to primed pluripotency demonstrates the cataltyic activity of histone-modifying enzymes can become necessary within different cellular contexts.
+Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
+Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression even more potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
+Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
+Additionally, we found _Kdm5c_-KOs accumulate H3K4me2/3 around the TSS of germline genes and impaired CpGme placement at CpG islands.
+KDM5C's shifting mode of germline gene repression as cells transition from na&#239;ve to primed pluripotency indicates chromatin regulators can employ both catalytic and non-catlytic repressive mechanisms at the same loci depending upon the developmental and cellular context.
 
@@
 	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
-	- Germline genes are methylated at promoter CGIs, which are typically not methylated in other gene types
+	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
 	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
-	- We found not all germline-enriched genes have CGIs, but still gain CpGme around TSS. Unclear what the recruitment mechanism is for these genes
-	- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they also participate in non-CGI TSS CpGme
-	- CGIs highly determine KDM5C recruitment to germline gene promoters and KDM5C germline gene CGI methylation.
+	- We found not all germline-enriched genes have CGIs, but many still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes
+	- CGIs highly determine KDM5C recruitment to germline gene promoters.
 			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
-	- Different classes of germline genes use distinct silencing mechanisms
+	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
+		- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they participate in non-CGI TSS CpGme
+	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
 
 The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development.
-However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
-In addition to germline genes, we also found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain.
-Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus.
-_Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
+However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
+In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain.
+Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that can impair learning and memory when overexpressed in the mouse brain[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
 Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.
-While the impact of ectopic germline transcripts upon neurodevelopment is currently unknown, some models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
+Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
+Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating they may promote genome instability and cellular de-differentiation.
+Intriguingly, select models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
 Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression.
-Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation, however further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
+Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation However, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
 

### 2024-08-21 submission/manuscript.Rmd
discussion small edits

@@
 # Discussion
-In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity and identified substantial misexpression of testis, liver, muscle, and ovary-enriched genes within the _Kdm5c_-KO brain.
-Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not somatic cells.
+In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
+We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes.
+Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testis somatic cells.
 _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis.
-We demonstrated that while _Kdm5c_ mutant sex did not influence the whether sperm or egg genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
-Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as KDM5C is enriched at germline gene promoters in EpiLCs but only bound to a subset of germline-enriched DEGs.
+We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
+Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as although KDM5C is enriched at germline gene promoters in EpiLCs, it is only bound to a subset of germline-enriched DEGs.
 Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands.
@@
 By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages.
-Furthermore, these results provide molecular footholds that can be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.
+Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.
 
@@
 For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-KDM5C may therefore negatively regulate totipotency, since KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].
-However, _Kdm5c_-KO misexpression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter.
-Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into EpiLCs[@mochizukiRepressionGermlineGenes2021].
-We found _Kdm5c_-KO EpiLCs expressed _Dazl_ but did not express 2-cell specific genes, suggesting the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
+KDM5C may therefore negatively regulate totipotency, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].
+However, misexpression of _Dazl_ in _Kdm5c_-KO ESCs was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter.
+
+Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
+We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes.
 Misexpression of germline genes in _Kdm5c_-KO EpiLCs could indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
 Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Altogether, these results demonstrate KDM5C repressess germline identity during embryogenesis and _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs.
+Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
 
-Although eggs and sperm intially express very similar sets of germline genes for PGC formation and meiosis, some germline genes are sexually dimorphic.
-We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of repression is independent of a given germline gene's sex.
-However, organismal sex did greatly influence the manifestation of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males.
-X chromosome inactivation (XCI)  defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022].
+Although eggs and sperm employ the same transcriptional programs for shared functions, e.g.
+PGC formation,  meiosis, and genome defense, some germline genes are sex specific.
+We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex.
+However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males.
+X chromosome inactivation (XCI) defects could explain why _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022].
 However, female germline DEGs were not significantly biased towards the X chromosome.
 Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
+Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
 Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity.
@@
 
-It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding.
+It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007].
 In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
-While MAX and E2F6 motifs were enriched at KDM5C-bound germline promoters in EpiLCs, only about one-third of promoters contained their consesus sequence.
+While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.
 Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters.
@@
 Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
-Together, this suggests once germline genes like _Dazl_ and _Rfx2_ are activated in _Kdm5c_-KO cells, they can then turn on other ectopic germline programs downstream and loosely mimic germ cell development.
+Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.
 
@@
 Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
-Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression even more potently as their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
+Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
 Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
 In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
-Additionally, we found _Kdm5c_-KOs accumulate H3K4me2/3 around the TSS of germline genes and impaired CpGme placement at CpG islands.
-KDM5C's shifting mode of germline gene repression as cells transition from na&#239;ve to primed pluripotency indicates chromatin regulators can employ both catalytic and non-catlytic repressive mechanisms at the same loci depending upon the developmental and cellular context.
+We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at germline CpG islands.
+KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates chromatin modifiers can switch between catalytic and non-catlytic gene regulatory mechanisms at the same loci over development.
 
@@
 	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
-	- We found not all germline-enriched genes have CGIs, but many still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes
+	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.
 	- CGIs highly determine KDM5C recruitment to germline gene promoters.
-			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function likely different from its somatic CGI function). 
+			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
 	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
-		- Other studies on germline gene repressors have shown they are importnat for CGIme, unclear if they participate in non-CGI TSS CpGme
+		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
 	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
 
-The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development.
+The above work provides the mechanistic foundation for KDM5C's repression of germline-specific transcription and offers novel insight into how dysregulation of _Kdm5c_-KO tissue identity changes over the course of development.
 However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
 In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain.
-Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that can impair learning and memory when overexpressed in the mouse brain[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].
-Thus, KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.
+Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022].
+_APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008].
+KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.
 Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
-Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating they may promote genome instability and cellular de-differentiation.
-Intriguingly, select models of other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
+Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating their dysregulation may promote genome instability and cellular de-differentiation.
+Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
 Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression.

### 2024-08-21 submission/manuscript.Rmd
abstract edits CGI

@@
 It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
-Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms behind their dysregulation and their functional consequences are largely unknown.
+While, loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, the mechanisms behind their dysregulation and their functional consequences are largely unknown.
 Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
-We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+We found male _Kdm5c_ knockout (-KO) mice that recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
 Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
@@
 Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males.
-We found KDM5C represses germline-specific transcription by binding to a subset of germline gene promoters in EpiLCs to facilitate DNA methylation of CpG islands.
-However, germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding.
-This suggests germline developmental programs can progress ectopically in the background of typical _Kdm5c_-KO development, due to downstream activation by germline transcription factors.
-These results define KDM5C's role in germline gene suppression and potentially link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.
-
-- Not sure if I like the last sentence, I think we should focus on what we did do versus what we could do in the future
-	- Add something that we deepened the characterization of germline gene classes that use different regulatory mechanisms (CpG islands vs not, meiotic vs late-stage, E2F6/MAX vs no)
+We found KDM5C binds to a subset of germline gene promoters in EpiLCs that contain promoter CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation.
+However, CGI-free germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding.
+This suggests ectopic germline transcriptional programs can progress in the background of typical _Kdm5c_-KO development, due to downstream activation by key drivers of germline gene expression.
+These results define KDM5C's role in germline gene suppression and link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.
 

### 2024-08-21 submission/manuscript.Rmd
results edit 1

@@
 Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Unexpectedly, human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
-Most studies investigating this relationship have explored their regulation of brain-specific genes and chromatin marks.
+Unexpectedly, human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and most studies investigating this relationship have explored their regulation of brain-specific genes and chromatin marks.
 However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
@@
 To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-We characterized lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3) found at active gene promoters[@iwaseXLinkedMentalRetardation2007].
+We characterized lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3) that is found at active gene promoters[@iwaseXLinkedMentalRetardation2007].
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
@@
 Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017], however it is unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development testis gene dysregulation begins, and if other tissue-specific genes are also aberrantly transcribed with KDM5C loss.
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
+However, it is unknown what types of testis genes are dysregulated, at what point in _Kdm5c_-KO development testis gene misexpression begins, and if other types of tissue-specific genes are aberrantly transcribed with KDM5C loss.
 
@@
 In mammals, chromatin regulators are crucial for decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
-The precise roles of KDM5C during this process remains unclear.
-Additionally, studies on germline gene repression have primarily focused on marker genes important for germ cell development rather than germline genes as a whole, given they lacked a curated list of germline-enriched genes.
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+The contribution of KDM5C to this process remains unclear.
+It is additionally unknown if KDM5C-mediated germline gene repression is impacted by chromosomal sex, given that females have a higher KDM5C dose than males due to its partial escape from X chromosome inactivation[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+Furthermore, studies on germline gene repression have primarily focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes.
 Therefore, it is unknown if the mechanism of repression differs for certain classes of germline genes, e.g.
 meiotic genes versus spermatid differentiation genes.
-
-It is also unknown if ectopic germline gene expression is influenced by chromosomal sex, as previous studies on germline gene repression have been exclusively in males.
-Sex is particularly pertinent in the case of KDM5C, which lies on the X chromosome and partially escapes X chromosome inactivation, resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 
@@
 
-Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the male _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017].
+Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some testis genes[@scandagliaLossKdm5cCauses2017].
 It is currently unclear if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus characterized dysregulation of _Kdm5c_-KO brain tissue identity by systematically assessing the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017], using our published mRNA-seq datasets of the adult amygdala and hippocampus for male mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
+We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017], using our published mRNA-seq datasets of the male adult amygdala and hippocampus for mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 
-We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are enriched in non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
-The majority of tissue-enriched differentially expressed genes (tissue-enriched DEGs) were testis genes (Figure 1A-C).
-Even though the testis has the largest total number of biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
-One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
+We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched within non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C).
+Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
+One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
 
-Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1D).
+Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B).
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Given the _Kdm5c_-KO mice we analyzed are male, this demonstrates ectopic expression of tissue-enriched genes is independent of organismal sex.
+Given that the _Kdm5c_-KO mice we analyzed are male, this demonstrates ectopic expression of tissue-enriched genes is independent of organismal sex.
 
-Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test).
+Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
 A liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] that has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
-For all tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
+
+For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
 Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
-Together, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
+Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
 

### 2024-08-21 submission/manuscript.Rmd
results edit 2

@@
 
-To further validate if these testis DEGs are truly germline genes, we assessed their expression in somatic versus germ cells within the testis.
-We first compared their expression in wild-type testes to those without germ cells[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
+To further validate if these testis DEGs are truly germline genes, we compared their expression in somatic and germ cells within the testis.
+We first evlauted their expression in wild-type testes versus those with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
 We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
-Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell types (e.g.
+Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g.
 spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C).

### 2024-08-21 submission/manuscript.Rmd
results edit 3

@@
 A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
-This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A, top)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
+This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
 While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-Therefore, we tested if KDM5C was necessary for silencing germline genes at this developmental stage by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
+Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
 
@@
 Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C).
-Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating KDM5C loss does not impair EpiLC formation.
+Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating the misexpression of tissue-enriched genes with KDM5C loss does not impair EpiLC formation.
 
-We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline DEGs are constitutively dysregulated or change over the course of development.
+We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
 We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
@@
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs do not revert back to a 2-cell-like state (Figure 3H).
+However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
 Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.

### 2024-08-21 submission/manuscript.Rmd
results edit 4

@@
 It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
+Sex is particularly pertinent in the case of KDM5C because it lies on the X chromosome and partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+We therefore explored the impact of chromosomal sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
 [@samantaActivationXistEvolutionarily2022].
 
-Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
-While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B), such as _Tktl2_ and _Esx1_.
-We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared).
-Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:0070585 protein localization to mitochondrion).
+Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
+While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs, such as _Tktl2_ and _Esx1_ (Figure 4B).
+We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared).
+Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion).
 Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).
@@
 The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D).
-This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated suppression of germline genes.
-Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-However, both shared and female-specific germline DEGs were not biased towards the X chromosome and the majority of female DEGs instead lie on autosomes (Figure 4G).
+This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression.
+Female sensitivity could be due to improper XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
+For both males and females, the majority of germline DEGs instead lie on autosomes (Figure 4G).
 Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.

### 2024-08-21 submission/manuscript.Rmd
results edit 4

@@
 While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
-Therefore, we wondered if _Kdm5c_ mutant males express sperm genes while mutant females express egg genes.
+Therefore, we wondered if _Kdm5c_ mutant males were more likely to express sperm genes and if mutant females would instead express egg genes.
 To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes.
@@
 We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J).
-Germline genes dysregulated exclusively in_Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex.
+Germline genes dysregulated exclusively in _Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex (Figure 4I).
 This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.
-These results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.
+Altogether, these results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.
 

### 2024-08-21 submission/manuscript.Rmd
results edits 5+6, combine edits in pdf

@@
 In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%), but 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX).
-In condordance with our gene ontology results, we did not observe KDM5C accumulation at any germline gene promtoers in PNCs (Figure 5D).
-Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
+In condordance with our gene ontology results, we did not observe KDM5C accumulation at any germline gene promtoers in PNCs (Figure 5D, Supplementary figure XXX).
+Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons.
 Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
@@
 RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5I), however it is not a direct target of KDM5C (Supplementary figure XXX).
+Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5I), however it is also not a direct target of KDM5C (Supplementary figure XXX).
 Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.
 
-## KDM5C promotes _de novo_ DNA methylation at germline genes (if there are differences, say at germline gene CpG islands)
+## KDM5C promotes _de novo_ DNA methylation at germline genes harboring CpG islands
 
@@
 KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], yet KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
-Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required later in development for long-term silencing of germline genes.
+Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required in the post-implantation embryo for long-term silencing of germline genes.
 In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
@@
 
-We then evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation.
+We next evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation.
 We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation into EpiLCs (Figure 6C).
@@
 
-- Germline genes are known to accumulate CpGme at (CGIs) during the transition from naive to primed pluripotency.
+- In wild-type cells, germline genes are known to accumulate CpGme at (CGIs) during the transition from naive to primed pluripotency.
 - We first identified the number of germline genes with CGIs neighboring their promoter.
-	- Found XXX% of germline genes had CGIs, XXX% of which were direct KDM5C targets in EpiLCs (Eulerr).
+	- Found XXX% of germline genes had CGIs, XXX% of which were direct KDM5C targets in EpiLCs.
+	- Gene ontology of CGI vs non-CGI germline genes (sperm vs meiotic?)
 
-- We then curated a list of germline genes that significantly gained CpGme in wild-type exEpiLCs compared to ESCs.
-	- Example gene bedgraph
-	- Majority of CGI germline genes?
+- We then identified germline genes that significantly gained CpGme, either within the gene promoter (TSS +/- 500bp) or at the promoter CGI, in wild-type exEpiLCs compared to ESCs (bismark and methylkit, q < 0.01).
+	- Volcano plot, no-CGI vs CGI, color with KDM5C binding
+	- Majority of germline gene promoters, regardless of CGI status, gained CpGme during WT differentiation
 
+- Then compared CpGme accumulation in wild-type vs Kdm5c-KO exEpiLCs. 
 	- Out of the CGI genes, which had significantly reduced CpGme in 5CKO (heatmap of % methylation)
 	- Highlight interesting genes affected vs unaffected by KDM5C (especially if same/different from E2F6, PRC1.6, Setdb1 targets)
-	- CGIs that never gain DNAme in WT
 	- Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 6F).
+		- CGI vs Non-CGI germline genes, any significant differences?
 
-- Non-CGI germline genes, any significant changes at their promoter? (Heatmap again?)
-
-- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via DNA methylation during early embryogenesis.
+- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via DNA methylation at CGIs during early embryogenesis.
 

### 2024-08-21 submission/manuscript.Rmd
update methods

@@
 Genomic DNA (gDNA) from na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells.
-gDNA was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
+gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
+- bismark
+- Methylkit
 

### 2024-08-21 code/ChIPseq_KDM5C_GO_Compare_BrainEpiLC.R
make GO longer

@@ -49,7 +49,7 @@ ck2 <- compareCluster(geneCluster = promo_compare, fun = enrichGO,  OrgDb = "org
 #ck <- setReadable(ck, OrgDb = "org.Mm.eg.db", keyType="ENSEMBL")
 head(ck2) 
 write.table(ck2, snakemake@output[[3]], row.names = FALSE, sep = ",")
-ggsave(snakemake@output[[4]], plot = dotplot(ck2, size = "Count"), width = 5.5, height = 5)
+ggsave(snakemake@output[[4]], plot = dotplot(ck2, size = "Count"), width = 5.5, height = 7.5)
 
 
 #get the overlap between promoters (euler plot)

### 2024-08-21 submission/manuscript.Rmd
new chipseq figure moving DEGs to supplement

@@
 Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).
-One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs (Figure 5K).
+One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs (Figure 5F).
 In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%), but 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX).
@@
 To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1).
-MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5F).
-One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5G).
-KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5H).
+MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G).
+One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H).
+KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I).
 RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5I), however it is also not a direct target of KDM5C (Supplementary figure XXX).
+Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure XXX).
 Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.

### 2024-08-21 code/WGBS_ESCvsEpiLC.R
annotate the loci for wgbs

@@ -5,6 +5,11 @@ KDM5C_binding <- read.csv(snakemake@input[[1]], sep = ",")
 
 print(head(KDM5C_binding))
 
+#get the gene coordinate information
+source("code/utilities/GeneTSSandTES.R")
+germTSS <- geneTSS_df(KDM5C_binding$ENSEMBL, 500)
+print(head(germTSS))
+
 
 samples <- c("all germ promoters", "germ CGI promoters")
 
@@ -14,43 +19,65 @@ txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
 library(ggplot2)
 library(dplyr)
 
-volcano_WGBS <- function(n){
-	df <- read.csv(snakemake@input[[n+1]], sep = ",")
+##### #make the dataframes for the volcano plot
+## TSS
+TSSdf <- read.csv(snakemake@input[[2]], sep = ",")
 
-	#rename chr to seqnames to match granges for merging later
-	names(df)[names(df) == "chr"]<- "seqnames"
-	print(paste("rows in", samples[n], "-", nrow(df)))
-	print(head(df))
+#rename chr to seqnames to match granges for merging later
+names(TSSdf)[names(TSSdf) == "chr"]<- "seqnames"
+print(paste("rows in all germ -", nrow(TSSdf)))
+print(head(TSSdf))
 	
+#get the germline genes for each coordinate
+plotTSS <- merge(germTSS, TSSdf)
+plotTSS <- merge(KDM5C_binding, plotTSS, by = "ENSEMBL")
+#remove any duplicated loci
+plotTSS <- plotTSS[!duplicated(plotTSS[6:7]),]
+print(paste("rows in merged all germ -", nrow(plotTSS)))
+print(head(plotTSS))
 
-	#get the germline genes for each coordinate
-		#make granges object with bed coordinates
-	peak <- makeGRangesFromDataFrame(df)
-	print(head(peak))
+names(plotTSS)[names(plotTSS) == "gene_name"] <- "SYMBOL"
 
-		#annotate locations
-	peakAnno <- annotatePeak(peak, tssRegion=c(-500, 500),
-                         TxDb=txdb, annoDb="org.Mm.eg.db")
+
+
+write.table(plotTSS, snakemake@output[[1]], sep = ",", row.names = FALSE)
+
+
+
+
+
+
+#####
+## CGIs
+CGIdf <- read.csv(snakemake@input[[3]], sep = ",")
+
+#rename chr to seqnames to match granges for merging later
[…]

### 2024-08-21 code/utilities/GeneTSSandTES.R

@@ -53,16 +53,19 @@ geneTSSwindow <- function(goi, n, window){
 
 }
 
-geneTSSdown <- function(goi, n, window){
+
+
+#return results for ROI as dataframe
+geneTSS_df <- function(goi, window){
 	#regions of interest
-	roi <- subset(gtf_df, ENSEMBL %in% goi, select = c(seqnames, TSS, TES))
+	roi <- subset(gtf_df, ENSEMBL %in% goi, select = c(gene_name, ENSEMBL, seqnames, TSS))
 	
 	#Add and subtract the window from the TSS
-	roi$TSS_down <- ifelse(roi$TSS > roi$TES, roi$TSS - window, ifelse(roi$TSS < roi$TES, roi$TSS + window, "oops"))
+	roi$start <- roi$TSS - window + 1
+	roi$end <- roi$TSS + window
 
-	roi$start <- ifelse(roi$TSS < roi$TSS_down, roi$TSS, ifelse(roi$TSS > roi$TSS_down, roi$TSS_down, "oops"))
-	roi$end <- ifelse(roi$TSS < roi$TSS_down, roi$TSS_down, ifelse(roi$TSS > roi$TSS_down, roi$TSS, "oops"))
-
-	write.table(roi[,c("seqnames","start", "end")], snakemake@output[[n]], sep = "\t", row.names = FALSE, col.names=FALSE, quote = FALSE)
+	return(roi)
 
 }
+
+

### 2024-08-25 code/WGBS_volcano.R
fixed writing output of WTvsKO

@@ -46,7 +46,7 @@ write.table(ESCvsEpiLC_TSS, snakemake@output[[1]], sep = ",", row.names = FALSE)
 
 ## TSS WT vs KO
 WTvsKO_TSS <- anno_TSS(read.csv(snakemake@input[[4]], sep = ","))
-write.table(ESCvsEpiLC_TSS, snakemake@output[[5]], sep = ",", row.names = FALSE)
+write.table(WTvsKO_TSS, snakemake@output[[5]], sep = ",", row.names = FALSE)
 
 
 ### TSS volcano plot
@@ -171,7 +171,7 @@ write.table(ESCvsEpiLC_CGI, snakemake@output[[2]], sep = ",",  row.names = FALSE
 
 
 WTvsKO_CGI <- anno_CGI(read.csv(snakemake@input[[5]], sep = ","))
-write.table(ESCvsEpiLC_CGI, snakemake@output[[6]], sep = ",",  row.names = FALSE)
+write.table(WTvsKO_CGI, snakemake@output[[6]], sep = ",",  row.names = FALSE)
 
 
 

### 2024-08-25 submission/manuscript.Rmd
drafting WGBS results

@@
 To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` +/- `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
+EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` &#177; `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
 
@@
 Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
-To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1).
+To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS &#177; 500 bp, q-value < 0.1).
 MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G).
@@
 We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
-
 We next evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation.
@@
 
+```{r, CGIs, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
+germgenes_CGI <- read.csv("../results/KDM5C_binding_allgerm_CGI.csv", header = TRUE)
+
+CGI <- germgenes_CGI[germgenes_CGI$Promo_CGI == "CGI",]
+
+ESCvsEpiLC <- read.csv("../results/WGBS/WGBS_restab_annotated_allgermTSS_500bp_WT_ESCvsEpiLC.csv")
+
+ESCvsEpiLC_25 <- ESCvsEpiLC[ESCvsEpiLC$meth.diff >= 25, ]
+
+WTvsKO <- read.csv("../results/WGBS/WGBS_restab_annotated_allgermTSS_500bp_WTvsKO.csv")
+
+WTvsKO_hypo <- WTvsKO[WTvsKO$meth.diff <= -25, ]
+
+#get the infomration for wgbs
+getWGBSinfo <- function(df, gene, column){
+	df2 <- df
+	info <- df2[df2$SYMBOL == gene,][[column]]
+	return(info)
+}
+
+```
+
+In wild-type cells, germline genes are known to accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012].
+We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) contained CGIs within their promoters (TSS &#177; 500 bp) (Figure 6F).
+CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) (Figure 6G).
+
 To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs).
+We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during ESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
+The majority of germline genes gained substantial CpGme at their promoter, regardless if their promoter contained a CGI (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%) (Figure 6H).
 
-- In wild-type cells, germline genes are known to accumulate CpGme at (CGIs) during the transition from naive to primed pluripotency.
-- We first identified the number of germline genes with CGIs neighboring their promoter.
-	- Found XXX% of germline genes had CGIs, XXX% of which were direct KDM5C targets in EpiLCs.
-	- Gene ontology of CGI vs non-CGI germline genes (sperm vs meiotic?)
-
-- We then identified germline genes that significantly gained CpGme, either within the gene promoter (TSS +/- 500bp) or at the promoter CGI, in wild-type exEpiLCs compared to ESCs (bismark and methylkit, q < 0.01).
-	- Volcano plot, no-CGI vs CGI, color with KDM5C binding
-	- Majority of germline gene promoters, regardless of CGI status, gained CpGme during WT differentiation
+We then assessed differentially methylated promoters in wild-type versus _Kdm5c_-KO exEpiLCs and found `r nrow(WTvsKO_hypo)` germline gene promtoers were significantly hypomethylated with _Kdm5c_ loss (Figure 6I).
+Hypomethylated promoters included genes consistently dysregulated in _Kdm5c_-KO cells, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6J).
 

### 2024-08-26 code/WGBS_volcano.R
kdm5c binding at WT vs KO DEGs

@@ -126,8 +126,81 @@ TSS_volcano <- function(n, TITLE, restab, maxy){
 
 }
 
+###for kdm5c binding
+TSS_volcano_5C <- function(n, TITLE, restab, maxy){
+
+	#restab <- methyl
+	#if the gene is tissue specific, the color key is the color for the tissue. Otherwise the gene will be gray
+	restab$keyvals <- ifelse(restab$qvalue > 0.01 & restab$KDM5C_binding == "Bound", 'black',
+						ifelse(restab$KDM5C_binding == "Bound" & restab$meth.diff < l2fcco & restab$meth.diff > -l2fcco, 'black',
+							ifelse(restab$qvalue <= 0.01 & restab$KDM5C_binding == "Bound", 'blue',
+								ifelse(restab$meth.diff >= l2fcco & restab$qvalue <= 0.01 & restab$KDM5C_binding == "Unbound", 'palevioletred1',
+                        			ifelse(restab$meth.diff <= -l2fcco & restab$qvalue <= 0.01 & restab$KDM5C_binding == "Unbound", 'lightseagreen',
+										"gray38")))))
+
+	#print(head(restab))
+
+	restab$names <- ifelse(restab$keyvals == 'blue', 'KDM5C-bound', 
+                         ifelse(restab$keyvals == 'gray38', 'n.s. KDM5C-unbound',
+						 	ifelse(restab$keyvals == 'black', 'n.s. KDM5C-bound',
+								ifelse(restab$keyvals == 'lightseagreen', 'KDM5C-unbound',
+									ifelse(restab$keyvals == 'palevioletred1', 'KDM5C-unbound',
+                            			"uh oh")))))
+
+
+
+	#for the figure legend
+	keyvals <- restab$keyvals
+	names(keyvals) <- restab$names 
+
+	#Make the axis range narrower so that there is space to see the data
+	neglog10 <- function(x){
+	  a = log10(x)
+	  return(-a)
+	}
+
+
+	
+	keyvals.shape <- ifelse(neglog10(restab$qvalue) > maxy, 9, 16)
+	#if the log2fc or the padj is greater than the maximum axis cut off, set it to the cut off and make the point an outlier shape
+	restab$qvalue[restab$qvalue< 10^(-maxy)] <- 10^(-maxy)
+
+	keyvals.shape[is.na(keyvals.shape)] <- 16
+	names(keyvals.shape)[keyvals.shape == 9] <- 'Outlier'
[…]

### 2024-08-26 submission/manuscript.Rmd
kdm5c binding at WT vs KO DEGs
draft wgbs figure

@@
 
-## KDM5C promotes _de novo_ DNA methylation at germline genes harboring CpG islands
+## KDM5C is recruited to CpG islands and facilitates _de novo_ DNA methylation at germline gene promoters
 
@@
 WTvsKO_hypo <- WTvsKO[WTvsKO$meth.diff <= -25, ]
+WTvsKO_hypo_5C <- WTvsKO_hypo[WTvsKO_hypo$KDM5C_binding == "Bound", ]
 
@@
 
-We then assessed differentially methylated promoters in wild-type versus _Kdm5c_-KO exEpiLCs and found `r nrow(WTvsKO_hypo)` germline gene promtoers were significantly hypomethylated with _Kdm5c_ loss (Figure 6I).
+We then assessed differentially methylated regions (DMRs) at germline gene promoters in wild-type versus _Kdm5c_-KO exEpiLCs.
+We found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6I).
+About half of promoter DMRs hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs).
 Hypomethylated promoters included genes consistently dysregulated in _Kdm5c_-KO cells, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6J).

### 2024-08-26 code/WGBS_germ_heatmap.R

@@ -128,3 +128,36 @@ p <- Heatmap(plotmatrix, column_title = "KDM5C-bound germline genes",show_row_na
 draw(p)
 	
 dev.off()
+
+
+#all germline DEGs
+EpiLC_DEGs <- read.csv(snakemake@input[[4]], sep = ",")[,"SYMBOL"]
+amy_DEGs <- read.csv(snakemake@input[[5]], sep = ",")[,"SYMBOL"]
+hip_DEGs <- read.csv(snakemake@input[[6]], sep = ",")[,"SYMBOL"]
+
+DEGs <- unique(c(hip_DEGs, EpiLC_DEGs, amy_DEGs))
+
+pdf(file = snakemake@output[[5]],   # The directory you want to save the file in
+    width = 5, # The width of the plot in inches
+    height = 10) # The height of the plot in inches
+
+plotmatrix <- avg_hclust_matrix[row.names(avg_hclust_matrix) %in% DEGs,]
+col_fun = colorRamp2(range(plotmatrix), hcl_palette = "Reds", reverse = TRUE)
+p <- Heatmap(plotmatrix, column_title = "EpiLC germline DEGs",show_row_names = TRUE, split = 2, column_order = c("avg_WT_esc","avg_KO_esc","avg_WT_96EpiLC", "avg_KO_96EpiLC"), heatmap_legend_param = list(title = "% CpGme"), col = col_fun)
+draw(p)
+	
+dev.off()
+
+
+
+pdf(file = snakemake@output[[6]],   # The directory you want to save the file in
+    width = 5, # The width of the plot in inches
+    height = 10) # The height of the plot in inches
+
+plotmatrix <- avg_hclust_matrix[row.names(avg_hclust_matrix) %in% DEGs,]
+plotmatrix <- plotmatrix[,c("avg_WT_96EpiLC", "avg_KO_96EpiLC")]
+col_fun = colorRamp2(range(plotmatrix), hcl_palette = "Reds", reverse = TRUE)
+p <- Heatmap(plotmatrix, column_title = "EpiLC germline DEGs",show_row_names = TRUE, split = 2, column_order = c("avg_WT_96EpiLC", "avg_KO_96EpiLC"), heatmap_legend_param = list(title = "% CpGme"), col = col_fun)
+draw(p)
+	
+dev.off()

### 2024-08-26 submission/manuscript.Rmd
formatting edits WGBS figure
results WGBS edits

--- 2024-08-25 submission/manuscript.Rmd
@@
 
-## KDM5C promotes _de novo_ DNA methylation at germline genes harboring CpG islands
+## KDM5C is recruited to germline gene promoters harboring CpG islands to facilitate _de novo_ DNA methylation
 
@@
 Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
-To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3) in our previously published ChIP-seq datasets of the wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
+To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3), at germline gene promoters in our previously published ChIP-seq datasets in male wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
 In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
 We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
-We next evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation.
-We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation into EpiLCs (Figure 6C).
+To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation (Figure 6C).
 While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6E).
+Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me at germline gene promoters.
 
@@
 
-In wild-type cells, germline genes are known to accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012].
+In wild-type cells, germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
 We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) contained CGIs within their promoters (TSS &#177; 500 bp) (Figure 6F).
-CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) (Figure 6G).
+CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) when compared to CGI-free genes (Figure 6G).
+Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline gene promoters (FISHER'S XXXX), with 79% of KDM5C-bound germline genes containing CGIs (Figure 6G).
 
-To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs).
+We then perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to assess how KDM5C loss impacted initial CpGme placement at germline gene promoters (Figure 6H).
 We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during ESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
-The majority of germline genes gained substantial CpGme at their promoter, regardless if their promoter contained a CGI (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%) (Figure 6H).
+In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6I).
 
-We then assessed differentially methylated promoters in wild-type versus _Kdm5c_-KO exEpiLCs and found `r nrow(WTvsKO_hypo)` germline gene promtoers were significantly hypomethylated with _Kdm5c_ loss (Figure 6I).
-Hypomethylated promoters included genes consistently dysregulated in _Kdm5c_-KO cells, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6J).
-
-- Then compared CpGme accumulation in wild-type vs Kdm5c-KO exEpiLCs. 
-	- Out of the CGI genes, which had significantly reduced CpGme in 5CKO (heatmap of % methylation)
-	- Highlight interesting genes affected vs unaffected by KDM5C (especially if same/different from E2F6, PRC1.6, Setdb1 targets)
-	- Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 6F).
-		- CGI vs Non-CGI germline genes, any significant differences?
-
-- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via DNA methylation at CGIs during early embryogenesis.
+We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J).
+Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs).
+We then evaluated promoter CpGme at germline genes ectopically transcibed in either _Kdm5c_-KO EpiLCs or within the brain and found promoter CpGme was substantially reduced in about half of germline DEGs (Figure 6K).
+Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
+Suprisingly, we only observed a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
+Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at select germline gene promoters during early embryogenesis, however other germline gene silencing mechansism can sufficiently compensate for KDM5C's loss at select germline gene promoters.
 

### 2024-08-27 submission/manuscript.Rmd
begin CGI discussion edits

@@
 In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
-We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at germline CpG islands.
-KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates chromatin modifiers can switch between catalytic and non-catlytic gene regulatory mechanisms at the same loci over development.
+We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters.
+KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.
 

### 2024-08-27 submission/manuscript.Rmd
abstract edits shorten

@@
 
-It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
-While, loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, the mechanisms behind their dysregulation and their functional consequences are largely unknown.
-Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
-We found male _Kdm5c_ knockout (-KO) mice that recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
+While, loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, the mechanisms and consequences are largely unknown.
+Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
 Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
-We then curated a list of germline-enriched genes for unbiased characterization of KDM5C in germline gene repression.
 Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
 Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males.
-We found KDM5C binds to a subset of germline gene promoters in EpiLCs that contain promoter CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation.
-However, CGI-free germline genes, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells independent of direct KDM5C binding.
-This suggests ectopic germline transcriptional programs can progress in the background of typical _Kdm5c_-KO development, due to downstream activation by key drivers of germline gene expression.
-These results define KDM5C's role in germline gene suppression and link impaired soma-vs-germline demarcation to a chromatin-based neurodevelopmental disorder.
+Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
+However, CGI-free germline gene promoters lacking KDM5C, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
+Thus, ectopic germline transcriptional programs can progress in the background of typical _Kdm5c_-KO development, due to downstream activation by key drivers of germline gene expression.
 

### 2024-08-27 submission/manuscript.Rmd
shigeki edits intro and res 1

@@
 
-A single genome holds the instructions to generate the myriad of cell types found within the adult organism.
-This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001].
-Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Unexpectedly, human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and most studies investigating this relationship have explored their regulation of brain-specific genes and chromatin marks.
-However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
-Very few studies have investigated this severe crisis in cellular identity in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if these ectopic genes contribute to neurodevelopmental impairments.
+A single genome holds the instructions to generate the myriad of cell types found an organism.
+This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
+Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance in regulating the expression of brain-specific genes.
+Loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.
 
-To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-We characterized lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3) that is found at active gene promoters[@iwaseXLinkedMentalRetardation2007].
+To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
+Here we focus on lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
@@
 Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017].
-However, it is unknown what types of testis genes are dysregulated, at what point in _Kdm5c_-KO development testis gene misexpression begins, and if other types of tissue-specific genes are aberrantly transcribed with KDM5C loss.
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
+However, it is unknown whether other types of testis genes are dysregulated, at what point in development germline gene misexpression begins, and if other tissue-specific genes are aberrantly transcribed with KDM5C loss.
 
@@
 The contribution of KDM5C to this process remains unclear.
-It is additionally unknown if KDM5C-mediated germline gene repression is impacted by chromosomal sex, given that females have a higher KDM5C dose than males due to its partial escape from X chromosome inactivation[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+It is additionally unknown if KDM5C-mediated germline gene repression is impacted by chromosomal sex, as females have a higher KDM5C dose than males due to its escape from X chromosome inactivation[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 Furthermore, studies on germline gene repression have primarily focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes.
@@
 
-To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We curated list of germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms.
+To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
 Based on the data presented below, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
@@
 
-Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some testis genes[@scandagliaLossKdm5cCauses2017].
+Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017].
 It is currently unclear if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017], using our published mRNA-seq datasets of the male adult amygdala and hippocampus for mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
+We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes specific to or enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017].
+We analyzed our published mRNA-seq datasets of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice[@vallianatosMutuallySuppressiveRoles2020].
 
@@
 
-We found a large proportion of genes that are significantly upregulated within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched within non-brain tissues in wild-type mice (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+We found a large proportion significantly upregulated genes within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
 For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C).
@@
 Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B).
-Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation and early zygote development[@chengMammalianOocytesStore2022] (Figure 1D).
-Given that the _Kdm5c_-KO mice we analyzed are male, this demonstrates ectopic expression of tissue-enriched genes is independent of organismal sex.
+Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D).
+Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of tissue-enriched genes is independent of organismal sex.
 
-Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
-A liver-biased DEG dysregulated in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] that has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
+Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
+Among liver-biased DEG derepressed in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] that has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
 
@@
 
-The testis contains both germ cells (meiotic cells, e.g.
-spermatogonia) and somatic cells (non-meiotic, e.g.
-Leydig cells) that support hormone production and germline functions.
-To determine if _Kdm5c_-KO testis-enriched DEGs are somatic or germline genes, we first evaluated their known functions through gene onotology analysis.
+In addition to germ cells, the testis also contains both somatic cells, such as Leydig cells that support hormone production and germline functions.
+To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated their known functions through gene onotology analysis.
 We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
-To further validate if these testis DEGs are truly germline genes, we compared their expression in somatic and germ cells within the testis.
-We first evlauted their expression in wild-type testes versus those with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
+To validate the cell-type origin of testicular DEGs, we then compared their expression in somatic and germ cells within the testis.
+We first evlauted their expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
@@
 
-We then wanted to characterize germline gene misexpression with _Kdm5c_ loss more deeply, but lacked a comprehensive list of mouse germline genes.
+As of yet, studies on germline gene silencing mechanisms have focused on a handful of key germline genes.
+The lack of a germline gene list prohibits a comprehensive, genome-wide analysis of their de-repression and silencing mechanisms.
 We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally assess germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
 

### 2024-08-28 submission/manuscript.Rmd
shigeki edits results

@@
 
-As of yet, studies on germline gene silencing mechanisms have focused on a handful of key germline genes.
-The lack of a germline gene list prohibits a comprehensive, genome-wide analysis of their de-repression and silencing mechanisms.
+As of yet, research on germline silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive mouse germline gene list.
 We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
@@
 
-We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and then assessed tissue-enriched gene expression.
+_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3A) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3B).
+We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
-Markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_, were unaltered in _Kdm5c_-KO EpiLCs (Figure 3C).
-Furthermore, _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3D), indicating the misexpression of tissue-enriched genes with KDM5C loss does not impair EpiLC formation.
 
-We next compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
+Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` were misexpressed in male _Kdm5c_-KO EpiLCs.
+We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
 We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
@@
 Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H).
-These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+These genes are typically expressed when PGCs commit to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see misexpression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
-Altogether, these results demonstrate that while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis.
+However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
+Altogether, these results demonstrate _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes.
 
@@
 While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs, such as _Tktl2_ and _Esx1_ (Figure 4B).
-We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared).
+We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
 Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion).
@@
 This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression.
-Female sensitivity could be due to improper XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+Female sensitivity could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
 However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
-For both males and females, the majority of germline DEGs instead lie on autosomes (Figure 4G).
-Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.
+The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G).
+Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of XCI defects.
 
@@
 This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.
-Altogether, these results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.
-
-- **note: the edited last sentence ended with "independent of germ cell sex", but I think if it's read out of context that sounds like we tested KDM5C in the four core genotypes.
-But I'm not sure if "germline gene sex" is confusing**
+Altogether, these results demonstrate that both male and female _Kdm5c_ mutant cells express both sperm and egg-biased genes.
 
@@
 KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017].
-However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters.
+However, the lack of a comprehensive list of germline-enriched genes prohibited comprehensive characterization of KDM5C binding at germline gene promoters.
 Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
@@
 Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
-Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
-There were no ontologies significantly enriched for genes only bound by KDM5C in PNCs.
+Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
+There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs.
 We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-In EpiLCs, we observed modest KDM5C signal at about half of all germline genes (Figure 5D).
+KDM5C binding around the TSS of many germline genes is evident in EpiLCs, but absent in PNCs (Figure 5D).
 Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).
-One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs (Figure 5F).
-In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%), but 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX).
-In condordance with our gene ontology results, we did not observe KDM5C accumulation at any germline gene promtoers in PNCs (Figure 5D, Supplementary figure XXX).
+
+ In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%).
+However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX).
+Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
 Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons.
 Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
+
+ One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs ().
 

### 2024-08-28 submission/manuscript.Rmd
shigeki discussion edits

@@
 
-Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` were misexpressed in male _Kdm5c_-KO EpiLCs.
+Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.
 We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
@@
 Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).
-
 In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%).
@@
 Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
-
- One notable gene that lacked KDM5C binding was _Stra8_, even though its mRNA is expressed in _Kdm5c_-KO EpiLCs ().
 
@@
 We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes.
-Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testis somatic cells.
+Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testicular somatic cells.
 _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis.
 We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
-Germline-enriched mRNAs can become aberrantly expressed in _Kdm5c_-KO cells indirect of KDM5C, as although KDM5C is enriched at germline gene promoters in EpiLCs, it is only bound to a subset of germline-enriched DEGs.
+Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, as KDM5C is only bound to a subset of germline-enriched DEGs.
 Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands.
@@
 For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-KDM5C may therefore negatively regulate totipotency, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023].
-However, misexpression of _Dazl_ in _Kdm5c_-KO ESCs was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023] and, unlike _Dazl_, KDM5C does not bind to the _Dux_ promoter.
-
-Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
+Based on the de-repression of _Dazl_ and _Zscan4c_ in _Kdm5c_-KO ESCs, KDM5C was thought to promote the 2-cell-to-ESC transition[[@guptaGenomewideScreenReveals2023].
+Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
 We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes.
-Misexpression of germline genes in _Kdm5c_-KO EpiLCs could indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
+Misexpression of many germline genes in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
 Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.
+Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflect KDM5C's primary role in germline gene repression.
 
@@
 However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males.
-X chromosome inactivation (XCI) defects could explain why _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004] and loss of _Kdm5c_ impairs XCI[@samantaActivationXistEvolutionarily2022].
-However, female germline DEGs were not significantly biased towards the X chromosome.
-Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022].
+Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
-Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity.
-Future studies are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.
+Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.
+
+Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
+KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+
+- H3K4me3 and CpGme typically do not colocalize
+	- https://pubmed.ncbi.nlm.nih.gov/17334365/
+	- https://www.nature.com/articles/s41594-017-0013-5
+	- https://pubmed.ncbi.nlm.nih.gov/23664763/
+	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2718496/
+	- https://pubmed.ncbi.nlm.nih.gov/19581485/
+
+- CpGme was imparied at some germline genes, including ones expressed throughout life
+-  Some regions, such as Dazl, were able to compensate for KDM5C loss, even though it gained H3K4me2 in EpiLCs. 
+	- Multiple repressive mechanisms
+We found _Kdm5c_-KO EpiLCs and the amygdala accumulate H3K4me2/3 around the TSS of germline genes, potential indicating KDM5C's cataltyic activity is necessary at later points in development.
+
+KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as
+
+as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters.
+Our study therefore indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.
+
+- DNAme and CpG islands
+	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
+	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
+	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
+	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.
+
+- Combine germline gene list, CGI, RFX information together. 
+	- Benefit of the list is finding CGI difference and kdm5c indirect mechanisms of de-repression
+    - List of genes enables us to propose this model
+    - CGI positive meiosis/germline formation regulators turned on → these turn on CGI-negative late stage regulators → These promote downstream dysregulation long term
+
+- CGIs highly determine KDM5C recruitment to germline gene promoters.
+			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
+	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
+		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
+	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
 
@@
 
-Unlike the previously characterized germline gene suppressors that place repressive histone modifications[@karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @auclairEHMT2DirectsDNA2016; @endohPCGF6PRC1SuppressesPremature2017; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021], KDM5C removes the active histone mark histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007].
-Yet, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
-Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that can influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
-Although dispensible in ESCs, KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
-In support of this, CpGme is significantly eroded at at least two germline promoters in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
-We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters.
-KDM5C's shifting role in germline gene repression during the transition from na&#239;ve to primed pluripotency indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.
-
-- DNAme and CpG islands
-	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
-	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
-	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
-	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.
-	- CGIs highly determine KDM5C recruitment to germline gene promoters.
-			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
-	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
-		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
-	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
-
-The above work provides the mechanistic foundation for KDM5C's repression of germline-specific transcription and offers novel insight into how dysregulation of _Kdm5c_-KO tissue identity changes over the course of development.
+The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes.
 However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
@@
 Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
-Ectopic germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and can drive brain turmor formation in _Drosophila_[@janicEctopicExpressionGermline2010], indicating their dysregulation may promote genome instability and cellular de-differentiation.
+Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012], indicating their dysregulation may promote genome instability and cellular de-differentiation.
 Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
 Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression.
-Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation However, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
+Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary.
+Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
 

### 2024-08-29 submission/manuscript.Rmd
discussion CGI drafts

@@
 
+- Our results indicate KDM5C's removal of H3K4me2/3 may be necessary at a subset of CGI-containing germline genes, given they fail to accumulate CpGme during EpiLC differentiation.
+These CpGme-sensitive germline genes, such as _D1Pas1_ and _Naa11_, are also de-repressed in the mature _Kdm5c_-KO brain and consistently dysregulated in multiple _Kdm5c_-KO datasets[@scandagliaLossKdm5cCauses2017].
+
 - H3K4me3 and CpGme typically do not colocalize
@@
 
-- CpGme was imparied at some germline genes, including ones expressed throughout life
--  Some regions, such as Dazl, were able to compensate for KDM5C loss, even though it gained H3K4me2 in EpiLCs. 
-	- Multiple repressive mechanisms
-We found _Kdm5c_-KO EpiLCs and the amygdala accumulate H3K4me2/3 around the TSS of germline genes, potential indicating KDM5C's cataltyic activity is necessary at later points in development.
+- Contrastingly, many germline genes had only a modest reduction in CpGme, including _Dazl_ even though its TSS accumulated substantial H3K4me2. 
+- Multiple regulators repress dazl, dazl is not expressed in the mature _Kdm5c_-KO brain
+	E2F6/MGA/MAX
+		- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
+		-  Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
 
-KDM5C's catalytic activity could be necessary to silence germline genes in EpiLCs, as
+- Altogether, our results indicates a given chromatin modifier's catalytic and non-catlytic regulation of gene expression can change depending upon the locus and developmental stage.
 
-as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
-We found _Kdm5c_-KO cells accumulate H3K4me2/3 around the TSS of germline genes and have impaired CpGme placement at CGIs for select germline gene promoters.
-Our study therefore indicates the requirement of catalytic and non-catlytic gene regulatory mechanisms can change depending upon the locus and developmental stage.
+Although the classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter.
+We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS.
+
+- CGIs highly determined KDM5C recruitment to germline gene promoters.
+
+- KDM5C loss also impacted CpGme at some KDM5C-unbound, CGI-free germline genes. Unclear what the repressive mechanism is for these genes.
+
+In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
+
+- Germline gene suppression has focused on the accumulation of CpGme at CGIs
 
 - DNAme and CpG islands
-	- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
-	- Germline genes are methylated at promoter CGIs, which are typically not methylated for other types of genes
-	- Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
-	- We found only 30% of germline-enriched genes have CGIs, but the majority of CGI-free germline genes still gain CpGme around the TSS. Unclear what the repressive mechanism is for these genes.
 
@@
 			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
-	- KDM5C loss impacts germline gene CGI methylation, but not really the non-CGI TSS methylation
 		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme

### 2024-08-29 submission/manuscript.Rmd
abstract edits

@@
 Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
-While, loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, the mechanisms and consequences are largely unknown.
-Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
+Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
 We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
 Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
-Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males.
+Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
+Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
 Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
-However, CGI-free germline gene promoters lacking KDM5C, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Thus, ectopic germline transcriptional programs can progress in the background of typical _Kdm5c_-KO development, due to downstream activation by key drivers of germline gene expression.
+However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
+Thus, distinct suppressive mechanisms are recruited to different germline gene classes and germline transcriptional programs can mirror germ cell differentiation in the background of typical somatic tissue development.
 

### 2024-08-30 submission/manuscript.Rmd
res 3 and 4 edits

@@
 However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Thus, distinct suppressive mechanisms are recruited to different germline gene classes and germline transcriptional programs can mirror germ cell differentiation in the background of typical somatic tissue development.
+Thus, distinct suppressive mechanisms are recruited to different germline gene classes and ectopic germline transcriptional programs can mirror germ cell development within somatic tissues.
 
@@
 
-A single genome holds the instructions to generate the myriad of cell types found an organism.
+A single genome holds the instructions to generate the myriad of cell types found within an organism.
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance in regulating the expression of brain-specific genes.
-Loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs.
+Loss of some chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
 However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.
@@
 To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-Here we focus on lysine demethylase 5C (KDM5C), also known as SMCX or JARID1C, which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
+Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
 Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
@@
 Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
-However, it is unknown whether other types of testis genes are dysregulated, at what point in development germline gene misexpression begins, and if other tissue-specific genes are aberrantly transcribed with KDM5C loss.
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
+However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.
 
 Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012].
-In mammals, chromatin regulators are crucial for decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency.
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
 The contribution of KDM5C to this process remains unclear.
-It is additionally unknown if KDM5C-mediated germline gene repression is impacted by chromosomal sex, as females have a higher KDM5C dose than males due to its escape from X chromosome inactivation[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Furthermore, studies on germline gene repression have primarily focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes.
-Therefore, it is unknown if the mechanism of repression differs for certain classes of germline genes, e.g.
+Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes.
+Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g.
 meiotic genes versus spermatid differentiation genes.
@@
 We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
-Based on the data presented below, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
 
@@
 Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017].
-It is currently unclear if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes specific to or enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017].
-We analyzed our published mRNA-seq datasets of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice[@vallianatosMutuallySuppressiveRoles2020].
+It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
+We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017].
+We analyzed tissue-specific differentially expressed genes (DEGs) in our published mRNA-seq datasets[@vallianatosMutuallySuppressiveRoles2020] of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1).
 
@@
 
-We found a large proportion significantly upregulated genes within the _Kdm5c_-KO brain (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) are typically enriched in non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
-For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C).
+We found a large proportion of significantly upregulated genes within the _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+For both the amygdala and hippocampus, the majority of tissue-enriched (DEGs) were testis genes (Figure 1A-C).
 Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
@@
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D).
-Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of tissue-enriched genes is independent of organismal sex.
+Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of gonad-enriched genes is independent of organismal sex.
 
 Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
-Among liver-biased DEG derepressed in both the hippocampus and amygdala is _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] that has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
+_Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
 
 For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
-Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
 Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
@@
 
-In addition to germ cells, the testis also contains both somatic cells, such as Leydig cells that support hormone production and germline functions.
-To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated their known functions through gene onotology analysis.
+_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017], however the testis also contains somatic cells that support hormone production and germline functions.
+To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology.
 We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
 
-To validate the cell-type origin of testicular DEGs, we then compared their expression in somatic and germ cells within the testis.
-We first evlauted their expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
+We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
@@
 
-As of yet, research on germline silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive mouse germline gene list.
-We therefore generated a list of germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
+As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive gene list.
+We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
@@
 
+getresval <- function(df, col, symbol){
+	val <- subset(df, SYMBOL == symbol)
+	val <- val[,col]
+	return(val)
+}
+
 ```
 
-Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
-A subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
-This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
-While some germline-enriched genes are also expressed in na&#239;ve embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
+Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], when they are silenced in epiblast stem cells that will  differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
+This developmental time point can be modeled _in vitro_ through differentiation of na&#239;ve embryonic stem cells (nESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
+While some germline-enriched genes are also expressed in nESCs and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
 Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
 
-_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3A) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3B).
+_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
 We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
-Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B).
+Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D).
+Using our list of mouse germline-enriched genes assembled above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.
 
-Using the list of mouse germline-enriched genes curated above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.
 We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
-We found the majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
+The majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
 EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
@@
 
-Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3H).
-These genes are typically expressed when PGCs commit to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H).
+These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
 However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
-Altogether, these results demonstrate _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the _Kdm5c_-KO brain primarily expresses spermiogenesis genes.
+Altogether, _Kdm5c_-KO EpiLCs expressing key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes indicate germline gene misexpression parallels germline development during proper _Kdm5c_-KO differentiation.
 
@@
 It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-Sex is particularly pertinent in the case of KDM5C because it lies on the X chromosome and partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-We therefore explored the impact of chromosomal sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
+Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
 [@samantaActivationXistEvolutionarily2022].
@@
 Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
-While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs, such as _Tktl2_ and _Esx1_ (Figure 4B).
-We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
+While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were male-specific and female-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
+We then compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
 Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion).
@@
 
-The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D).
+The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D-F).
 This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression.
@@
 
-While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm.
-Therefore, we wondered if _Kdm5c_ mutant males were more likely to express sperm genes and if mutant females would instead express egg genes.
-To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes.
+Although many germline genes have shared functions in the male and female germline, some have unique or sex-biased expression.
+Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females primarily expressed egg genes.
+To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H).
 We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H).
-We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J).
-Germline genes dysregulated exclusively in _Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex (Figure 4I).
-This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.
-Altogether, these results demonstrate that both male and female _Kdm5c_ mutant cells express both sperm and egg-biased genes.
+This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes.
+We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J).
+Furthermore, germline genes dysregulated exclusively in either male or female mutants were also not biased towards their corresponding germ cell sex (Figure 4I).
+Altogether, these results demonstrate sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.
 

### 2024-08-30 submission/manuscript.Rmd
res 5

@@
 KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017].
-However, the lack of a comprehensive list of germline-enriched genes prohibited comprehensive characterization of KDM5C binding at germline gene promoters.
+However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters.
 Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
@@
 To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-EpiLCs had a higher total number of KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1, removal of _Kdm5c_-KO false positive peaks) and KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` &#177; `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
+EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1).
+KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` &#177; `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
 
-The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 5B).
+The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3,677 EpiLC only promoters) (Figure 5B).
 Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
@@
 There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs.
-We next evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes.
-KDM5C binding around the TSS of many germline genes is evident in EpiLCs, but absent in PNCs (Figure 5D).
-Based on our ChIP-seq peak cut-off criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 5E).
-In EpiLCs, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%).
-However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs (Supplementary figure XXX).
+Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D).
+Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288, 3)`% of all germline genes) (Figure 5E).
+Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%) (Supplementary figure 1A-C).
+However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs.
 Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
-Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including enrichment of meiotic genes, but does not directly regulate germline genes in neurons.
+Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
 Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
@@
 Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
-To elucidate if KDM5C is recruited to germline gene promoters by a similar mechanism, we used HOMER to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters[@heinzSimpleCombinationsLineageDetermining2010] (TSS &#177; 500 bp, q-value < 0.1).
+To elucidate if KDM5C is recruited to germline genes by a similar mechanism, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
 MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G).
 One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H).
-KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I).
+KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I, Supplementary figure 1D).
 RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure XXX).
+Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure 1E).
 Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.

### 2024-08-30 submission/manuscript.Rmd
add log2fc to 2-cell genes

@@
 
-zscan4d <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]
-duxf3 <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['padj'][,1]
+zscan4d_padj <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]
+duxf3_padj <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['padj'][,1]
+
+zscan4d_l2fc <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['log2FoldChange'][,1]
+duxf3_l2fc <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['log2FoldChange'][,1]
 
@@
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
+However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
 Altogether, _Kdm5c_-KO EpiLCs expressing key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes indicate germline gene misexpression parallels germline development during proper _Kdm5c_-KO differentiation.

### 2024-08-30 submission/supplement.Rmd
make supplementary figure file

--- 2024-08-30 submission/manuscript.Rmd
@@
 ---
-title:  Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
+title: Supplement - Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
 output: pdf_document
@@
 header-includes:
+   - \renewcommand{\figurename}{Supplementary Figure}
    - \usepackage{caption}
@@
 ---
-```{r, knitrcommas, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-#adds commas to large numbers
-library(huxtable)
-
-knitr::knit_hooks$set(inline = function(x) {
-  prettyNum(x, big.mark=",")
-})
-
-```
-Katherine M. Bonefas, Ilakkiya Venkatachalam, and Shigeki Iwase.
-
-# Abstract
-
-Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
-Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
-We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
-Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
-Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
-Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
-However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Thus, distinct suppressive mechanisms are recruited to different germline gene classes and ectopic germline transcriptional programs can mirror germ cell development within somatic tissues.
-
-# Introduction
-
-A single genome holds the instructions to generate the myriad of cell types found within an organism.
-This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs.
-Loss of some chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
-However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.
-
-To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
-Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
-MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
-Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
-However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.
-
-Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012].
-In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
-The contribution of KDM5C to this process remains unclear.
-Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes.
-Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g.
-meiotic genes versus spermatid differentiation genes.
-
-To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
-Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-
-# Results
-
-## Tissue-enriched genes are aberrantly expressed in the _Kdm5c_-KO brain
-
-```{r, DESeq2_params, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-#get the parameters used repeatedly, like the log2fold change cut off
-source("../code/utilities/parameters.R")
-```
-
-Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017].
-It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017].
-We analyzed tissue-specific differentially expressed genes (DEGs) in our published mRNA-seq datasets[@vallianatosMutuallySuppressiveRoles2020] of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1).
-
-```{r, tissue_genes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-#get the number and stats for tissue-enriched gene expression
-tissue_res <- read.csv("../data/processed/TissueSpecific_AMYHIP_numberofgenesDEGs.csv", header = TRUE)
-
-#get the value you want to print 
-#col is which data column, tiss is which tissue
-gettissval <- function(col, tiss){
-	val <- subset(tissue_res, tissue == tiss)
-	val <- val[,col]
-	return(val)
-}
-
-```
-
-We found a large proportion of significantly upregulated genes within the _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
-For both the amygdala and hippocampus, the majority of tissue-enriched (DEGs) were testis genes (Figure 1A-C).
-Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
-One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
-
-Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B).
-Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D).
-Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of gonad-enriched genes is independent of organismal sex.
-
-Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
-_Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
-
-For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
-Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
-Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
-
-```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-germgenes <- read.csv("../data/processed/germGENES20.csv", header = TRUE)
-
-#testis gene ontology results
-testisGO <- read.csv("../results/GO_Brain_TestisDEGs.csv")
-
-#function to get the GO ID
-getGO <- function(x, df, column, cluster){
-	df_test <- df
-
-	if (cluster == "no") {
-		out <- df_test[df_test$Description == x,][column][,1]
-	} else {
-		out <- df_test[df_test$Description == x & df_test$Cluster == cluster,][column][,1]
-	}
-	outformat <- ifelse(column == "ID", toString(out), out) #an attempt to remove the commas from GO ID numbers, didn't work
-	return(outformat) 
-}
-
-```
-### Germline genes are misexpressed in the _Kdm5c_-KO brain
-
-_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017], however the testis also contains somatic cells that support hormone production and germline functions.
-To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology.
-We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
-
-We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
-Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
-We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018].
-Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g.
-spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C).
-Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.
-
-As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive gene list.
-We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].
-We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
-
-## _Kdm5c_-KO epiblast-like cells aberrantly express key regulators of germline identity
-
-```{r, epilcs, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-
-#testis gene ontology results
-epigerm <- read.csv("../results/DESeq2/germDEGs/germDEGs_EpiLC_XY5cKO.csv")
-GO_EpiBrain <- read.csv("../results/GO_EpiLC_vs_Brain.csv")
-
-epi_restab <- read.csv("../data/processed/restable_EpiLC_XY5cKO.csv", row.names = 1)
-epi_restab$ENSEMBL <- rownames(epi_restab)
-
-zscan4d <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]
-duxf3 <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['padj'][,1]
-
-getresval <- function(df, col, symbol){
-	val <- subset(df, SYMBOL == symbol)
-	val <- val[,col]
-	return(val)
-}
-
-```
-
-Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], when they are silenced in epiblast stem cells that will  differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
-This developmental time point can be modeled _in vitro_ through differentiation of na&#239;ve embryonic stem cells (nESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
-While some germline-enriched genes are also expressed in nESCs and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
-
-_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
-We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
-Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D).
-Using our list of mouse germline-enriched genes assembled above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.
-
-We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
-The majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
-EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
-While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).
-
-Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H).
-These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
-Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (q = `r signif(duxf3, 3)`) and _Zscan4d_ (q = `r signif(zscan4d, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
-Altogether, _Kdm5c_-KO EpiLCs expressing key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes indicate germline gene misexpression parallels germline development during proper _Kdm5c_-KO differentiation.
-
-## Female epiblast-like cells have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
-```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-#get the parameters used repeatedly, like the log2fold change cut off
-source("../../KDM5C_Germ_Dev_Sex_RA/code/utilities/parameters.R")
-```
-
-It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
-[@samantaActivationXistEvolutionarily2022].
-
-Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
-While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were male-specific and female-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
-We then compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
-Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion).
-Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).
-
-The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D-F).
-This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression.
-Female sensitivity could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
-The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G).
-Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of XCI defects.
-
-### Germline gene misexpression in _Kdm5c_ mutants is independent of germ cell sex
-
-Although many germline genes have shared functions in the male and female germline, some have unique or sex-biased expression.
-Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females primarily expressed egg genes.
-To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H).
-We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes.
-We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J).
-Furthermore, germline genes dysregulated exclusively in either male or female mutants were also not biased towards their corresponding germ cell sex (Figure 4I).
-Altogether, these results demonstrate sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.
-
-## KDM5C binds to a subset of germline gene promoters during early embryogenesis
-
-KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017].
-However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters.
-Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
-
-```{r, KDM5C_ChIPseq, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-KDM5C_ChIP_EpiLC_promo <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_EpiLC.csv")
-
-KDM5C_ChIP_PNC_promo <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_PNC.csv")
-
-KDM5C_ChIP_EpiLC <- read.csv("../data/raw/ChIPseq_mm10_EpiLC_WTnoKO_consensus_peaks.bed", sep = "\t")
-
-KDM5C_ChIP_PNC <- read.csv("../data/raw/ChIPseq_mm10_PNC_WTnoKO_consensus_peaks.bed", sep = "\t")
-
-GO_KDM5C_ChIP <- read.csv("../results/ChIPseq_KDM5C_GO_EpiLC_vs_PNC_unique.csv")
-
-```
-
-To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1).
-KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` &#177; `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
-
-The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3,677 EpiLC only promoters) (Figure 5B).
-Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
-Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
-There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs.
-Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D).
-Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288, 3)`% of all germline genes) (Figure 5E).
-Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%) (Supplementary figure 1A-C).
-However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs.
-Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
-Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
-Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
-
-Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
-To elucidate if KDM5C is recruited to germline genes by a similar mechanism, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
-MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G).
-One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H).
-KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I, Supplementary figure 1D).
-RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure 1E).
-Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.
-
-## KDM5C is recruited to germline gene promoters harboring CpG islands to facilitate _de novo_ DNA methylation
-
-In the early embryo, germline gene promoters are initially decorated with repressive histone modifications and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-Our results above indicate KDM5C also acts at germline gene promoters during this time period.
-However, how KDM5C interacts with other germline gene silencing mechanisms is currently unclear.
-KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], yet KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
-Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required in the post-implantation embryo for long-term silencing of germline genes.
-In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
-
-Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
-To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3), at germline gene promoters in our previously published ChIP-seq datasets in male wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
-In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
-We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
-To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation (Figure 6C).
-While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6E).
-Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me at germline gene promoters.
-
-```{r, CGIs, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
-germgenes_CGI <- read.csv("../results/KDM5C_binding_allgerm_CGI.csv", header = TRUE)
-
-CGI <- germgenes_CGI[germgenes_CGI$Promo_CGI == "CGI",]
-
-ESCvsEpiLC <- read.csv("../results/WGBS/WGBS_restab_annotated_allgermTSS_500bp_WT_ESCvsEpiLC.csv")
-
-ESCvsEpiLC_25 <- ESCvsEpiLC[ESCvsEpiLC$meth.diff >= 25, ]
-
-WTvsKO <- read.csv("../results/WGBS/WGBS_restab_annotated_allgermTSS_500bp_WTvsKO.csv")
-
-WTvsKO_hypo <- WTvsKO[WTvsKO$meth.diff <= -25, ]
-WTvsKO_hypo_5C <- WTvsKO_hypo[WTvsKO_hypo$KDM5C_binding == "Bound", ]
-
-#get the infomration for wgbs
-getWGBSinfo <- function(df, gene, column){
-	df2 <- df
-	info <- df2[df2$SYMBOL == gene,][[column]]
-	return(info)
-}
-
-```
-
-In wild-type cells, germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
-We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) contained CGIs within their promoters (TSS &#177; 500 bp) (Figure 6F).
-CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) when compared to CGI-free genes (Figure 6G).
-Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline gene promoters (FISHER'S XXXX), with 79% of KDM5C-bound germline genes containing CGIs (Figure 6G).
-
-We then perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to assess how KDM5C loss impacted initial CpGme placement at germline gene promoters (Figure 6H).
-We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during ESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
-In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6I).
-
-We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J).
-Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs).
-We then evaluated promoter CpGme at germline genes ectopically transcibed in either _Kdm5c_-KO EpiLCs or within the brain and found promoter CpGme was substantially reduced in about half of germline DEGs (Figure 6K).
-Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
-Suprisingly, we only observed a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
-Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at select germline gene promoters during early embryogenesis, however other germline gene silencing mechansism can sufficiently compensate for KDM5C's loss at select germline gene promoters.
-
-# Discussion
-In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
-We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes.
-Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testicular somatic cells.
-_Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis.
-We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
-Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, as KDM5C is only bound to a subset of germline-enriched DEGs.
-Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands.
-Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages.
-Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.
-
-It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal.
-For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-Based on the de-repression of _Dazl_ and _Zscan4c_ in _Kdm5c_-KO ESCs, KDM5C was thought to promote the 2-cell-to-ESC transition[[@guptaGenomewideScreenReveals2023].
-Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes.
-Misexpression of many germline genes in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
-Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflect KDM5C's primary role in germline gene repression.
-
-Although eggs and sperm employ the same transcriptional programs for shared functions, e.g.
-PGC formation,  meiosis, and genome defense, some germline genes are sex specific.
-We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex.
-However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males.
-The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022].
-Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
-Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.
-
-Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
-KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
-
-- Our results indicate KDM5C's removal of H3K4me2/3 may be necessary at a subset of CGI-containing germline genes, given they fail to accumulate CpGme during EpiLC differentiation.
-These CpGme-sensitive germline genes, such as _D1Pas1_ and _Naa11_, are also de-repressed in the mature _Kdm5c_-KO brain and consistently dysregulated in multiple _Kdm5c_-KO datasets[@scandagliaLossKdm5cCauses2017].
-
-- H3K4me3 and CpGme typically do not colocalize
-	- https://pubmed.ncbi.nlm.nih.gov/17334365/
-	- https://www.nature.com/articles/s41594-017-0013-5
-	- https://pubmed.ncbi.nlm.nih.gov/23664763/
-	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2718496/
-	- https://pubmed.ncbi.nlm.nih.gov/19581485/
-
-- Contrastingly, many germline genes had only a modest reduction in CpGme, including _Dazl_ even though its TSS accumulated substantial H3K4me2. 
-- Multiple regulators repress dazl, dazl is not expressed in the mature _Kdm5c_-KO brain
-	E2F6/MGA/MAX
-		- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
-		-  Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
-
-- Altogether, our results indicates a given chromatin modifier's catalytic and non-catlytic regulation of gene expression can change depending upon the locus and developmental stage.
-
-Although the classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter.
-We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS.
-
-- CGIs highly determined KDM5C recruitment to germline gene promoters.
-
-- KDM5C loss also impacted CpGme at some KDM5C-unbound, CGI-free germline genes. Unclear what the repressive mechanism is for these genes.
-
-In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
-
-- Germline gene suppression has focused on the accumulation of CpGme at CGIs
-
-- DNAme and CpG islands
-
-- Combine germline gene list, CGI, RFX information together. 
-	- Benefit of the list is finding CGI difference and kdm5c indirect mechanisms of de-repression
-    - List of genes enables us to propose this model
-    - CGI positive meiosis/germline formation regulators turned on → these turn on CGI-negative late stage regulators → These promote downstream dysregulation long term
-
-- CGIs highly determine KDM5C recruitment to germline gene promoters.
-			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
-		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
-	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
-
-It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007].
-In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
-While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.
-Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters.
-Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
-Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.
-
-The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes.
-However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
-In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain.
-Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022].
-_APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008].
-KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.
-Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
-Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012], indicating their dysregulation may promote genome instability and cellular de-differentiation.
-Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
-Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression.
-Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary.
-Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
-
-# Materials and Methods
-
-## Classifying tissue-enriched and germline-enriched genes
-Tissue-enriched differentially expressd genes (DEGs) were determined by their classification in a previously published dataset from 17 male and female mouse tissues[@liComprehensiveMouseTranscriptomic2017].
-This study defined tissue expression as greather than 1 Fragments Per Kilobase of transcript per Million mapped read (FPKM) and tissue enrichment as at least 4-fold higher expression than any other tissue.
-
-We curated a list of germline-enriched genes using an RNA-seq dataset from wild-type and germline-depleted (Kit^W/Wv^) male and female mouse embryos from embryonic day 12, 14, and 16 [@sohGeneRegulatoryProgram2015], as well as adult male testes[@muellerIndependentSpecializationHuman2013].
-Germline-enriched genes met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type somatic tissues[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted (Kit^W/Wv^) germline, for any sex or time point, does not exceed 20% of maximum expression in wild-type germline.
-
-## Cell culture
-We utilized our previously established cultures of male wild-type and _Kdm5c_ knockout (-KO) embryonic stem cells[@samantaActivationXistEvolutionarily2022].
-Sex was confirmed by genotyping _Uba1_/_Uba1y_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'.
-Deletion of _Kdm5c_ was confirmed through the primers 5'-ATGCCCATATTAAGAGTCCCTG-3', 5'-TCTGCCTTGATGGGACTGTT-3', and 5'-GGTTCTCAACACTCACATAGTG-3'.
-
-Emrbyonic stem cells (ESCs) and epiblast-like cells were cultured using previously established methods[@samantaGeneratingPrimedPluripotent2020].
-Briefly, ESCs were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
-They were then transitioned into ground-state, "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
-Both pESC and nESC media were supplemented with 3 $\mu$M GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), 1 $\mu$M MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and 1,000 units/mL leukemia inhibitory factor (LIF, Millipore#ESG1107).
-
-nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF), as previously described [@samantaGeneratingPrimedPluripotent2020].
-
-## Immunocytochemistry (ICC)
-ICC of DAZL in EpiLCs was performed by first growing cells on fibronectin-coated coverslips.
-Cells were then washed thrice with phosphobuffered saline (PBS), fixed in 4% paraformaldehyde, washed thrice in PBS, and blacked in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour.
-They were then washed thrice with PBS and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4°C with gentle rocking.
-The next day, cells were rinsed thrice with PBS, and incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1,000) with DAPI (1:1,000) in blocking buffer for 1 hour at room temperature.
-Coverslips were then rinsed thrice in PBS, and mounted onto slides using Prolong Gold (Invitrogen #P36930).
-Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and quantified via ImageJ before unblinding.
-
-## RNA sequencing (RNA-seq)
-After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads.
-Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser.
-RStudio (v3.6.0) was then used to analyze counts files by DESeq2 (v1.26.0)[@loveModeratedEstimationFold2014a] to identify differentially expressed genes (DEGs) with a q-value (p-adjusted via FDR/Benjamini–Hochberg correction) less than 0.1 and a log2 fold change greater than 0.5.
-For all DESeq2 analyses, log2 fold changes were calculated with lfcShrink using the ashr package[@stephensFalseDiscoveryRates2016].
-MA-plots were generated by ggpubr (v0.6.0), and Eulerr diagrams were generated by eulerr (v6.1.1).
-Boxplots and scatterplots were generated by ggpubr (v0.6.0) and ggplot2 (v3.3.2).
-The Upset plot was generated via the package UpSetR (v1.4.0)[@conwayUpSetRPackageVisualization2017].
-Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
-
-## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
-ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches.
-Only uniquely mapped reads were used for analysis.
-Peaks were called using MACS2 software (v2.2.9.1) using input BAM files for normalization, with filters for a q-value < 0.1 and a fold enrichment > 1.
-We removed “black-listed” genomic regions that often give aberrant signals.
-Common peak sets were obtained in R via DiffBind (v3.6.5).
-In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (v2.25.0).
-Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1).
-Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
-Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010].
-Average binding across the genome was visualized using deeptools (v3.1.3).
-Bigwigs were visualized using the UCSC genome browser.
-
-## Whole genome bisulfite sequencing (WGBS)
-Genomic DNA (gDNA) from na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells.
-gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
-- bismark
-- Methylkit
-
-## Data availability
-### Published datasets
-All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo.
-Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797).
-
-Previously published ChIP-seq experiments included KDM5C in wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797) and mouse primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036).
-ChIP-seq of histone 3 lysine 4 dimethylation in male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
-ChIP-seq of histone 3 lysine 4 trimethylation in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
-
-### Data analysis
-Scripts used to generate the results, tables, and figures of this study are available via a GitHub repository: XXX
-
-# Acknowledgements
-- Jacob Mueller for providing insight in germline gene regulation. 
-- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast-like cells
-- Ilakkiya
-- Funding sources
-
-# References
-<div id="refs"></div>
-
-\newpage
-# Figures and Tables
-- Supplementary table 1: list of all germline genes. 
-	- Columns to include: 
-		- KDM5C bound vs not
-		- DEG in EpiLC, brain, both, neither (separate columns?)
-
-\begin{figure}
-  \centering 
-  \includegraphics{../submission/compiled_figs/Brain_tissue_genes.pdf}
-  \caption[Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
-Left - MA plot of mRNA-sequencing.
-Right - Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
-Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
-\textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
-Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
-\textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
-Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
-  \label{figurelabel}
-\end{figure}
-
-\begin{figure}
-  \centering 
-  \includegraphics{../submission/compiled_figs/What_are_germline_genes.pdf}
-  \caption[Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain]{\textbf{Aberrant transcription of germline genes in the \textit{Kdm5c}-KO in the brain.} \textbf{A.} enrichPlot gene ontology (GO) of \textit{Kdm5c}-KO amygdala and hippocampus testis-enriched DEGs \textbf{B.} Expression of testis DEGs in wild-type (WT) testis versus germ cell-depleted (W/Wv) testis (Mueller et al 2013).
-Expression is in Fragments Per Kilobase of transcript per Million mapped reads (FPKM).
-\textbf{C.} Number of testis DEGs that were classified as cell-type specific markers in a single cell RNA-seq dataset of the testis (Green et al 2018).
-Germline cell types are highlighted in green, somatic cell types in black.
-\textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and germline-depleted mice and in adult mouse non-gonadal tissues (Li et al 2017).}
-  \label{figurelabel}
-\end{figure}
-
-\begin{figure}
-  \centering 
-  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
-  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells.
-Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
-\textbf{B.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs.
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
-\textbf{C.}  Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs.
-\textbf{D.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs.
-Welch's t-test, expression in transcripts per million (TPM).
-\textbf{E.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation.
-Brightfield images taken at 20X.\textbf{F.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
-\textbf{G.} enrichPlot comparing enriched biological process gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
-\textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs.
-Top right - Example 2-cell genes that are not dysregulated in \textit{Kdm5c}-KO EpiLCs.
-p-values for Welch's t-test.
-Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{I.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
-Percentage of DAZL-positive cells normalized to number of DAPI-positive cells, p-value for Welch's t-test.}
-  \label{figurelabel}
-\end{figure}
-
-\begin{figure}
-  \centering 
-  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
-  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
-Left - MA plot of mRNA-sequencing.
-Right - Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
-Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
-\textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
-Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
-\textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
-Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
-  \label{figurelabel}
-\end{figure}
-
-\begin{figure}
-  \centering 
-  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
-  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom).
-\textbf{B.} Overlap of genes with KDM5C bound to their promoters in EpiLCs (purple) and PNCs (blue).
-\textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
-Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared).
-\textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right).
-\textbf{E.} KDM5C binding at the promoters of RNA-seq germline DEGs.
-Genes were classified as either only dysregulated in EpiLCs (EpiLC only), genes dyresgulated in the hippocampus or amygdala but not EpiLCs (brain only), or genes dysregulated in both EpiLCs and the brain (common).
-(Legend continued on next page.)}
-  \label{figurelabel}
-\end{figure}
-\addtocounter{figure}{-1}
-\begin{figure} [t!]
-  \caption[]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} (Legend continued.) \textbf{F.} Example KDM5C ChIP-seq bigwigs of DEGs unqiue to EpiLCs.
-Although both are expressed in \textit{Kdm5c}-KO EpiLCs, KDM5C is only bound to the \textit{Dazl} promoter and not the \textit{Stra8} promoter in EpiLCs.
-\textbf{G.} Example KDM5C ChIP-seq bigwigs of RNA-seq DEGs common between the brain and EpiLCs.
-KDM5C is bound to the \textit{D1Pas1} promoter but not the \textit{XXX} promoter in EpiLCs.
-\textbf{H.} Example KDM5C ChIP-seq bigwigs of RNA-seq DEGs unique to the brain.
-KDM5C is bound to the \textit{XXX} promoter but not the \textit{Meig1} promoter}
-\end{figure}
 
@@
   \centering
-  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
+  \includegraphics{../submission/compiled_figs/Supplement_KDM5C_ChIPseq.pdf}
   \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
@@
 \end{figure}
-
-\clearpage
-# Notes
-## Things to do
-- Move dazl to new figure if other staining works
-- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
-- Motif analysis
-	- Discussion - talk about motifs
-
-## Dazl
-We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019].
-A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
-This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function.
-We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I).
-We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020].
-Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
-
-- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
-**note: maybe just put in results**.
-Could move around depending upon if I get pheno working.
-
-## Discussion notes
-
-- For other paper:
-	- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
-	- - Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654
-
-- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
-	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
-- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
-	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
-
-- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
-	- then transition into the long term silencing mechanism paragraph
-
-- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
-
-- Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
-- Studies on other chromatin regulators have focused on key marker genes to identify germline gene misexpression, such as _Dazl_. 
-- Using a germline-depleted mouse model, we curated a list of mouse germline-enriched genes to globally assess germline gene dysregulation. 
-- Many of the genes dysregulated in the mature _Kdm5c_-KO brain were important for late stages of spermatogenesis, such as those important for sperm flagellar structure.
-Contrastingly, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germ cell identity, including _Dazl_ and _Stra8_.
-
-- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
-- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
-
-- The including the demarcation between soma and germline fates.
-
-	the _Kdm5c_-KO brain also 
-	Previous work identified testis-enriched genes in _Kdm5c_ genes
-	- 
-	- However unlike the gonadal-biased DEGs,
-
-- The demarcation of the germline versus soma is a key feature of multicellularity and sexually dimorphic reproduction
-
-- Anything known about tissue-biased gene expression in other H3K4me regulators?
-- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
-- tissue-biased gene expression:
-
-- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
-
-- Papers to read/reference:
-	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
-	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
-
-## Figure outline:
-
-**Figure 1: Misexpression of tissue-specific genes in the _Kdm5c_-KO brain**
-	* MA-plot and bar graphs of tissue-enriched genes
-	* Example testis-specific genes (NCBI and bigwigs)
-	* An example ovary tissue-specific gene 
-	* An example muscle/liver tissue-specific gene (NCBI and bigwigs)
-
-**Figure 2: The male _Kdm5c_-KO brain expresses male and female germline-enriched genes**
-	* Gene ontology of testis DEGs in the amygdala and hippocampus - ontologies are germline ontologies
-	* Expression of testis DEGs in germline-depleted testis (this is adult testis data)
-	* scRNAseq of testis - # of testis DEGs that are germline-specific markers
-	* Although far fewer, 5CKO brain also expresses ovary-enriched genes(NCBI and bigwigs of Zar1) 
-	* These ovary enriched genes are also germline specific (NCBI/Li tissues are in adult ovary, so it would be best to show they're oocyte-specific in adult ovary.
-But I don't think there's a published adult female W/Wv dataset.
-Could try looking at scRNAseq or just do TPM in embryonic W/Wv data since oocytes are developed at this point?
-Or both?)
-	* Defining what is/isn't a germline gene, and which are male/female biased using embryonic W/Wv data
-
-**Figure 3: Kdm5c-KO epiblast-like cells express key drivers regulators of germline identity**
-	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 
-	* B) 5CKO EpiLCs express EpiLC differentiation genes similar to WT lvls
-	* C) Male EpiLCs express germline genes (example Cyct again)
-	* Overlap between brain and EpiLC germline genes - show they're mostly unique
-	* GO of Brain and EpiLC germline genes (meiotic enriched) 
-	* Bigwigs or TPM of master regulators
-		* Show that while some are also 2-cell genes (Dazl), 2-cell specific genes aren't dysregulated (Zscan4).
-Important point because published KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I can tell only genes shared between germline and 2-cell are dysregulated.
-	* Staining of Dazl (+ Stra8 if I can get it to work)
-
-**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
-	* Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
-	* Increase in H3K4me2 in EpiLCs at germline genes 
-	* Kdm5c binding in EpiLCs vs PNCs to show that germline repression is happening in early embryo
-	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later).
-KDM5C RNA and protein ESC --> EpiLC (increasing then decreasing)
-	* RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
-	* DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)
-
-**Figure 5: Ectopic, germline-like phenotypes in Kdm5c-KO ESCs/EpiLCs**
-	* Sycp3 staining
-	* DDX4 staining and repression of retrotransposons
-	* Cilia??
-
-Gaps in knowledge addressed: 
-* Are other tissue-enriched genes dysregulated, or only testis, germline genes?
-* Curating a robust list of male and female germline genes
-	* Should talk about 2-cell genes vs germline genes - way to systematically categorize? 
-* Mechanism behind long-term germline gene misexpression
-	* Recent evidence suggests loss of KDM5C in ESCs express some germline genes
-		* Unclear if catalytic activity is required for long-term silencing
-		* Unclear if their dysregulation lasts throughout life or  the same between brain or not
-	* When in development does it begin?
-- Recent evidence suggests some germline genes expressed in 5CKO ESCs but unclear if their dysregulation lasts throughout differentiation and if the identity of germline genes are different compared to the brain
-* Are there functional consequences to germline gene misexpression?
-
-Introduction:
-* Chromatin regulators are important for cellular identity
-	* H3K4me1-3 linked to active gene promoters and enhancers
-	* Surprisingly, mutations in Chromatin regulators lead to many NDDs (including many H3K4 regulators)
-		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
-	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
-		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
-	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention
-
-* Loss of KDM5C can result in the misexpression of genes typically only found in the testis
-	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
-	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
-	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
-	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
-		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
-		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
-		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
-	* Systematic characterization of ectopic germline genes hasn't been done
-		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
-		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
-
-**Germline gene repression background:**
-
-Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017].
-Unlike somatic cells, germ cells (e.g.
-sperm and eggs) undergo meiosis and pass on their genetic material to the next generation.
-The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline.
-Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
-KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017].
-In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development.
-However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.

### 2024-08-30 code/utilities/paper_render.R

@@ -3,4 +3,5 @@
 library('tinytex')
 library('rmarkdown')
 render('submission/manuscript.Rmd')
+render('submission/supplement.Rmd')
 

### 2024-08-30 submission/manuscript.Rmd
add supplementary figure
remove notes

@@
 \clearpage
-# Notes
-## Things to do
-- Move dazl to new figure if other staining works
-- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
-- Motif analysis
-	- Discussion - talk about motifs
-
-## Dazl
-We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019].
-A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test).
-This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function.
-We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I).
-We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020].
-Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.
-
-- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].
-**note: maybe just put in results**.
-Could move around depending upon if I get pheno working.
-
-## Discussion notes
-
-- For other paper:
-	- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
-	- - Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654
-
-- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
-	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
-- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
-	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
-
-- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
-	- then transition into the long term silencing mechanism paragraph
-
-- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.
-
-- Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
-- Studies on other chromatin regulators have focused on key marker genes to identify germline gene misexpression, such as _Dazl_. 
-- Using a germline-depleted mouse model, we curated a list of mouse germline-enriched genes to globally assess germline gene dysregulation. 
-- Many of the genes dysregulated in the mature _Kdm5c_-KO brain were important for late stages of spermatogenesis, such as those important for sperm flagellar structure.
-Contrastingly, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germ cell identity, including _Dazl_ and _Stra8_.
-
-- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
-- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
-
-- The including the demarcation between soma and germline fates.
-
-	the _Kdm5c_-KO brain also 
-	Previous work identified testis-enriched genes in _Kdm5c_ genes
-	- 
-	- However unlike the gonadal-biased DEGs,
-
-- The demarcation of the germline versus soma is a key feature of multicellularity and sexually dimorphic reproduction
-
-- Anything known about tissue-biased gene expression in other H3K4me regulators?
-- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
-- tissue-biased gene expression:
-
-- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
-
-- Papers to read/reference:
-	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
-	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/
-
-## Figure outline:
-
-**Figure 1: Misexpression of tissue-specific genes in the _Kdm5c_-KO brain**
-	* MA-plot and bar graphs of tissue-enriched genes
-	* Example testis-specific genes (NCBI and bigwigs)
-	* An example ovary tissue-specific gene 
-	* An example muscle/liver tissue-specific gene (NCBI and bigwigs)
-
-**Figure 2: The male _Kdm5c_-KO brain expresses male and female germline-enriched genes**
-	* Gene ontology of testis DEGs in the amygdala and hippocampus - ontologies are germline ontologies
-	* Expression of testis DEGs in germline-depleted testis (this is adult testis data)
-	* scRNAseq of testis - # of testis DEGs that are germline-specific markers
-	* Although far fewer, 5CKO brain also expresses ovary-enriched genes(NCBI and bigwigs of Zar1) 
-	* These ovary enriched genes are also germline specific (NCBI/Li tissues are in adult ovary, so it would be best to show they're oocyte-specific in adult ovary.
-But I don't think there's a published adult female W/Wv dataset.
-Could try looking at scRNAseq or just do TPM in embryonic W/Wv data since oocytes are developed at this point?
-Or both?)
-	* Defining what is/isn't a germline gene, and which are male/female biased using embryonic W/Wv data
-
-**Figure 3: Kdm5c-KO epiblast-like cells express key drivers regulators of germline identity**
-	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 
-	* B) 5CKO EpiLCs express EpiLC differentiation genes similar to WT lvls
-	* C) Male EpiLCs express germline genes (example Cyct again)
-	* Overlap between brain and EpiLC germline genes - show they're mostly unique
-	* GO of Brain and EpiLC germline genes (meiotic enriched) 
-	* Bigwigs or TPM of master regulators
-		* Show that while some are also 2-cell genes (Dazl), 2-cell specific genes aren't dysregulated (Zscan4).
-Important point because published KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I can tell only genes shared between germline and 2-cell are dysregulated.
-	* Staining of Dazl (+ Stra8 if I can get it to work)
-
-**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
-	* Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
-	* Increase in H3K4me2 in EpiLCs at germline genes 
-	* Kdm5c binding in EpiLCs vs PNCs to show that germline repression is happening in early embryo
-	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later).
-KDM5C RNA and protein ESC --> EpiLC (increasing then decreasing)
-	* RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
-	* DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)
-
-**Figure 5: Ectopic, germline-like phenotypes in Kdm5c-KO ESCs/EpiLCs**
-	* Sycp3 staining
-	* DDX4 staining and repression of retrotransposons
-	* Cilia??
-
-Gaps in knowledge addressed: 
-* Are other tissue-enriched genes dysregulated, or only testis, germline genes?
-* Curating a robust list of male and female germline genes
-	* Should talk about 2-cell genes vs germline genes - way to systematically categorize? 
-* Mechanism behind long-term germline gene misexpression
-	* Recent evidence suggests loss of KDM5C in ESCs express some germline genes
-		* Unclear if catalytic activity is required for long-term silencing
-		* Unclear if their dysregulation lasts throughout life or  the same between brain or not
-	* When in development does it begin?
-- Recent evidence suggests some germline genes expressed in 5CKO ESCs but unclear if their dysregulation lasts throughout differentiation and if the identity of germline genes are different compared to the brain
-* Are there functional consequences to germline gene misexpression?
-
-Introduction:
-* Chromatin regulators are important for cellular identity
-	* H3K4me1-3 linked to active gene promoters and enhancers
-	* Surprisingly, mutations in Chromatin regulators lead to many NDDs (including many H3K4 regulators)
-		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
-	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
-		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
-	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention
-
-* Loss of KDM5C can result in the misexpression of genes typically only found in the testis
-	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
-	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
-	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
-	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
-		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
-		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
-		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
-	* Systematic characterization of ectopic germline genes hasn't been done
-		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
-		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.
-
-**Germline gene repression background:**
-
-Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017].
-Unlike somatic cells, germ cells (e.g.
-sperm and eggs) undergo meiosis and pass on their genetic material to the next generation.
-The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline.
-Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters.
-KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017].
-In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023].
-In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development.
-However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.

### 2024-08-30 submission/manuscript.Rmd
germline gene methylation results

@@
 
-## KDM5C is recruited to germline gene promoters harboring CpG islands to facilitate _de novo_ DNA methylation
+## KDM5C is recruited to CpG islands within germline gene promoters to facilitate _de novo_ DNA methylation
 
@@
 Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
-To test this hypothesis, we first characterized KDM5C's substrates, histone 3 lysine 4 di- and trimethylation (H3K4me2/3), at germline gene promoters in our previously published ChIP-seq datasets in male wild type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
-In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the transcription start site (TSS) of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
-We additionally found a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
+To test this hypothesis, we first characterized KDM5C's substrates (H3K4me2/3) at germline gene promoters in our published ChIP-seq datasets of male wild-type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
+In congruence with the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the TSS of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
+There was also a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
 To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation (Figure 6C).
@@
 CGI <- germgenes_CGI[germgenes_CGI$Promo_CGI == "CGI",]
+KDM5C_bound_CGI  <- germgenes_CGI[germgenes_CGI$KDM5C_binding == "Bound" & germgenes_CGI$Promo_CGI == "CGI",]
+
+KDM5C_bound  <- germgenes_CGI[germgenes_CGI$KDM5C_binding == "Bound",]
 
@@
 
-In wild-type cells, germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
-We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) contained CGIs within their promoters (TSS &#177; 500 bp) (Figure 6F).
+Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
+We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (TSS &#177; 500 bp) (Figure 6F).
 CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) when compared to CGI-free genes (Figure 6G).
-Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline gene promoters (FISHER'S XXXX), with 79% of KDM5C-bound germline genes containing CGIs (Figure 6G).
+Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (FISHER'S XXXX), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline genes containing CGIs (Figure 6G).
 
-We then perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to assess how KDM5C loss impacted initial CpGme placement at germline gene promoters (Figure 6H).
-We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during ESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
+To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) (Figure 6H).
+We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
 In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6I).
 
-We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J).
+We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J).
 Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs).
+Promoters that showed the most robust loss of CpGme (low q-values) harbored CGIs (Figure 6J).
 We then evaluated promoter CpGme at germline genes ectopically transcibed in either _Kdm5c_-KO EpiLCs or within the brain and found promoter CpGme was substantially reduced in about half of germline DEGs (Figure 6K).
-Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets, such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
-Suprisingly, we only observed a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
-Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at select germline gene promoters during early embryogenesis, however other germline gene silencing mechansism can sufficiently compensate for KDM5C's loss at select germline gene promoters.
+Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
+Suprisingly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
+Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters during early embryogenesis, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me.
 

### 2024-08-30 code/WGBS_germ_hist.R
histograms of CpG me

@@ -1,0 +1,187 @@
+# 2024.08.30 - histogram of germline gene average % methylation
+
+#want a histogram of germline gene methylation, split by CGI vs not, WT vs KO
+
+#read in the KDM5C bound genes/CGI
+germ <- read.csv(snakemake@input[[1]], sep = ",")
+
+print(head(germ))
+
+#get the gene coordinate information
+source("code/utilities/GeneTSSandTES.R")
+germTSS <- geneTSS_df(germ$ENSEMBL, 500)
+print(head(germTSS))
+
+
+## TSS
+TSSdf <- read.csv(snakemake@input[[2]], sep = ",")
+
+#rename chr to seqnames to match granges for merging later
+names(TSSdf)[names(TSSdf) == "chr"]<- "seqnames"
+print(paste("rows in all germ -", nrow(TSSdf)))
+print(head(TSSdf))
+	
+#get the germline genes for each coordinate
+plotTSS <- merge(germTSS, TSSdf)
+plotTSS <- merge(germ, plotTSS, by = "ENSEMBL")
+#remove any duplicated gene names
+# plotTSS <- plotTSS[!duplicated(plotTSS$SYMBOL),]
+print(paste("rows in merged all germ -", nrow(plotTSS)))
+print(head(plotTSS))
+
+write.table(plotTSS, snakemake@output[[1]], sep = ",", row.names = FALSE)
+
+
+#a dataframe that has the gene symbols and the % methylation for each sample as columns
+percmeth <- data.frame(ENSEMBL = plotTSS$ENSEMBL, plotTSS[,(ncol(plotTSS)-7):ncol(plotTSS)]) 
+
+avg_percmeth <- data.frame(ENSEMBL = plotTSS$ENSEMBL, "avg_WT_esc" = rowMeans(subset(percmeth, select = c("WT_A_esc", "WT_B_esc"))), 
+"avg_KO_esc" = rowMeans(subset(percmeth, select = c("KO_A_esc", "KO_B_esc"))),
+"avg_WT_96EpiLC" = rowMeans(subset(percmeth, select = c("WT_A_96EpiLC", "WT_B_96EpiLC"))),
+"avg_KO_96EpiLC" = rowMeans(subset(percmeth, select = c("KO_A_96EpiLC", "KO_B_96EpiLC"))))
+
+
+### subset for just genes that have a significant increase CpGme in WT ESC to exEpiLCs
[…]

### 2024-08-30 code/WGBS_germ_hist.R
add stats

@@ -69,13 +69,19 @@ for(s in 1:length(CGI_status)){
 
 	plot_df <- data.frame(ENSEMBL = rep(avg_percmeth_p$ENSEMBL, 2), genotype = c(rep("WT", nrow(avg_percmeth_p)), rep("KO", nrow(avg_percmeth_p))), AVGpercentMeth = c(avg_percmeth_p$avg_WT_96EpiLC, avg_percmeth_p$avg_KO_96EpiLC))
 
-	plotlist[[s]] <- gghistogram(plot_df, x = "AVGpercentMeth", add = "mean", rug = TRUE, color = "genotype", fill = "genotype", add_density = TRUE, title = paste("TSS -", CGI_status[s]), palette = c(EpiLC96_XY_KO, EpiLC96_XY_WT))
+		#do wt vs KO statistics
+	# stats[[s]] <- ks.test(plot_df[plot_df$genotype == "WT",][['AVGpercentMeth']], plot_df[plot_df$genotype == "KO",][['AVGpercentMeth']])
+
+	#compare means
+	sig <- wilcox.test(plot_df[plot_df$genotype == "WT",][['AVGpercentMeth']], plot_df[plot_df$genotype == "KO",][['AVGpercentMeth']])
+	stats[[s]] <- sig
+
+	p <- gghistogram(plot_df, x = "AVGpercentMeth", add = "mean", rug = TRUE, color = "genotype", fill = "genotype", add_density = TRUE, xlab = "average % CpGme",title = paste("exEpiLC % CpGme TSS", CGI_status[s]), palette = c(EpiLC96_XY_KO, EpiLC96_XY_WT))
 
 
-	#do wt vs KO statistics - Kolmogorov-Smirnov test
-	stats[[s]] <- ks.test(plot_df[plot_df$genotype == "WT",][['AVGpercentMeth']], plot_df[plot_df$genotype == "KO",][['AVGpercentMeth']])
 
-
+	plotlist[[s]] <- ggpar(p, subtitle = paste("Mann-Whitney p =", signif(sig$p.value, 3)), xlim = c(0, 100), font.subtitle = 10)
+	
 
 
 }
@@ -83,7 +89,7 @@ for(s in 1:length(CGI_status)){
 
 library("gridExtra")
 
-ggsave(snakemake@output[[2]], grid.arrange(grobs = plotlist, nrow = 1), width = 7, height = 3)
[…]

### 2024-09-02 submission/manuscript.Rmd
editing discussion

@@
 In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
-We first characterized the misexpression of tissue-enriched genes within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes.
-Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to germ cells and not expressed within testicular somatic cells.
-_Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expresses genes important for late spermatogenesis.
-We demonstrated that while _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
-Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, as KDM5C is only bound to a subset of germline-enriched DEGs.
-Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes long-term germline gene silencing through DNA methylation at CpG islands.
+We first characterized tissue-enriched genes expressed within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes.
+Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells.
+_Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis.
+We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
+Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation via ectopic RFX transcription factors.
+Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands.
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled unique repressive mechansisms governing distinct classes of germline gene in somatic lineages.
-Furthermore, these data provide molecular footholds that can then be exploited to test the ultimate contribution of ectopic germline gene expression upon neurodevelopment.
+By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled repressive mechansisms governing distinct classes of germline gene in somatic lineages.
+Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
 
-It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal.
-For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific mRNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-Based on the de-repression of _Dazl_ and _Zscan4c_ in _Kdm5c_-KO ESCs, KDM5C was thought to promote the 2-cell-to-ESC transition[[@guptaGenomewideScreenReveals2023].
-Although expressed in na&#239;ve ESCs, _Dazl_ and other "self-renewal" germline genes are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-We found that while _Kdm5c_-KO EpiLCs also expressed _Dazl_, they did not express 2-cell specific genes.
-Misexpression of many germline genes in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
-Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-Altogether, these data suggest _Kdm5c_-KO germline gene misexpression occurs ectopically in conjunction with typical developmental programs and the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflect KDM5C's primary role in germline gene repression.
-
+**make edits**
 Although eggs and sperm employ the same transcriptional programs for shared functions, e.g.
@@
 
-Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
-KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
+For example, while _Dazl_ is primarily known for regulating the translation of germline-specific mRNAs, it is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
+KDM5C was thought to promote the 2-cell-to-ESC transition, given _Dazl_ and _Zscan4c_ are de-repressed in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023].
+Although expressed in na&#239;ve ESCs, "self-renewal" germline genes like _Dazl_ are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
+We found that while _Kdm5c_-KO EpiLCs also express _Dazl_, they did not express 2-cell-specific genes.
+These data suggest the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflects KDM5C's primary role in germline gene repression.
+Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs), rather than de-differentiating into 2-cell-like cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
+Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
+In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated.
+Thus, ectopic germline gene expression occurs in parallel with proper _Kdm5c_-KO differentiation and somatic development.
+
+- In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
+- The classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter, which is facilitated by E2F6/DP1 and MAX/MGA binding sites
+- KDM5C does not have sequence specific binding, associates with PRC1.6 but likely not the major mechanisms of recruitment 
+- KDM5C highly enriched at meiotic, promoter CGI germline genes
+- A plausible mechanism for KDM5C-mediated germline gene silencing is KDM5C's removal of H3K4me2/3 at germline gene CGIs[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given that (H3K4me2/3) oppose DNMTs and primarily do no colocalize with CpGme.
+- Indeed we observed H3K4me2/3 accumulation at germline TSS and impaired CGI methylation at select germline genes. Loss of KDM5C strongly impairs CpGme at a subset of CGI-containing germline genes.
+- However, Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
+- KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+In EpiLCs, Dazl still gained CpGme even with elevated H3K4me2.
+This suggests other redundant silencing mechanisms to ensure eventual CpGme of dazl, however other germline CGIs require KDM5C to ensure sufficient methylation
+- Furthermore, our results indicates a histone modifier's catalytic and non-catlytic roles can change depending upon the locus and developmental stage.
+
+- It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007].
+In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
+While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.
+
+- Our list of germline genes
+- We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS. 
+	- CGI-containing = meiotic
+- Loss of KDM5C binding at a subset of CGI-containing germline genes in EpiLCs, e.g.
+D1Pas1, Naa11, strongly impairs CpGme, resulting in their ectopic transcription in both EpiLCs and in the mature brain.
+- Our germline gene list enabled us to identify germline-enriched genes activated independent of
+
+- This analysis revealed some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. 
+- Germline genes unbound by KDM5C primarily late-stage spermatogenesis genes that lacked CGIs and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
+
+Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
+Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.
+
+- KDM5C loss also activates RFX2 transctiption factors whose binding sites are enriched at non-CGI spermiogenesis genes.
+
+- CGIs are resistant
+	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5001583/
+	- https://pubmed.ncbi.nlm.nih.gov/14504655/
 
@@
 
-- Altogether, our results indicates a given chromatin modifier's catalytic and non-catlytic regulation of gene expression can change depending upon the locus and developmental stage.
-
-Although the classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter.
-We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS.
-
-- CGIs highly determined KDM5C recruitment to germline gene promoters.
-
-- KDM5C loss also impacted CpGme at some KDM5C-unbound, CGI-free germline genes. Unclear what the repressive mechanism is for these genes.
-
-In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
-
-- Germline gene suppression has focused on the accumulation of CpGme at CGIs
-
-- DNAme and CpG islands
+-
 
@@
 - CGIs highly determine KDM5C recruitment to germline gene promoters.
-			- KDM5C is known to be enriched at CGIs (in neurons? are these methylated? or is its germline CGI function different from its somatic CGI function?). 
 		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
-	- Altogether demonstrates different classes of germline genes use distinct silencing mechanisms
-
-It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007].
-In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
-While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.
-Additionally, some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters.
-Germline genes unbound by KDM5C were primarily late-stage spermatogenesis genes and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
-Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.
 
@@
 nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF), as previously described [@samantaGeneratingPrimedPluripotent2020].
-
-## Immunocytochemistry (ICC)
-ICC of DAZL in EpiLCs was performed by first growing cells on fibronectin-coated coverslips.
-Cells were then washed thrice with phosphobuffered saline (PBS), fixed in 4% paraformaldehyde, washed thrice in PBS, and blacked in PBS containing 0.3% Triton X-100, and 5% fetal bovine serum for 1 hour.
-They were then washed thrice with PBS and incubating in primary antibody (Rabbit anti DAZL, abcam ab34139, 1:200) in the blocking solution overnight at 4°C with gentle rocking.
-The next day, cells were rinsed thrice with PBS, and incubated in secondary antibody (Alexafluor 488 Invitrogen #710369, 1:1,000) with DAPI (1:1,000) in blocking buffer for 1 hour at room temperature.
-Coverslips were then rinsed thrice in PBS, and mounted onto slides using Prolong Gold (Invitrogen #P36930).
-Images were taken blinded for genotype, chosen based on similar levels of DAPI signal, and quantified via ImageJ before unblinding.
 

### 2024-09-02 code/WGBS_volcano.R
edit aesthetics volcano

@@ -99,18 +99,19 @@ TSS_volcano <- function(n, TITLE, restab, maxy){
 
 	
 	#labdesc <- ifelse(labels == "yes", restab$SYMBOL, ifelse(labels == "no", NA, "oop"))
+	labels <- c("D1Pas1", "Cyct", "Stra8", "Zar1", "Dazl", "Naa11", "Tex15", "Tex14", "Tex11", "Rpl10l")
 
 	library(EnhancedVolcano)
 	p <- EnhancedVolcano(restab, lab = restab$SYMBOL,
     	                 x = 'meth.diff', y = 'qvalue',
-            	         #selectLab = labels,
+            	         selectLab = labels,
                 	     xlim=c(-100,100),
     	                 xlab ="% methylation difference",
         	             ylab = bquote(~-Log[10]~qvalue),
             	         pCutoff = 0.01, FCcutoff = l2fcco,
     	                 title = TITLE,
         	             subtitle = " ",
-            	         labSize = 4.0,
+            	         labSize = 5.5,
                 	     colAlpha = .6,
 						 pointSize = 3,
                     	 colCustom = keyvals,
@@ -118,8 +119,8 @@ TSS_volcano <- function(n, TITLE, restab, maxy){
             	         legendPosition = 'right',
                 	     gridlines.major = FALSE,
                     	 gridlines.minor = FALSE,
-						 drawConnectors = TRUE, widthConnectors = 1.0, colConnectors = 'black',
-                	     legendLabSize = 10, legendIconSize = 3.0)
+						 drawConnectors = TRUE, widthConnectors = .5, colConnectors = 'black',
+                	     legendLabSize = 12, legendIconSize = 3.0)
 
 	ggsave(snakemake@output[[n]], plot = p, width = 9, height = 7)
 
@@ -294,12 +295,12 @@ CGI_volcano <- function(n, TITLE, restab){
 	# names(keyvals.shape)[keyvals.shape == 16] <- 'Regular'
 
 	#select label
-	labels <- c("D1Pas1", "Stra8", "Zar1", "Dazl")
+	#labels <- c("D1Pas1", "Cyct", "Stra8", "Zar1", "Dazl", "Naa11", "Tex15", "Tex14", "Tex11", "Rpl10l")
 
 	library(EnhancedVolcano)
 	p <- EnhancedVolcano(restab, lab = restab$SYMBOL,
     	                 x = 'meth.diff', y = 'qvalue',
-            	         selectLab = labels,
+            	         #selectLab = labels,
[…]

### 2024-09-02 submission/manuscript.Rmd
discusssion edits

@@
 
-Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021].
-To elucidate if KDM5C is recruited to germline genes by a similar mechanism, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
+Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+PRC1.6 members have been proposed to recruit KDM5C to germline gene promoters, given their association in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023] and that KDM5C itself does not contain sequence-specific binding domains[@iwaseXLinkedMentalRetardation2007].
+To elucidate how KDM5C is recruited to germline genes, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
 MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G).
@@
 
-**make edits**
 Although eggs and sperm employ the same transcriptional programs for shared functions, e.g.
@@
 
-- In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
-- The classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter, which is facilitated by E2F6/DP1 and MAX/MGA binding sites
-- KDM5C does not have sequence specific binding, associates with PRC1.6 but likely not the major mechanisms of recruitment 
-- KDM5C highly enriched at meiotic, promoter CGI germline genes
-- A plausible mechanism for KDM5C-mediated germline gene silencing is KDM5C's removal of H3K4me2/3 at germline gene CGIs[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given that (H3K4me2/3) oppose DNMTs and primarily do no colocalize with CpGme.
-- Indeed we observed H3K4me2/3 accumulation at germline TSS and impaired CGI methylation at select germline genes. Loss of KDM5C strongly impairs CpGme at a subset of CGI-containing germline genes.
-- However, Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
-- KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
-In EpiLCs, Dazl still gained CpGme even with elevated H3K4me2.
-This suggests other redundant silencing mechanisms to ensure eventual CpGme of dazl, however other germline CGIs require KDM5C to ensure sufficient methylation
-- Furthermore, our results indicates a histone modifier's catalytic and non-catlytic roles can change depending upon the locus and developmental stage.
+While promoter CGIs are typically resistant to CpGme[@longProtectionCpGIslands2016], CGIs at germline promoters are hightly methylated in somatic cells[@weberDistributionSilencingPotential2007].
+In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g.
+_D1Pas1_ and _Naa11_, strongly impaired CGI methylation, which resulted in their long-lasting de-repression into adulthood.
+Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 primarily do not colocalize with CpGme[@weberDistributionSilencingPotential2007] and can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
+Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023].
+In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2.
+_Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+Together, this suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme.
+Furthermore, these results indicate the requirement for catalytic activity for the same class of genes can change depending upon the locus and developmental stage.
 
-- It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007].
-In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018].
-While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.
-
-- Our list of germline genes
-- We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS. 
-	- CGI-containing = meiotic
-- Loss of KDM5C binding at a subset of CGI-containing germline genes in EpiLCs, e.g.
-D1Pas1, Naa11, strongly impairs CpGme, resulting in their ectopic transcription in both EpiLCs and in the mature brain.
-- Our germline gene list enabled us to identify germline-enriched genes activated independent of
-
-- This analysis revealed some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. 
-- Germline genes unbound by KDM5C primarily late-stage spermatogenesis genes that lacked CGIs and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-
-Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006].
-Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.
-
-- KDM5C loss also activates RFX2 transctiption factors whose binding sites are enriched at non-CGI spermiogenesis genes.
-
-- CGIs are resistant
-	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5001583/
-	- https://pubmed.ncbi.nlm.nih.gov/14504655/
-
-- Our results indicate KDM5C's removal of H3K4me2/3 may be necessary at a subset of CGI-containing germline genes, given they fail to accumulate CpGme during EpiLC differentiation.
-These CpGme-sensitive germline genes, such as _D1Pas1_ and _Naa11_, are also de-repressed in the mature _Kdm5c_-KO brain and consistently dysregulated in multiple _Kdm5c_-KO datasets[@scandagliaLossKdm5cCauses2017].
-
-- H3K4me3 and CpGme typically do not colocalize
-	- https://pubmed.ncbi.nlm.nih.gov/17334365/
-	- https://www.nature.com/articles/s41594-017-0013-5
-	- https://pubmed.ncbi.nlm.nih.gov/23664763/
-	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2718496/
-	- https://pubmed.ncbi.nlm.nih.gov/19581485/
-
-- Contrastingly, many germline genes had only a modest reduction in CpGme, including _Dazl_ even though its TSS accumulated substantial H3K4me2. 
-- Multiple regulators repress dazl, dazl is not expressed in the mature _Kdm5c_-KO brain
-	E2F6/MGA/MAX
-		- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
-		-  Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci
-
--
-
-- Combine germline gene list, CGI, RFX information together. 
-	- Benefit of the list is finding CGI difference and kdm5c indirect mechanisms of de-repression
-    - List of genes enables us to propose this model
-    - CGI positive meiosis/germline formation regulators turned on → these turn on CGI-negative late stage regulators → These promote downstream dysregulation long term
-
-- CGIs highly determine KDM5C recruitment to germline gene promoters.
-		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme
+By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct repressive mechanisms governing early versus late-stage germline developmental programs.
+In EpiLCs, KDM5C was highly enriched at germline promoters containing CGIs, many of which had roles in early germ cell formation and meiosis.
+However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells.
+CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
+These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_ and _Dazl_ turn on downstream germline programs that can loosely mimic germ cell development, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
+Therefore, we propose KDM5C is recruited to germline genes that shape germ cell formation via their promoter CGIs to act as break against runaway activation of germline-specific programs.
+Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation and acts as break against runaway activation of germline-specific programs.
 

### 2024-09-02 code/WGBS_volcano.R

@@ -99,7 +99,7 @@ TSS_volcano <- function(n, TITLE, restab, maxy){
 
 	
 	#labdesc <- ifelse(labels == "yes", restab$SYMBOL, ifelse(labels == "no", NA, "oop"))
-	labels <- c("D1Pas1", "Cyct", "Stra8", "Zar1", "Dazl", "Naa11", "Tex15", "Tex14", "Tex11", "Rpl10l")
+	labels <- c("D1Pas1", "Cyct", "Ddx4", "Stra8", "Zar1", "Dazl", "Naa11", "Tex15", "Tex14", "Tex11", "Rpl10l")
 
 	library(EnhancedVolcano)
 	p <- EnhancedVolcano(restab, lab = restab$SYMBOL,

### 2024-09-03 submission/manuscript.Rmd
fix order of last fig and add in results

@@
 Promoters that showed the most robust loss of CpGme (low q-values) harbored CGIs (Figure 6J).
-We then evaluated promoter CpGme at germline genes ectopically transcibed in either _Kdm5c_-KO EpiLCs or within the brain and found promoter CpGme was substantially reduced in about half of germline DEGs (Figure 6K).
-Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
-Suprisingly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
+
+We then compared average percent CpGme at germline gene promtoers with or without CGIs.
+While the degree of CpGme was largely unchanged for non-CGI promoters, CGI promoters on average had a significant reduction in CpGme with KDM5C loss (Figure 6K) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test).
+Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
+Suprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
 Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters during early embryogenesis, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me.

### 2024-09-09 submission/supplement.Rmd
supplement figure legend

@@
   \includegraphics{../submission/compiled_figs/Supplement_KDM5C_ChIPseq.pdf}
-  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
-Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala.
-\textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs.
-Right - Average H3K4me2 at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs.
-\textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein.
-\textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}).
-* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test.
-\textbf{E.} KDM5C protein expression normalized to DAXX.
-Quantified intensity using ImageJ (artificial units - au).
-Right - representative lanes of Western blot for KDM5C and DAXX.
-* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} XXX {G.} XXX.}
+  \caption[KDM5C binds to a subset of germline RNA-seq differentially expressed genes]{\textbf{KDM5C binds to a subset of germline RNA-seq differentially expressed genes.} \textbf{A.} Bar graph of the number of germline-enriched DEGs with promoter KDM5C ChIP-seq peaks in wild-type EpiLCs (Top) and PNCs (Bottom).
+RNA-seq DEGs were classified as shared betwen EpiLCs and the brain (Common), unique to EpiLCs (EpiLC Only), or unique to one or multiple brain regions (Brain Only).
+\textbf{B.} Average bigwigs of two example RNA-seq DEGs dysregulated in EpiLCs but not the brain, \textit{Dazl} and \textit{Stra8}.
+Top is the RNA-seq tracks for wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs, bottom is the KDM5C ChIP-seq tracks, with the annotated transcription start site (TSS) for each gene.
+\textbf{C.} Same as B but for two example DEGs only dysregualted in the brain and not expressed in EpiLCs, \textit{Ccnb1ip1} and \textit{Spata18}.
+\textbf{D.} HOMER motif analysis of all KDM5C-unbound germline DEGs shows signficant enrichment of multiple RFX members and their X-box motif.
+\textbf{E.} KDM5C ChIP-seq shows no KDM5C accumulation at the \textit{Rfx2} promoter in EpiLCs or PNCs.}
   \label{figurelabel}

### 2024-09-09 submission/manuscript.Rmd
update fig3 caption

--- 2024-09-02 submission/manuscript.Rmd
@@
 Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
-\textbf{B.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs.
+\textbf{B.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation.
+Brightfield images taken at 20X.
+\textbf{C.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs.
+Welch's t-test, expression in transcripts per million (TPM).
+\textbf{D.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs.
 * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
-\textbf{C.}  Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs.
-\textbf{D.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs.
-Welch's t-test, expression in transcripts per million (TPM).
-\textbf{E.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation.
-Brightfield images taken at 20X.\textbf{F.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
+\textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
+\textbf{F.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue).
 \textbf{G.} enrichPlot comparing enriched biological process gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
@@
 p-values for Welch's t-test.
-Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{I.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs.
-Percentage of DAZL-positive cells normalized to number of DAPI-positive cells, p-value for Welch's t-test.}
+Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs.}
   \label{figurelabel}

### 2024-09-09 submission/manuscript.Rmd
fig 4 legend

@@
 \textbf{F.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue).
-\textbf{G.} enrichPlot comparing enriched biological process gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
+\textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
 \textbf{H.} Top left - Example germline identity DEGs unique to EpiLCs.
@@
   \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
-  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B).
-Left - MA plot of mRNA-sequencing.
-Right - Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
-Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
-\textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
-Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
-\textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
-Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
+  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs.
+\textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs.
+Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only).
+\textbf{D.} Heatmap of germline DEGs shared between male and female mutants.
+Color is the average log 2 fold change from sex-matched wild-type.
+\textbf{E.} Number of genes within each cluster from D.
+Clusters with higher expression in females compared to males (XX-biased) highlighted in orange.
+\textbf{F.} Example average bigwigs of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants.
+\textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome.
+Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant.
+X chromosome highlighted in black.
+\textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes.
+\textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants.
+\textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) dysregulated in both male and female \textit{Kdm5c} mutants.}
   \label{figurelabel}

### 2024-09-09 submission/manuscript.Rmd
fig 5 legend

@@
   \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom).
-\textbf{B.} Overlap of genes with KDM5C bound to their promoters in EpiLCs (purple) and PNCs (blue).
+\textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS &#177; 500) in EpiLCs (purple) and PNCs (blue).
 \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
@@
 \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right).
-\textbf{E.} KDM5C binding at the promoters of RNA-seq germline DEGs.
-Genes were classified as either only dysregulated in EpiLCs (EpiLC only), genes dyresgulated in the hippocampus or amygdala but not EpiLCs (brain only), or genes dysregulated in both EpiLCs and the brain (common).
-(Legend continued on next page.)}
+\textbf{E.} Eulerr of number of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple).
+\textbf{F.} Example KDM5C ChIP-seq bigwigs of KDM5C binding at the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs.
+{G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs.
+\textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence.
+\textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs.
+\textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs.
+P-value of Welch's t-test, expression in transcripts per million (TPM).}
   \label{figurelabel}
 \end{figure}
-\addtocounter{figure}{-1}
-\begin{figure} [t!]
-  \caption[]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} (Legend continued.) \textbf{F.} Example KDM5C ChIP-seq bigwigs of DEGs unqiue to EpiLCs.
-Although both are expressed in \textit{Kdm5c}-KO EpiLCs, KDM5C is only bound to the \textit{Dazl} promoter and not the \textit{Stra8} promoter in EpiLCs.
-\textbf{G.} Example KDM5C ChIP-seq bigwigs of RNA-seq DEGs common between the brain and EpiLCs.
-KDM5C is bound to the \textit{D1Pas1} promoter but not the \textit{XXX} promoter in EpiLCs.
-\textbf{H.} Example KDM5C ChIP-seq bigwigs of RNA-seq DEGs unique to the brain.
-KDM5C is bound to the \textit{XXX} promoter but not the \textit{Meig1} promoter}
-\end{figure}
 
-\newpage
 \begin{figure}

### 2024-09-09 submission/manuscript.Rmd
fig 6 figure legend

@@
 \textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein.
-\textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} RNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}).
+\textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}).
 * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test.
@@
 Right - representative lanes of Western blot for KDM5C and DAXX.
-* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} XXX {G.} XXX.}
+* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters, based on UCSC annotation.
+Comparing all germline-enriched genes, KDM5C-bound germline genes, or KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of CGI-promoter versus non-CGI promoter germline genes.
+\textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS &#177; 500) in wild-type ESCs versus exEpiLCs.
+Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink \textbf{J.} Volcano plot of WGBS of wild-type versus \textit{Kdm5c}-KO exEpiLCs.
+Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink, hypomethylated genes lackig a promoter CGI in blue.
+\textbf{K.} Example UCSC browser shots of germline genes of CpG methylation (CpGme) in wild-type and \textit{Kdm5c}-KO ESCs and exEpiLCs.
+\textbf{L.} Histogram of average percent CpGme at the promoter for germline genes with or without promoter CGIs.
+Wild-type in navy and \textit{Kdm5c}-KO in light blue.
+Dashed lines are average methylation for each genotype, p-values for Mann−Whitney U test.}
   \label{figurelabel}

### 2024-09-09 submission/manuscript.Rmd
acknowledgements and contributions

@@
 
+## CpG island (CGI) analysis
+Locations of CpG islands were determined through the mm10 UCSC genome browser CpG island track.
+CGI coordinates were then annotated using ChIPseeker (v1.32.1) and filtered for promoters of germline genes (TSS &#177; 500).
+
 ## Whole genome bisulfite sequencing (WGBS)
@@
 gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
-- bismark
-- Methylkit
+Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark (v0.22.1).
+Analysis of differential methylation at germline gene promoters was perfomed using methylKit (v1.28.0) with a minimum coverage of 3 paired reads, a percentage cut-off of 25%, and q-value of 0.01.
+Average percentage methylation at germline gene promoters was determined via methylKit (v1.28.0).
+Methylation bedgraph tracks were generated via Bismark and visualized using the UCSC genome browser.
 
@@
 # Acknowledgements
-- Jacob Mueller for providing insight in germline gene regulation. 
-- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast-like cells
-- Ilakkiya
-- Funding sources
+We thank Drs.
+Sundeep Kalantry, Milan Samanta, and Rebecca Malcore for providing protocols and expertise in culturing mouse ESCs and EpiLCs, as well as providing wild-type and _Kdm5c_-KO ESCs used in this study.
+We thank Dr.
+Jacob Mueller for his insight in germline gene regulation and directing us to the germline-depleted mouse models.
+We also thank Drs.
+Stephanie Bielas, Michael Sutton, Donna Martin, and the members of the Iwase, Sutton, Bielas, and Martin labs for helpful discussions and critiques of the data.
+We thank members of the University of Michigan Reproductive Sciences Program for providing feedback throughout the development of this work.
+This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B), and the NIH Early Stage Training in the Neurosciences Training Grant (T32-NS076401 to K.M.B).
+
+# Author contributions
+K.M.B.
+and S.I.
+conceived the study and designed the experiments.
+I.V.
+generated the ESC and exEpiLC WGBS data.
+K.M.B performed the data analysis and all other experiments.
+K.M.B and S.I.
+wrote and edited the manuscript.
 

### 2024-09-09 submission/manuscript.Rmd
render figure legends and fix plus minus issue

@@
   \centering 
+  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
+  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs.
+\textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs.
+Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only).
+\textbf{D.} Heatmap of germline DEGs shared between male and female mutants.
+Color is the average log 2 fold change from sex-matched wild-type.
+\textbf{E.} Number of genes within each cluster from D.
+Clusters with higher expression in females compared to males (XX-biased) highlighted in orange.
+\textbf{F.} Example average bigwigs of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants.
+\textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome.
+Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant.
+X chromosome highlighted in black.
+\textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes.
+\textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants.
+\textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) dysregulated in both male and female \textit{Kdm5c} mutants.}
+  \label{figurelabel}
+\end{figure}
+
+\begin{figure}
+  \centering 
   \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
-  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis A.} ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom).
-\textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS &#177; 500) in EpiLCs (purple) and PNCs (blue).
+  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis.]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.}  ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom).
+\textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue).
 \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
@@
 Comparing all germline-enriched genes, KDM5C-bound germline genes, or KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of CGI-promoter versus non-CGI promoter germline genes.
-\textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS &#177; 500) in wild-type ESCs versus exEpiLCs.
+\textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS $\pm$ 500) in wild-type ESCs versus exEpiLCs.
 Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink \textbf{J.} Volcano plot of WGBS of wild-type versus \textit{Kdm5c}-KO exEpiLCs.
@@
 Wild-type in navy and \textit{Kdm5c}-KO in light blue.
-Dashed lines are average methylation for each genotype, p-values for Mann−Whitney U test.}
+Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.}
   \label{figurelabel}

### 2024-09-09 code/DESeq2_EpiLC.R
switch EpiLC to XX vs XY results

@@ -1,110 +1,0 @@
-#############################################################################################
-#                                                                                           #
-# 2022 12 14 DESeq2 using snakemake of Kdm5c mutant EpiLCs                                  #
-# Inputs:
-#       cts = counts file (.txt)                                                            #
-#       sampleinfo = sample information sheet #MAKE SURE THE SAMPLES MATCH THE CTS FILE NAMES
-# Libraries: DESeq2, ggplot2                                                                #
-# Outputs: DESeq2 results tables (.csv)                             #
-#                                                                                           #
-#############################################################################################
-#load the color palette utilities first
-source('code/utilities/colorpalettes.R')
-
-###1) format the raw counts table so that all values are integers and there are no gene variant numbers
-cts <- read.csv(snakemake@input[["cts"]], sep ="\t", row.names = 1)
-    #convert counts to integers
-cts[1:ncol(cts)] <- lapply(cts[1:ncol(cts)], as.integer)
-
-###2) Make the sample sheet (a csv file that has the genotype and sex for each sample)
-#get the column (sample) names
-samples <-  colnames(cts)
-
-#get the sex of each sample
-sex <- c()
-ct <- 1
-for(r in samples){
-		#get the first two characters of the string. If they're XX, they're female
-	if (substr(r, 1, 2) == "XX"){
-		rei <- "XX"
-	    } else if (substr(r, 1, 2) == "XY"){ #if they're male
-		rei <- "XY"
-	    } else {
-		rei <- "ERROR"
-	}
-	sex[ct] <- rei
-	ct <- ct + 1
-	
-}
-
-#get the genotype of each sample
-genotype <- c()
-cnt <- 1
-for(g in samples){
-		#get the end fo the string
-	if (substr(g, 10, 11) == "WT"){
-		geno <- "WT"
-        } else if (substr(g, 10, 12) == "HET") {
-        geno <- "5cHET"
-        } else if (substr(g, 10, 11) == "KO"){
-        geno <- "5cKO"
-        } else { #the rest of the samples have 7 characters
-		geno <- "ERROR"
-	}
-	genotype[cnt] <- geno
-	cnt <- cnt + 1
-}
-
-
-SampleInfo <- data.frame(Sample = colnames(cts), Genotype = genotype, Sex = sex, Type = rep("paired.end", length(genotype)))
[…]

### 2024-09-09 code/Upset_Brain_EpiLC.R

@@ -3,7 +3,7 @@
 ##1) Read in the germline genes list
 germ <- read.csv(snakemake@input[[1]], sep = ",")
 
-##2) Read in the EpiLC and brain XX and XY germline DEGs, make sure files are in same order
+##2) Read in the EpiLC and brain XY germline DEGs, make sure files are in same order
 source("code/utilities/parameters.R") #for the log2fc cut off (l2fcco)
 samples <- c("EpiLC XY 5CKO", "AMY 5CKO", "HIP 5CKO")
 DEGs <- list()

### 2024-09-09 submission/manuscript.Rmd
update XY EpiLC to XX vs XY deseq2 results

@@
 We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development.
-The majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F).
+The majority of germline DEGs were unique to either EpiLCs or the brain, with only _D1Pas1_ and _Cyct_ shared across all tissue/cell types (Figure 3E-F).
 EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
@@
 \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
-\textbf{F.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue).
+\textbf{F.} Average bigwigs of an example germline gene, \textit{D1Pas1}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue).
 \textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
@@
 \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants.
-\textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) dysregulated in both male and female \textit{Kdm5c} mutants.}
+\textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants.}
   \label{figurelabel}

### 2024-09-11 code/WGBS_germ_CGI.R
add comments

@@ -1,5 +1,5 @@
 
-### 24.04.26 compare binding locations of KDM5C ChIp-seq peaks and number of germline genes bound
+### 24.04.26 compare binding locations of KDM5C ChIp-seq peaks and number of germline genes bound for CGI and CGI-free germline genes
 
 
 # Visualize genomic locations using ChIPseeker

### 2024-09-11 code/serverscripts/WGBS/methylkit_mincov3.R
clarify WGBS instructions

@@ -76,14 +76,14 @@ for (i in cells){
 
 
 for (n in cells){
-	#get the methylation for cgis
-	regions <- regionCounts(mylists[[n]],bed)
+	# #get the methylation for cgis
+	# regions <- regionCounts(mylists[[n]],bed)
 
-	united_regions <- unite(regions, destrand=FALSE)
-	print("united regions")
-	print(head(united_regions))
+	# united_regions <- unite(regions, destrand=FALSE)
+	# print("united regions")
+	# print(head(united_regions))
 
-	write.table(united_regions, paste0("../results/methylKit/regionCounts_CGI_3min_", n, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
+	# write.table(united_regions, paste0("../results/methylKit/regionCounts_CGI_3min_", n, ".csv"), sep = ',', row.names = FALSE, quote = FALSE)
 
 
     #combine methylation calls into one

### 2024-09-11 code/WGBS_germ_regions.R
add all ensembl genes bed file for WGBS analysis

@@ -8,3 +8,6 @@ WIND <- 500
 germ <- read.csv(snakemake@input[[1]], sep = ",")
 geneTSSwindow(germ$ENSEMBL, 1, WIND)
 
+#all gene promoters using EpiLC gene expected counts Ensembl list
+all_genes <- read.csv(snakemake@input[[2]], sep = ",")
+geneTSSwindow(all_genes$gene_id, 1, WIND)

### 2024-09-11 code/serverscripts/WGBS/methyl_diff_allpromo_exEpiLC_WTvsKO_min3.R

@@ -1,0 +1,125 @@
+# 24.08.19 DNA methylation analysis of bismark bam files at germline CGI and promoter regions using methylKit
+# 5CKO vs WT exEpiLCs
+#https://www.bioconductor.org/packages/release/bioc/vignettes/methylKit/inst/doc/methylKit.html
+#https://nbis-workshop-epigenomics.readthedocs.io/en/latest/content/tutorials/methylationSeq/Seq_Tutorial.html
+#https://compgenomr.github.io/book/extracting-interesting-regions-differential-methylation-and-segmentation.html
+#run from command line by 'Rscript [nameofscript].R'
+
+#qvalue cut off
+qval <- 0.01
+
+
+library(methylKit)
+library(ggplot2)
+library(genomation)
+
+
+
+#regional summary analysis
+library(ChIPseeker)
+library(dplyr)
+
+#need granges as input for the bedfile
+#read in the bed file of CGIs for germline gene promoters
+
+
+
+#get list of WT files
+difffiles <- Sys.glob("../results/methylKit/*EpiLC_CpG.txt")
+difffiles
+#make as a list
+samp_list <- list()
+for (s in 1:length(difffiles)){
+    samp_list[[s]] <- difffiles[s]
+}
+  
+#get the sample names (IDs)   
+ID <- gsub("../results/methylKit/", "", difffiles)
+ID <- gsub("_CpG.txt", "", ID)
+#format as a list
+ID_list <- list()
+for (d in 1:length(ID)){
+	ID_list[[d]] <- ID[d]
+}
+
+print("sample ID_list")
+print(ID_list)
+
+
+#get the sample cell type (first two characters) - vector format
+cell <- c("KO", "KO", "WT", "WT")
+
+# treatment is 1 or 0, 1 for KO, 0 for WT
+treat <- c()
+for (g in 1:length(cell)) {
+treat[g] <- ifelse(cell[g] == "KO", 1, ifelse(cell[g] == "WT", 0, NA))
+}
+  
+  #read in the files
+myobj <- methRead(location = samp_list, sample.id = ID_list, assembly="mm10", treatment = treat,
+                   context="CpG", mincov = 3)
+#filter based on read coverage
+myobj <- filterByCoverage(myobj,lo.count=3,lo.perc=NULL,
+                    hi.count=NULL,hi.perc=99.9)
+  
+
+#get the differential methylation for the bedfile regions of interest
+#outname is the output file name
[…]

### 2024-09-11 code/WGBS_germ_regions.R
get TSS window of all ensembl genes

@@ -9,5 +9,6 @@ germ <- read.csv(snakemake@input[[1]], sep = ",")
 geneTSSwindow(germ$ENSEMBL, 1, WIND)
 
 #all gene promoters using EpiLC gene expected counts Ensembl list
-all_genes <- read.csv(snakemake@input[[2]], sep = ",")
-geneTSSwindow(all_genes$gene_id, 1, WIND)
+all_genes <- read.csv(snakemake@input[[2]], sep = "\t")
+# head(all_genes)
+geneTSSwindow(all_genes$gene_id, 2, WIND)

### 2024-09-11 code/WGBS_germ_CGI.R
list of CGI and non-CGI genes for homer

@@ -1,5 +1,5 @@
 
-### 24.04.26 compare binding locations of KDM5C ChIp-seq peaks and number of germline genes bound for CGI and CGI-free germline genes
+### 24.07.26 compare binding locations of KDM5C ChIp-seq peaks and number of germline genes bound for CGI and CGI-free germline genes
 
 
 # Visualize genomic locations using ChIPseeker
@@ -124,4 +124,10 @@ head(ck)
 
 write.table(ck, snakemake@output[[4]], row.names = FALSE, sep = ",")
 
-ggsave(snakemake@output[[5]], plot = dotplot(ck, showCategory = 10), width = 8, height = 8)
+ggsave(snakemake@output[[5]], plot = dotplot(ck, showCategory = 10), width = 4, height = 8)
+
+
+
+### Save the gene names for HOMER motif analysis
+write.table(subset(germ, Promo_CGI == "CGI")[,"SYMBOL"], snakemake@output[[6]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)
+write.table(subset(germ, Promo_CGI == "no")[,"SYMBOL"], snakemake@output[[7]], sep = "\t", col.names = FALSE, row.names = FALSE, quote = FALSE)

### 2024-09-11 code/serverscripts/HOMER/MotifEnrich_germCGIs.py

@@ -1,0 +1,30 @@
+# 24.07.19 HOMER to find motifs enriched in KDM5C bound vs unbound germline DEGs to see what is promoting their dysregulation independent of KDM5C
+
+import sys
+import os
+import glob
+
+#where HOMER is located
+HOMER = "../../HOMER/"
+
+#output directory
+OUT = "../results"
+
+#window of bp that you want to search aroudn the TSS
+WINDOW = "500"
+
+#input files of gene list
+#input file needs to be a .txt file with the gene IDs in the first column, and no quotes around them.
+GENES = glob.glob("../data/raw/*CGI_HOMER.txt")
+
+for i in GENES:
+	#clean the file name 
+	clean = i.replace("../data/raw/", "")
+	clean = clean.replace("_HOMER.txt", "")
+
+	#make a directory for the outputs for that sample because for some reason that's what homer does
+	OUTDIR = "{OUT}/{CLEAN}".format(OUT = OUT, CLEAN = clean)
+	os.system("mkdir {OUTDIR}".format(OUTDIR = OUTDIR))
+
+	#run the HOMER command for that gene list
+	os.system("nohup perl {HOMER}bin/findMotifs.pl {GENES} mouse {OUTDIR} -start -{WINDOW} -end {WINDOW} &".format(HOMER = HOMER, OUT = OUT, GENES = i, OUTDIR = OUTDIR, WINDOW = WINDOW))

### 2024-09-11 submission/manuscript.Rmd
clarify WGBS instructions
add cgi description

@@
 Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
-We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023] and found out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (TSS &#177; 500 bp) (Figure 6F).
+We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023], which classified CGIs as regions that have greater than 50% GC content, are larger than 200 base pairs, and the ratio of CG dinucleotides observed over the expected amount based on the number of Gs and Cs is greater than 0.6.
+Out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (TSS &#177; 500 bp) (Figure 6F).
 CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) when compared to CGI-free genes (Figure 6G).
@@
 		- KDM5C bound vs not
-		- DEG in EpiLC, brain, both, neither (separate columns?)
+		- Log2fc in EpiLC, brain (separate columns?)
+    - CGI vs non
 

### 2024-09-11 submission/manuscript.Rmd
shigeki edits

@@
 Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
+Here, we examine the roles of the NDD-associated lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
 We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
@@
 Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
-Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
+Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
 However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
@@
 Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
 However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.
@@
 It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017].
-We analyzed tissue-specific differentially expressed genes (DEGs) in our published mRNA-seq datasets[@vallianatosMutuallySuppressiveRoles2020] of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1).
+We systematically tested whether other tissue-specific genes are misexpressed in the brain with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
 
@@
 
-We found a large proportion of significantly upregulated genes within the _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
-For both the amygdala and hippocampus, the majority of tissue-enriched (DEGs) were testis genes (Figure 1A-C).
+We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C).
 Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
@@
 
-Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], when they are silenced in epiblast stem cells that will  differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020].
+Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], when germline genes are silenced in epiblast stem cells that will form the somatic tissues[@bardotMouseGastrulationCoordination2020].
 This developmental time point can be modeled _in vitro_ through differentiation of na&#239;ve embryonic stem cells (nESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020].
 While some germline-enriched genes are also expressed in nESCs and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
+Therefore, we tested if KDM5C was necessary for the initial silencing of germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.
 
 _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
-We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
+We then identified tissue-enriched DEGs in a RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D).
@@
 Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H).
-These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate and later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state.
-Altogether, _Kdm5c_-KO EpiLCs expressing key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes indicate germline gene misexpression parallels germline development during proper _Kdm5c_-KO differentiation.
+However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell state.
+Altogether, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes, indicating germline gene misexpression mirrors germline development during the progression of somatic development.
 
@@
 
-It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on male cells[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
 Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
@@
 Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
-While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were male-specific and female-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
+While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were sex-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
 We then compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
@@
 
-The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D-F).
+The majority of germline genes expressed in both sexes were more highly dysregulated in females compared to males (Figure 4D-F).
 This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression.
@@
 
-The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3,677 EpiLC only promoters) (Figure 5B).
+The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were bound by KDM5C only in EpiLCs (3,677 EpiLC only promoters) (Figure 5B).
 Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
-Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
-There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs.
+Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`)) (Figure 5C).
+There were no ontologies significantly enriched for PNC-specific KDM5C target genes.
 Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D).
-Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288, 3)`% of all germline genes) (Figure 5E).
-Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%) (Supplementary figure 1A-C).
-However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs.
+Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E).
+Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 22/63*100`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
+However, KDM5C did bind the promoter of 4 out of the 5 genes dysregulated in both the brain and EpiLCs.
 Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
@@
 Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-PRC1.6 members have been proposed to recruit KDM5C to germline gene promoters, given their association in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023] and that KDM5C itself does not contain sequence-specific binding domains[@iwaseXLinkedMentalRetardation2007].
-To elucidate how KDM5C is recruited to germline genes, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
-MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G).
+PRC1.6 members may recruit KDM5C to germline gene promoters, given their association with KDM5C in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023].
+We thus used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
+MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917), but not at germline genes unbound by KDM5C (Figure 5G).
 One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H).
@@
 Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
-Suprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
+Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
 Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters during early embryogenesis, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me.

### 2024-09-12 code/WGBS_volcano.R
draft volcano plot for all promoters

@@ -59,21 +59,16 @@ TSS_volcano <- function(n, TITLE, restab, maxy){
 
 	#restab <- methyl
 	#if the gene is tissue specific, the color key is the color for the tissue. Otherwise the gene will be gray
-	restab$keyvals <- ifelse(restab$qvalue > 0.01 & restab$Promo_CGI == "CGI", 'black',
-						ifelse(restab$Promo_CGI == "CGI" & restab$meth.diff < l2fcco & restab$meth.diff > -l2fcco, 'black',
-							ifelse(restab$qvalue <= 0.01 & restab$Promo_CGI == "CGI", 'red',
-								ifelse(restab$meth.diff >= l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", 'palevioletred1',
-                        			ifelse(restab$meth.diff <= -l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", 'lightseagreen',
-										"gray38")))))
+	restab$keyvals <- ifelse(restab$qvalue <= 0.01 & restab$Promo_CGI == "CGI", 'red',
+							ifelse(restab$meth.diff >= l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", 'lightseagreen',
+                        		ifelse(restab$meth.diff <= -l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", 'lightseagreen',
+									"gray38")))
 
 	#print(head(restab))
 
-	restab$names <- ifelse(restab$keyvals == 'red', 'promoter CGI', 
-                         ifelse(restab$keyvals == 'gray38', 'n.s. non-CGI',
-						 	ifelse(restab$keyvals == 'black', 'n.s. promoter CGI',
-								ifelse(restab$keyvals == 'lightseagreen', 'non-CGI hypomethylated',
-									ifelse(restab$keyvals == 'palevioletred1', 'non-CGI hypermethylated',
-                            			"uh oh")))))
+	restab$names <- ifelse(restab$keyvals == 'red', 'CGI promoter', 
+                         ifelse(restab$keyvals == 'gray38', 'n.s.',
+								ifelse(restab$keyvals == 'lightseagreen', 'non-CGI promoter', "uh oh")))
 
 
 
[…]

### 2024-09-12 submission/manuscript.Rmd
editing CGI results

@@
 
-## KDM5C is recruited to CpG islands within germline gene promoters to facilitate _de novo_ DNA methylation
+## KDM5C is recruited to CpG islands at germline promoters to facilitate _de novo_ DNA methylation
 
-In the early embryo, germline gene promoters are initially decorated with repressive histone modifications and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021].
-Our results above indicate KDM5C also acts at germline gene promoters during this time period.
-However, how KDM5C interacts with other germline gene silencing mechanisms is currently unclear.
-KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], yet KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
-Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required in the post-implantation embryo for long-term silencing of germline genes.
-In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
+Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
+Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes.
+However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023].
+To reconcile these observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
 
-Based on the above observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
-To test this hypothesis, we first characterized KDM5C's substrates (H3K4me2/3) at germline gene promoters in our published ChIP-seq datasets of male wild-type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
+To test this hypothesis, we first characterized KDM5C's expression as na&#239;ve ESCs differentiate into EpiLCs (Figure 6A).
+While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6B), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6C).
+
+We then characterized KDM5C's substrates at germline gene promoters with loss of _Kdm5c_ using ChIP-seq datasets of H3K4me2/3 in EpiLCs[@samantaActivationXistEvolutionarily2022] and the amygdal[@vallianatosMutuallySuppressiveRoles2020].
+Germline genes had a marked increase in H3K4me2 in EpiLCs (Figure 6D) and H3K4me3 in the _Kdm5c_-KO amygdala (Figure 6E) around the TSS in _Kdm5c_-KO samples.
+
+in our published ChIP-seq datasets of male wild-type and _Kdm5c_-KO amygdala and EpiLCs.
+
+_Kdm5c_-KO
+
+KDM5C's substrates (H3K4me2/3) at germline gene promoters in our published ChIP-seq datasets of male wild-type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
+
+Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me at germline gene promoters.
+
   In congruence with the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the TSS of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
-There was also a marked increase in H3K4me2 around the TSS of germline genes in _Kdm5c_-KO EpiLCs (Figure 6B).
-To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation (Figure 6C).
-While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6E).
-Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me at germline gene promoters.
+
+   To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation
 

### 2024-09-12 submission/manuscript.Rmd
shigeki edits

@@
 While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6B), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6C).
-
-We then characterized KDM5C's substrates at germline gene promoters with loss of _Kdm5c_ using ChIP-seq datasets of H3K4me2/3 in EpiLCs[@samantaActivationXistEvolutionarily2022] and the amygdal[@vallianatosMutuallySuppressiveRoles2020].
-Germline genes had a marked increase in H3K4me2 in EpiLCs (Figure 6D) and H3K4me3 in the _Kdm5c_-KO amygdala (Figure 6E) around the TSS in _Kdm5c_-KO samples.
-
-in our published ChIP-seq datasets of male wild-type and _Kdm5c_-KO amygdala and EpiLCs.
-
-_Kdm5c_-KO
-
-KDM5C's substrates (H3K4me2/3) at germline gene promoters in our published ChIP-seq datasets of male wild-type and _Kdm5c_-KO amygdala[@vallianatosMutuallySuppressiveRoles2020] and EpiLCs[@samantaActivationXistEvolutionarily2022].
-
-Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me at germline gene promoters.
-
-  In congruence with the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed aberrant accumulation of H3K4me3 around the TSS of germline genes in the _Kdm5c_-KO amygdala (Figure 6A).
-
-   To elucidate KDM5C's embyonic role, we then characterized KDM5C's mRNA and protein expression during male ESC to EpiLC differentiation
+We then characterized KDM5C's substrates (H3K4me2/3) at germline gene promoters with _Kdm5c_ loss using published ChIP-seq datasets[@samantaActivationXistEvolutionarily2022; @vallianatosMutuallySuppressiveRoles2020].
+_Kdm5c_-KO samples showed a marked increase in H3K4me2 in EpiLCs (Figure 6D) and H3K4me3 in the amygdala (Figure 6E) around the TSS of germline genes.
+Together, these data suggest KDM5C acts during the transition between ESCs and EpiLCs to remove H3K4me2/3 at germline gene promoters.
 
@@
 
+#get all promoters 5CKO vs WT WGBS information
+exEpi_WTvsKO_all <- read.csv("../results/WGBS/WGBS_restab_annotated_allgenes_WTvsKO.csv")
+
 ```
@@
 Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
-We first identified how many germline genes contained CGIs using the UCSC genome browser[@nassarUCSCGenomeBrowser2023], which classified CGIs as regions that have greater than 50% GC content, are larger than 200 base pairs, and the ratio of CG dinucleotides observed over the expected amount based on the number of Gs and Cs is greater than 0.6.
-Out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (TSS &#177; 500 bp) (Figure 6F).
-CGI-containing germline genes were enriched for meiotic gene ontologies, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) when compared to CGI-free genes (Figure 6G).
+We first examined how many of our germline-enriched genes had promoter CGIs (TSS &#177; 500 bp) using the UCSC genome browser[@nassarUCSCGenomeBrowser2023].
+Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F).
+CGI-containing germline genes had substantial enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) (Figure 6G).
 Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (FISHER'S XXXX), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline genes containing CGIs (Figure 6G).
 
-To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) (Figure 6H).
+To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and exEpiLCs (Figure 6H).
 We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
@@
 
-We then identified germline gene promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp) and found `r nrow(WTvsKO_hypo)` germline promoters were significantly hypomethylated with _Kdm5c_ loss (Figure 6J).
-Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated DMRs).
-Promoters that showed the most robust loss of CpGme (low q-values) harbored CGIs (Figure 6J).
+We then identified promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp).
+Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary figure 2).
+Hypomethylated promoters were significantly enriched for germline genes (`r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters) (Figure 6J).
 
-We then compared average percent CpGme at germline gene promtoers with or without CGIs.
-While the degree of CpGme was largely unchanged for non-CGI promoters, CGI promoters on average had a significant reduction in CpGme with KDM5C loss (Figure 6K) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test).
+Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated promoters).
+Promoters that showed the most robust loss of CpGme (lowest q-values) harbored CGIs (Figure 6J).
+CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6K) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test).
 Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
 Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
-Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters during early embryogenesis, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me.
+Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters.
+This suggests KDM5C's cataltyic activity is required for germline gene repression in EpiLCs, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me around the TSS.
 
@@
 Germline-enriched genes met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type somatic tissues[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted (Kit^W/Wv^) germline, for any sex or time point, does not exceed 20% of maximum expression in wild-type germline.
+We defined sperm and egg-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
+Genes that did not meet this threshold for either sex were classified as 'unbiased'.
 
@@
 nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF), as previously described [@samantaGeneratingPrimedPluripotent2020].
+
+## RT-qPCR
+XXX
+
+## Western Blot
+XXX
 
@@
 ## CpG island (CGI) analysis
-Locations of CpG islands were determined through the mm10 UCSC genome browser CpG island track.
-CGI coordinates were then annotated using ChIPseeker (v1.32.1) and filtered for promoters of germline genes (TSS &#177; 500).
+Locations of CpG islands were determined through the mm10 UCSC genome browser CpG island track[@nassarUCSCGenomeBrowser2023], which classified CGIs as regions that have greater than 50% GC content, are larger than 200 base pairs, and have a ratio of CG dinucleotides observed over the expected amount greater than 0.6.
+CGI genomic coordinates were then annotated using ChIPseeker (v1.32.1) and filtered for ones that lie within promoters of our germline-enriched genes (TSS &#177; 500).
 

### 2024-09-12 submission/manuscript.Rmd
shigeki edits

@@
 In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
 The contribution of KDM5C to this process remains unclear.
@@
 We then identified promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp).
-Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary figure 2).
-Hypomethylated promoters were significantly enriched for germline genes (`r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters) (Figure 6J).
+Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary figure 2A).
+Hypomethylated promoters were significantly enriched for germline genes (`r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters) (Supplementary figure 2B).
 
@@
 In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
-We first characterized tissue-enriched genes expressed within the _Kdm5c_-KO brain and identified substantial dysregulation of testis, liver, muscle, and ovary-enriched genes.
+We first characterized tissue-enriched genes expressed within the _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes.
 Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells.
@@
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, including its interaction with known silencing mechanisms, we unveiled repressive mechansisms governing distinct classes of germline gene in somatic lineages.
+By systematically characterizing KDM5C's role in germline gene repression, we unveiled derepressive mechansisms governing distinct classes of germline gene in somatic lineages.
 Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
@@
 The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022].
-Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
+Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+KDM5C's Y-chromsome homolog, KDM5D, exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and it is currently unknown to regulate germline gene expression.
 Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.
 
+We found KDM5C is largely dispensible for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
 It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
-For example, while _Dazl_ is primarily known for regulating the translation of germline-specific mRNAs, it is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015].
-KDM5C was thought to promote the 2-cell-to-ESC transition, given _Dazl_ and _Zscan4c_ are de-repressed in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023].
 Although expressed in na&#239;ve ESCs, "self-renewal" germline genes like _Dazl_ are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
 We found that while _Kdm5c_-KO EpiLCs also express _Dazl_, they did not express 2-cell-specific genes.
-These data suggest the 2-cell-like state observed in _Kdm5c_-KO ESCs likely reflects KDM5C's primary role in germline gene repression.
+These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression.
 Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs), rather than de-differentiating into 2-cell-like cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
-Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which is known to repress EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
+Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
 In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated.
-Thus, ectopic germline gene expression occurs in parallel with proper _Kdm5c_-KO differentiation and somatic development.
+Thus, ectopic germline gene expression occurs along with proper somatic development in _Kdm5c_-KO animals.
 
-While promoter CGIs are typically resistant to CpGme[@longProtectionCpGIslands2016], CGIs at germline promoters are hightly methylated in somatic cells[@weberDistributionSilencingPotential2007].
+Our work provides novel insight into the cross-talk between H3K4me and CpGme.
 In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g.
-_D1Pas1_ and _Naa11_, strongly impaired CGI methylation, which resulted in their long-lasting de-repression into adulthood.
+_D1Pas1_ and _Naa11_, strongly impaired CGI methylation, and resulted in their long-lasting de-repression into adulthood.
 Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 primarily do not colocalize with CpGme[@weberDistributionSilencingPotential2007] and can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
@@
 _Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-Together, this suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme.
-Furthermore, these results indicate the requirement for catalytic activity for the same class of genes can change depending upon the locus and developmental stage.
+Together, this suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007].
+Furthermore, these results indicate the requirement for catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
 
-By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct repressive mechanisms governing early versus late-stage germline developmental programs.
-In EpiLCs, KDM5C was highly enriched at germline promoters containing CGIs, many of which had roles in early germ cell formation and meiosis.
+By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct derepressive mechanisms governing early versus late-stage germline developmental programs.
+Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs.
 However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells.
 CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_ and _Dazl_ turn on downstream germline programs that can loosely mimic germ cell development, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
-Therefore, we propose KDM5C is recruited to germline genes that shape germ cell formation via their promoter CGIs to act as break against runaway activation of germline-specific programs.
+These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
 Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation and acts as break against runaway activation of germline-specific programs.
@@
 Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
-Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012], indicating their dysregulation may promote genome instability and cellular de-differentiation.
-Intriguingly, some models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012].
-Like KDM5C, the chromatin regulators underlying these conditions - DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2) - primarily silence gene expression.
+Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
+Intriguingly, mouse and human models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
 Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary.

### 2024-09-13 code/WGBS_volcano.R
update volcano aes and add GO

@@ -42,10 +42,14 @@ anno_TSS <- function(TSSdf){
 
 ## TSS WT ESC vs EpiLCs
 ESCvsEpiLC_TSS <- anno_TSS(read.csv(snakemake@input[[2]], sep = ","))
+ESCvsEpiLC_TSS <- subset(ESCvsEpiLC_TSS, select = -c(strand))
+
 write.table(ESCvsEpiLC_TSS, snakemake@output[[1]], sep = ",", row.names = FALSE)
 
 ## TSS WT vs KO
 WTvsKO_TSS <- anno_TSS(read.csv(snakemake@input[[4]], sep = ","))
+WTvsKO_TSS <- subset(WTvsKO_TSS, select = -c(strand))
+
 write.table(WTvsKO_TSS, snakemake@output[[5]], sep = ",", row.names = FALSE)
 
 
@@ -342,6 +346,8 @@ print(head(WTvsKO_all_plot))
 
 names(WTvsKO_all_plot)[names(WTvsKO_all_plot) == "gene_name"] <- "SYMBOL"
 
+
+WTvsKO_all_plot <- subset(WTvsKO_all_plot, select = -c(strand))
 write.table(WTvsKO_all_plot, snakemake@output[[9]], sep = ",", row.names = FALSE)
 
 #TSS_volcano(10, "WT vs KO exEpiLC - all promoters", WTvsKO_all_plot, 200)
@@ -353,17 +359,17 @@ TSS_germ <- function(n, TITLE, restab, maxy){
 
 	#restab <- methyl
 	#if the gene is tissue specific, the color key is the color for the tissue. Otherwise the gene will be gray
-	restab$keyvals <- ifelse(restab$qvalue <= 0.01 & restab$ENSEMBL %in% KDM5C_binding$ENSEMBL & restab$meth.diff >= l2fcco, '#2fa10d',
-						ifelse(restab$qvalue <= 0.01 & restab$ENSEMBL %in% KDM5C_binding$ENSEMBL & restab$meth.diff <= l2fcco, '#2fa10d',
-							ifelse(restab$meth.diff >= l2fcco & restab$qvalue <= 0.01 & !(restab$ENSEMBL %in% KDM5C_binding$ENSEMBL), 'lightseagreen',
-                        		ifelse(restab$meth.diff <= -l2fcco & restab$qvalue <= 0.01 & !(restab$ENSEMBL %in% KDM5C_binding$ENSEMBL), 'lightseagreen',
[…]

### 2024-09-13 code/WGBS_volcano.R
fix volcano colors

@@ -63,16 +63,17 @@ TSS_volcano <- function(n, TITLE, restab, maxy){
 
 	#restab <- methyl
 	#if the gene is tissue specific, the color key is the color for the tissue. Otherwise the gene will be gray
-	restab$keyvals <- ifelse(restab$qvalue <= 0.01 & restab$Promo_CGI == "CGI", 'red',
-							ifelse(restab$meth.diff >= l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", 'lightseagreen',
-                        		ifelse(restab$meth.diff <= -l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", 'lightseagreen',
-									"gray38")))
+	restab$keyvals <- ifelse(restab$qvalue <= 0.01 & restab$Promo_CGI == "CGI" & restab$meth.diff >= l2fcco, 'red',
+						ifelse(restab$qvalue <= 0.01 & restab$Promo_CGI == "CGI" & restab$meth.diff <= -l2fcco, 'red',
+							ifelse(restab$meth.diff >= l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", '#86dbd7',
+                        		ifelse(restab$meth.diff <= -l2fcco & restab$qvalue <= 0.01 & restab$Promo_CGI == "no", '#86dbd7',
+									"gray38"))))
 
 	#print(head(restab))
 
 	restab$names <- ifelse(restab$keyvals == 'red', 'CGI promoter', 
                          ifelse(restab$keyvals == 'gray38', 'n.s.',
-								ifelse(restab$keyvals == 'lightseagreen', 'non-CGI promoter', "uh oh")))
+								ifelse(restab$keyvals == '#86dbd7', 'non-CGI promoter', "uh oh")))
 
 
 

### 2024-09-13 code/Germ_CGI_stage.R
germ cell stages CGI

@@ -1,0 +1,47 @@
+#24.09.13 Germline stages of development for CGI genes
+
+#list of germline genes
+germ <- read.csv(snakemake@input[[1]], sep = ",")
+
+
+#expression of genes in spermatogenesis stages
+greengerm <- read.csv(snakemake@input[[2]], sep = ",")
+
+
+
+#plotting df columns:
+	#germ cell stage
+	#transient or silent
+	#average expression (for all transient or silent)
+green_plot <- data.frame()
+expression <- c("CGI", "no")
+for(i in expression){
+	symbols <- subset(germ, Promo_CGI == i)[,"SYMBOL"]
+
+	#for every symbol, get the expression value and add it to the dataframe
+	for(k in symbols){
+		expr <- subset(greengerm, greengerm$SYMBOL == k)
+	
+		germexpr <- c(t(expr[1,2:ncol(expr)])) #values of the expression
+		
+		#get the name of the germ cell stage and the means of expression, skipping the first column (gene symbols)
+		df <- data.frame(Stage = colnames(greengerm)[2:ncol(greengerm)], GermExpr = germexpr, CGI_status = rep(i, ncol(greengerm)-1))
+
+		green_plot <- rbind(green_plot, df)
+
+	}
+		
+}
+
+print(head(green_plot))
+
+library('ggpubr')
+
+my_comparisons <- expression
+p <- ggboxplot(green_plot, "Stage", "GermExpr", fill = "CGI_status",  palette = c("#f93a0b", "#ff8a7a"), title = "Average expression in germ cell stages", ylab = "log(Avg of Normalized Expression + 1)", xlab = "Stage of Sperm Development") +
+	stat_compare_means(aes(label=..p.signif.., group=CGI_status), method="wilcox.test", label.y = 6.5, size = 8) +
+	font("xy.text", size = 24) + font("title", size = 35, face = "bold") + font("xlab", size = 24) + font("ylab", size = 24)
+
+p <- ggpar(p, legend = "top", legend.title = "CGI at promoter")
+
+ggsave(snakemake@output[[1]], p, width = 15, height = 10)

### 2024-09-13 code/Germ_CGI_stage.R
heat maps of stages expression

@@ -45,3 +45,39 @@ p <- ggboxplot(green_plot, "Stage", "GermExpr", fill = "CGI_status",  palette = 
 p <- ggpar(p, legend = "top", legend.title = "CGI at promoter")
 
 ggsave(snakemake@output[[1]], p, width = 15, height = 10)
+
+
+## heatmap of expression across stages
+	#multiple heatmaps of each developmental stage stacked on top 
+#make a hclust matrix for each stage, have the genes expression in CGI vs no. 
+
+#for every germ cell stage, get the germ cell expression and for CGI and CGI no genes
+
+#make 2 heatmaps (CGI vs no), split by germ cell stage. Use the same scale for both so they can be compared to each other
+
+#each column is a germ cell stage, each row is a gene, values are the expression
+library(ComplexHeatmap)
+library(circlize)
+
+expression <- c("CGI", "no")
+#setting the range of colors
+heat_color = colorRamp2(range(subset(greengerm, select = c(-SYMBOL))), hcl_palette = "Reds", reverse = TRUE)
+
+
+for(i in 1:length(expression)){
+	symbols <- subset(germ, Promo_CGI == expression[i])[,"SYMBOL"]
+
+	df <- subset(greengerm, greengerm$SYMBOL %in% symbols)
+	row.names(df) <- df$SYMBOL
+	plotmatrix <- subset(df, select = c(-SYMBOL))
+	p <- Heatmap(plotmatrix, show_row_names = FALSE, column_title = paste("Germline genes with", expression[i]), cluster_columns = FALSE, heatmap_legend_param = list(title = "log(Avg of Normalized Expression + 1)"), col = heat_color)
+	
+
+	pdf(file = snakemake@output[[i+1]],   # The directory you want to save the file in
+    	width = 10, # The width of the plot in inches
+    	height = 10) # The height of the plot in inches
+		draw(p)
+	dev.off()
+		
+}
+

### 2024-09-14 code/WGBS_germ_CGI.R
rearrange wgbs figure

@@ -95,7 +95,7 @@ library("ggpubr")
 q <- ggbarplot(plotdf, "Kdm5c_binding", "Percent", fill = "CpG_island", color = "CpG_island", palette = c("no" = "aquamarine4", "CGI" = "aquamarine3"),
 		title = "CpG islands at germline gene promoters", label = TRUE, lab.col = "black", lab.vjust = 1, xlab = " ", ylab = "% of genes", orientation = "vert") 
 
-ggsave(snakemake@output[[3]], q, width = 6, height = 4)
+ggsave(snakemake@output[[3]], q, width = 5, height = 4)
 
 
 #gene ontology of CGI germline genes vs non
@@ -124,7 +124,7 @@ head(ck)
 
 write.table(ck, snakemake@output[[4]], row.names = FALSE, sep = ",")
 
-ggsave(snakemake@output[[5]], plot = dotplot(ck, showCategory = 10), width = 4, height = 8)
+ggsave(snakemake@output[[5]], plot = dotplot(ck, showCategory = 10), width = 5, height = 9)
 
 
 

### 2024-09-14 submission/manuscript.Rmd
rearrange wgbs figure
CGI GO info and fisher

@@
 
+#Gene ontology CGI
+GO_CGI <- read.csv("../results/germline_CGI_GO.csv")
+
+#Fisher's exact test KDM5C binding at CGI promoters
+noCGI <- c(bound = nrow(germgenes_CGI[germgenes_CGI$KDM5C_binding == "Bound" & germgenes_CGI$Promo_CGI == "no",]), unbound = nrow(germgenes_CGI[germgenes_CGI$KDM5C_binding == "Unbound" & germgenes_CGI$Promo_CGI == "no",]))
+yesCGI <- c(bound = nrow(KDM5C_bound_CGI), unbound = nrow(germgenes_CGI[germgenes_CGI$KDM5C_binding == "Unbound" & germgenes_CGI$Promo_CGI == "CGI",]))
+CGI_5C <- data.frame(row.names = c("bound", "unbound"), yesCGI, noCGI)
+
+CGI_5C_fish <- fisher.test(CGI_5C)
+
 ```
@@
 Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F).
-CGI-containing germline genes had substantial enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:XXXX, p.adj) and meiosis I (GO:XXXX, p.adj) (Figure 6G).
-Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (FISHER'S XXXX), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline genes containing CGIs (Figure 6G).
+CGI-containing germline genes had substantial enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G).
+Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes, with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test) of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).
 
@@
 Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary figure 2A).
-Hypomethylated promoters were significantly enriched for germline genes (`r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters) (Supplementary figure 2B).
+Hypomethylated promoters were significantly enriched for germline gene ontologies, such as meiotic nuclear division (GO:0140013, p.adjust = 0.012)(Supplementary figure 2B), with `r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonging to germline genes.
 
@@
   \centering 
-  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
-  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs.
-\textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs.
-Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only).
-\textbf{D.} Heatmap of germline DEGs shared between male and female mutants.
-Color is the average log 2 fold change from sex-matched wild-type.
-\textbf{E.} Number of genes within each cluster from D.
-Clusters with higher expression in females compared to males (XX-biased) highlighted in orange.
-\textbf{F.} Example average bigwigs of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants.
-\textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome.
-Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant.
-X chromosome highlighted in black.
-\textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes.
-\textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants.
-\textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) dysregulated in both male and female \textit{Kdm5c} mutants.}
-  \label{figurelabel}
-\end{figure}
-
-\begin{figure}
-  \centering 
   \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}

### 2024-09-14 submission/manuscript.Rmd
talk about germ stages

@@
 
-Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012], reaching peak methylation levels when differentiated into EpiLCs for 96 hours (extended EpiLCs, exEpiLCs)[@mochizukiRepressionGermlineGenes2021].
+Germline genes accumulate CpG methylation (CpGme) at CpG islands (CGIs) during the transition from na&#239;ve to primed pluripotency[@meissnerGenomescaleDNAMethylation2008; @borgelTargetsDynamicsPromoter2010; @hackettPromoterDNAMethylation2012].
 We first examined how many of our germline-enriched genes had promoter CGIs (TSS &#177; 500 bp) using the UCSC genome browser[@nassarUCSCGenomeBrowser2023].
@@
 CGI-containing germline genes had substantial enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G).
-Although a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes, with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test) of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).
+Germline genes with promoter CGIs were more highly expressed than CGI-free genes across spermatogenesis stages, with highest expression in meiotic spermatocytes (Figure 6H).
+Contrastingly, CGI-free genes only displaying substantial expression in post-meiotic round spermatids.
+Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes, with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test) of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).
 
-To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and exEpiLCs (Figure 6H).
+To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96-hour extend EpiLCs (exEpiLCs), when germline genes reach peak methylation level[@mochizukiRepressionGermlineGenes2021] (Figure 6I).
 We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
-In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6I).
+In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6J).
 
@@
 Hypomethylated promoters were significantly enriched for germline gene ontologies, such as meiotic nuclear division (GO:0140013, p.adjust = 0.012)(Supplementary figure 2B), with `r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonging to germline genes.
-
 Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated promoters).
-Promoters that showed the most robust loss of CpGme (lowest q-values) harbored CGIs (Figure 6J).
-CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6K) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test).
-Significantly hypomethylated promoters included genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6L).
-Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6M).
+Promoters that showed the most robust loss of CpGme (lowest q-values) harbored CGIs (Figure 6K).
+CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6L) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test).
+Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6M).
+Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N).
 Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters.

### 2024-09-14 code/Germ_CGI_stage.R
fix CGI figs

@@ -38,13 +38,13 @@ print(head(green_plot))
 library('ggpubr')
 
 my_comparisons <- expression
-p <- ggboxplot(green_plot, "Stage", "GermExpr", fill = "CGI_status",  palette = c("#f93a0b", "#ff8a7a"), title = "Average expression in germ cell stages", ylab = "log(Avg of Normalized Expression + 1)", xlab = "Stage of Sperm Development") +
+p <- ggboxplot(green_plot, "Stage", "GermExpr", fill = "CGI_status",  palette = c("#f93a0b", "#ff8a7a"), title = "Average expression in germ cell stages", ylab = "log(Avg of Normalized Expression + 1)", xlab = "Stage of Spermatogenesis") +
 	stat_compare_means(aes(label=..p.signif.., group=CGI_status), method="wilcox.test", label.y = 6.5, size = 8) +
 	font("xy.text", size = 24) + font("title", size = 35, face = "bold") + font("xlab", size = 24) + font("ylab", size = 24)
 
 p <- ggpar(p, legend = "top", legend.title = "CGI at promoter")
 
-ggsave(snakemake@output[[1]], p, width = 15, height = 10)
+ggsave(snakemake@output[[1]], p, width = 15, height = 8)
 
 
 ## heatmap of expression across stages

### 2024-09-15 submission/manuscript.Rmd
define 5CKO in main body

@@
 
-We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO (5CKO) brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
 For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C).
@@
 Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs.
-[@samantaActivationXistEvolutionarily2022].
+We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO, XY 5CKO), female homozygous knockout (XX _Kdm5c_-KO, XX 5CKO), and female heterozygous knockout (XX _Kdm5c_-HET, XX 5CHET) EpiLCs [@samantaActivationXistEvolutionarily2022].
 

### 2024-09-15 submission/manuscript.Rmd
add rt-qpcr and western to methods

@@
 
-## RT-qPCR
-XXX
+## Real time quantitative PCR (RT-qPCR)
+nESCs were differentiated into EpiLCs as described above.
+Cells were lysed with Tri-reagent BD (Sigma #T3809) at 0, 24, and 48 hours of differentiation.
+RNA was phase separated with 0.1 uL/uL 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516).
+For each sample, 2 ug of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT.
+Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP using Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).
 
 ## Western Blot
-XXX
+Total protein was extracted during nESC to EpiLC differentiation at 0, 24, and 48 hours by sonicating cells at 20% amplitude for 15 seconds in 2X SDS sample buffer, then boiling at 100°C for 10 minutes.
+Proteins were separated on a 7.5% SDS page gel, transferred overnight onto a fluorescent membrane, blotted for rabbit anti-KDM5C (in house, 1:500) and mouse anti-DAXX (Santa Cruz #(H-7): sc-8043, 1:500) imaged using the LiCor Odyssey CLx system.
+Band intensitity was quantified using ImageJ.
 
-## RNA sequencing (RNA-seq)
+## RNA sequencing (RNA-seq) data analysis
 After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads.
@@
 
-## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq)
+## Chromatin immunoprecipitation followed by DNA sequecing (ChIP-seq) data analysis
 ChIP-seq reads were aligned to mm10 using Bowtie1 (v1.1.2) allowing up to two mismatches.

### 2024-09-15 submission/manuscript.Rmd
edit figure legends

@@
 For each sample, 2 ug of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT.
-Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP using Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).
+Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP with the Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).
 
@@
 gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
+All samples had greater than 99% bisulfite conversion rates.
 Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark (v0.22.1).
@@
 Right - Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal).
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - UCSC browser view of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal) (Average, n = 4).
 Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
-\textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
+\textbf{D.} Left - UCSC browser view of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}.
 Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red.
-\textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
+\textbf{E.} Left - UCSC browser view of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}.
 Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
@@
 Germline cell types are highlighted in green, somatic cell types in black.
-\textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and germline-depleted mice and in adult mouse non-gonadal tissues (Li et al 2017).}
+\textbf{D.} Sankey diagram of mouse genes filtered for germline enrichment based on their expression in wild-type and W/Wv mice and in adult mouse non-gonadal tissues (Li et al 2017).}
   \label{figurelabel}
@@
 Bottom - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs).
-\textbf{B.} Representative images of wild-type (WT) and \textit{Kdm5c}-KO cells during ESC to EpiLC differentiation.
+\textbf{B.} Representative images of male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) cells during ESC to EpiLC differentiation.
 Brightfield images taken at 20X.
@@
 \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.
-\textbf{F.} Average bigwigs of an example germline gene, \textit{D1Pas1}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple), amygdala (middle, red), and hippocampus (bottom, blue).
+\textbf{F.} UCSC browser view of an example germline gene, \textit{D1Pas1}, that is dysregulated \textit{Kdm5c}-KO EpiLCs (top, purple.
+Average, n = 3), amygdala (middle, red.
+Average, n = 4), and hippocampus (bottom, blue.
+Average, n = 4).
 \textbf{G.} enrichPlot gene ontology analysis comparing enriched biological processes for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs.
@@
 p-values for Welch's t-test.
-Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs.}
+Bottom - UCSC browser view of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs (Average, n = 3).}
   \label{figurelabel}
@@
 Clusters with higher expression in females compared to males (XX-biased) highlighted in orange.
-\textbf{F.} Example average bigwigs of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants.
+\textbf{F.}  UCSC browser view of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants (Average, n = 3).
 \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome.
@@
 \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right).
-\textbf{E.} Eulerr of number of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple).
-\textbf{F.} Example KDM5C ChIP-seq bigwigs of KDM5C binding at the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs.
-{G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs.
+\textbf{E.} Eulerr overlap of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple).
+\textbf{F.} Example KDM5C ChIP-seq signal around the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs.
+\textbf{G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs.
 \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence.
@@
   \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
-  \caption[KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation]{\textbf{KDM5C’s catalytic activity promotes long-term silencing of germline genes via DNA methylation.} \textbf{A.} Left - Bigwigs of representative histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq peaks at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
-Right - Average H3K4me3 at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala.
-\textbf{B.} Left - Bigwigs of representative histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq peaks at representative germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs.
-Right - Average H3K4me2 at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs.
-\textbf{C.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation protocol and collection time points for RNA and protein.
-\textbf{D.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression, calculated in comparision to TBP expression (2\textsuperscript{-deltaCT}).
+  \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points.
+\textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs.Expression calculated in comparision to TBP mRNA expression (2\textsuperscript{-deltaCT}).
 * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test.
-\textbf{E.} KDM5C protein expression normalized to DAXX.
+\textbf{C.} KDM5C protein expression normalized to DAXX.
 Quantified intensity using ImageJ (artificial units - au).
 Right - representative lanes of Western blot for KDM5C and DAXX.
-* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters, based on UCSC annotation.
-Comparing all germline-enriched genes, KDM5C-bound germline genes, or KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of CGI-promoter versus non-CGI promoter germline genes.
-\textbf{H.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{I.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation at germline gene promoters (TSS $\pm$ 500) in wild-type ESCs versus exEpiLCs.
-Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink \textbf{J.} Volcano plot of WGBS of wild-type versus \textit{Kdm5c}-KO exEpiLCs.
-Promoter CGI genes highlighted in red, hypermethylated genes lackig a promoter CGI in pink, hypomethylated genes lackig a promoter CGI in blue.
-\textbf{K.} Example UCSC browser shots of germline genes of CpG methylation (CpGme) in wild-type and \textit{Kdm5c}-KO ESCs and exEpiLCs.
-\textbf{L.} Histogram of average percent CpGme at the promoter for germline genes with or without promoter CGIs.
-Wild-type in navy and \textit{Kdm5c}-KO in light blue.
-Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.}
+* p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test.
+\textbf{D.} Top - Representative UCSC browser view of histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq signal at two germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs.
+Bottom - Average H3K4me2 signal at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs.
+\textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
+Bottom - Average H3K4me3 signal at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala.
+\textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation.
+Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter.
+\textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018.
+SPG - spermatogonia, Pre-Lep - preleptotene spermatocytes, S-Cytes - meiotic spermatocytes, S-Tids - post-meiotic haploid round spermatids, E.Sp - elongating spermatids.
+Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001.
+\textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs.
+Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters.
+Promoters with CGIs in red, CGI-free promoters in light blue.
+\textbf{L.} Histogram of average percent CpGme at the promoter of germline genes with or without CGIs.
+Wild-type in navy and \textit{Kdm5c}-KO (KO) in light blue.
+Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.
+\textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs}
   \label{figurelabel}

### 2024-09-15 submission/manuscript.Rmd
add citations to methods

@@
 We removed “black-listed” genomic regions that often give aberrant signals.
-Common peak sets were obtained in R via DiffBind (v3.6.5).
+Common peak sets were obtained in R via DiffBind[estrogen@ross-innesDifferentialOestrogenReceptor2012] (v3.6.5).
 In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (v2.25.0).
-Peak proximity to genome annotations was determined by ChIPSeeker (v1.32.1).
+Peak proximity to genome annotations was determined by ChIPSeeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1).
 Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
@@
 Locations of CpG islands were determined through the mm10 UCSC genome browser CpG island track[@nassarUCSCGenomeBrowser2023], which classified CGIs as regions that have greater than 50% GC content, are larger than 200 base pairs, and have a ratio of CG dinucleotides observed over the expected amount greater than 0.6.
-CGI genomic coordinates were then annotated using ChIPseeker (v1.32.1) and filtered for ones that lie within promoters of our germline-enriched genes (TSS &#177; 500).
+CGI genomic coordinates were then annotated using ChIPseeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1) and filtered for ones that lie within promoters of our germline-enriched genes (TSS &#177; 500).
 
@@
 All samples had greater than 99% bisulfite conversion rates.
-Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark (v0.22.1).
-Analysis of differential methylation at germline gene promoters was perfomed using methylKit (v1.28.0) with a minimum coverage of 3 paired reads, a percentage cut-off of 25%, and q-value of 0.01.
-Average percentage methylation at germline gene promoters was determined via methylKit (v1.28.0).
+Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark[@kruegerBismarkFlexibleAligner2011] (v0.22.1).
+Analysis of differential methylation at germline gene promoters was perfomed using methylKit[@akalinMethylKitComprehensivePackage2012] (v1.28.0) with a minimum coverage of 3 paired reads, a percentage cut-off of 25%, and q-value of 0.01.
+methylKit was also used to calculate average percentage methylation at germline gene promoters.
 Methylation bedgraph tracks were generated via Bismark and visualized using the UCSC genome browser.
@@
 ### Data analysis
-Scripts used to generate the results, tables, and figures of this study are available via a GitHub repository: XXX
+Scripts used to generate the results, tables, and figures of this study are available via the GitHub repository: XXX
 

### 2024-09-15 submission/manuscript.Rmd
fix formatting on fig6 caption

@@
 \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue).
-\textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
+\textbf{C.} Gene ontology (GO) comparison of genes with KDM5C bound to their promoter in EpiLCs and PNCs.
 Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared).
@@
   \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points.
-\textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs.Expression calculated in comparision to TBP mRNA expression (2\textsuperscript{-deltaCT}).
+\textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs.
+Expression calculated in comparison to TBP mRNA expression (2\textsuperscript{-deltaCT}).
 * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test.
@@
 \textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala.
-Bottom - Average H3K4me3 signal at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala.
+Bottom - Average H3K4me3 signal at the transcription start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala.
 \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation.
-Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter.
+Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes.
+(Legend continued on next page.)}
+  \label{figurelabel}
+\end{figure}
+\newpage
+\addtocounter{figure}{-1}
+\begin{figure} [t!]
+  \caption[]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} (Legend continued.) \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter.
 \textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018.
@@
 Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001.
-\textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs.
-Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters.
+\textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation.
+\textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comparing CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs.
+Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25\%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters.
 Promoters with CGIs in red, CGI-free promoters in light blue.
@@
 Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.
-\textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs}
-  \label{figurelabel}
+\textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs.}
 \end{figure}

### 2024-09-16 submission/manuscript.Rmd
adjust labels and figure legend

@@
 Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.
-\textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs.}
+\textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI with number of CpGs, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs, and KDM5C ChIP-seq in EpiLCs.}
 \end{figure}

### 2024-09-16 submission/manuscript.Rmd
order wgbs output

--- 2024-09-15 submission/manuscript.Rmd
@@
 Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test.
-\textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs.}
+\textbf{M.} UCSC browser view of germline genes, showing UCSC annotated CGI with number of CpGs, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs, and KDM5C ChIP-seq signal in EpiLCs.}
 \end{figure}

### 2024-09-16 code/WGBS_volcano.R

@@ -349,6 +349,7 @@ names(WTvsKO_all_plot)[names(WTvsKO_all_plot) == "gene_name"] <- "SYMBOL"
 
 
 WTvsKO_all_plot <- subset(WTvsKO_all_plot, select = -c(strand))
+WTvsKO_all_plot <- WTvsKO_all_plot[order(WTvsKO_all_plot$meth.diff), ]
 write.table(WTvsKO_all_plot, snakemake@output[[9]], sep = ",", row.names = FALSE)
 
 #TSS_volcano(10, "WT vs KO exEpiLC - all promoters", WTvsKO_all_plot, 200)

### 2024-09-16 submission/supplement.Rmd
supplementary fig 2

@@
 \end{figure}
+
+\newpage
+\begin{figure}
+  \centering
+  \includegraphics{../submission/compiled_figs/Supplement_WGBS.pdf}
+  \caption[Loss of KDM5C impairs CpG methylation of germline gene promoters]{\textbf{[Loss of KDM5C impairs CpG methylation of germline gene promoters.} \textbf{A.} Volcano plot of whole genome bisulfite sequencing (WGBS) for all gene promoters in wild-type (WT) versus \textit{Kdm5c}-KO extended EpiLCs (exEpiLCs).
+Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25\%).
+Germline promoters highlighted in green, non-germline promoters in light blue, non-significant promoters in gray.
+\textbf{B.} enrichPlot gene ontology of all promoters significantly hypomethylated in \textit{Kdm5c}-KO exEpiLCs.}
+  \label{figurelabel}
+\end{figure}

### 2024-09-16 submission/manuscript.Rmd
abstract edits

@@
 
-Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
+Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms.
 Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we examine the roles of the NDD-associated lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity.
-We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
-Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary.
-Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
+Here, we examine the role of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) mutated in Claes-Jensen X-linked intellectual disability, in tissue identity.
+We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary.
+Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
 Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
@@
 However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Thus, distinct suppressive mechanisms are recruited to different germline gene classes and ectopic germline transcriptional programs can mirror germ cell development within somatic tissues.
+Thus, distinct mechanisms govern the misexpression of germline gene classes, including activation by ectopic germline programs that mirror germ cell development within somatic tissues.
 
@@
 Briefly, ESCs were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
-They were then transitioned into ground-state, "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax, Anti-Anti, N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
+They were then transitioned into ground-state, "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
 Both pESC and nESC media were supplemented with 3 $\mu$M GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), 1 $\mu$M MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and 1,000 units/mL leukemia inhibitory factor (LIF, Millipore#ESG1107).

### 2024-09-16 submission/manuscript.Rmd
intro edits

@@
 The contribution of KDM5C to this process remains unclear.
-Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes.
+Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated germline-enriched gene list.
 Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g.
@@
 To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
+We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
 Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

### 2024-09-16 submission/manuscript.Rmd
res 1 edits

@@
 It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We systematically tested whether other tissue-specific genes are misexpressed in the brain with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
+We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
 
@@
 
-We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO (5CKO) brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
-For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C).
+We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B).
 Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
-One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
+An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
 
@@
 
-Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
+Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
 _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

### 2024-09-16 submission/manuscript.Rmd
res 2 edits

@@
 spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C).
-Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.
+Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.
 
-As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive gene list.
+As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide, due to the lack of a comprehensive gene list.
 We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013].

### 2024-09-16 submission/manuscript.Rmd
res 3 edits

@@
 The majority of germline DEGs were unique to either EpiLCs or the brain, with only _D1Pas1_ and _Cyct_ shared across all tissue/cell types (Figure 3E-F).
-EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+EpiLC germline DEGs had particularly high enrichment of meiosis-related gene ontologies when compared to the brain (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
 While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

### 2024-09-16 submission/manuscript.Rmd
res 4 edits

@@
 
-Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
+In EpiLCs, homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
 While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were sex-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
-We then compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
-Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion).
-Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).
+We then compared the known functions of germline genes dysregulated only in females (XX only - unique to XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
+Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion).
+Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 - meiotic nuclear division) and egg-specific functions (GO:0007292 - female gamete generation).
 
@@
 Although many germline genes have shared functions in the male and female germline, some have unique or sex-biased expression.
-Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females primarily expressed egg genes.
+Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females would primarily express egg genes.
 To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H).
 We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes.
-We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J).
+This criteria yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes.
+We found regardless of sex, egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J).
 Furthermore, germline genes dysregulated exclusively in either male or female mutants were also not biased towards their corresponding germ cell sex (Figure 4I).

### 2024-09-16 submission/manuscript.Rmd
begin discussion edits

@@
 
-Although many germline genes have shared functions in the male and female germline, some have unique or sex-biased expression.
+Although many germline genes have shared functions in the male and female germline, e.g.
+PGC formation,  meiosis, and genome defense, some have unique or sex-biased expression.
 Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females would primarily express egg genes.
@@
 In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity.
-We first characterized tissue-enriched genes expressed within the _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes.
+We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes.
 Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells.
@@
 We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
-Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation via ectopic RFX transcription factors.
+Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation through ectopic RFX transcription factors.
 Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands.
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, we unveiled derepressive mechansisms governing distinct classes of germline gene in somatic lineages.
+By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression distinct germline gene classes within somatic lineages.
 Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
 
-Although eggs and sperm employ the same transcriptional programs for shared functions, e.g.
-PGC formation,  meiosis, and genome defense, some germline genes are sex specific.
-We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex.
-However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males.
-The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022].
+By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic.
+Although sex did not impact whether egg or sperm-specific genes were dysregulated, organismal sex did greatly influence the degree of germline gene dysregulation.
+Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
-Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-KDM5C's Y-chromsome homolog, KDM5D, exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and it is currently unknown to regulate germline gene expression.
-Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.
+Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D, which exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007].
+However, KDM5D has not been reported to regulate germline gene expression.
+Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
 
 We found KDM5C is largely dispensible for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
-It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
-Although expressed in na&#239;ve ESCs, "self-renewal" germline genes like _Dazl_ are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-We found that while _Kdm5c_-KO EpiLCs also express _Dazl_, they did not express 2-cell-specific genes.
+It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
+These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
+We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_.
 These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression.
-Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs), rather than de-differentiating into 2-cell-like cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].
+Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
 Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
 In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated.
-Thus, ectopic germline gene expression occurs along with proper somatic development in _Kdm5c_-KO animals.
+Thus, ectopic germline gene expression occurs along with overall proper somatic differentiation in _Kdm5c_-KO animals.
 
-Our work provides novel insight into the cross-talk between H3K4me and CpGme.
+Our work provides novel insight into the cross-talk between H3K4me and CpGme, which are often mutually exclusive[@weberDistributionSilencingPotential2007].
 In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g.
 _D1Pas1_ and _Naa11_, strongly impaired CGI methylation, and resulted in their long-lasting de-repression into adulthood.
-Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 primarily do not colocalize with CpGme[@weberDistributionSilencingPotential2007] and can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
 However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
@@
 _Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-Together, this suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007].
+This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007].
 Furthermore, these results indicate the requirement for catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
+Further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.
 
@@
 These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
-Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation and acts as break against runaway activation of germline-specific programs.
+Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation to act as a break against runaway activation of germline-specific programs.
 
@@
 However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
-In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain.
+In addition to germline genes, we also identified significant enrichment of muscle and liver-biased transcripts within the _Kdm5c_-KO brain.
 Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022].
@@
 KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.
+
 Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.

### 2024-09-16 submission/manuscript.Rmd
discussion edits

@@
 Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
-Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
-Intriguingly, mouse and human models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
-Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary.
+In _C.
+elegans_, misexpression of germline genes due to loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012].
+Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
+Intriguingly, mouse models and human cells for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
+Recently, the transcription factor ZMYM2 (ZNF198), which is mutated in the NDD neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and DNA methylation[@graham-paquinZMYM2EssentialMethylation2023].
+Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with similar erosion of the germline versus soma boundary.
 Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

### 2024-09-16 submission/manuscript.Rmd
results edits ChIPseq

@@
 
-KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017].
+KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023].
 However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters.
@@
 EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1).
-KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` &#177; `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).
+KDM5C was primarily localized to gene promoters in both cell types (Promoters = transcription start site (TSS) &#177; `r promorange`bp, EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)`), although PNCs showed increased localization to non-promoter regions (Figure 5A).
 
@@
 Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
-Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`)) (Figure 5C).
+Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
 There were no ontologies significantly enriched for PNC-specific KDM5C target genes.
@@
 Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E).
-Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 22/63*100`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
-However, KDM5C did bind the promoter of 4 out of the 5 genes dysregulated in both the brain and EpiLCs.
-Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
+Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
+However, KDM5C did bind the promoter of 4 out of the 5 genes dysregulated in both the brain and EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_).
+Representative examples of KDM5C-bound and unbound EpiLC DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
 Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
-Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.
+Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their gene promoters.
 
 Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-PRC1.6 members may recruit KDM5C to germline gene promoters, given their association with KDM5C in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023].
+PRC1.6 members may recruit KDM5C to germline gene promoters[@scandagliaLossKdm5cCauses2017], given their association with KDM5C in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023].
 We thus used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
@@
 RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure 1E).
-Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.
+Although _Rfx2_ is also not a direct target of KDM5C (Supplementary figure 1E), RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J).
+Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of many KDM5C-unbound germline genes in _Kdm5c_-KO cells.
 
@@
 However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023].
-To reconcile these observations, we hypothesized KDM5C erases H3K4me3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
+To reconcile these observations, we hypothesized KDM5C erases H3K4me2/3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
 
@@
 Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F).
-CGI-containing germline genes had substantial enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G).
+CGI-containing germline genes had higher enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G).
 Germline genes with promoter CGIs were more highly expressed than CGI-free genes across spermatogenesis stages, with highest expression in meiotic spermatocytes (Figure 6H).
-Contrastingly, CGI-free genes only displaying substantial expression in post-meiotic round spermatids.
-Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes, with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test) of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).
+Contrastingly, CGI-free genes only displayed substantial expression in post-meiotic round spermatids (Figure 6H).
+Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).
 
-To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96-hour extend EpiLCs (exEpiLCs), when germline genes reach peak methylation level[@mochizukiRepressionGermlineGenes2021] (Figure 6I).
-We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| >= 25%, TSS &#177; 500 bp).
+To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96-hour extend EpiLCs (exEpiLCs), when germline genes reach peak methylation levels[@mochizukiRepressionGermlineGenes2021] (Figure 6I).
+We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp).
 In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6J).
@@
 Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary figure 2A).
+Many promoters hyper- and hypomethylated in _Kdm5c_-KO exEpiLCs belonged to genes with unknown functions.
 Hypomethylated promoters were significantly enriched for germline gene ontologies, such as meiotic nuclear division (GO:0140013, p.adjust = 0.012)(Supplementary figure 2B), with `r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonging to germline genes.
 Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated promoters).
-Promoters that showed the most robust loss of CpGme (lowest q-values) harbored CGIs (Figure 6K).
+
+Promoters that showed the most robust loss of CpGme in _Kdm5c_-KO exEpiLCs (lowest q-values) harbored CGIs (Figure 6K).
 CGI promoters, but not CGI-free promoters, had a significant reduction in CpGme with KDM5C loss as a whole (Figure 6L) (Non-CGI promoters p = 0.0846, CGI promoters p = 0.0081, Mann-Whitney U test).
-Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _Naa11_ and _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) (Figure 6M).
+Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M).
 Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N).
-Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs to promote CpGme at germline gene promoters.
-This suggests KDM5C's cataltyic activity is required for germline gene repression in EpiLCs, however some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me around the TSS.
+Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters.
+Furthermore, this suggests while KDM5C's cataltyic activity is required for repression of some germline genes, some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me2/3 around the TSS.
 
@@
 Intriguingly, mouse models and human cells for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
-Recently, the transcription factor ZMYM2 (ZNF198), which is mutated in the NDD neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and DNA methylation[@graham-paquinZMYM2EssentialMethylation2023].
+Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and DNA methylation[@graham-paquinZMYM2EssentialMethylation2023].
 Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with similar erosion of the germline versus soma boundary.
@@
 We thank members of the University of Michigan Reproductive Sciences Program for providing feedback throughout the development of this work.
-This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B), and the NIH Early Stage Training in the Neurosciences Training Grant (T32-NS076401 to K.M.B).
+This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B), the NIH Early Stage Training in the Neurosciences Training Grant (T32-NS076401 to K.M.B), and
 

### 2024-09-16 code/ChIPseq_KDM5C_EpiLC_PNC.R

@@ -100,6 +100,7 @@ for (k in samples){
 	#read in the dataframe
 	a <- germGroups[[s]]
 	KDM5C_bound <- a[a %in% peakENSEMBL[[k]]]
+	print(KDM5C_bound)
 	KDM5C_unbound <- a[!a %in% peakENSEMBL[[k]]]
 	df <- data.frame(ChIP =  rep(k, 2), GeneCategory = rep(s, 2), KDM5C_binding = c("Bound", "Unbound"), Count = c(length(KDM5C_bound), length(KDM5C_unbound)))
 	qlotdf <- rbind(qlotdf, df)

### 2024-09-16 submission/manuscript.Rmd
acknowledgements

@@
 ## Data availability
+### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
+XXXX
+
 ### Published datasets
@@
 We also thank Drs.
-Stephanie Bielas, Michael Sutton, Donna Martin, and the members of the Iwase, Sutton, Bielas, and Martin labs for helpful discussions and critiques of the data.
+Kenneth Kwan, Natalie Tronson, Michael Sutton, Stephanie Bielas, Donna Martin, and the members of the Iwase, Sutton, Bielas, and Martin labs for helpful discussions and critiques of the data.
 We thank members of the University of Michigan Reproductive Sciences Program for providing feedback throughout the development of this work.
-This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B), the NIH Early Stage Training in the Neurosciences Training Grant (T32-NS076401 to K.M.B), and
+This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B.), the NIH Early Stage Training in the Neurosciences Training Grant (T32-NS076401 to K.M.B.), and the Michigan Predoctoral Training in Genetics Grant (NIH T32GM007544, to I.V.)
 

### 2024-09-16 submission/manuscript.Rmd
abstract edit

@@
 Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we examine the role of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) mutated in Claes-Jensen X-linked intellectual disability, in tissue identity.
+Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity.
 We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.

### 2024-09-17 submission/manuscript.Rmd
fix acknowledgements

@@
 We also thank Drs.
-Kenneth Kwan, Natalie Tronson, Michael Sutton, Stephanie Bielas, Donna Martin, and the members of the Iwase, Sutton, Bielas, and Martin labs for helpful discussions and critiques of the data.
+Gabriel Corfas, Kenneth Kwan, Natalie Tronson, Michael Sutton, Stephanie Bielas, Donna Martin, and the members of the Iwase, Sutton, Bielas, and Martin labs for helpful discussions and critiques of the data.
 We thank members of the University of Michigan Reproductive Sciences Program for providing feedback throughout the development of this work.

### 2024-09-17 submission/manuscript.Rmd
fix acknowledgements

@@
 We thank Drs.
-Sundeep Kalantry, Milan Samanta, and Rebecca Malcore for providing protocols and expertise in culturing mouse ESCs and EpiLCs, as well as providing wild-type and _Kdm5c_-KO ESCs used in this study.
+Sundeep Kalantry, Milan Samanta, and Rebecca Malcore for providing protocols and expertise in culturing mouse ESCs and EpiLCs, as well as providing the wild-type and _Kdm5c_-KO ESCs used in this study.
 We thank Dr.
@@
 We thank members of the University of Michigan Reproductive Sciences Program for providing feedback throughout the development of this work.
-This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B.), the NIH Early Stage Training in the Neurosciences Training Grant (T32-NS076401 to K.M.B.), and the Michigan Predoctoral Training in Genetics Grant (NIH T32GM007544, to I.V.)
+This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B.), the NIH Early Stage Training in the Neurosciences Training Grant (NIH T32NS076401 to K.M.B.), and the Michigan Predoctoral Training in Genetics Grant (NIH T32GM007544, to I.V.)
 

### 2024-09-17 submission/manuscript.Rmd
TBP sequence

@@
 For each sample, 2 ug of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT.
-Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP with the Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).
+Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP using the primers 5'-TTCAGAGGATGCTCTAGGGAAGA-3' 5'-CTGTGGAGTAAGTCCTGTGCC-3' with the Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).
 

### 2024-09-18 submission/manuscript.Rmd
affiliations and formatting

@@
 ```
-Katherine M. Bonefas, Ilakkiya Venkatachalam, and Shigeki Iwase.
+Katherine M. Bonefas^1,2^, Ilakkiya Venkatachalam^2,3^, and Shigeki Iwase^2^.
+
+1. Neuroscience Graduate Program, University of Michigan Medical School, Ann Arbor, MI, 48109, USA.
+2. Department of Human Genetics, Michigan Medicine, University of Michigan Medical School, Ann Arbor, MI, 48109, USA.
+3. Genetics and Genomics Graduate Program, University of Michigan, Ann Arbor, MI, 48109, USA.
+
+\newpage
 
@@
 Thus, distinct mechanisms govern the misexpression of germline gene classes, including activation by ectopic germline programs that mirror germ cell development within somatic tissues.
+
+\newpage
 

### 2024-09-18 submission/manuscript.Rmd
abstract edits

@@
 However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Thus, distinct mechanisms govern the misexpression of germline gene classes, including activation by ectopic germline programs that mirror germ cell development within somatic tissues.
+Together, these data indicate that KDM5C decommissions germline genes by facilitating CGI methylation to prevent runaway activation of germline developmental programs in somatic lineages.
 

### 2024-09-18 submission/manuscript.Rmd
small edits

@@
 Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs.
-Loss of some chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+Loss of chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
 However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.
@@
 
-We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
 For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B).
@@
 While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were sex-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
-We then compared the known functions of germline genes dysregulated only in females (XX only - unique to XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C).
+We then compared the known functions of germline genes dysregulated only in females, only in males, or in all samples (Figure 4C).
 Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion).
@@
 
-KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023].
+KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but not in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023].
 However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters.
@@
 Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E).
-Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
-However, KDM5C did bind the promoter of 4 out of the 5 genes dysregulated in both the brain and EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_).
+Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
 Representative examples of KDM5C-bound and unbound EpiLC DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
+However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary figure 1A).
 Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
-Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their gene promoters.
+Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their gene promoters, but genes dysregulated across _Kdm5c_-KO development are often direct KDM5C targets.
 
@@
 Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes.
-However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023].
+However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
 To reconcile these observations, we hypothesized KDM5C erases H3K4me2/3 to promote the initial placement of CpGme at germline gene promoters in EpiLCs.
@@
 Contrastingly, CGI-free genes only displayed substantial expression in post-meiotic round spermatids (Figure 6H).
-Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6G).
+Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6F).
 
@@
 Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters.
-Furthermore, this suggests while KDM5C's cataltyic activity is required for repression of some germline genes, some loci can compensate for KDM5C loss through other silencing mechansisms, even when retaining H3K4me2/3 around the TSS.
+Furthermore, this suggests while KDM5C's cataltyic activity is required for the repression of some germline genes, CpGme can be placed at others even with elevated H3K4me2/3 around the TSS.
 
@@
 We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes.
-Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells.
+Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and absent in somatic cells.
 _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis.
-We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression.
-Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation through ectopic RFX transcription factors.
+We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression with KDM5C loss.
+Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors.
 Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands.
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression distinct germline gene classes within somatic lineages.
-Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
+By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression of distinct germline gene classes within somatic lineages.
+Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
 
 By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic.
-Although sex did not impact whether egg or sperm-specific genes were dysregulated, organismal sex did greatly influence the degree of germline gene dysregulation.
+While sex did not determine whether egg or sperm-specific genes were dysregulated, it greatly influenced the degree of germline gene dysregulation.
 Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males.
-Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D, which exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007].
-However, KDM5D has not been reported to regulate germline gene expression.
+Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+Intriguingly, heterozygous knockout females also had over double the number of germline DEGs than hemizygous knockout males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males.
+Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D.
+However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression.
 Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
 
-We found KDM5C is largely dispensible for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
-It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
+We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
+It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
 These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
@@
 Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated.
-Thus, ectopic germline gene expression occurs along with overall proper somatic differentiation in _Kdm5c_-KO animals.
+In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus.
+Thus, ectopic germline gene expression occurs in conjunction with overall proper somatic differentiation in _Kdm5c_-KO animals.
 
-Our work provides novel insight into the cross-talk between H3K4me and CpGme, which are often mutually exclusive[@weberDistributionSilencingPotential2007].
+Our work provides novel insight into the cross-talk between H3K4me2/3 and CpGme, which are generally mutually exclusive[@weberDistributionSilencingPotential2007].
 In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g.
-_D1Pas1_ and _Naa11_, strongly impaired CGI methylation, and resulted in their long-lasting de-repression into adulthood.
+_D1Pas1_, strongly impaired CGI methylation and resulted in their long-lasting de-repression into adulthood.
 Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
@@
 In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2.
-_Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+_Dazl_ and a few other germline genes employ multiple repressive mechanisms to facilitate CpGme, such as DNMT3A/B recruitment via E2F6 and MGA[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
 This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007].
-Furthermore, these results indicate the requirement for catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
-Further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.
+Furthermore, these results indicate the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
+However, further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.
 
@@
 Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs.
-However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells.
+However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that are de-repressed in _Kdm5c_-KO cells.
 CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
 These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
-Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation to act as a break against runaway activation of germline-specific programs.
+Therefore, we propose KDM5C is recruited via promoter CGIs to act as a break against runaway activation of germline-specific programs.
 
@@
 In _C.
-elegans_, misexpression of germline genes due to loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012].
+elegans_, somatic misexpression of germline genes via loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling and ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012].
 Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
-Intriguingly, mouse models and human cells for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
-Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and DNA methylation[@graham-paquinZMYM2EssentialMethylation2023].
-Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with similar erosion of the germline versus soma boundary.
-Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
+Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
+Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023].
+Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary.
+Further research is required to determine the impact of these germline genes upon neuronal functions and the extent to which this phenomenon occurs in humans.
 
@@
 # Figures and Tables
-- Supplementary table 1: list of all germline genes. 
-	- Columns to include: 
-		- KDM5C bound vs not
-		- Log2fc in EpiLC, brain (separate columns?)
-    - CGI vs non
 
@@
 \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants.
-\textbf{J.} Example bigwigs of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants.}
+\textbf{J.} UCSC browser view of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants (Average of n = 3).}
   \label{figurelabel}

### 2024-09-19 submission/manuscript.Rmd
small edits

@@
 Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity.
+Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) that is mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity.
 We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
@@
 Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
+Germline gene repression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
 Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
 However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Together, these data indicate that KDM5C decommissions germline genes by facilitating CGI methylation to prevent runaway activation of germline developmental programs in somatic lineages.
+Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a break against runaway activation of germline developmental programs in somatic lineages.
 
@@
 MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
-Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017].
+Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023].
 RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
@@
 The contribution of KDM5C to this process remains unclear.
-Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated germline-enriched gene list.
+Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for germ cell development, given the lack of a comprehensive list for germline-enriched genes.
 Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g.
-meiotic genes versus spermatid differentiation genes.
+meiotic versus spermatid differentiation genes.
 
@@
 
-Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B).
+Interestingly, we also observed significant enrichment of ovary-biased genes in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B).
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D).
@@
 
-Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
+Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
 _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
 
-For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E).
+Analysis of oligo(dT)-primed libraries[@vallianatosMutuallySuppressiveRoles2020] demonstrates aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts in the _Kdm5c_-KO brain (Figure 1C-E).
 Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
@@
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D).
-Using our list of mouse germline-enriched genes assembled above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.
+Using our list of mouse germline-enriched genes assembled above, we identified `r nrow(epigerm)` germline genes misexpressed in male _Kdm5c_-KO EpiLCs.
 
@@
 Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H).
-These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate and later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate and again later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
@@
 This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression.
-Female sensitivity could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
+Heightened germline gene dysregulation in females could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+However, female germline DEGs were not biased towards the X chromosome and females had a a similar overall proportion of germline DEGs belonging to the X chromosome as males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
 The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G).
@@
 
-To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016].
-EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1).
-KDM5C was primarily localized to gene promoters in both cell types (Promoters = transcription start site (TSS) &#177; `r promorange`bp, EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)`), although PNCs showed increased localization to non-promoter regions (Figure 5A).
+To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016] (MACS2 q < 0.1, fold enrichment > 1, and removal of false-positive peaks in _Kdm5c_-KO).
+EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`).
+KDM5C was primarily localized to gene promoters in both cell types (promoters = transcription start site (TSS) &#177; `r promorange` bp, EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)`), although PNCs showed increased localization to non-promoter regions (Figure 5A).
 
@@
 Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
-Germline-specific ontologies were enriched only in EpiLC-specific KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
-There were no ontologies significantly enriched for PNC-specific KDM5C target genes.
+Germline ontologies were enriched only in EpiLC-specific, KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
+There were no significant ontologies for PNC-specific KDM5C target genes.
 Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D).
@@
 Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
-Representative examples of KDM5C-bound and unbound EpiLC DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
+Representative examples of EpiLC DEGs bound and unbound by KDM5C in EpiLCs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
 However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary figure 1A).
 Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
-Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their gene promoters, but genes dysregulated across _Kdm5c_-KO development are often direct KDM5C targets.
+Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their gene promoters, however genes dysregulated across _Kdm5c_-KO development are often direct KDM5C targets.
 
@@
 _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis.
-We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression with KDM5C loss.
+We demonstrated that although sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs have heightened germline gene de-repression with KDM5C loss.
 Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors.
@@
 Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-Intriguingly, heterozygous knockout females also had over double the number of germline DEGs than hemizygous knockout males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males.
+Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) .
 Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D.
 However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression.
-Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
+Altogether, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
 
 We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
-It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
+It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/the inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
 These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
@@
 In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g.
-_D1Pas1_, strongly impaired CGI methylation and resulted in their long-lasting de-repression into adulthood.
+_D1Pas1_, strongly impaired promoter CGI methylation and resulted in their long-lasting de-repression into adulthood.
 Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
@@
 This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007].
-Furthermore, these results indicate the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
+These results also suggest the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
 However, further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.
@@
 
-Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
+Given that germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown.
 In _C.
@@
 Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
-Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021].
+Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2).
 Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023].
-Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary.
+Thus, KDM5C is among a growing cohort of neurodevelopmental disorders that have erosion of the germline-soma boundary.
 Further research is required to determine the impact of these germline genes upon neuronal functions and the extent to which this phenomenon occurs in humans.
@@
 Sex was confirmed by genotyping _Uba1_/_Uba1y_ on the X and Y chromosomes with the following primers: 5'-TGGATGGTGTGGCCAATG-3', 5'-CACCTGCACGTTGCCCTT-3'.
-Deletion of _Kdm5c_ was confirmed through the primers 5'-ATGCCCATATTAAGAGTCCCTG-3', 5'-TCTGCCTTGATGGGACTGTT-3', and 5'-GGTTCTCAACACTCACATAGTG-3'.
+Deletion of _Kdm5c_ exons 11 and 12, which destabilize KDM5C protein[@iwaseMouseModelXlinked2016], was confirmed through the primers 5'-ATGCCCATATTAAGAGTCCCTG-3', 5'-TCTGCCTTGATGGGACTGTT-3', and 5'-GGTTCTCAACACTCACATAGTG-3'.
 
@@
 Briefly, ESCs were initially cultured in primed ESC (pESC) media consisting of KnockOut DMEM (Gibco#10829–018), fetal bovine serum (Gibco#A5209501), KnockOut serum replacement (Invitrogen#10828–028), Glutamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), MEM Non-essential amino acids (Gibco#11140-050), and beta-mercaptoethanol (Sigma#M7522).
-They were then transitioned into ground-state, "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol.
+They were then transitioned into ground-state, "na&#239;ve" ESCs (nESCs) by culturing for four passages in N2B27 media containing DMEM/F12 (Gibco#11330–032), Neurobasal media (Gibco#21103–049), Gluamax (Gibco#35050-061), Anti-Anti (Gibco#15240-062), N2 supplement (Invitrogen#17502048), and B27 supplement without vitamin A (Invitrogen#12587-010), and beta-mercaptoethanol (Sigma#M7522).
 Both pESC and nESC media were supplemented with 3 $\mu$M GSK3 inhibitor CHIR99021 (Sigma #SML1046-5MG), 1 $\mu$M MEK inhibitor PD0325901 (Sigma #PZ0162-5MG), and 1,000 units/mL leukemia inhibitory factor (LIF, Millipore#ESG1107).
 
-nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), beta-mercaptoethanol, fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and activin A (R&D Biotechne 338AC050CF), as previously described [@samantaGeneratingPrimedPluripotent2020].
+nESCs were differentiated into epiblast-like cells (EpiLCs, 48 hours) and extendend EpiLCs (exEpiLCs, 96 hours) by culturing in N2B27 media containing DMEM/F12, Neurobasal media, Gluamax, Anti-Anti, N2 supplement, B27 supplement (Invitrogen#17504044), and beta-mercaptoethanol supplemented with 10 ng/mL fibroblast growth factor 2 (FGF2, R&D Biotechne 233-FB), and 20 ng/mL activin A (R&D Biotechne 338AC050CF), as previously described [@samantaGeneratingPrimedPluripotent2020].
 
@@
 Cells were lysed with Tri-reagent BD (Sigma #T3809) at 0, 24, and 48 hours of differentiation.
-RNA was phase separated with 0.1 uL/uL 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516).
-For each sample, 2 ug of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT.
+RNA was phase separated with 0.1 uL/uL 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516) and ethanol purified.
+For each sample, 2 $\mu$g of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT.
 Expression of _Kdm5c_ was detected using the primers 5'-CCCATGGAGGCCAGAGAATAAG-3' 5'-CTCAGCGGATAAGAGAATTTGCTAC-3' and normalized to TBP using the primers 5'-TTCAGAGGATGCTCTAGGGAAGA-3' 5'-CTGTGGAGTAAGTCCTGTGCC-3' with the Power SYBR™ Green PCR Master Mix (ThermoFisher #4367659).
@@
 Total protein was extracted during nESC to EpiLC differentiation at 0, 24, and 48 hours by sonicating cells at 20% amplitude for 15 seconds in 2X SDS sample buffer, then boiling at 100°C for 10 minutes.
-Proteins were separated on a 7.5% SDS page gel, transferred overnight onto a fluorescent membrane, blotted for rabbit anti-KDM5C (in house, 1:500) and mouse anti-DAXX (Santa Cruz #(H-7): sc-8043, 1:500) imaged using the LiCor Odyssey CLx system.
+Proteins were separated on a 7.5% SDS page gel, transferred overnight onto a fluorescent membrane, blotted for rabbit anti-KDM5C (in house, 1:500) and mouse anti-DAXX (Santa Cruz #(H-7): sc-8043, 1:500) and imaged using the LiCor Odyssey CLx system.
 Band intensitity was quantified using ImageJ.
@@
 After ensuring read quality via FastQC (v0.11.8), reads were then mapped to the mm10 _Mus musculus_ genome (Gencode) using STAR (v2.5.3a), during which we removed duplicates and kept only uniquely mapped reads.
-Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser.
+Count files were generated by FeatureCounts (Subread v1.5.0), and BAM files were converted to bigwigs using deeptools (v3.1.3) and visualized by the UCSC genome browser[@nassarUCSCGenomeBrowser2023].
 RStudio (v3.6.0) was then used to analyze counts files by DESeq2 (v1.26.0)[@loveModeratedEstimationFold2014a] to identify differentially expressed genes (DEGs) with a q-value (p-adjusted via FDR/Benjamini–Hochberg correction) less than 0.1 and a log2 fold change greater than 0.5.
@@
 We removed “black-listed” genomic regions that often give aberrant signals.
-Common peak sets were obtained in R via DiffBind[estrogen@ross-innesDifferentialOestrogenReceptor2012] (v3.6.5).
-In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks were then subtracted from wild-type samples using bedtools (v2.25.0).
-Peak proximity to genome annotations was determined by ChIPSeeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1).
+Common peak sets were obtained in R via DiffBind[@ross-innesDifferentialOestrogenReceptor2012] (v3.6.5).
+In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks false-positive peaks were then removed from wild-type samples using bedtools (v2.25.0).
+Peak proximity to genomic loci was determined by ChIPSeeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1).
 Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
-Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010].
-Average binding across the genome was visualized using deeptools (v3.1.3).
-Bigwigs were visualized using the UCSC genome browser.
+Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010] to search for known motifs within 500 base pairs of the transcription start site.
+Average binding across genes was visualized using deeptools (v3.1.3).
+Bigwigs were visualized using the UCSC genome browser[@nassarUCSCGenomeBrowser2023].
 
@@
 Locations of CpG islands were determined through the mm10 UCSC genome browser CpG island track[@nassarUCSCGenomeBrowser2023], which classified CGIs as regions that have greater than 50% GC content, are larger than 200 base pairs, and have a ratio of CG dinucleotides observed over the expected amount greater than 0.6.
-CGI genomic coordinates were then annotated using ChIPseeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1) and filtered for ones that lie within promoters of our germline-enriched genes (TSS &#177; 500).
+CGI genomic coordinates were then annotated using ChIPseeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1) and filtered for ones that lie within promoters of germline-enriched genes (TSS &#177; 500).
 
@@
 Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark[@kruegerBismarkFlexibleAligner2011] (v0.22.1).
-Analysis of differential methylation at germline gene promoters was perfomed using methylKit[@akalinMethylKitComprehensivePackage2012] (v1.28.0) with a minimum coverage of 3 paired reads, a percentage cut-off of 25%, and q-value of 0.01.
+Analysis of differential methylation at germline gene promoters was perfomed using methylKit[@akalinMethylKitComprehensivePackage2012] (v1.28.0) with a minimum coverage of 3 paired reads, a percentage greater than 25% or less than -25%, and q-value less than 0.01.
 methylKit was also used to calculate average percentage methylation at germline gene promoters.
-Methylation bedgraph tracks were generated via Bismark and visualized using the UCSC genome browser.
+Methylation bedgraph tracks were generated via Bismark and visualized using the UCSC genome browser[@nassarUCSCGenomeBrowser2023].
 
@@
 ### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
-XXXX
+Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject XXX
 
@@
 All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo.
-Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722) and male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797).
+Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722).
+Male and female wild-type, _Kdm5c_-KO, and _Kdm5c_-HET EpiLCs[@samantaActivationXistEvolutionarily2022] are available at GEO: GSE96797.
 
-Previously published ChIP-seq experiments included KDM5C in wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797) and mouse primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036).
-ChIP-seq of histone 3 lysine 4 dimethylation in male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
-ChIP-seq of histone 3 lysine 4 trimethylation in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
+Previously published ChIP-seq experiments included KDM5C binding in wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (available at GEO: GSE96797) and mouse primary neuron cultures (PNCs) from the cortex and hippocampus[@iwaseMouseModelXlinked2016] (available at GEO: GSE61036).
+ChIP-seq of histone 3 lysine 4 dimethylation (H3K4me2) in male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
+ChIP-seq of histone 3 lysine 4 trimethylation (H3K4me3) in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
 

### 2024-09-25 submission/supplement.Rmd
description of supplementary tables

@@
 
+# Supplementary Tables
+
+**Supplementary table 1: Misexpression of tissue-enriched genes within the _Kdm5c_-KO brain.** 1) DESeq2 results table of the _Kdm5c_-KO hippocampus, with tissue-enriched genes annotated 2) Same as 1 but for the _Kdm5c_-KO amygdala 3) Results of the Fisher exact test for enrichment tissue-specific genes.
+4) enrichPlot gene ontology results of testis-enriched genes misexpressed in the amygdala or hippocampus.
+
+**Supplementary table 2: Mouse germline-enriched genes.** List of mouse germline-enriched genes identified in this study (see Methods).
+Includes whether germline gene promoters have KDM5C ChIP-seq peaks and CpG islands.
+
+**Supplementary table 3: Germline gene misexpression in _Kdm5c_ mutant EpiLCs.** 1) DESeq2 results table for male _Kdm5c_-KO (_Kdm5c^-/y^_) epiblast-like cells (EpiLCs), with annotations for tissue-enriched genes.
+2) Results of Fisher exact test on XY EpiLC tissue-enriched genes.
+3) enrichPlot gene ontology results of germline-enriched genes misexpressed in male _Kdm5c_-KO EpiLCs, amygdala, and hippocampus.
+4) Germline genes misexpressed in XY _Kdm5c_-KO (_Kdm5c^-/y^_), XX _Kdm5c_-KO (_Kdm5c^-/-^_), and XX _Kdm5c_-HET (_Kdm5c^-/+^_) EpiLCs.
+5) enrichPlot gene ontology results of germline genes misexpressed in male versus female _Kdm5c_ mutant EpiLCs
+
+**Supplementary table 4: KDM5C ChIP-seq in EpiLCs and PNCs.** 1) enrichPlot gene ontology results of KDM5C-bound promoters (see Methods) in epiblast-like cells (EpiLCs) and forebrain primary neuron cultures (PNCs).
+2) HOMER known motif analysis of germline gene promoters (TSS &#177; 500 bp) bound by KDM5C .
+3) HOMER known motif analysis of germline genes not bound by KDM5C.
+
+**Supplementary table 5: Germline gene CpG islands and promoter CpG methylation.** 1) enrichPlot gene ontology results of germline genes with and without CpG islands (CGIs) within their promoter (TSS &#177; 500 bp).
+2) methylKit whole genome bisulfite sequencing (WGBS) results table comparing germline gene promoter CpG methylation (TSS &#177; 500 bp) in male wild-type ESCs versus wild-type extended EpiLCs (exEpiLCs).
+3) methylKit WGBS results table comparing germline gene promoter CpG methylation in male _Kdm5c_-KO versus wild-type exEpiLCs.
+
+# Figures
 \newpage
@@
   \includegraphics{../submission/compiled_figs/Supplement_WGBS.pdf}
-  \caption[Loss of KDM5C impairs CpG methylation of germline gene promoters]{\textbf{[Loss of KDM5C impairs CpG methylation of germline gene promoters.} \textbf{A.} Volcano plot of whole genome bisulfite sequencing (WGBS) for all gene promoters in wild-type (WT) versus \textit{Kdm5c}-KO extended EpiLCs (exEpiLCs).
+  \caption[Loss of KDM5C impairs CpG methylation of germline gene promoters]{\textbf{Loss of KDM5C impairs CpG methylation of germline gene promoters.} \textbf{A.} Volcano plot of whole genome bisulfite sequencing (WGBS) for all gene promoters in wild-type (WT) versus \textit{Kdm5c}-KO extended EpiLCs (exEpiLCs).
 Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25\%).

### 2024-09-25 submission/manuscript.Rmd
clarify z-score in figure legend
ilakkiya edits and calculate tissue degs

@@
 Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-Germline gene repression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression.
+Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression.
 Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
@@
 Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
-Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
+Pathogenic mutations in the X chromosome gene _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
 MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
@@
 In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
 The contribution of KDM5C to this process remains unclear.
-Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for germ cell development, given the lack of a comprehensive list for germline-enriched genes.
-Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g.
+Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for early germ cell development, given the lack of a comprehensive list for germline-enriched genes.
+Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g.
 meiotic versus spermatid differentiation genes.
@@
 We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
+To illuminate the impact of sex upon germline gene suppression, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants.
 Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
@@
 
+amy_res <- read.csv("../data/processed/TissueDEGs_amy5CKO.csv", header = TRUE)
+amy_up = nrow(amy_res[amy_res$Direction == "Up",])
+amy_tissue = nrow(amy_res[amy_res$Direction == "Up" & amy_res$tissue %in% tissue_res$tissue,])
+
+hip_res <- read.csv("../data/processed/TissueDEGs_hip5CKO.csv", header = TRUE)
+hip_up = nrow(hip_res[hip_res$Direction == "Up",])
+hip_tissue = nrow(hip_res[hip_res$Direction == "Up" & hip_res$tissue %in% tissue_res$tissue,])
+
 ```
 
-We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B).
+We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: `r paste0(amy_tissue, "/", amy_up)` up DEGs, `r signif(amy_tissue/amy_up *100, 4)`% ; Hippocampus: `r paste0(hip_tissue, "/", hip_up)` up DEGs, `r signif(hip_tissue/hip_up *100, 4)`%) (Figure 1A-B).
 For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B).
@@
 \textbf{D.} Heatmap of germline DEGs shared between male and female mutants.
-Color is the average log 2 fold change from sex-matched wild-type.
+Color is the average log 2 fold change from sex-matched wild-type, z-scored across rows.
 \textbf{E.} Number of genes within each cluster from D.

### 2024-09-25 submission/manuscript.Rmd
Ilakkiya edits

@@
 Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) that is mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity.
-We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
+We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly express many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
 Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary.
@@
 However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a break against runaway activation of germline developmental programs in somatic lineages.
+Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.
 
@@
 For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B).
-Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
+Even though the testis has the largest total number of tissue-enriched genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
 An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
@@
 
-We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
+We then evaluated _Kdm5C_-KO testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979].
 Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B).
@@
 
-_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
+_Kdm5c_-KO cell morphology during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
 We then identified tissue-enriched DEGs in a RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
@@
 Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H).
-These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate and again later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+These genes are typically expressed when a subset of epiblast cells become primordial germ cells (PGCs) and again later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
@@
 Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion).
-Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 - meiotic nuclear division) and egg-specific functions (GO:0007292 - female gamete generation).
 
@@
 
-Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs within germline gene promoters[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
 PRC1.6 members may recruit KDM5C to germline gene promoters[@scandagliaLossKdm5cCauses2017], given their association with KDM5C in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023].
@@
 
-To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96-hour extend EpiLCs (exEpiLCs), when germline genes reach peak methylation levels[@mochizukiRepressionGermlineGenes2021] (Figure 6I).
+To assess how KDM5C loss impacts initial CpGme placement at germline gene promoters, we perfomed whole genome bisulfite sequencing (WGBS) in male wild-type and _Kdm5c_-KO ESCs and 96-hour extend EpiLCs (exEpiLCs), when germline genes reach peak methylation levels[@mochizukiRepressionGermlineGenes2021] (Figure 6I).
 We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp).
@@
 Many promoters hyper- and hypomethylated in _Kdm5c_-KO exEpiLCs belonged to genes with unknown functions.
-Hypomethylated promoters were significantly enriched for germline gene ontologies, such as meiotic nuclear division (GO:0140013, p.adjust = 0.012)(Supplementary figure 2B), with `r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonging to germline genes.
-Approximately half of germline promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated promoters).
+`r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonged to germline genes and germline-relevant ontologies are significantly enriched, such as meiotic nuclear division (GO:0140013, p.adjust = 0.012) (Supplementary figure 2B).
+Approximately half of all germline gene promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated promoters).
 
@@
 Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M).
-Surprisingly, we found only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N).
+Surprisingly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N).
 Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters.
@@
 By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic.
-While sex did not determine whether egg or sperm-specific genes were dysregulated, it greatly influenced the degree of germline gene dysregulation.
 Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
@@
 Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) .
-Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D.
+Males could partially compensate for KDM5C's loss via the Y-chromosome homolog, KDM5D.
 However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression.
@@
 We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
-It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/the inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
+It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in the inner cell mass/na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
 These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
@@
 These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
-Therefore, we propose KDM5C is recruited via promoter CGIs to act as a break against runaway activation of germline-specific programs.
+Therefore, we propose KDM5C is recruited via promoter CGIs to act as a brake against runaway activation of germline-specific programs.
 
@@
 ChIP-seq of histone 3 lysine 4 dimethylation (H3K4me2) in male wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] is also available at GEO: GSE96797.
-ChIP-seq of histone 3 lysine 4 trimethylation (H3K4me3) in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are avilable at GEO: GSE127817.
+ChIP-seq of histone 3 lysine 4 trimethylation (H3K4me3) in wild-type and _Kdm5c_-KO male amygdala[@vallianatosMutuallySuppressiveRoles2020] are available at GEO: GSE127817.
 

### 2024-09-25 submission/manuscript.Rmd
edits

@@
 ---
-title:  Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
+title:  Sexually dimorphic repression of germline identity by the X-linked intellectual disability gene KDM5C
 output: pdf_document
@@
 Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) that is mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity.
+Here, we examine how the X-linked intellectual disability gene lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), contributes to tissue identity.
 We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly express many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
@@
 Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
-However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
+However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
 Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.
@@
 To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
-Pathogenic mutations in the X chromosome gene _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
+Here we focus on the X chromosome gene lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
+Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
 MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
@@
 We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
-To illuminate the impact of sex upon germline gene suppression, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants.
+Additionally, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants to illuminate the impact of sex upon germline gene suppression.
 Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
@@
 
+tiss_list <- tissue_res$tissue
+tiss_list <- tiss_list[tiss_list != "Brain"]
 amy_res <- read.csv("../data/processed/TissueDEGs_amy5CKO.csv", header = TRUE)
 amy_up = nrow(amy_res[amy_res$Direction == "Up",])
-amy_tissue = nrow(amy_res[amy_res$Direction == "Up" & amy_res$tissue %in% tissue_res$tissue,])
+amy_tissue = nrow(amy_res[amy_res$Direction == "Up" & amy_res$tissue %in% tiss_list,])
 
@@
 hip_up = nrow(hip_res[hip_res$Direction == "Up",])
-hip_tissue = nrow(hip_res[hip_res$Direction == "Up" & hip_res$tissue %in% tissue_res$tissue,])
+hip_tissue = nrow(hip_res[hip_res$Direction == "Up" & hip_res$tissue %in% tiss_list,])
 
@@
 
-We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: `r paste0(amy_tissue, "/", amy_up)` up DEGs, `r signif(amy_tissue/amy_up *100, 4)`% ; Hippocampus: `r paste0(hip_tissue, "/", hip_up)` up DEGs, `r signif(hip_tissue/hip_up *100, 4)`%) (Figure 1A-B).
+`r print(tiss_list)`
+
+We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: `r paste0(amy_tissue, "/", amy_up)` up DEGs, `r signif(amy_tissue/amy_up *100, 4)`% ; Hippocampus: `r paste0(hip_tissue, "/", hip_up)` up DEGs, `r signif(hip_tissue/hip_up *100, 4)`%) (Figure 1A-B, Supplementary Table 1).
 For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B).
-Even though the testis has the largest total number of tissue-enriched genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
+Even though the testis has the largest total number of tissue-enriched genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-enriched DEGs were significantly enriched in both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test).
 An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).
 
-Interestingly, we also observed significant enrichment of ovary-biased genes in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B).
+Interestingly, we also observed significant enrichment of ovary-enriched genes in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact Test) (Figure 1A-B).
 Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D).
@@
 
-Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
-_Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
+Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
+_Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene, is among the liver-enriched DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
 
-Analysis of oligo(dT)-primed libraries[@vallianatosMutuallySuppressiveRoles2020] demonstrates aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts in the _Kdm5c_-KO brain (Figure 1C-E).
-Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
+Our analysis of oligo(dT)-primed libraries[@vallianatosMutuallySuppressiveRoles2020] indicates aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts in the _Kdm5c_-KO brain (Figure 1C-E).
+Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact Test), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes).
 Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.
@@
 To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology.
-We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).
+We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A, Supplementary Table 1).
 
@@
 We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline.
-These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
+These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary Table 2).
 
@@
 _Kdm5c_-KO cell morphology during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C).
-We then identified tissue-enriched DEGs in a RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1).
+We then identified tissue-enriched DEGs in a RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1, Supplementary Table 3).
 Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D).
@@
 The majority of germline DEGs were unique to either EpiLCs or the brain, with only _D1Pas1_ and _Cyct_ shared across all tissue/cell types (Figure 3E-F).
-EpiLC germline DEGs had particularly high enrichment of meiosis-related gene ontologies when compared to the brain (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
+EpiLC germline DEGs had particularly high enrichment of meiosis-related gene ontologies when compared to the brain (Figure 3G, Supplementary Table 3), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).
 While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).
@@
 Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H).
-These genes are typically expressed when a subset of epiblast cells become primordial germ cells (PGCs) and again later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
+These genes are typically expressed when a subset of epiblast stem cells become primordial germ cells (PGCs) and then again in mature germ cells to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].
 Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023].
-However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell state.
+However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H, Supplementary Table 3), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell state.
 Altogether, _Kdm5c_-KO EpiLCs express key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes, indicating germline gene misexpression mirrors germline development during the progression of somatic development.
 
-## Female epiblast-like cells have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
+## Female epiblast-like cells have heightened germline gene misexpression with _Kdm5c_ loss
 ```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
@@
 Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
-We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO, XY 5CKO), female homozygous knockout (XX _Kdm5c_-KO, XX 5CKO), and female heterozygous knockout (XX _Kdm5c_-HET, XX 5CHET) EpiLCs [@samantaActivationXistEvolutionarily2022].
+We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (_Kdm5c^-/y^_, XY _Kdm5c_-KO, XY 5CKO), female homozygous knockout (_Kdm5c^-/-^_, XX _Kdm5c_-KO, XX 5CKO), and female heterozygous knockout (_Kdm5c^-/+^_, XX _Kdm5c_-HET, XX 5CHET) EpiLCs [@samantaActivationXistEvolutionarily2022].
 
-In EpiLCs, homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A).
+In EpiLCs, homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A, Supplementary Table 3).
 While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were sex-specific, such as _Tktl2_ and _Esx1_ (Figure 4B).
-We then compared the known functions of germline genes dysregulated only in females, only in males, or in all samples (Figure 4C).
+We then compared the known functions of germline genes dysregulated uniquely in males and females or misexpressed in all samples (Figure 4C, Supplementary Table 3).
 Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion).
@@
 Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females would primarily express egg genes.
-To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H).
+To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4, Supplementary Table 2).
 We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex.
-This criteria yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes.
+This criteria yielded `r nrow(subset(germgenes, sexBias == "Egg-biased"))` egg-biased, `r nrow(subset(germgenes, sexBias == "Sperm-biased"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes.
 We found regardless of sex, egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J).
@@
 
-To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016] (MACS2 q < 0.1, fold enrichment > 1, and removal of false-positive peaks in _Kdm5c_-KO).
+To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016] (MACS2 q < 0.1, fold enrichment > 1, and removal of false-positive _Kdm5c_-KO peaks).
 EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`).
@@
 The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were bound by KDM5C only in EpiLCs (3,677 EpiLC only promoters) (Figure 5B).
-Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C).
+Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C, Supplementary Table 4).
 Germline ontologies were enriched only in EpiLC-specific, KDM5C-bound promoters, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C).
@@
 Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D).
-Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E).
-Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary figure 1A-C).
+Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288*100, 3)`% of all germline genes) (Figure 5E, Supplementary Table 2).
+Of note, KDM5C was only bound to about one third of RNA-seq DEG promoters unique to EpiLCs or the brain (EpiLC only DEGs: `r signif(22/63*100, 3)`%, Brain only DEGs: `r signif(6/20*100,3)`%) (Supplementary Figure 1A-C).
 Representative examples of EpiLC DEGs bound and unbound by KDM5C in EpiLCs are _Dazl_ and _Stra8_, repsectively (Figure 5F).
-However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary figure 1A).
+However, the four of the five germline genes dysregulated in both EpiLCs and the brain were bound by KDM5C in EpiLCs (_D1Pas1_, _Hsf2bp_, _Cyct_, and _Stk31_) (Supplementary Figure 1A).
 Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.
@@
 
-Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs within germline gene promoters[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
+Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains the transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs within germline gene promoters[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
 PRC1.6 members may recruit KDM5C to germline gene promoters[@scandagliaLossKdm5cCauses2017], given their association with KDM5C in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023].
-We thus used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1).
+We thus used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1, Supplementary Table 4).
 MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917), but not at germline genes unbound by KDM5C (Figure 5G).
@@
 RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-Although _Rfx2_ is also not a direct target of KDM5C (Supplementary figure 1E), RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J).
+Although _Rfx2_ is also not a direct target of KDM5C (Supplementary Figure 1E), RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J).
 Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of many KDM5C-unbound germline genes in _Kdm5c_-KO cells.
@@
 
-Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].
+Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
 Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes.
@@
 To test this hypothesis, we first characterized KDM5C's expression as na&#239;ve ESCs differentiate into EpiLCs (Figure 6A).
-While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6B), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 6C).
+While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 6B), KDM5C protein initially increased from 0 to 24 hours and then decreased to near knockout levels by 48 hours (Figure 6C).
 We then characterized KDM5C's substrates (H3K4me2/3) at germline gene promoters with _Kdm5c_ loss using published ChIP-seq datasets[@samantaActivationXistEvolutionarily2022; @vallianatosMutuallySuppressiveRoles2020].
@@
 We first examined how many of our germline-enriched genes had promoter CGIs (TSS &#177; 500 bp) using the UCSC genome browser[@nassarUCSCGenomeBrowser2023].
-Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F).
-CGI-containing germline genes had higher enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G).
+Notably, out of `r nrow(germgenes)` germline-enriched genes, only `r nrow(CGI)` (`r signif(nrow(CGI)/nrow(germgenes)*100, 4)`%) had promoter CGIs (Figure 6F, Supplementary Table 2).
+CGI-containing germline genes had higher enrichment of meiotic gene ontologies compared to CGI-free genes, including meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_CGI, 'p.adjust', "CGI"), 3)`) and meiosis I (GO:0007127, p.adjust = `r signif(getGO("meiosis I", GO_CGI, 'p.adjust', "CGI"), 3)`) (Figure 6G, Supplementary Table 5).
 Germline genes with promoter CGIs were more highly expressed than CGI-free genes across spermatogenesis stages, with highest expression in meiotic spermatocytes (Figure 6H).
 Contrastingly, CGI-free genes only displayed substantial expression in post-meiotic round spermatids (Figure 6H).
-Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's exact test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6F).
+Although only a minor portion of germline gene promoters contained CGIs, CGIs strongly determined KDM5C's recruitment to germline genes (p = `r signif(CGI_5C_fish$p.value, 3)`, Odds Ratio = `r signif(CGI_5C_fish$estimate, 3)`, Fisher's Exact Test), with `r signif(nrow(KDM5C_bound_CGI)/nrow(KDM5C_bound)*100, 4)`% of KDM5C-bound germline gene promoters harboring CGIs (Figure 6F).
 
@@
 We first identified which germline gene promoters siginificantly gained CpGme in wild-type cells during nESC to exEpiLCs differentiation (methylKit[@akalinMethylKitComprehensivePackage2012], q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp).
-In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6J).
+In wild-type cells, the majority of germline genes gained substantial CpGme at their promoter during differentiation (`r signif(nrow(ESCvsEpiLC_25)/nrow(ESCvsEpiLC)*100, 4)`%), regardless if their promoter contained a CGI (Figure 6J, Supplementary Table 5).
 
-We then identified promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp).
-Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary figure 2A).
+We then identified promoters differentially methylated in wild-type versus _Kdm5c_-KO exEpiLCs (methylKit, q < 0.01, |methylation difference| > 25%, TSS &#177; 500 bp, Supplementary Table 5).
+Of the `r nrow(exEpi_WTvsKO_all)` promoters assessed, `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypomethylated and `r nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff >= 25 & exEpi_WTvsKO_all$qvalue <= 0.01, ])` promoters were significantly hypermethylated with KDM5C loss (Supplementary Figure 2A).
 Many promoters hyper- and hypomethylated in _Kdm5c_-KO exEpiLCs belonged to genes with unknown functions.
-`r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonged to germline genes and germline-relevant ontologies are significantly enriched, such as meiotic nuclear division (GO:0140013, p.adjust = 0.012) (Supplementary figure 2B).
+However, `r signif((nrow(WTvsKO_hypo)/nrow(exEpi_WTvsKO_all[exEpi_WTvsKO_all$meth.diff <= -25 & exEpi_WTvsKO_all$qvalue <= 0.01, ]))*100, 4)`% of hypomethylated promoters belonged to germline genes and germline-relevant ontologies like meiotic nuclear division (GO:0140013, p.adjust = 0.012) are significantly enriched (Supplementary Figure 2B, Supplementary Table 5).
 Approximately half of all germline gene promoters hypomethylated in _Kdm5c_-KO exEpiLCs are direct targets of KDM5C in EpiLCs (`r nrow(WTvsKO_hypo_5C)` out of `r nrow(WTvsKO_hypo)` hypomethylated promoters).
@@
 By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic.
-Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
+Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the large number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
 Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
@@
 However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression.
-Altogether, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
+Nevertheless, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
 
@@
 However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
-In addition to germline genes, we also identified significant enrichment of muscle and liver-biased transcripts within the _Kdm5c_-KO brain.
-Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022].
+In addition to germline genes, we also identified significant enrichment of muscle and liver-enriched transcripts within the _Kdm5c_-KO brain.
+Intriguingly, select liver and muscle-enriched DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022].
 _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008].
@@
 Right - Number of tissue-enriched differentially expressed genes (DEGs).
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - UCSC browser view of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal) (Average, n = 4).
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's Exact Test \textbf{C.} Left - UCSC browser view of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal) (Average, n = 4).
 Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red.
@@
 \textbf{D.} Number of tissue-enriched differentially expressed genes (DEGs) in \textit{Kdm5c}-KO EpiLCs.
-* p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test.
+* p<0.05, ** p<0.01, *** p<0.001, Fisher's Exact Test.
 \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets.

### 2024-09-25 submission/supplement.Rmd
edits

@@
 ---
-title: Supplement - Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
+title: Supplement - Sexually dimorphic repression of germline identity by the X-linked intellectual disability gene KDM5C
 output: pdf_document
@@
 3) methylKit WGBS results table comparing germline gene promoter CpG methylation in male _Kdm5c_-KO versus wild-type exEpiLCs.
+4) enrichPlot gene ontology results of promoters hypomethylated in _Kdm5c_-KO exEpiLCs
 

### 2024-09-26 submission/manuscript.Rmd
small edits

@@
 Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the large number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-Increased germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
+Heightened germline gene misexpression in females may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008].
 Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) .
@@
 
-By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct derepressive mechanisms governing early versus late-stage germline developmental programs.
+By generating a comprehensive list of mouse germline-enriched genes, we revealed distinct derepressive mechanisms governing early versus late-stage germline programs.
 Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs.
@@
 CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].
-These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
+These data suggest that once activated during early embryogenesis, drivers of germline gene expression like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain.
 Therefore, we propose KDM5C is recruited via promoter CGIs to act as a brake against runaway activation of germline-specific programs.
@@
 Cells were lysed with Tri-reagent BD (Sigma #T3809) at 0, 24, and 48 hours of differentiation.
-RNA was phase separated with 0.1 uL/uL 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516) and ethanol purified.
+RNA was phase separated with 0.1 $\mu$L/$\mu$L 1-bromo-3-chloropropane (Sigma #B9673) and then precipitated with with isopropanol (Sigma #I9516) and ethanol purified.
 For each sample, 2 $\mu$g of RNA was reverse transcribed using the ProtoScript II Reverse transcriptase kit from New England Biolabs (NEB #M0368S) and primed with oligo dT.
@@
 Total protein was extracted during nESC to EpiLC differentiation at 0, 24, and 48 hours by sonicating cells at 20% amplitude for 15 seconds in 2X SDS sample buffer, then boiling at 100°C for 10 minutes.
-Proteins were separated on a 7.5% SDS page gel, transferred overnight onto a fluorescent membrane, blotted for rabbit anti-KDM5C (in house, 1:500) and mouse anti-DAXX (Santa Cruz #(H-7): sc-8043, 1:500) and imaged using the LiCor Odyssey CLx system.
+Proteins were separated on a 7.5% SDS page gel, transferred overnight onto a fluorescent membrane, blotted for rabbit anti-KDM5C (in house, 1:500) and mouse anti-DAXX (Santa Cruz #(H-7): sc-8043, 1:500), and then imaged using the LiCor Odyssey CLx system.
 Band intensitity was quantified using ImageJ.
@@
 Common peak sets were obtained in R via DiffBind[@ross-innesDifferentialOestrogenReceptor2012] (v3.6.5).
-In the case of KDM5C ChIP-seq, _Kdm5c_-KO peaks false-positive peaks were then removed from wild-type samples using bedtools (v2.25.0).
+In the case of KDM5C ChIP-seq, _Kdm5c_-KO false-positive peaks were then removed from wild-type samples using bedtools (v2.25.0).
 Peak proximity to genomic loci was determined by ChIPSeeker[@yuChIPseekerBioconductorPackage2015] (v1.32.1).
 Gene ontology (GO) analyses were performed by the R package enrichPlot (v1.16.2) using the biological processes setting and compareCluster.
-Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010] to search for known motifs within 500 base pairs of the transcription start site.
+Enriched motifs were identified using HOMER[@heinzSimpleCombinationsLineageDetermining2010] to search for known motifs within 500 base pairs up and downstream of the transcription start site.
 Average binding across genes was visualized using deeptools (v3.1.3).
@@
 ## Whole genome bisulfite sequencing (WGBS)
-Genomic DNA (gDNA) from na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells.
-gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150bp paired-end reads (PE150).
+Genomic DNA (gDNA) from male na&#239;ve ESCs and extended EpiLCs was extracted using the Wizard Genomic DNA Purification Kit (Promega A1120), following the instructions for Tissue Culture Cells.
+gDNA from two wild-types and two _Kdm5c_-KOs of each cell type was sent to Novogene for WGBS using the Illumina NovaSeq X Plus platform and sequenced for 150 bp paired-end reads (PE150).
 All samples had greater than 99% bisulfite conversion rates.
 Reads were adapter and quality trimmed with Trim Galore (v0.6.10) and aligned to the mm10 genome using Bismark[@kruegerBismarkFlexibleAligner2011] (v0.22.1).
-Analysis of differential methylation at germline gene promoters was perfomed using methylKit[@akalinMethylKitComprehensivePackage2012] (v1.28.0) with a minimum coverage of 3 paired reads, a percentage greater than 25% or less than -25%, and q-value less than 0.01.
+Analysis of differential methylation at gene promoters was perfomed using methylKit[@akalinMethylKitComprehensivePackage2012] (v1.28.0) with a minimum coverage of 3 paired reads, a percentage greater than 25% or less than -25%, and q-value less than 0.01.
 methylKit was also used to calculate average percentage methylation at germline gene promoters.
@@
 ### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
-Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject XXX
+Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject PRJNA1165148.
 
@@
 All published datasets are available at the Gene Expression Omnibus (GEO) https://www.ncbi.nlm.nih.gov/geo.
-Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020] (available at GEO: GSE127722).
+Published RNA-seq datasets analyzed in this study included the male wild-type and _Kdm5c_-KO adult amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020], available at GEO: GSE127722.
 Male and female wild-type, _Kdm5c_-KO, and _Kdm5c_-HET EpiLCs[@samantaActivationXistEvolutionarily2022] are available at GEO: GSE96797.
@@
 ### Data analysis
-Scripts used to generate the results, tables, and figures of this study are available via the GitHub repository: XXX
+Scripts used to generate the results, tables, and figures of this study are available via the GitHub repository: https://github.com/kbonefas/KDM5C_Germ_Mechanism
 
@@
 generated the ESC and exEpiLC WGBS data.
-K.M.B performed the data analysis and all other experiments.
-K.M.B and S.I.
-wrote and edited the manuscript.
+K.M.B performed all data analysis and all other experiments.
+The manuscript was written by K.M.B and S.I.
+and edited by K.M.B, S.I., and I.V.
 

### 2024-09-26 submission/manuscript.Rmd
new title

@@
 ---
-title:  Sexually dimorphic repression of germline identity by the X-linked intellectual disability gene KDM5C
+title: The X-linked intellectual disability gene KDM5C is a sexually dimorphic repressor of germline identity
 output: pdf_document

### 2024-09-26 submission/supplement.Rmd
new title

@@
 ---
-title: Supplement - Sexually dimorphic repression of germline identity by the X-linked intellectual disability gene KDM5C
+title: Supplement - The X-linked intellectual disability gene KDM5C is a sexually dimorphic repressor of germline identity
 output: pdf_document

### 2024-09-26 submission/manuscript.Rmd
change title again

--- 2024-09-26 submission/manuscript.Rmd
@@
 ---
-title:  Sexually dimorphic repression of germline identity by the X-linked intellectual disability gene KDM5C
+title: The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages
 output: pdf_document

### 2024-09-26 submission/supplement.Rmd
change title again

--- 2024-09-25 submission/supplement.Rmd
@@
 ---
-title: Supplement - Sexually dimorphic repression of germline identity by the X-linked intellectual disability gene KDM5C
+title: Supplement - The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages
 output: pdf_document

### 2024-09-26 code/EpiLC_XXvsXY.R
fisher test chromosome enrichment

@@ -122,16 +122,49 @@ chromo_hist(xxonly_chr, "DEG_ratio", "XX only germline DEGs", 4)
 #get histogram of all XX DEGs
 chromo_hist(XX_all_chr, "DEG_ratio", "all XX germline DEGs", 9)
 
+#fisher exact test for each chromosome
+ct = 1
+pvalues <- c()
+oddsratio <- c()
+for (m in XX_all_chr$chr){
 
-#save the chromosome results
+	#get that chromosome
+	df <- subset(XX_all_chr, chr == m)
+	df$DEGs
+	total_DEGs <- sum(XX_all_chr$DEGs)
+	total_germline <-sum(XX_all_chr$germline)
+
+	# for germline genes that are on each chromosome, how many are XX DEGs
+	# fisherdf <- data.frame("chr_yes" = c(chr_yes_DEG_yes, chr_yes_DEG_no), "chr_no" = c(chr_no_DEG_yes, chr_no_DEG_no), row.names = c("DEG_yes", "DEG_no"))
+
+	fisherdf <- data.frame("chr_yes" = c(df$DEGs, df$germline-df$DEGs), "chr_no" = c(total_DEGs-df$DEGs, total_germline-total_DEGs-(df$germline-df$DEGs)), row.names = c("DEG_yes", "DEG_no"))
+	print(fisherdf)
+
+	fishtest <- fisher.test(fisherdf)
+
+
+	pvalues[ct] <- fishtest$p.value
+	oddsratio[ct] <- fishtest$estimate[[1]]
+	ct <- ct + 1
+
+}
+		
+
+
+#add as column for fisher test pvalues and odds ratio
+XX_all_chr$pvalues <- pvalues
+XX_all_chr$oddsratio <- oddsratio
+
+
+#save the chromosome results for how many XX specific genes are on each chromosome
 write.table(XX_all_chr, snakemake@output[[5]], sep = ",", row.names = F)
 
 
 
 
-#percentage of germline DEGs on each chromsome for each sample
+#percentage of germline DEGs on each chromosome for each sample
 
-# number of DEGs on that chromsome/total number of germline DEGs for that sample * 100
+# number of DEGs on that chromosome/total number of germline DEGs for that sample * 100
 #need a dataframe with
 # sample chromosome percentage
 
@@ -179,7 +212,7 @@ germhist$samples <- factor(germhist$samples, levels = samples)
 print(germhist)
 
 library('ggpubr')
[…]

### 2024-09-26 code/EpiLC_XXvsXY.R
fix histogram height

@@ -212,7 +212,7 @@ germhist$samples <- factor(germhist$samples, levels = samples)
 print(germhist)
 
 library('ggpubr')
-q <- ggbarplot(germhist, x = "samples", y = "count", color = "samples", fill = "samples", ylim = c(0, 0.5), palette = EpiLCpalette, label = TRUE, lab.pos = "in", lab.col = "black")
+q <- ggbarplot(germhist, x = "samples", y = "count", color = "samples", fill = "samples", palette = EpiLCpalette, label = TRUE, lab.pos = "in", lab.col = "black")
 
 ggsave(snakemake@output[[6]], plot = q, width = 3, height = 3)
 

### 2024-09-26 code/utilities/germ_chromo.R

@@ -106,7 +106,7 @@ chromo_DEG_count_list <- function(genies){
 library('ggpubr')
 chromo_hist <- function(chr_cnt_df, y_col_name, title_name, out){
 
-	p <- ggbarplot(chr_cnt_df, "chr", y_col_name, fill = "chr", color = "chr") + rotate_x_text(45)
+	p <- ggbarplot(chr_cnt_df, "chr", y_col_name, fill = "chr", color = "chr", ylim = c(0, 0.5)) + rotate_x_text(45)
 	pl <- ggpar(p, legend = "none", ylab = "# of genes/total", xlab = "Chromosome #", title = paste0("Location of ", title_name) )
 	ggsave(snakemake@output[[out]], plot = pl, width = 5, height = 4)
 

### 2024-09-27 submission/supplement.Rmd
fix supplement tables

@@
 
-**Supplementary table 1: Misexpression of tissue-enriched genes within the _Kdm5c_-KO brain.** 1) DESeq2 results table of the _Kdm5c_-KO hippocampus, with tissue-enriched genes annotated 2) Same as 1 but for the _Kdm5c_-KO amygdala 3) Results of the Fisher exact test for enrichment tissue-specific genes.
+**Supplementary table 1: Misexpression of tissue-enriched genes within the _Kdm5c_-KO brain.** 1) Genes upregulated in the _Kdm5c_-KO hippocampus, with tissue-enriched genes annotated 2) Same as 1 but for the _Kdm5c_-KO amygdala 3) Results of the Fisher exact test for enrichment tissue-specific genes.
 4) enrichPlot gene ontology results of testis-enriched genes misexpressed in the amygdala or hippocampus.
@@
 
-**Supplementary table 3: Germline gene misexpression in _Kdm5c_ mutant EpiLCs.** 1) DESeq2 results table for male _Kdm5c_-KO (_Kdm5c^-/y^_) epiblast-like cells (EpiLCs), with annotations for tissue-enriched genes.
+**Supplementary table 3: Germline gene misexpression in _Kdm5c_ mutant EpiLCs.** 1) Genes upregulated in male _Kdm5c_-KO (_Kdm5c^-/y^_) epiblast-like cells (EpiLCs), with annotations for tissue-enriched genes.
 2) Results of Fisher exact test on XY EpiLC tissue-enriched genes.

### 2024-09-27 submission/manuscript.Rmd
remove extraneous print
shigeki edits

@@
 
+Correspondence should be addressed to K. Bonefas and S. Iwase (siwase@umich.edu)
+
 \newpage
@@
 
-Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms.
+Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
 Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
@@
 Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression.
-Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation.
+Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation.
 However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
@@
 The contribution of KDM5C to this process remains unclear.
-Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for early germ cell development, given the lack of a comprehensive list for germline-enriched genes.
+Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes, given the lack of a comprehensive list for germline-enriched genes.
 Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g.
@@
 It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
+We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO in figures)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
 
@@
 
-`r print(tiss_list)`
-
 We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: `r paste0(amy_tissue, "/", amy_up)` up DEGs, `r signif(amy_tissue/amy_up *100, 4)`% ; Hippocampus: `r paste0(hip_tissue, "/", hip_up)` up DEGs, `r signif(hip_tissue/hip_up *100, 4)`%) (Figure 1A-B, Supplementary Table 1).
@@
 Although not consistent across brain regions, we also found significant enrichment of genes biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B).
-_Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene, is among the liver-enriched DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
+These include _Apolipoprotein C-I (Apoc1)_, a lipoprotein metabolism and transport gene[@roulandRoleApolipoproteinC12022] (Figure 1E,see Discussion).
 
@@
 
-_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017], however the testis also contains somatic cells that support hormone production and germline functions.
-To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology.
+_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017] (Figure 1), however the testis also contains somatic cells that support hormone production and germline functions.
+To determine if _Kdm5c_-KO results in ectopic expression of testicular somatic genes, we first evaluated the known functions of testicular DEGs through gene onotology.
 We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A, Supplementary Table 1).
@@
 KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but not in neurons[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023].
-However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters.
-Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
+However, due to the lack of a comprehensive list of germline-enriched genes, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.
 
@@
 Previous work found two germline gene promoters have a marked reduction in DNA CpG methylation (CpGme) in the adult _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017].
-Since histone 3 lysine 4 di- and trimethylation (H3K4me2/3) impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes.
+Since histone H3K4me2/3 impede _de novo_ CpGme[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's removal of H3K4me2/3 may be required to suppress germline genes.
 However, KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023].
@@
 Significantly hypomethylated promoters included germline genes consistently dysregulated across multiple _Kdm5c_-KO RNA-seq datasets[@scandagliaLossKdm5cCauses2017], such as _D1Pas1_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "D1Pas1", "qvalue"),3)`) and _Naa11_ (methlyation difference = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO_hypo, "Naa11", "qvalue"),3)`) (Figure 6M).
-Surprisingly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N).
-Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at germline gene promoters.
-Furthermore, this suggests while KDM5C's cataltyic activity is required for the repression of some germline genes, CpGme can be placed at others even with elevated H3K4me2/3 around the TSS.
+Unexpectedly, we observed only a modest reduction in CpGme at _Dazl_'s promoter (methlyation difference = `r signif(getWGBSinfo(WTvsKO, "Dazl", "meth.diff"),4)`%, q-value = `r signif(getWGBSinfo(WTvsKO, "Dazl", "qvalue"),3)`) (Figure 6N).
+Altogether, these results demonstrate KDM5C is recruited to germline gene CGIs in EpiLCs to promote CpGme at those promoters.
+Furthermore, our data suggest while KDM5C's cataltyic activity is required for the repression of some germline genes, CpGme can be placed at others even with elevated H3K4me2/3 around the TSS.
 
@@
 Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via indirect mechanisms, such as activation through ectopic RFX transcription factors.
-Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands.
+Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through CGI DNA methylation.
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression of distinct germline gene classes within somatic lineages.
+By systematically characterizing KDM5C's role in germline gene repression, we unveiled distinct mechansisms governing the misexpression of distinct germline gene classes in somatic lineages.
 Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
@@
 Intriguingly, heterozygous knockout females (_Kdm5c^-/+^_) also had over double the number of germline DEGs than hemizygous knockout males (_Kdm5c^-/y^_), even though their expression of KDM5C should be roughly equivalent to that of wild-type males (_Kdm5c^+/y^_) .
-Males could partially compensate for KDM5C's loss via the Y-chromosome homolog, KDM5D.
-However, KDM5D exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and has not been reported to regulate germline gene expression.
+Males could partially compensate for KDM5C's loss via the Y-chromosome homolog, KDM5D[@iwaseXLinkedMentalRetardation2007].
+However, KDM5D has not been reported to regulate germline gene expression.
 Nevertheless, these results demonstrate germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.
@@
 We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
-It is important to note that while we highlighted KDM5C's repression of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in the inner cell mass/na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023].
-These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021].
-We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_.
-These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression.
+While some germline genes, such as _Dazl_, are also expressed in the 2-cell stage, inner cell mass, and na&#239;ve ESCs, they are silenced in epiblast stem cells/EpiLCs[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023; @mochizukiRepressionGermlineGenes2021].
+Our data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression (Figure 3).
 Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014].
@@
 _D1Pas1_, strongly impaired promoter CGI methylation and resulted in their long-lasting de-repression into adulthood.
-Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
+Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 repell DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015].
 However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
@@
 _Dazl_ and a few other germline genes employ multiple repressive mechanisms to facilitate CpGme, such as DNMT3A/B recruitment via E2F6 and MGA[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021].
-This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007].
-These results also suggest the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.
-However, further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.
+Thus, while some germline CGIs require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007], others do not.
+These results also suggest the requirement for KDM5C's catalytic activity can change depending upon the locus and developmental stage.
+Further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.
 
@@
 Therefore, we propose KDM5C is recruited via promoter CGIs to act as a brake against runaway activation of germline-specific programs.
+Future studies should adress how KDM5C is targeted to CGIs.
 
 The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes.
-However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown.
+However, the contribution of these ectopic, tissue-specific genes towards neurological impairments is still unknown.
 In addition to germline genes, we also identified significant enrichment of muscle and liver-enriched transcripts within the _Kdm5c_-KO brain.
@@
 Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2).
-Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes neurodevelopmental-craniofacial syndrome with variable renal and cardiac abnormalities (OMIM #619522), was also shown to repress germline genes by promoting H3K4 methylation removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023].
-Thus, KDM5C is among a growing cohort of neurodevelopmental disorders that have erosion of the germline-soma boundary.
+Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes an NDD (OMIM #619522), was also shown to repress germline genes by promoting H3K4me removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023].
+Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary.
 Further research is required to determine the impact of these germline genes upon neuronal functions and the extent to which this phenomenon occurs in humans.

### 2024-09-27 submission/manuscript.Rmd
small edits

@@
 We found KDM5C is largely dispensable for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs.
-While some germline genes, such as _Dazl_, are also expressed in the 2-cell stage, inner cell mass, and na&#239;ve ESCs, they are silenced in epiblast stem cells/EpiLCs[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023; @mochizukiRepressionGermlineGenes2021].
+While some germline genes, such as _Dazl_, are also expressed in the 2-cell stage, the inner cell mass, and na&#239;ve ESCs, they are silenced in epiblast stem cells/EpiLCs[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023; @mochizukiRepressionGermlineGenes2021].
 Our data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression (Figure 3).
@@
 Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018].
-In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus.
+In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated in the amygdala and hippocampus (Figure 1).
 Thus, ectopic germline gene expression occurs in conjunction with overall proper somatic differentiation in _Kdm5c_-KO animals.
@@
 Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2).
-Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes an NDD (OMIM #619522), was also shown to repress germline genes by promoting H3K4me removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023].
+Recently, the transcription factor ZMYM2 (ZNF198), whose mutation causes a NDD (OMIM #619522), was also shown to repress germline genes by promoting H3K4me removal and CpGme[@graham-paquinZMYM2EssentialMethylation2023].
 Thus, KDM5C is among a growing cohort of neurodevelopmental disorders with erosion of the germline-soma boundary.

### 2024-09-27 submission/manuscript.Rmd
x chromosome fisher results

@@
 Heightened germline gene dysregulation in females could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004].
-However, female germline DEGs were not biased towards the X chromosome and females had a a similar overall proportion of germline DEGs belonging to the X chromosome as males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
+However, female germline DEGs were not biased towards the X chromosome (p = 1, Odds Ratio = 0.96, Fisher's Exact Test) and females had a a similar overall proportion of germline DEGs belonging to the X chromosome as males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G).
 The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G).
@@
 \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome.
+P-values for Fisher Exact Test, ** p < 0.01, n.s.
+non-significant.
+Germline DEGs were only significant for chromosome 2, in which they were significantly depleted.
 Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant.

### 2024-09-27 submission/manuscript.Rmd
add new grants

@@
 We thank members of the University of Michigan Reproductive Sciences Program for providing feedback throughout the development of this work.
-This work was supported by grants from the National Institutes of Health (NIH) (National Institute of Neurological Disorders and Stroke: NS089896, 5R21NS104774, and NS116008 to S.I.), Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B.), the NIH Early Stage Training in the Neurosciences Training Grant (NIH T32NS076401 to K.M.B.), and the Michigan Predoctoral Training in Genetics Grant (NIH T32GM007544, to I.V.)
+This work was supported by grants from the National Institutes of Health (NIH) National Institute of Neurological Disorders and Stroke (NS089896, 5R21NS104774, and NS116008 to S.I.), National institute of Mental Health (1R21MH135290 to S.I.), the Simons Foundation Autism Research Initiative (SFARI, SFI-AN-AR-Pilot-00005721 to S.I.), the Farrehi Family Foundation Grant (to S.I.), the University of Michigan Career Training in Reproductive Biology (NIH T32HD079342, to K.M.B.), the NIH Early Stage Training in the Neurosciences Training Grant (NIH T32NS076401 to K.M.B.), and the Michigan Predoctoral Training in Genetics Grant (NIH T32GM007544, to I.V.)
 

### 2024-09-27 code/TissueGenes_AmyHip.R
deseq2 results table with tissue annotations

@@ -8,7 +8,7 @@ fisherdfs <- list()
 # get the testis-specific DEGs for each samples for later plotting in wt and germline-depleted testis
 testisDEGs <- data.frame()
 for(i in 1:length(titles)){
-	tissDEG <- tissue_genes(i)
+	tissDEG <- tissue_genes(i) 
 	#testis DEGs
 	TEdeg <- subset(tissDEG, tissue == "Testis")
 	TEdeg <- subset(TEdeg, select = c("ENSEMBL", "SYMBOL"))
@@ -54,6 +54,19 @@ ggsave(snakemake@output[[9]], plot = pl, width = 5, height = 5)
 
 ### save tissue-specific DEGs in a dataframe
 for(i in 1:length(titles)){
+	#tissue DEGs
 	tissDEG <- tissue_genes(i)
-	write.table(tissDEG, snakemake@output[[9+i]], sep = ",", row.names = FALSE)
+	#all DEGs
+	DEGs <- read.csv(snakemake@input[[i]], sep =",")
+	# print(paste(n, "tissue DEGs:", nrow(tissue_DEGs)))
+	print(paste("all DEGs", nrow(DEGs)))
+	print(tail(DEGs))
+	#all DEGs together annotated
+	annot <- merge(DEGs, tissDEG, by = "ENSEMBL", all.x = T)
+	print(paste("merge DEGs", nrow(annot)))
+	print(tail(annot))
+	#annot <- subset(annot, select = c("ENSEMBL", "baseMean.x", "log2FoldChange.x", "lfcSE.x ", "pvalue.x", "padj.x", "Direction.x", "SYMBOL.x", "tissue"))
+
+	write.table(annot, snakemake@output[[9+i]], sep = ",", row.names = FALSE)
 	}
+

### 2024-09-27 code/TissueGenes_EpiLC.R

@@ -34,5 +34,16 @@ write.table(fisherdfs, snakemake@output[[4]], sep = ",", row.names = FALSE)
 ### save tissue-specific DEGs in a dataframe
 for(i in 1:length(titles)){
 	tissDEG <- tissue_genes(i)
-	write.table(tissDEG, snakemake@output[[i+4]], sep = ",", row.names = FALSE)
+	#all DEGs
+	DEGs <- read.csv(snakemake@input[[i]], sep =",")
+	# print(paste(n, "tissue DEGs:", nrow(tissue_DEGs)))
+	print(paste("all DEGs", nrow(DEGs)))
+	print(tail(DEGs))
+	#all DEGs together annotated
+	annot <- merge(DEGs, tissDEG, by = "ENSEMBL", all.x = T)
+	annot <- annot[!duplicated(annot), ]
+	print(paste("merge DEGs", nrow(annot)))
+	print(tail(annot))
+	#annot <- subset(annot, select = c("ENSEMBL", "baseMean.x", "log2FoldChange.x", "lfcSE.x ", "pvalue.x", "padj.x", "Direction.x", "SYMBOL.x", "tissue"))
+	write.table(annot, snakemake@output[[i+4]], sep = ",", row.names = FALSE)
 	}

### 2024-09-27 code/utilities/tissueSPECIFICgenes.R

@@ -41,12 +41,15 @@ source("code/utilities/parameters.R") #get cut off of log2fc
 tissue_genes <- function(n){
 	#first read in the DEG table 
 	DEGs <- read.csv(snakemake@input[[n]], sep =",")
-	DEGs <- subset(DEGs, log2FoldChange > l2fcco)
 	names(DEGs)[names(DEGs) == 'X'] <- 'ENSEMBL'
+	print(head(DEGs))
+
+	DEGs_up <- subset(DEGs, log2FoldChange > l2fcco)
 	#merge together, keeping all columns. Genes that aren't tissue specific will be NAs
-	tissue_DEGs <- merge(Tissues, DEGs, by = "ENSEMBL", all=T)
+	tissue_DEGs <- merge(Tissues, DEGs_up, by = "ENSEMBL", all=T)
 	tissue_DEGs <- subset(tissue_DEGs, ENSEMBL %in% DEGs$ENSEMBL)
-	# print("tissue DEGs:")
+	# tissue_DEGs <- merge(tissue_DEGs, DEGs, by = c("ENSEMBL", "SYMBOL", "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj", "Direction"), all=T)
+	# print(paste(n, "tissue DEGs:", nrow(tissue_DEGs)))
 	# print(head(tissue_DEGs))
 	return(tissue_DEGs)
 

### 2024-09-27 submission/supplement.Rmd
small edits
update supplement table description

--- 2024-09-26 submission/supplement.Rmd
@@
 
-**Supplementary table 3: Germline gene misexpression in _Kdm5c_ mutant EpiLCs.** 1) DESeq2 results table for male _Kdm5c_-KO (_Kdm5c^-/y^_) epiblast-like cells (EpiLCs), with annotations for tissue-enriched genes.
+**Supplementary table 3: Germline gene misexpression in _Kdm5c_ mutant EpiLCs.** 1) DESeq2 results table of male _Kdm5c_-KO (_Kdm5c^-/y^_) epiblast-like cells (EpiLCs), with annotations for tissue-enriched genes.
 2) Results of Fisher exact test on XY EpiLC tissue-enriched genes.
 3) enrichPlot gene ontology results of germline-enriched genes misexpressed in male _Kdm5c_-KO EpiLCs, amygdala, and hippocampus.
-4) Germline genes misexpressed in XY _Kdm5c_-KO (_Kdm5c^-/y^_), XX _Kdm5c_-KO (_Kdm5c^-/-^_), and XX _Kdm5c_-HET (_Kdm5c^-/+^_) EpiLCs.
+4) DESeq2 results table of XX _Kdm5c_-HET (_Kdm5c^-/+^_) EpiLCs 5) DESeq2 results table of XX _Kdm5c_-KO (_Kdm5c^-/-^_) EpiLCs 6) Germline genes misexpressed in XY _Kdm5c_-KO, XX _Kdm5c_-KO, and XX _Kdm5c_-HET EpiLCs.
 5) enrichPlot gene ontology results of germline genes misexpressed in male versus female _Kdm5c_ mutant EpiLCs
@@
 
-# Figures
+# Supplementary Figures
 \newpage

### 2024-09-27 submission/manuscript.Rmd
add XX vs XY GO to supplement and pvalues to text

@@
 We then compared the known functions of germline genes dysregulated uniquely in males and females or misexpressed in all samples (Figure 4C, Supplementary Table 3).
-Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle) and flagellar (GO:0003341 - cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion).
+Female-specific germline DEGs were enriched for meiotic (GO:0051321 - meiotic cell cycle, p.adjust = 7.81E-14) and flagellar (GO:0003341 - cilium movement, p.adjust = 4.87E-06) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 - protein localization to mitochondrion, p.adjust = 0.025).
 

### 2024-10-26 submission/manuscript.Rmd
add competing interests
add bioproject link
abstract update

@@
 
-Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs).
-Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown.
-Here, we examine how the X-linked intellectual disability gene lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), contributes to tissue identity.
-We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly express many liver, muscle, ovary, and testis genes within the amygdala and hippocampus.
-Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary.
-Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
+A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including the the distinction of the germline from the soma.
+In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications.
+Failure to repress germline genes in somatic lineages is common signature of cancer and observed in select neurodevelopmental disorders, however it is currently unclear how factors like development and sex influence their repression and somatic misexpression.
+Here, we examine how cellular context influences the development of somatic tissue identity in mice with loss of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain.
+Late-stage spermatogenesis genes, but not somatic testicular genes, were highly enriched within the _Kdm5c_-KO brain, indicating an erosion of soma-germline boundary.
+Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
 Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression.
@@
 Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.
+
+**notes**
+
+- Distinguishing the germline from the soma is a key step in the evolution of multicellularity and sexual reproduction. 
+- Germline gene repression is orchestrated by chromatin regulators and transcription factors. 
+- Much of these discoveries have been made looking at the repression of key marker genes for germ cell specification in the early embryo,
+
+- While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in late oogenesis and spermatogenesis, employ the same silencing mechanisms and how silencing changes over the course of development.
+- In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
+- Germline genes have not been comprehensively assessed as a whole
 
@@
 ### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
-Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject PRJNA1165148.
+Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject PRJNA1165148. https://www.ncbi.nlm.nih.gov/bioproject/PRJNA1165148
 
@@
 
-# Author contributions
+# Author Contributions
 K.M.B.
@@
 and edited by K.M.B, S.I., and I.V.
+
+# Declaration of Interest
+S.I. is a member of the Scientific Advisory Board of KDM5C Advocacy, Research, Education & Support (KARES). Other authors declare no conflict of interest.
 

### 2024-10-29 submission/manuscript.Rmd
new intro draft

@@
 
-A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including the the distinction of the germline from the soma.
+A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including distinguishing the germline from the soma.
 In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications.
@@
 We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain.
-Late-stage spermatogenesis genes, but not somatic testicular genes, were highly enriched within the _Kdm5c_-KO brain, indicating an erosion of soma-germline boundary.
+By developing a comprehensive list of mouse germline-enriched genes, we found late-stage spermatogenesis genes and not somatic testicular genes were highly enriched within the _Kdm5c_-KO brain.
 Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
 Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression.
-Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation.
-However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
+Although many germline mRNAs are expressed in the _Kdm5c_-KO brain and EpiLCs, KDM5C recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation.
+Late-stage spermatogenesis genes lacking KDM5C and devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
 Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.
@@
 
-- While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in late oogenesis and spermatogenesis, employ the same silencing mechanisms and how silencing changes over the course of development.
 - In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
@@
 
-A single genome holds the instructions to generate the myriad of cell types found within an organism.
+A single genome holds the instructions for generating the myriad of cell types found within an organism.
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs.
-Loss of chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
-However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.
+Human genetic studies revealed impaired chromatin regulation commonly occurs in both cancer[@zhaoLanguageChromatinModification2021; @chiCovalentHistoneModificationsmiswritten2010; @berdascoAberrantEpigeneticLandscape2010] and neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
+While many studies have identified their importance for regulating tumor suppressor genes and brain-specific transcriptional programs, loss of chromatin regulators can also cause ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of testis genes in colon tumors[@feichtingerMetaanalysisClinicalData2012] or liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
+However, the mechanisms driving ectopic gene expression and its impact upon cancer and neurodevelopment are still poorly understood.
 
-To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression.
-Here we focus on the X chromosome gene lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
-Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
+Separation of germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[XXX, @devlinPanmetazoanViewGermlinesoma2023].
+In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency.
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms.
+Furthermore, because many studies have focused on the silencing of key marker genes during early male embryogenesis, much is unknown about how cellular context influences the manifestation of germline gene misexpression.
+Given that impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[XXX; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021], elucidating the impact of factors like sex and development is crucial for determining the mechanisms governing long-term misexpression of germline genes.
+
+Here, we employed a genome-wide analyses to explore the impact of cellular context upon loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
+KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
+Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
 MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].
 Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023].
-RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017].
-However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.
+RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some testis germline genes within the brain[@scandagliaLossKdm5cCauses2017].
+However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, what mechanisms underlie their dysregulation, and how KDM5C interacts with other known germline silencing mechanisms.
 
-Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012].
-In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
-The contribution of KDM5C to this process remains unclear.
-Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes, given the lack of a comprehensive list for germline-enriched genes.
-Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g.
-meiotic versus spermatid differentiation genes.
-
-To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
-We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time.
+To illuminate KDM5C's role in tissue identity, we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.
+We curated a list of mouse germline-enriched genes, enabling genome-wide analysis of germline gene silencing mechanisms for the first time.
 Additionally, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants to illuminate the impact of sex upon germline gene suppression.

### 2024-10-29 submission/manuscript.Rmd
intro edits

@@
 
-A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including distinguishing the germline from the soma.
-In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications.
-Failure to repress germline genes in somatic lineages is common signature of cancer and observed in select neurodevelopmental disorders, however it is currently unclear how factors like development and sex influence their repression and somatic misexpression.
-Here, we examine how cellular context influences the development of somatic tissue identity in mice with loss of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
-We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain.
-By developing a comprehensive list of mouse germline-enriched genes, we found late-stage spermatogenesis genes and not somatic testicular genes were highly enriched within the _Kdm5c_-KO brain.
-Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression.
-Although many germline mRNAs are expressed in the _Kdm5c_-KO brain and EpiLCs, KDM5C recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation.
-Late-stage spermatogenesis genes lacking KDM5C and devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
-Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.
+The division labor among cellular lineages is a pivotal step in the evolution of multicellularity.
+In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI).
+Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders.
+However it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation.
+Here, we examine how cellular context influences the development of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain.
+By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline identity and meiosis, such as _Dazl_ and _Stra8_.
+KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.
+KDM5C represses germline genes during ESC to EpiLC differentiation, during which it is recruited to a subset of CGI-containing germline identity gene promoters to facilitate CpG methylation.
+However the majority of germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
+Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.
 
 **notes**
+- How does sex and development influence germline gene repression?
+- How does KDM5C interact with known silencing machinery? 
+- Do all germline genes use teh same repressive mechanisms
+
+  aberrantly expressed , including _Dazl_ and _Stra8_.
+  - Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
 
@@
 
-A single genome holds the instructions for generating the myriad of cell types found within an organism.
+A single genome holds the instructions to generate the myriad of cell types found within an organism.
 This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
@@
 
-Separation of germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[XXX, @devlinPanmetazoanViewGermlinesoma2023].
+Separation between germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023].
 In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
-While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms.
-Furthermore, because many studies have focused on the silencing of key marker genes during early male embryogenesis, much is unknown about how cellular context influences the manifestation of germline gene misexpression.
-Given that impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[XXX; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021], elucidating the impact of factors like sex and development is crucial for determining the mechanisms governing long-term misexpression of germline genes.
+Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in the later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms.
+Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e.
+sex and tissue environment) influences the manifestation of germline gene misexpression.
+Intriguingly, impaired soma-germline demarcation is a signature of aggressive cancers and observed in select neurodevelopmental disorders (NDDs)[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @berdascoAberrantEpigeneticLandscape2010; @bonefasSomatogermlineTransformationChromatinlinked2021].
+Thus, elucidating how cell context contributes to germline gene dysregulation will reveal novel mechanisms governing these pathologies.
 
-Here, we employed a genome-wide analyses to explore the impact of cellular context upon loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
+Here, we employed genome-wide analyses to explore loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
 KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
@@
 It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain.
-We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO in figures)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
+We first systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO in figures)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].
 
@@
 elegans_, somatic misexpression of germline genes via loss of _Retinoblastoma (Rb)_ homologs results in enhanced piRNA signaling and ectopic P granule formation in neurons[@wangSomaticMisexpressionGermline2005; @wuRepressionGermlineRNAi2012].
-Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017] and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
+Ectopic testicular germline transcripts have also been observed in a variety of cancers, including brain turmors in _Drosophila_ and mammals and shown to promote cancer progression[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017; @janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023].
 Intriguingly, mouse models and human cells for other chromatin-linked NDDs also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], such as mutations in DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, and methyl-CpG -binding protein 2 (MECP2).

### 2024-10-29 submission/manuscript.Rmd
intro and abstract edits

@@
 
-The division labor among cellular lineages is a pivotal step in the evolution of multicellularity.
+The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity.
 In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI).
 Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders.
-However it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation.
-Here, we examine how cellular context influences the development of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
+However, it is currently unclear if all germline genes use the same repressive mechanisms and if factors like development and sex influence their dysregulation.
+Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
 We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain.
-By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline identity and meiosis, such as _Dazl_ and _Stra8_.
+By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline fate, such as _Dazl_ and _Stra8_.
 KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.
-KDM5C represses germline genes during ESC to EpiLC differentiation, during which it is recruited to a subset of CGI-containing germline identity gene promoters to facilitate CpG methylation.
-However the majority of germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors.
+KDM5C represses germline genes during ESC to EpiLC differentiation and is recruited to a subset of CGI-containing germline gene promoters to facilitate DNA CpG methylation.
+However, the majority of mouse germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells through ectopic activation by RFX transcription factors.
 Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.
-
-**notes**
-- How does sex and development influence germline gene repression?
-- How does KDM5C interact with known silencing machinery? 
-- Do all germline genes use teh same repressive mechanisms
-
-  aberrantly expressed , including _Dazl_ and _Stra8_.
-  - Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_.
-
-- Distinguishing the germline from the soma is a key step in the evolution of multicellularity and sexual reproduction. 
-- Germline gene repression is orchestrated by chromatin regulators and transcription factors. 
-- Much of these discoveries have been made looking at the repression of key marker genes for germ cell specification in the early embryo,
-
-- In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
-- Germline genes have not been comprehensively assessed as a whole
 
@@
 Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
-While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in the later stages of oogenesis and spermatogenesis, employ the same silencing mechanisms.
+While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes employ the same silencing mechanisms, such as those involved in the later stages of oogenesis and spermatogenesis.
 Furthermore, because many studies have focused on the silencing of key marker genes during early male embryonic development, much is unknown about how cellular context (i.e.
@@
 
-Here, we employed genome-wide analyses to explore loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
+Here, we employed genome-wide analyses to explore the loss of tissue identity in mice lacking the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C).
 KDM5C lies on the X chromosome and erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007].
-Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
+Somatic loss of KDM5C promotes tumorigenicity in a variety of cancer types[@shenSuppressionEnhancerOveractivation2016; @chenE6ProteinExpressed2018; @zhengDeficiencyXinactivationEscaping2021], while pathogenic germline mutations cause the NDD Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534).
 MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020].

### 2024-11-05 submission/manuscript.Rmd
abstract edits

@@
 Here, we examine how cellular context influences the formation of somatic tissue identity in mice lacking lysine demethylase 5c (KDM5C), an X chromosome eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3).
-We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, and germline genes within the brain.
-By developing a comprehensive list of mouse germline-enriched genes, we found the _Kdm5c_-KO brain aberrantly expresses late-stage spermatogenesis genes while _Kdm5c_-KO epiblast-like cells (EpiLCs) primarily expressed key regulators of germline fate, such as _Dazl_ and _Stra8_.
-KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.
-KDM5C represses germline genes during ESC to EpiLC differentiation and is recruited to a subset of CGI-containing germline gene promoters to facilitate DNA CpG methylation.
-However, the majority of mouse germline gene promoters did not harbor CGIs and were unbound by KDM5C, including late-stage spermatogenesis genes that are expressed in _Kdm5c_-KO cells through ectopic activation by RFX transcription factors.
-Altogether, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a sex-biased brake against runaway activation of germline developmental programs in somatic lineages.
+We found male _Kdm5c_ knockout (-KO) mice aberrantly express many tissue-specific genes within the brain, the majority of which are unique to the germline.
+By developing a comprehensive list of mouse germline-enriched genes, we observed _Kdm5c_-KO cells aberrantly express key drivers of germline fate during early embryogenesis but late-stage spermatogenesis genes within the mature _Kdm5c_-KO brain.
+KDM5C binds germline gene promoter CGIs to facilitate DNA CpG methylation as embryonic stem cells differentiate into epiblast-like cells (EpiLCs).
+However, the majority of late-stage spermatogenesis genes expressed within the _Kdm5c_-KO brain did not harbor promoter CGIs.
+These CGI-free germline genes were not bound by KDM5C and instead expressed through ectopic activation by RFX transcription factors.
+Furthermore, KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.
+
+Altogether, these data revealed distinct regulatory classes of germline genes and sex-biased silencing mechanisms in somatic cells.
 

### 2024-11-05 submission/manuscript.Rmd
shigeki abstract edits
discussion small edit

@@
 The division of labor among cellular lineages is a pivotal step in the evolution of multicellularity.
-In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGI).
+In mammals, the soma-germline boundary is formed during early embryogenesis, when genes that drive germline identity are repressed in somatic lineages through DNA and histone modifications at promoter CpG islands (CGIs).
 Somatic misexpression of germline genes is a signature of cancer and observed in select neurodevelopmental disorders.
@@
 We found male _Kdm5c_ knockout (-KO) mice aberrantly express many tissue-specific genes within the brain, the majority of which are unique to the germline.
-By developing a comprehensive list of mouse germline-enriched genes, we observed _Kdm5c_-KO cells aberrantly express key drivers of germline fate during early embryogenesis but late-stage spermatogenesis genes within the mature _Kdm5c_-KO brain.
-KDM5C binds germline gene promoter CGIs to facilitate DNA CpG methylation as embryonic stem cells differentiate into epiblast-like cells (EpiLCs).
+By developing a comprehensive list of mouse germline-enriched genes, we observed _Kdm5c_-KO cells aberrantly express key drivers of germline fate during early embryogenesis but late-stage spermatogenesis genes within the mature brain.
+KDM5C binds CGIs within germline gene promoters to facilitate DNA CpG methylation as embryonic stem cells differentiate into epiblast-like cells (EpiLCs).
 However, the majority of late-stage spermatogenesis genes expressed within the _Kdm5c_-KO brain did not harbor promoter CGIs.
 These CGI-free germline genes were not bound by KDM5C and instead expressed through ectopic activation by RFX transcription factors.
-Furthermore, KDM5C-mediated germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.
-
+Furthermore, germline gene repression is sexually dimorphic, as female EpiLCs require a higher dose of KDM5C to maintain germline silencing.
 Altogether, these data revealed distinct regulatory classes of germline genes and sex-biased silencing mechanisms in somatic cells.
@@
 
-A single genome holds the instructions to generate the myriad of cell types found within an organism.
-This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017].
-Human genetic studies revealed impaired chromatin regulation commonly occurs in both cancer[@zhaoLanguageChromatinModification2021; @chiCovalentHistoneModificationsmiswritten2010; @berdascoAberrantEpigeneticLandscape2010] and neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018].
-While many studies have identified their importance for regulating tumor suppressor genes and brain-specific transcriptional programs, loss of chromatin regulators can also cause ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of testis genes in colon tumors[@feichtingerMetaanalysisClinicalData2012] or liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009].
-However, the mechanisms driving ectopic gene expression and its impact upon cancer and neurodevelopment are still poorly understood.
-
-Separation between germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023].
+The separation of germline and somatic cellular identity is a pivotal step in the evolution of multicellularity and sexual reproduction[@hanschenMulticellularityDrivesEvolution2018; @michodEvolutionaryTransitionsIndividuality2011; @chenMulticellularitySexHelped2021; @devlinPanmetazoanViewGermlinesoma2023].
 In mammals, chromatin regulators decommission germline genes in somatic lineages when the early embryo transitions from na&#239;ve to primed pluripotency.
-Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
+Germline gene promoters initially gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) at their CpG islands (CGIs) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021].
 While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes employ the same silencing mechanisms, such as those involved in the later stages of oogenesis and spermatogenesis.
@@
 We curated a list of mouse germline-enriched genes, enabling genome-wide analysis of germline gene silencing mechanisms for the first time.
-Additionally, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants to illuminate the impact of sex upon germline gene suppression.
-Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
+We identified two classes of germline genes based on their promoter CpG island content, which are dysregulated with KDM5C loss by distinct mechanisms and in a sex-biased manner.
 
@@
 Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.
-By systematically characterizing KDM5C's role in germline gene repression, we unveiled distinct mechansisms governing the misexpression of distinct germline gene classes in somatic lineages.
-Ultimately, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.
+By systematically characterizing KDM5C's role in germline gene repression, we unveiled divergent mechanisms governing the misexpression of distinct germline gene classes in somatic lineages.
 

### 2024-11-05 submission/manuscript.Rmd
update title

@@
 ---
-title: The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages
+title: KDM5C is a sex-biased brake against germline gene expression programs in somatic lineages
 output: pdf_document

### 2024-11-05 submission/manuscript.Rmd
genome research formatting spacing and refs

@@
 ---
-title: KDM5C is a sex-biased brake against germline gene expression programs in somatic lineages
+title: The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages
 output: pdf_document
@@
    - \usepackage{siunitx}
-   - \spacing{1.5}
+   - \spacing{2}
    - \pagenumbering{arabic}
    - \usepackage{indentfirst}
+   - \usepackage[style=alphabetic]{biblatex}
 

### 2024-11-08 submission/manuscript.Rmd
add competing interest section

@@
 
-## Data availability
+## Data access
 ### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
@@
 Scripts used to generate the results, tables, and figures of this study are available via the GitHub repository: https://github.com/kbonefas/KDM5C_Germ_Mechanism
+
+# Competing Interest
+S.I. is a member of the Scientific Advisory Board of KDM5C Advocacy, Research, Education & Support (KARES). All other authors declare no conflict of interest.
 

### 2024-11-08 submission/supplement.Rmd
fix supplement numbering
fix supplement title

@@
 ---
-title: Supplement - The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages
+title: Supplement - KDM5C is a sex-biased brake against germline gene expression programs in somatic lineages
 output: pdf_document
@@
 4) DESeq2 results table of XX _Kdm5c_-HET (_Kdm5c^-/+^_) EpiLCs 5) DESeq2 results table of XX _Kdm5c_-KO (_Kdm5c^-/-^_) EpiLCs 6) Germline genes misexpressed in XY _Kdm5c_-KO, XX _Kdm5c_-KO, and XX _Kdm5c_-HET EpiLCs.
-5) enrichPlot gene ontology results of germline genes misexpressed in male versus female _Kdm5c_ mutant EpiLCs
+7) enrichPlot gene ontology results of germline genes misexpressed in male versus female _Kdm5c_ mutant EpiLCs
 
